---
title: "Eight Views of a Playtest"
subtitle: "What you will be able to build by the end of these notes"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show the code"
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

## Ten questions R can answer

These are the questions that come up in the week after a playtest. None of them is
exotic. All of them are painful to answer with a spreadsheet, and each one is
answered below by a single figure and a few dozen lines of R.

| The question you are actually asking | Answered by |
| :- | :- |
| I have 120 rows of results and no idea what is in them. Where do I start? | [View 1](#view-1-the-table-you-can-actually-read) |
| All four characters average about the same score. Does that mean they are balanced? | [View 2](#view-2-distributions-not-averages) |
| I designed four classes to play differently. Did they stay different once real people played them? | [View 3](#view-3-six-numbers-as-a-shape) |
| I want players to score higher. Which ability is actually worth buffing? | [View 4](#view-4-what-actually-drives-score) |
| Testers keep giving up partway through. When do I lose them, and is it the same moment every time? | [View 5](#view-5-when-people-stop-playing) |
| The survey came back with 960 ratings across eight questions. Which complaint do I fix first? | [View 6](#view-6-what-players-said-ranked-by-how-badly-it-went) |
| My frame-rate counter says 60 FPS. My testers say the game stutters. Who is lying? | [View 7](#view-7-the-frame-rate-counter-is-lying-to-you) |
| How can I find where players are loitering, sniping, or being funnelled into a kill zone? | [View 8](#view-8-where-players-actually-stand) |
| Which parts of my level is nobody using? | [View 8](#view-8-where-players-actually-stand) |
| How did one particular player move through the level — did they learn the route, or wander? | [View 8b](#following-one-player) |

Every one of these is a question about **your own game**, asked of data your own
engine can already record. The only thing missing is the ability to read it.

## Before you read any code

This document is a **showcase**, not a lesson. Nothing here is examinable and you
are not expected to understand the code yet. Its only job is to show you what three
weeks of R can do to a pile of raw playtest numbers.

```{r line-counts}
#| include: false
src    <- readLines("index.qmd")
starts <- grep("^```\\{r view-", src)
ends   <- grep("^```$", src)
view_lines <- sapply(starts, function(s) {
  body <- src[(s + 1):(min(ends[ends > s]) - 1)]
  sum(!grepl("^\\s*$|^#\\|", body))
})
n_csv <- length(grep('read_csv\\("data/', src))
```

Every figure below was produced from `r n_csv` CSV files, in chunks of between
`r min(view_lines)` and `r max(view_lines)` lines of code. By the end of these
notes you will be able to answer the first seven questions with base R, dplyr and
ggplot2, producing plainer, static versions of the plots shown here. The three
movement questions in View 8 use position data that stays in this showcase. The
polish on this page is the destination, not the syllabus.

::: {.callout-note}
## The data
*Tidebreak* is a fictional four-player top-down arena, used here so that the
example is self-contained. The playtest recorded four things:

| File | One row per | Holds |
| :- | :- | :- |
| `tidebreak_sessions.csv` | player, per session | Six end-of-match ability ratings, eleven match statistics, session length and quit reason, for 120 sessions |
| `tidebreak_survey.csv` | player, per session | Eight agreement items rated 1 to 5 |
| `tidebreak_frame_times.csv` | rendered frame | Frame time in milliseconds, by build and level zone |
| `tidebreak_positions.csv` | player, per second | Top-down position for eight players across one 180-second match |

All four are synthetic, but they behave like the real thing: the numbers are
messy, the interesting findings are not the obvious ones, and at least two widely
held assumptions about the game turn out to be wrong.
:::

```{r setup}
library(tidyverse)
library(plotly)
library(DT)

sessions    <- read_csv("data/tidebreak_sessions.csv")
survey      <- read_csv("data/tidebreak_survey.csv")
frame_times <- read_csv("data/tidebreak_frame_times.csv")
positions   <- read_csv("data/tidebreak_positions.csv")
level       <- read_csv("data/tidebreak_level.csv")

# A small, fixed palette used by every figure in this document.
pal <- c(blue = "#2a78d6", orange = "#eb6834", aqua = "#1baf7a", yellow = "#eda100")
ink <- c(primary = "#0b0b0b", secondary = "#52514e", grid = "#e6e5e1")

# Each archetype keeps the same colour in every figure, so that a colour always
# means the same thing no matter which chart you are looking at.
arch_pal <- c(Scout = pal[["blue"]],  Sentinel = pal[["orange"]],
              Engineer = pal[["aqua"]], Duelist = pal[["yellow"]])

# Likewise for the two builds, which appear in Views 5 and 7.
build_pal <- c(A = pal[["orange"]], B = pal[["blue"]])

theme_tb <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title      = element_text(face = "bold", colour = ink["primary"], size = 13),
      plot.subtitle   = element_text(colour = ink["secondary"], size = 11),
      axis.title      = element_text(colour = ink["secondary"], size = 10),
      axis.text       = element_text(colour = ink["secondary"]),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = ink["grid"], linewidth = 0.3),
      legend.position = "none"
    )
}
```

## View 1 — The table you can actually read

**Question: what happened in each session?**

Before you analyse anything you need to be able to look at your data. A raw CSV in
Excel is awkward at 120 rows and unusable at 120,000. A rendered table is
searchable and sortable, and it goes straight into a report.

Sort by **Score**, or type `Sentinel` into the search box, and you will already
see the finding that View 2 makes formal.

```{r view-1}
sessions %>%
  select(Session = session_id, Player = player_id, Cohort = cohort,
         Archetype = archetype, Build = build, Score = score,
         Kills = kills, Deaths = deaths, `Alive (s)` = time_alive_s,
         Awareness = awareness) %>%
  datatable(
    rownames = FALSE,
    options  = list(pageLength = 8, dom = "ftip", scrollX = TRUE),
    caption  = "Tidebreak playtest: 120 sessions"
  ) %>%
  formatStyle("Score",
    background = styleColorBar(range(sessions$score), "#cde2fb"),
    backgroundSize = "98% 60%", backgroundRepeat = "no-repeat",
    backgroundPosition = "center")
```

## View 2 — Distributions, not averages

**Question: which archetype performs best?**

The tempting answer is to compare four average scores. Do that here and the
archetypes look almost identical, which would be a design decision made on a
misreading.

A violin plot draws the whole distribution. Read the **Sentinel** shape: it is not
one hump but two. Sentinel runs either anchor the objective and score well, or
collapse in the opening seconds and score badly. There is almost no middle. That
is a balance problem an average would have hidden completely.

```{r view-2}
sessions %>%
  mutate(archetype = fct_reorder(archetype, score, .fun = median)) %>%
  ggplot(aes(archetype, score, fill = archetype)) +
  geom_violin(colour = NA, alpha = 0.85, width = 0.9) +
  geom_boxplot(width = 0.12, fill = "white", colour = ink[["secondary"]],
               outlier.shape = NA, linewidth = 0.4) +
  scale_fill_manual(values = arch_pal) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Sentinel is bimodal: it either anchors the round or dies in it",
    subtitle = "Score distribution by archetype, 30 sessions each. Box shows median and interquartile range.",
    x = NULL, y = "Score"
  ) +
  theme_tb()
```

## View 3 — Six numbers as a shape

**Question: are the four archetypes actually playing differently?**

Each session ends with six ability ratings on a 0–100 scale. Comparing four
archetypes across six measures means twenty-four numbers, which no one can hold in
their head. Drawn as a radar, each archetype becomes a shape you can recognise at a
glance, and the grey outline behind each one is the all-player average for
reference.

The shapes are distinct and they are roughly the shapes the design intended. That
is the result you want from this chart: confirmation that the archetypes are not
quietly converging on the same playstyle.

```{r view-3, fig.width=9, fig.height=3.4}
metrics <- c("accuracy", "mobility", "survivability",
             "objective_control", "support", "awareness")

pretty <- c(accuracy = "Accuracy", mobility = "Mobility",
            survivability = "Survival", objective_control = "Objective",
            support = "Support", awareness = "Awareness")

long <- sessions %>%
  select(archetype, all_of(metrics)) %>%
  pivot_longer(-archetype, names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(pretty[metric], levels = unname(pretty)))

by_arch <- long %>% group_by(archetype, metric) %>% summarise(value = mean(value))
overall <- long %>% group_by(metric) %>% summarise(value = mean(value))

ggplot() +
  geom_polygon(data = overall, aes(metric, value, group = 1),
               fill = NA, colour = ink[["grid"]], linewidth = 0.8) +
  geom_polygon(data = by_arch,
               aes(metric, value, group = archetype, fill = archetype, colour = archetype),
               alpha = 0.35, linewidth = 0.9) +
  geom_point(data = by_arch, aes(metric, value, colour = archetype), size = 1.8) +
  coord_polar(clip = "off") +
  facet_wrap(~archetype, nrow = 1) +
  scale_fill_manual(values = arch_pal) +
  scale_colour_manual(values = arch_pal) +
  scale_y_continuous(limits = c(0, 100), breaks = c(25, 50, 75)) +
  labs(
    title    = "Four archetypes, four distinct ability profiles",
    subtitle = "Mean end-of-match rating per ability. Grey outline is the all-player average.",
    x = NULL, y = NULL
  ) +
  theme_tb() +
  theme(axis.text.y   = element_blank(),
        axis.text.x   = element_text(size = 7),
        strip.text    = element_text(face = "bold", colour = ink[["primary"]]),
        panel.spacing = unit(1.6, "lines"),
        plot.margin   = margin(10, 20, 10, 20))
```

## View 4 — What actually drives score

**Question: which abilities and behaviours move the scoreboard?**

A correlogram puts every pair of variables in one picture. Blue means the two rise
together, red means one rises as the other falls, and pale means no relationship
at all.

Three things to read off it:

- **Awareness** is the strongest positive correlate of score. Map knowledge beats
  mechanical skill in this build.
- **Shots fired** is essentially uncorrelated with score. Players who shoot more do
  not win more, which contradicts what the team assumed.
- **Mobility** and **survivability** pull against each other, which is the intended
  trade-off working as designed.

```{r view-4, fig.height=6}
vars <- c("accuracy", "mobility", "survivability", "objective_control",
          "support", "awareness", "shots_fired", "distance_m",
          "kills", "deaths", "objectives", "score")

nice <- c(accuracy = "Accuracy", mobility = "Mobility", survivability = "Survival",
          objective_control = "Objective ctrl", support = "Support",
          awareness = "Awareness", shots_fired = "Shots fired",
          distance_m = "Distance", kills = "Kills", deaths = "Deaths",
          objectives = "Objectives", score = "Score")

cm <- cor(sessions[vars])
ord <- hclust(as.dist(1 - cm))$order
lv  <- nice[vars[ord]]

cm_long <- as.data.frame(as.table(cm)) %>%
  setNames(c("a", "b", "r")) %>%
  mutate(a = factor(nice[as.character(a)], levels = lv),
         b = factor(nice[as.character(b)], levels = lv)) %>%
  filter(as.integer(a) < as.integer(b))   # lower triangle only; the diagonal is always 1

ggplot(cm_long, aes(a, b, fill = r)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = sprintf("%.2f", r),
                colour = abs(r) > 0.55), size = 2.9) +
  scale_fill_gradient2(low = "#e34948", mid = "#f0efec", high = "#2a78d6",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = ink[["secondary"]]),
                      guide = "none") +
  coord_fixed() +
  labs(
    title    = "Awareness drives score; shots fired does not",
    subtitle = "Pearson correlation between every pair of measures, 120 sessions.",
    x = NULL, y = NULL
  ) +
  theme_tb() +
  theme(axis.text.x       = element_text(angle = 45, hjust = 1),
        legend.position   = "right",
        panel.grid.major  = element_blank(),
        panel.grid.minor  = element_blank())
```

## View 5 — When people stop playing

**Question: we changed how the game explains its objective. Did it keep anyone?**

A retention curve shows what fraction of participants were still playing at each
minute of the test. It is the shape that matters, not the endpoint: a smooth decay
means people drifted away, a cliff means something specific pushed them out.

Build A has a cliff. Between minutes 11 and 13 it drops from 88% to 68%, and 15 of
the 17 people who left in that window are logged as frustration quits. Build B,
which explains its objective differently, is still at 90% at minute 13 and loses
five people to frustration across the whole test against Build A's twenty.

Now look at what a summary table would have told you: median session length is 23.2
minutes for Build A and 27.1 for Build B. That reads as a modest improvement. The
curve shows something else — one specific, fixable moment where a fifth of the
Build A testers gave up.

```{r view-5, fig.height=5}
minutes <- seq(0, 32, by = 0.25)

retention <- map_dfr(c("A", "B"), function(b) {
  lengths <- sessions %>% filter(build == b) %>% pull(session_minutes)
  tibble(build = b, minute = minutes,
         pct = map_dbl(minutes, ~ 100 * mean(lengths >= .x)))
})

ends <- retention %>% group_by(build) %>% slice_max(minute, n = 1)

ggplot(retention, aes(minute, pct, colour = build)) +
  annotate("rect", xmin = 11, xmax = 13, ymin = 0, ymax = 100,
           fill = ink[["grid"]], alpha = 0.55) +
  annotate("text", x = 13.4, y = 24, hjust = 0, size = 3.2,
           colour = ink[["secondary"]],
           label = "88% to 68% in two minutes,\nalmost all frustration quits") +
  geom_step(linewidth = 1.1, direction = "hv") +
  geom_text(data = ends, aes(label = paste0("Build ", build)),
            hjust = 0, nudge_x = 0.6, size = 3.6, fontface = "bold",
            show.legend = FALSE) +
  scale_colour_manual(values = build_pal) +
  scale_x_continuous(limits = c(0, 37), breaks = seq(0, 30, 5),
                     labels = function(x) paste0(x, " min")) +
  scale_y_continuous(limits = c(0, 100), labels = function(y) paste0(y, "%")) +
  labs(
    title    = "Build A loses people at a cliff; Build B does not",
    subtitle = "Percentage of participants still playing, by minute of the test session.",
    x = NULL, y = "Still playing"
  ) +
  theme_tb()
```

## View 6 — What players said, ranked by how badly it went

**Question: the survey has eight questions and 120 responses. Which one should we
act on first?**

Eight averages tell you almost nothing, because a mean of 3.0 can mean everyone
was indifferent or that half loved it and half hated it. A diverging bar chart
splits each item at the neutral response: disagreement runs left, agreement runs
right, and the items sort by the net.

The result is uncomfortable in the way real survey results usually are. The art is
the best-rated thing in the game by a distance. The worst-rated item is whether
players ever knew what they were supposed to be doing — the same thing that drives
the cliff in View 5. People praise the visuals and quit over the design.

```{r view-6, fig.height=5}
item_label <- c(
  objective_clear     = "I always knew what my objective was",
  death_understood    = "I understood why I died",
  difficulty_fair     = "The difficulty felt fair",
  level_navigable     = "The level was easy to navigate",
  would_play_again    = "I would play this again",
  match_length        = "The match length felt right",
  controls_responsive = "The controls felt responsive",
  visual_style        = "The visual style is appealing"
)

resp_label <- c("Strongly disagree", "Disagree", "Neutral", "Agree", "Strongly agree")

props <- survey %>%
  pivot_longer(-session_id, names_to = "item", values_to = "resp") %>%
  count(item, resp) %>%
  complete(item, resp = 1:5, fill = list(n = 0)) %>%
  group_by(item) %>% mutate(pct = 100 * n / sum(n)) %>% ungroup()

# Net agreement decides the order; the neutral block is split so the bars
# pivot on a true zero rather than on the left edge.
ranking <- props %>%
  group_by(item) %>%
  summarise(net = sum(pct[resp >= 4]) - sum(pct[resp <= 2])) %>%
  arrange(net)

item_levels <- unname(item_label[ranking$item])

as_bars <- function(d, sign) {
  d %>%
    mutate(signed = sign * if_else(resp == 3, pct / 2, pct),
           item   = factor(unname(item_label[item]), levels = item_levels),
           resp   = factor(resp_label[resp], levels = resp_label))
}

# The two sides stack in opposite directions, so that the strongest opinion
# always sits furthest from the centre line.
bars_neg <- props %>% filter(resp <= 3) %>% as_bars(-1)
bars_pos <- props %>% filter(resp >= 3) %>% as_bars(+1)

ranking <- ranking %>%
  mutate(item = factor(unname(item_label[item]), levels = item_levels))

ggplot(mapping = aes(signed, item, fill = resp)) +
  geom_col(data = bars_neg, width = 0.72, position = position_stack()) +
  geom_col(data = bars_pos, width = 0.72, position = position_stack(reverse = TRUE)) +
  geom_vline(xintercept = 0, colour = "#fcfcfb", linewidth = 0.9) +
  geom_text(data = ranking, aes(x = 106, y = item, label = sprintf("%+.0f", net)),
            inherit.aes = FALSE, size = 3.3, fontface = "bold",
            colour = ink[["secondary"]]) +
  scale_fill_manual(values = c("Strongly disagree" = "#e34948",
                               "Disagree"          = "#f2a9a8",
                               "Neutral"           = "#d9d8d4",
                               "Agree"             = "#9ec5f4",
                               "Strongly agree"    = "#2a78d6")) +
  # coord_cartesian zooms; scale limits would DROP the segments that fall
  # outside them, silently truncating the longest bars.
  scale_x_continuous(breaks = seq(-60, 90, 30),
                     labels = function(x) paste0(abs(x), "%")) +
  coord_cartesian(xlim = c(-68, 112)) +
  labs(
    title    = "They love the art and cannot tell what the game wants",
    subtitle = "Share of 120 responses. Bars pivot on neutral; the number is net agreement.",
    x = NULL, y = NULL, fill = NULL
  ) +
  theme_tb() +
  theme(legend.position = "top",
        legend.text     = element_text(size = 9),
        panel.grid.major.y = element_blank())
```

## View 7 — The frame-rate counter is lying to you

**Question: both builds average about 59 FPS. Testers say Build B stutters. Who is
right?**

Both are. Average frame rate is a mean, and a mean is exactly the wrong summary for
a quantity where the rare bad values are the entire user experience. Build B
renders almost every frame slightly faster than Build A. It also renders one frame
in every twenty-five — about twice a second — in 35 to 55 ms instead of 17, and
only in one zone of the level: the corridor that View 8 identifies as the choke
point.

The trace shows it. The table shows why the FPS readout missed it: the two builds
are separated by 0.05 ms at the mean and by 20 ms at the 99th percentile.

```{r view-7, fig.height=4.2}
ggplot(frame_times %>% filter(zone == "Corridor", frame <= 400),
       aes(frame, ms, colour = build)) +
  geom_hline(yintercept = 1000 / 60, linetype = "22",
             colour = ink[["secondary"]], linewidth = 0.4) +
  geom_line(linewidth = 0.45) +
  facet_wrap(~build, ncol = 1, labeller = labeller(build = function(b) paste("Build", b))) +
  scale_colour_manual(values = build_pal) +
  scale_y_continuous(limits = c(0, 60), breaks = c(1000/60, 30, 45, 60),
                     labels = c("16.7\n(60 FPS)", "30", "45", "60")) +
  labs(
    title    = "Identical average, one of them unplayable",
    subtitle = "Frame time in the Corridor zone, first 400 frames. Dashed line is the 60 FPS budget.",
    x = "Frame", y = "Frame time (ms)"
  ) +
  theme_tb() +
  theme(strip.text = element_text(face = "bold", colour = ink[["primary"]]))
```

```{r view-7-table}
frame_times %>%
  group_by(Build = build) %>%
  summarise(
    `Mean FPS`     = round(1000 / mean(ms), 1),
    `Mean (ms)`    = round(mean(ms), 2),
    `Median (ms)`  = round(median(ms), 1),
    `95th (ms)`    = round(quantile(ms, 0.95), 1),
    `99th (ms)`    = round(quantile(ms, 0.99), 1),
    `Worst (ms)`   = round(max(ms), 1)
  ) %>%
  knitr::kable(caption = "Every column agrees the builds are the same until the last two.")
```

## View 8 — Where players actually stand

**Question: where are the choke points, the camping spots and the dead space?**

This is the figure that changes level layouts. The blue field is total time spent
per square metre across the whole match. The moving dots are the eight players,
one frame per second, coloured by whether they are moving, loitering or holding
position. Dark blocks are cover; the block in the north-east is the raised gantry.

Press play, or drag the slider.

Three readings, all of them actionable:

- The gap between the central core and the east block carries almost all lateral
  traffic, and **14 of the 20 deaths in this match happened inside it**. That is a
  choke point.
- Two players spend the entire match holding the **north-east gantry**. Long dwell,
  no movement, high kill share: a sniper perch.
- The **west corner** is never visited by anyone. That is floor space the level is
  paying for and not using.

```{r view-8, fig.height=6}
# Time spent per 2m x 2m cell, for the background field.
gx_seq <- seq(1, 59, 2)
gy_seq <- seq(1, 39, 2)

field <- positions %>%
  mutate(gx = floor(x / 2) * 2 + 1, gy = floor(y / 2) * 2 + 1) %>%
  count(gx, gy) %>%
  complete(gx = gx_seq, gy = gy_seq, fill = list(n = 0)) %>%
  arrange(gy, gx)

z <- matrix(field$n, nrow = length(gy_seq), ncol = length(gx_seq), byrow = TRUE)

# Level geometry drawn as rectangles behind the data.
shapes <- pmap(level, function(block, xmin, xmax, ymin, ymax, kind) {
  list(type = "rect", x0 = xmin, x1 = xmax, y0 = ymin, y1 = ymax,
       fillcolor = if (kind == "elevated") "#6f6e6a" else "#2f2e2c",
       # The gantry is walkable, so it stays translucent and the dwell field
       # underneath it still reads.
       opacity = if (kind == "elevated") 0.35 else 0.92,
       line = list(width = 0), layer = "above")
})

# The animated layer starts at the first second where all three states occur, so
# that the legend is complete from the first frame.
first_full <- positions %>%
  group_by(t) %>% summarise(k = n_distinct(state)) %>%
  filter(k == 3) %>% slice_min(t) %>% pull(t)

deaths <- positions %>% filter(event == "death")

plot_ly(height = 560) %>%
  add_heatmap(
    x = gx_seq, y = gy_seq, z = z, zmin = 0, zmax = 60,
    colorscale = list(c(0, "#fcfcfb"), c(0.25, "#cde2fb"),
                      c(0.6, "#5598e7"), c(1, "#104281")),
    showscale = TRUE, hoverinfo = "z",
    colorbar = list(title = "Seconds\nspent", len = 0.55, thickness = 12,
                    y = 0.72, ticksuffix = "  ")
  ) %>%
  add_markers(
    data = deaths, x = ~x, y = ~y, name = "Death",
    marker = list(symbol = "x", size = 11, color = "#e34948",
                  line = list(color = "#fcfcfb", width = 1)),
    text = ~paste0("death · ", player_id, " · t=", t, "s"), hoverinfo = "text"
  ) %>%
  add_markers(
    data = positions %>% filter(t >= first_full),
    x = ~x, y = ~y, frame = ~t, color = ~state,
    colors = c(Holding = pal[["orange"]], Loitering = pal[["aqua"]],
               Moving = pal[["blue"]]),
    marker = list(size = 13, line = list(color = "#fcfcfb", width = 2)),
    text = ~paste0(player_id, " · t=", t, "s · ", state),
    hoverinfo = "text"
  ) %>%
  layout(
    title  = list(text = "<b>One corridor carries the traffic and most of the deaths</b>",
                  x = 0, font = list(size = 14)),
    shapes = shapes,
    xaxis  = list(title = "x (m)", range = c(0, 60), constrain = "domain"),
    yaxis  = list(title = "y (m)", range = c(0, 40), scaleanchor = "x"),
    legend = list(orientation = "h", x = 0, y = 1.02, title = list(text = "")),
    margin = list(t = 80)
  ) %>%
  animation_opts(frame = 90, transition = 45, redraw = FALSE) %>%
  animation_slider(currentvalue = list(prefix = "t = ", suffix = " s"))
```

### Following one player

The map above answers questions about the *level*. To ask questions about a
*player* — did they learn the route, did they loop, did they ever leave their
corner — you need to follow one of them.

First, all eight routes at once. The line darkens as the match progresses, so you
can read direction without playing anything. P03 covers the most ground (268 m).
P07 travels almost as far (190 m) without ever leaving a two-metre-wide lane — it
shuttles up and down the same corridor for the whole match. P01 and P02 manage 51 m
and 35 m between them across three minutes.

```{r view-8b-overview, fig.width=9, fig.height=4.6}
walls <- level %>% filter(kind != "elevated")
raised <- level %>% filter(kind == "elevated")

ggplot() +
  geom_rect(data = walls, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#2f2e2c", alpha = 0.9) +
  geom_rect(data = raised, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#6f6e6a", alpha = 0.55) +
  geom_path(data = positions, aes(x, y, colour = t), linewidth = 0.7) +
  facet_wrap(~player_id, nrow = 2) +
  scale_colour_gradient(low = "#cde2fb", high = "#104281",
                        name = "Time (s)", breaks = c(1, 90, 180)) +
  coord_fixed(xlim = c(0, 60), ylim = c(0, 40), expand = FALSE) +
  labs(
    title    = "Eight routes through the same three minutes",
    subtitle = "Line darkens as the match progresses. Two players never leave the gantry.",
    x = NULL, y = NULL
  ) +
  theme_tb() +
  theme(axis.text         = element_blank(),
        panel.grid.major  = element_blank(),
        panel.grid.minor  = element_blank(),
        strip.text        = element_text(face = "bold", colour = ink[["primary"]]),
        legend.position   = "right",
        legend.key.height = unit(0.9, "lines"))
```

Now one player, animated, with the trail drawn behind them. Change `focus_player`
to any ID from the panel above and re-run the chunk. The orange dots mark every
second this player spent holding position, so you can see where they will stop
before they get there.

```{r view-8b-animated, fig.height=6}
focus_player <- "P06"

path <- positions %>% filter(player_id == focus_player) %>% arrange(t)

# For each second, every position up to that second: this is what makes the
# trail grow rather than jump.
trail <- map_dfr(path$t, function(k) path %>% filter(t <= k) %>% mutate(f = k))

plot_ly(height = 560) %>%
  add_markers(
    data = path %>% filter(state == "Holding"), x = ~x, y = ~y,
    name = "Held position",
    marker = list(size = 7, color = pal[["orange"]], opacity = 0.7),
    text = ~paste0("holding · t=", t, "s"), hoverinfo = "text"
  ) %>%
  add_markers(
    data = path %>% slice(1), x = ~x, y = ~y, name = "Start",
    marker = list(symbol = "circle-open", size = 16, color = ink[["primary"]],
                  line = list(width = 2.5)),
    hoverinfo = "text", text = "start"
  ) %>%
  add_paths(
    data = trail, x = ~x, y = ~y, frame = ~f,
    line = list(color = pal[["blue"]], width = 2.5),
    opacity = 0.8, showlegend = FALSE, hoverinfo = "skip"
  ) %>%
  add_markers(
    data = path, x = ~x, y = ~y, frame = ~t,
    marker = list(size = 16, color = pal[["blue"]],
                  line = list(color = "#fcfcfb", width = 2.5)),
    showlegend = FALSE,
    text = ~paste0(focus_player, " · t=", t, "s · ", state), hoverinfo = "text"
  ) %>%
  layout(
    title  = list(text = paste0("<b>", focus_player, "'s route through the match</b>"),
                  x = 0, font = list(size = 14)),
    shapes = shapes,
    xaxis  = list(title = "x (m)", range = c(0, 60), constrain = "domain"),
    yaxis  = list(title = "y (m)", range = c(0, 40), scaleanchor = "x"),
    legend = list(orientation = "h", x = 0, y = 1.02, title = list(text = "")),
    margin = list(t = 80)
  ) %>%
  animation_opts(frame = 70, transition = 0, redraw = FALSE) %>%
  animation_slider(currentvalue = list(prefix = "t = ", suffix = " s"))
```

## What each view cost

| View | Chart type | Lines of R | Skills it needs |
| :- | :- | :- | :- |
| 1 | Interactive table | 14 | Reading data; selecting and renaming columns |
| 2 | Violin with boxplot | 14 | Describing distributions |
| 3 | Radar small multiples | 34 | Reshaping data; grouped summaries |
| 4 | Correlogram | 35 | Relationships between variables |
| 5 | Retention step curve | 27 | Building a summary from scratch; annotation |
| 6 | Diverging Likert bars | 55 | Reshaping; ordering by a computed value |
| 7 | Frame-time trace + summary table | 16 + 11 | Filtering; percentiles; rendered tables |
| 8 | Animated position map | 53 | Grouping, counting, interactive output |
| 8b | Route small multiples | 24 | Drawing paths; faceting by player |
| 8b | Animated single-player trail | 39 | Filtering to one case; building frames |

None of this is difficult, and most of the length in Views 3 onwards is labelling
and styling rather than analysis. All of it is repeatable: the same code runs unchanged
on next week's build, which is the real reason to write analysis as code rather
than clicking through a spreadsheet.

The rest of these notes teach you the parts, in order, starting with the question
that matters most: **what are you actually trying to find out?**
