---
title: "Lab 04 — Relationships in Your Data"
subtitle: "Stage 3B clinic: correlation on Tidebreak, then a correlation matrix of your own measures"
---

::: {.callout-note appearance="simple" icon="false" title="Get the files"}
- **Data:** [tidebreak_sessions.csv](../data/tidebreak_sessions.csv){download="tidebreak_sessions.csv"}, [tidebreak_survey.csv](../data/tidebreak_survey.csv){download="tidebreak_survey.csv"}. Save them in the `data/` folder of your project.
:::

# Overview

This is a clinic, not a full lab. It is short, and it ends with your team's own
data. The aim is one thing:

> **A correlation matrix of 3–5 meaningful measures from your own playtest, read
> and checked, with one relationship written up as a finding.**

Part A is a warm-up on *Tidebreak*: measure relationships, choose between Pearson
and Spearman, read a matrix, relate telemetry to self-report, and catch a
relationship that reverses within groups. Part B is your project. Bring your
team's latest playtest export, in your project's `data/` folder.

| Function | What it does |
| :- | :- |
| `cor(x, y)` | Pearson correlation, −1 to +1 |
| `cor(x, y, method = "spearman")` | Spearman correlation: for ratings, curves and outliers |
| `select()` then `cor()` | A correlation matrix of the columns you chose |
| `pairs()` | Every pair as a small scatterplot, to check the matrix |
| `left_join()` | Add columns from another table, matched on a key |
| `group_by()` + `summarise(r = cor(...))` | The correlation within each group |

# Learning outcomes

By the end of this clinic you can:

- measure a relationship with `cor()`, and choose between Pearson and Spearman;
- build and read a correlation matrix of 3–5 measures, and check it with
  scatterplots;
- relate a telemetry measure to a self-report measure, and say what it means when
  they disagree;
- check a relationship within groups before trusting it;
- write a relationship as a finding without claiming a cause.

# Ground rules

- Draw before you quote. Every correlation stands in for a scatterplot.
- Choose your measures for a question. A matrix of every column finds luck.
- Write "is associated with", never "causes", unless someone changed something on
  purpose.
- Type the code. Open the solution **after** your attempt.

# Setup (run this first)

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
survey   <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)

nrow(sessions)
nrow(survey)
```

---

# Part A — Warm-up on Tidebreak

Notes for this part:
[Relationships and Correlation](../notes/10-relationships-and-correlation.qmd).

## Exercise 01 — What goes with score? `[Do]`

Use `cor()` to measure how strongly `score` is related to `awareness`, and to
`shots_fired`.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex01
cor(sessions$awareness, sessions$score)
cor(sessions$shots_fired, sessions$score)
```

`cor()` gives a number from −1 to +1. Near 0 means no straight-line relationship;
the sign gives the direction.
:::

**Write 2 sentences:** a designer wants to reward players who fire more shots,
"because they are the ones scoring". What do you tell them?

## Exercise 02 — Pearson or Spearman? `[Apply]`

Do players who die more rate the game as harder? Draw `difficulty` against
`deaths`, then compute both the Pearson and the Spearman correlation. Which would
you report, and why?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex02
#| fig-alt: "Jittered scatterplot of difficulty rating against deaths, rising from lower left to upper right with a lot of scatter."
ggplot(sessions, aes(x = deaths, y = difficulty)) +
  geom_jitter(width = 0.15, height = 0.15, alpha = 0.6) +
  labs(x = "Deaths", y = "Difficulty rating (1-7)")

cor(sessions$deaths, sessions$difficulty)
cor(sessions$deaths, sessions$difficulty, method = "spearman")
```

The two agree closely, because the relationship is roughly straight with no wild
values. Report Spearman anyway: `difficulty` is a 1–7 rating, so the gaps between
its values are not known to be equal, and Spearman only uses their order.
`geom_jitter()` nudges the points so that sessions with the same values do not hide
each other.
:::

**Write 1 sentence:** describe the relationship in words a designer would use,
including how loose it is.

## Exercise 03 — A correlation matrix `[Apply]`

Build a Spearman correlation matrix of `deaths`, `wrong_turns`, `difficulty`,
`enjoyment` and `session_minutes`, rounded to 2 places. Then draw every pair with
`pairs()`. Which pair goes together most strongly, and which measure goes with
nothing?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex03
m <- sessions |>
  select(deaths, wrong_turns, difficulty, enjoyment, session_minutes)

m |> cor(method = "spearman") |> round(2)
```

```{r}
#| label: ex03b
#| fig-alt: "A five by five grid of small scatterplots. Deaths against difficulty shows a rising band; wrong turns show no pattern against anything."
#| fig-height: 6
pairs(m, pch = 19, cex = 0.4, col = adjustcolor("grey30", alpha.f = 0.4))
```

```{r}
#| label: ex03-check
#| include: false
cm <- cor(m, method = "spearman")
stopifnot(cm["deaths", "difficulty"] > 0.5,
          all(abs(cm["wrong_turns", -2]) < 0.15))
```

Ignore the diagonal of 1s and read one triangle. `deaths` and `difficulty` go
together most strongly, `r round(cm["deaths", "difficulty"], 2)`. `wrong_turns` is
close to 0 with every other measure. The `pairs()` panels agree: one rising band
for deaths and difficulty, shapeless clouds for wrong turns.
:::

**Write 2 sentences:** wrong turns differ hugely between the builds, yet correlate
with nothing here. Why might that be, and what would you check next?

## Exercise 04 — Telemetry against self-report `[Apply]`

Do players who got lost more rate the level as harder to navigate? Join the survey
onto the sessions with `left_join()`, matching on `session_id`. Then find the
Spearman correlation between `wrong_turns` and `level_navigable`, and each build's
mean of both.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex04
joined <- sessions |>
  left_join(survey, by = "session_id")

cor(joined$wrong_turns, joined$level_navigable, method = "spearman")

joined |>
  group_by(build) |>
  summarise(mean_wrong_turns = mean(wrong_turns),
            mean_navigable   = mean(level_navigable))
```

```{r}
#| label: ex04-check
#| include: false
r_nav <- cor(joined$wrong_turns, joined$level_navigable, method = "spearman")
nav_b <- tapply(joined$level_navigable, joined$build, mean)
stopifnot(abs(r_nav) < 0.15, abs(diff(nav_b)) < 0.2)
```

The correlation is `r round(r_nav, 2)`: no relationship. Build A's players made far
more wrong turns but rated navigation almost exactly as Build B's did,
`r round(nav_b[["A"]], 2)` against `r round(nav_b[["B"]], 2)`. The telemetry saw a
navigation problem; the players did not report one.
:::

**Write 3 sentences:** what does this mean for a team that only asks players
whether the level was easy to navigate? What would **observation** add?

## Exercise 05 — The trade-off `[Apply]`

Draw `survivability` against `mobility` as a scatterplot with a straight-line
trend, and measure the relationship with `cor()`.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex05
#| fig-alt: "Scatterplot of survivability against mobility with a falling black trend line."
ggplot(sessions, aes(x = mobility, y = survivability)) +
  geom_point(colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  labs(title = "Mobility against survivability",
       x = "Mobility rating", y = "Survivability rating")

cor(sessions$mobility, sessions$survivability)
```

`geom_smooth(method = "lm")` draws the straight line that best fits the points;
`se = FALSE` leaves out the shaded band. The line and `cor()` describe the same
relationship, one as a picture and one as a number.
:::

**Write 1 sentence:** describe the relationship in words a balance designer would
use.

## Exercise 06 — Faster players are more fragile? `[Decide]`

```{r}
#| label: ex06-setup
#| include: false
r_all  <- cor(sessions$mobility, sessions$survivability)
r_arch <- sapply(split(sessions, sessions$archetype),
                 function(d) cor(d$mobility, d$survivability))
stopifnot(r_all < -0.5, all(r_arch > 0))
```

The balance designer reads Exercise 05 and proposes a patch: *"Faster players are
more fragile, so when a Scout player gets faster, cut their health to match."*
Redraw the scatterplot with each archetype in its own colour and its own trend
line, compute the correlation within each archetype, and decide whether the patch
follows from the data.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex06
#| fig-alt: "Survivability against mobility with points coloured by archetype, each with its own rising trend line. The four clusters sit along a falling diagonal, Sentinel at the top left and Scout at the bottom right."
ggplot(sessions, aes(x = mobility, y = survivability, colour = archetype)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Mobility against survivability, by archetype",
       x = "Mobility rating", y = "Survivability rating", colour = "Archetype")

sessions |>
  group_by(archetype) |>
  summarise(r = cor(mobility, survivability))
```

**One defensible answer: no, the patch reads the data backwards.** Across all
players the correlation is `r round(r_all, 2)`. Within every archetype it is
positive, from `r round(min(r_arch), 2)` to `r round(max(r_arch), 2)`. The
downward trend is the gap *between* archetypes: Scouts are built fast and fragile,
Sentinels slow and tough. Within any one archetype, better players are better at
both. **Archetype is the confounding variable**: it drives both measures, and
ignoring it reverses the direction of the relationship.

**What makes it defensible.** It checks the relationship inside the groups the
design is made of before drawing a conclusion about individuals, and it says what
the overall correlation *does* show: the archetypes are balanced against each other
as intended.

**What would make it indefensible.** Applying the patch because the correlation is
strong. A strong correlation across groups says nothing about what happens to one
player within a group.
:::

**Write 3 sentences:** name the confounder, say what it does to the conclusion, and
say which section of the report (Results or Discussion) this belongs in.

---

# Part B — Your project `[Decide]`

This is the part that matters. Allow at least half the clinic for it.

Work through these steps on your team's own playtest data, in a Quarto document in
your project.

1. **Write the design question** the relationships bear on, in one sentence.
2. **Choose 3–5 numeric measures** that matter to it. Aim for at least one from
   telemetry and one from self-report, so you can see whether they agree. Join
   files on a shared ID with `left_join()` if they are separate.
3. **Build the matrix**: `select()` the measures, then
   `cor(method = "spearman")` if any is a rating, and `round(2)`.
4. **Check it with `pairs()`**. Any correlation whose scatterplot does not look
   like its number is not to be trusted.
5. **Pick the one relationship that matters most** to your question, and draw it
   properly with `ggplot2`.
6. **Check it within groups**: build, level, character or experience, whichever
   your game has. Does it hold inside each?

```r
our_data |>
  select(measure_1, measure_2, measure_3, measure_4) |>
  cor(method = "spearman") |>
  round(2)

our_data |>
  group_by(build) |>
  summarise(n = n(), r = cor(measure_1, measure_2, method = "spearman"))
```

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answer"}
There is no single right matrix: it depends on your game and your question. Check
your work against this list.

- [ ] Every measure in the matrix bears on the question; none is there "in case".
- [ ] At least one telemetry and one self-report measure, if your playtest
  collected both.
- [ ] Spearman if any measure is a rating; Pearson only if the scatterplots are
  straight and free of extreme values.
- [ ] You drew the pairs before quoting any number.
- [ ] The relationship you chose was checked within groups.
- [ ] Your write-up says "is associated with", not "causes".
- [ ] If telemetry and self-report disagree, you have said so: that is a finding,
  not a failure.

Exercises 03, 04 and 06 are worked versions of steps 3 to 6 on *Tidebreak*.
:::

**Write 3–4 sentences**, using **What → Pattern → Meaning → Action**, for the one
relationship that matters most:

- **What** did you measure, on whom, and how many?
- **Pattern**: how strongly do the two measures go together, and does it hold
  within groups?
- **Meaning**: what might explain it? Name at least one explanation other than
  cause.
- **Action**: what would the team test next to find out which explanation is
  right?

---

# Extension — How much, not just how strongly `[Apply]`

For anyone who finishes early. A correlation says how tightly two measures go
together; a regression line says *how much* one changes with the other. This is
Going Further material, explained on [Regression](../going-further/regression.qmd).

Fit a regression of `score` on `awareness` with `lm()`. Report the slope with its
95% confidence interval, then list the seven sessions that scored furthest below
what their awareness predicted.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ext
fit <- lm(score ~ awareness, data = sessions)
coef(fit)
confint(fit)

worst <- order(resid(fit))[1:7]
sessions[worst, c("session_id", "archetype", "awareness", "score")]
```

```{r}
#| label: ext-numbers
#| include: false
b1    <- coef(fit)[["awareness"]]
b1_ci <- confint(fit)["awareness", ]
n_sen <- sum(sessions$archetype[order(resid(fit))[1:7]] == "Sentinel")
stopifnot(n_sen >= 5)
```

Each extra point of awareness goes with about `r round(b1, 1)` more points of
score on average (95% CI `r round(b1_ci[1], 1)` to `r round(b1_ci[2], 1)`).
`resid(fit)` is actual minus predicted, so `order()` puts the biggest
underperformers first. `r n_sen` of the seven are Sentinels: the collapsing games,
which awareness does not explain at all.
:::

**Write 2 sentences:** turn the slope into a sentence a balance designer could use,
then say what the regression cannot tell them.

---

# Appendix — Common errors

::: {.callout-warning collapse="true" title="Show the error guide"}

| Message or symptom | What it usually means |
| :- | :- |
| `'x' must be numeric` | A text column got into `cor()`. `select()` only numeric measures |
| `NA` in the matrix | A column has missing values. Add `use = "complete.obs"` to `cor()`, and say how many rows it left out |
| `the standard deviation is zero` | A measure has the same value in every row: it cannot correlate with anything |
| A matrix of every column you have | Choose 3–5 measures for your question instead |
| `left_join()` gives more rows than you started with | The key is repeated in the second table. Check it with `count(key) \|> filter(n > 1)` |

:::
