---
title: "Lab 03 — A Figure for Your Report"
subtitle: "Stage 3A clinic: ggplot2 on Tidebreak, then on your own playtest data"
---

::: {.callout-note appearance="simple" icon="false" title="Get the files"}
- **Data:** [tidebreak_sessions.csv](../data/tidebreak_sessions.csv){download="tidebreak_sessions.csv"}. Save them in the `data/` folder of your project.
:::

# Overview

This is a clinic, not a full lab. It is short, and it ends with your team's own
data. The aim is one thing:

> **One figure from your own playtest data that you would put in the final
> report.**

Part A is a warm-up on *Tidebreak*: rebuild two plots you already know in
`ggplot2`, then build and fix the kind of figure a report needs. Part B is your
project. Bring your team's latest playtest export, in your project's `data/`
folder.

| Function | What it does |
| :- | :- |
| `ggplot()`, `aes()` | Start a figure: the data, and which column drives which axis or colour |
| `geom_histogram()`, `geom_bar()`, `geom_boxplot()`, `geom_point()` | The marks: bars, boxes, points |
| `facet_wrap()` | One panel per group, on shared axes |
| `labs()` | Title, axis and legend labels |

# Learning outcomes

By the end of this clinic you can:

- rebuild a base R plot in `ggplot2`, and point to the data, mapping and geometry;
- split a figure by group in the way that makes the intended comparison easiest;
- tell mapping an aesthetic from setting one, and fix a figure that confuses them;
- produce one report-quality figure from your own playtest data, with a caption
  that states the takeaway.

# Ground rules

- One figure, one question. Write the question before you write the code.
- Look with base R first if you need to; build the report figure in `ggplot2`.
- Type the code. Open the solution **after** your attempt.
- Check your own figure against the list in
  [A Figure for Your Report](../notes/09-visualising-with-ggplot2.qmd#a-figure-for-your-report).

# Setup (run this first)

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
nrow(sessions)
```

If `library(ggplot2)` fails, run `install.packages("ggplot2")` once in the Console.

---

# Part A — Warm-up on Tidebreak

Notes for this part: [Visualising with ggplot2](../notes/09-visualising-with-ggplot2.qmd).

## Exercise 01 — A histogram, rebuilt `[Do]`

Redraw Lab 01's histogram of `score` in `ggplot2`, with bins 100 points wide and a
title and axis labels a reader could follow.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex01
#| fig-alt: "ggplot2 histogram of score per session: most sessions between 1,200 and 1,800, with a small separate group near 500 to 700."
ggplot(sessions, aes(x = score)) +
  geom_histogram(binwidth = 100, fill = "grey60", colour = "white") +
  labs(title = "Score per session", x = "Score", y = "Sessions")
```

`binwidth` is in the units of the variable: 100 points of score per bar.
:::

**Write 1 sentence:** point to the data, the mapping and the geometry in your code.

## Exercise 02 — A boxplot, rebuilt `[Do]`

Redraw Lab 01's boxplot of `score` by `archetype` in `ggplot2`, with labels.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex02
#| fig-alt: "ggplot2 boxplots of score by archetype. Sentinel's box is much taller than the others, reaching down to about 1,000."
ggplot(sessions, aes(x = archetype, y = score)) +
  geom_boxplot(fill = "grey85") +
  labs(title = "Score by archetype", x = NULL, y = "Score")
```

Group on x, numeric on y: the mapping replaces base R's `score ~ archetype`.
:::

**Write 2 sentences:** the Sentinel box is the widest. Draw the Sentinel scores as a
histogram and say what the box was hiding.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answer"}

```{r}
#| label: ex02-check
#| fig-alt: "Histogram of Sentinel scores only: a group of low scores between about 500 and 700, an empty gap, then the main group between 1,200 and 1,900."
sessions |>
  filter(archetype == "Sentinel") |>
  ggplot(aes(x = score)) +
  geom_histogram(binwidth = 100, fill = "grey60", colour = "white") +
  labs(title = "Sentinel score", x = "Score", y = "Sessions")
```

Two humps, not one wide spread: Sentinels either anchor the round or collapse
early. A boxplot summarises one distribution and cannot show two. Note that a pipe
can feed straight into `ggplot()`.
:::

## Exercise 03 — The navigation figure `[Apply]`

Build a report figure for the navigation result: the distribution of `wrong_turns`
in each build, in panels stacked so the x axes line up, labelled for a level
designer.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex03
#| fig-alt: "Faceted histograms of wrong turns per session. Build A, above, peaks around 9; Build B, below, sits around 3 to 7."
#| fig-height: 5
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```

`facet_wrap(~ build, ncol = 1)` gives one panel per build on a shared x axis, so
the shift between builds is the first thing the reader sees. The fill is set, not
mapped, because the panel labels already name the builds.
:::

**Write 2 sentences:** what does this figure show that the two means from Lab 02
did not?

## Exercise 04 — Fix this figure `[Apply]`

A teammate made this figure for the report to show how sessions ended in each
build. Find three things wrong with it, then fix them.

```{r}
#| label: ex04-bad
#| fig-alt: "A stacked bar chart of quit reasons with build stacked inside each bar, bars filled in a single salmon colour with a legend reading grey, and raw column names as axis labels."
ggplot(sessions, aes(x = quit_reason)) +
  geom_bar(aes(fill = "grey")) +
  labs(x = "quit_reason")
```

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex04
#| fig-alt: "Two panels side by side, one per build, of horizontal bars counting how sessions ended. Build A's frustrated bar is about four times as long as Build B's."
ggplot(sessions, aes(y = quit_reason)) +
  geom_bar(fill = "grey60") +
  facet_wrap(~ build) +
  labs(title = "How sessions ended, by build",
       x = "Sessions", y = NULL)
```

Three problems, three fixes:

1. **It does not split by build**, so it cannot answer the question it is for.
   `facet_wrap(~ build)` puts the builds side by side.
2. **`fill = "grey"` is inside `aes()`**, so `ggplot2` treats "grey" as data: the
   bars come out in a default colour, with a meaningless legend. A fixed colour
   goes outside `aes()`.
3. **The labels are for R, not for a reader.** There is no title and the axis says
   `quit_reason`. `y = NULL` removes a label that would only repeat the category
   names, and putting the categories on `y` makes the long names readable.
:::

**Write 1 sentence:** which of the three problems would most mislead a producer who
glanced at the figure for five seconds?

## Exercise 05 — Which figure leads? `[Decide]`

The report has room for one figure about enjoyment. Draw two candidates: a boxplot
of `enjoyment` by build, and a bar chart of `enjoyment` faceted by build. Decide
which one goes in the report.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex05a
#| fig-alt: "Two boxplots of enjoyment rating by build, almost identical, both with medians at 4."
ggplot(sessions, aes(x = build, y = enjoyment)) +
  geom_boxplot(fill = "grey85") +
  labs(title = "Enjoyment by build", x = "Build", y = "Enjoyment (1-7)")
```

```{r}
#| label: ex05b
#| fig-alt: "Two faceted bar charts of enjoyment rating, Build A above and Build B below, with very similar shapes peaking at 4."
ggplot(sessions, aes(x = enjoyment)) +
  geom_bar(fill = "grey60") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Enjoyment rating, by build", x = "Enjoyment (1-7)", y = "Sessions")
```

**One defensible answer: the faceted bar chart.** Enjoyment is a 1–7 rating, so it
has only seven possible values. The bars show exactly how many players chose each
one, and the two builds' shapes can be compared directly. The boxplot reduces a
seven-value rating to five numbers, several of which land on the same value, and
looks more precise than the data is.

**What makes it defensible.** It matches the geometry to the variable type
([rule 3](../notes/ten-rules.qmd)) and to the comparison the reader is meant to
make.

**What would make it indefensible.** Choosing the boxplot because it looks more
"statistical", or putting both in the report so the reader has to work out which
one matters.
:::

**Write 2 sentences:** write the caption for the figure you chose. State the
takeaway, not the name of the chart.

---

# Part B — Your project `[Decide]`

This is the part that matters. Allow at least half the clinic for it.

Work through these steps on your team's own playtest data. Do it in a Quarto
document in your project, so the figure is already in the format the report needs.

1. **Write the question** the figure will answer, in one sentence a designer would
   recognise.
2. **Load your data** with `read_csv()` and look at it with `glimpse()`. Check the
   types of the columns you need ([rule 3](../notes/ten-rules.qmd)).
3. **Look first, quickly.** One `hist()` or `boxplot()` to see the shape before you
   design anything.
4. **Draft the figure** in `ggplot2`: data, mapping, geometry.
5. **Split it for the comparison you mean**: facet or dodge by build, level or
   player group.
6. **Label it for a reader**: title, axis labels with units, n somewhere visible.
7. **Put it in a chunk with a caption** that states the takeaway, and render the
   document.

````markdown
```{{r}}
#| label: fig-our-finding
#| fig-cap: "The takeaway, as a sentence, with n."
#| echo: false
ggplot(our_data, aes(x = ...)) +
  geom_...() +
  labs(title = "...", x = "...", y = "...")
```
````

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answer"}
There is no single right figure: it depends on your game and your question. Check
yours against this list before you show it to your team.

- [ ] I can say the question in one sentence, and the figure makes that comparison
  the easiest thing to see.
- [ ] The geometry suits the variable types.
- [ ] Groups are faceted or dodged, not stacked, when the groups are the point.
- [ ] Every colour carries information; styling colours are set outside `aes()`.
- [ ] Title, axis labels and units are in words a reader understands; no raw
  column names.
- [ ] n is visible.
- [ ] The caption states the takeaway.
- [ ] The document renders.

Exercise 03 is a worked version of the whole process on *Tidebreak*.
:::

**Write 3–4 sentences**, using **What → Pattern → Meaning → Action**, for the
finding your figure shows:

- **What** did you measure, on whom, and how many?
- **Pattern**: what does the figure show?
- **Meaning**: what might that mean for the player?
- **Action**: what will the team change, keep or test next?

---

# Appendix — Common ggplot2 errors

::: {.callout-warning collapse="true" title="Show the error guide"}

| Message or symptom | What it usually means |
| :- | :- |
| `could not find function "ggplot"` | You have not run `library(ggplot2)` |
| ``Cannot use `+` with a single argument`` | A line starts with `+`. Move it to the end of the line above |
| `object 'build' not found` | A column name used outside `aes()` |
| A legend that names a colour, such as "grey" | A fixed colour inside `aes()`. Move it outside |
| A smooth colour gradient where you wanted two colours | You mapped a number to colour. Make it a category first |
| `stat_bin() using bins = 30` message | You did not set `binwidth`. Choose one in the variable's units |

:::
