---
title: "Lab 02 — Comparing Builds"
subtitle: "Stage 2: cleaning, t-tests, and whether a difference matters"
---

::: {.callout-note appearance="simple" icon="false" title="Get the files"}
- **Data:** [tidebreak_sessions_raw.csv](../data/tidebreak_sessions_raw.csv){download="tidebreak_sessions_raw.csv"}, [tidebreak_sessions.csv](../data/tidebreak_sessions.csv){download="tidebreak_sessions.csv"}, [tidebreak_paired.csv](../data/tidebreak_paired.csv){download="tidebreak_paired.csv"}, [tidebreak_frame_times.csv](../data/tidebreak_frame_times.csv){download="tidebreak_frame_times.csv"}, [tidebreak_survey.csv](../data/tidebreak_survey.csv){download="tidebreak_survey.csv"}. Save them in the `data/` folder of your project.
:::

# Overview

Build B added lighting and signage to guide players through *Tidebreak*. This lab
asks the question the team built it to answer:

> **Does improved environmental guidance help players navigate the level?**

Part A takes the raw export the team received and gets it into a shape that can
answer that question, keeping a written record of every decision. Part B puts a
size, an interval and a p-value on the differences, and asks the question no
function answers: does this difference matter? By the end you will have answered
the Stage 2 question, *did the change work?*, and written the recommendation.

| Function | What it does |
| :- | :- |
| `distinct()` | Drop rows that repeat an earlier row exactly |
| `is.na()` | Find missing values |
| `trimws()`, `tolower()` | Make text labels consistent |
| `\|>` | Pass the result on its left into the function on its right |
| `filter()`, `select()`, `mutate()`, `arrange()` | Keep rows, keep columns, change a column, sort |
| `count()`, `group_by()`, `summarise()` | Counts, and one row of summaries per group |
| `t.test()` | Independent-samples and paired t-tests, with confidence intervals |

# Learning outcomes

By the end of this lab you can:

- find duplicates, gaps and inconsistent labels in a raw export, and fix them in
  one readable pipeline;
- write a cleaning log that records every row dropped, why, and what it did to *n*;
- explain a p-value by reading a shuffle of the build labels;
- run an independent-samples and a paired t-test, and pull the estimate, interval
  and p-value out of the result;
- tell "no meaningful difference" apart from "cannot tell" using a confidence
  interval;
- decide whether a statistically significant difference justifies a design change,
  and write the recommendation.

# Ground rules

- Clean before you describe. Every change you make to the data goes in your
  cleaning log, including changes that remove no rows.
- Look at the plot before running the test. Every Part B exercise assumes you have
  looked at the data first.
- Type the code. Open the solution **after** your attempt, even when your attempt
  worked.
- Every exercise ends with a short written answer. Write it. The Challenge is
  judged on the writing, not the code.
- When you report a difference, give its size and its interval, not only the
  p-value.

# Setup (run this first)

```{r}
#| label: setup
library(readr)
library(dplyr)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
paired   <- read_csv("data/tidebreak_paired.csv", show_col_types = FALSE)
frames   <- read_csv("data/tidebreak_frame_times.csv", show_col_types = FALSE)
raw      <- read_csv("data/tidebreak_sessions_raw.csv",
                     trim_ws = FALSE, show_col_types = FALSE)

nrow(raw)
nrow(sessions)
nrow(paired)
```

`raw` is the export as delivered, and Part A cleans it. `trim_ws = FALSE` stops
`read_csv()` quietly stripping spaces from text values, so you see the file the way
Excel or a database import would. `sessions` has one row per player per session,
with each player on one build only. `paired` is a separate study in which every
player played both builds.

---

# Part A — Clean the export

Notes for this part: [Cleaning and Transforming](../notes/06-cleaning-and-transforming.qmd).

`raw` is the export the team actually received. Every problem in it is silent:
nothing throws an error, and every problem changes an answer. Part A finds them,
fixes them, and ends with the cleaning log that goes in your report.

## Exercise 01 — Rows exported twice `[Do]`

How many rows of `raw` are exact copies of an earlier row? Then list any
`session_id` that appears more than once.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex01
nrow(raw) - nrow(distinct(raw))

raw |> count(session_id) |> filter(n > 1)
```

`distinct()` drops exact repeats, so the difference in rows is the number of
copies. Counting `session_id` and keeping counts above 1 finds every key that
appears twice.
:::

**Write 2 sentences:** the number of repeated keys equals the number of exact
copies. What would it mean if there were more repeated keys, and why could you not
simply delete the extra rows?

## Exercise 02 — Where the gaps are `[Do]`

Count the missing values in every column of `raw` at once.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex02
colSums(is.na(raw))
```

`is.na()` on a data frame gives a TRUE/FALSE table the same shape as the data, and
`colSums()` counts the TRUEs in each column.
:::

**Write 1 sentence:** which column's gaps would you investigate first, and why
that one?

## Exercise 03 — Clean it in one pipeline `[Apply]`

Build `clean` from `raw` in one pipeline: drop the duplicated rows, then fix
`cohort` so that it has exactly three values. Count `cohort` before and after to
prove the fix worked, and note the number of rows.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex03
raw |> count(cohort)

clean <- raw |>
  distinct() |>
  mutate(cohort = tolower(trimws(cohort)))

clean |> count(cohort)
nrow(clean)
```

`trimws()` removes the stray spaces and `tolower()` the inconsistent capitals.
Counting before and after is the only proof that the fix caught every variant.
:::

**Write 2 sentences:** the relabelling removed no rows. Why does it still belong in
the cleaning log?

## Exercise 04 — Who is missing? `[Apply]`

List every row of `clean` with a blank `session_minutes` **or** a blank `deaths`.
Keep only `session_id`, `build`, `quit_reason`, `session_minutes` and `deaths`, and
sort by `build` then `quit_reason`.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex04
clean |>
  filter(is.na(session_minutes) | is.na(deaths)) |>
  select(session_id, build, quit_reason, session_minutes, deaths) |>
  arrange(build, quit_reason)
```

`|` means "or", so a row is kept if either value is missing. Sorting puts rows
with the same build and outcome next to each other, which is what makes the
pattern visible.
:::

**Write 2 sentences:** describe the pattern in the blank `session_minutes` rows,
and say how it differs from the pattern in the blank `deaths` rows.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answer"}

```{r}
#| label: ex04-check
#| echo: false
blank_min <- clean[is.na(clean$session_minutes), ]
blank_dth <- clean[is.na(clean$deaths), ]
stopifnot(all(blank_min$build == "A"), all(blank_min$quit_reason == "Frustrated"),
          length(unique(blank_dth$quit_reason)) > 1)
```

All `r nrow(blank_min)` blank session lengths are Build A frustration quits. The
`r nrow(blank_dth)` blank death counts are spread across
`r length(unique(blank_dth$build))` builds and
`r length(unique(blank_dth$quit_reason))` different outcomes. One set looks like
something in the game failed to log; the other looks like noise.
:::

## Exercise 05 — Write the cleaning log `[Decide]`

Decide what to do with the rows that have gaps, then write the cleaning log for
`clean`: every step, the number of rows after it, and why. Your log is the record
a reader of the report would use to check your *n*
([rule 6](../notes/ten-rules.qmd)).

Before you decide, compare each build's median `session_minutes` with the blanks
left out, against the median of the Build A frustration quits that *were*
recorded.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex05
clean |>
  group_by(build) |>
  summarise(median_minutes = median(session_minutes, na.rm = TRUE),
            missing        = sum(is.na(session_minutes)))

clean |>
  filter(build == "A", quit_reason == "Frustrated") |>
  summarise(recorded = sum(!is.na(session_minutes)),
            median_minutes = median(session_minutes, na.rm = TRUE))

sum(complete.cases(clean))   # what "drop every incomplete row" would leave
```

```{r}
#| label: ex05-numbers
#| include: false
recorded_a_frustrated <- clean$session_minutes[clean$build == "A" &
                                                 clean$quit_reason == "Frustrated"]
```

**One defensible answer: keep every row, and handle the gaps per analysis.**

| Step | Rows after | Why |
| :- | -: | :- |
| Raw export | `r nrow(raw)` | As delivered |
| Remove exact duplicates | `r nrow(clean)` | `r sum(duplicated(raw))` sessions exported twice; no key duplicates with different values |
| Standardise `cohort` | `r nrow(clean)` | `r sum(!raw$cohort[!duplicated(raw)] %in% c("Novice", "Regular", "Veteran"))` labels had stray spaces or inconsistent capitals |
| Keep blank `session_minutes` | `r nrow(clean)` | All `r nrow(blank_min)` are Build A frustration quits. Summaries of session length exclude them and say so |
| Keep blank `deaths` | `r nrow(clean)` | `r nrow(blank_dth)` rows, no pattern. Excluded from deaths summaries only |

**What makes it defensible.** The blank session lengths are not random. Recorded
Build A frustration quits have a median of
`r round(median(recorded_a_frustrated, na.rm = TRUE), 1)` minutes, so the
missing ones are almost certainly short, and Build A's median of
`r round(median(clean$session_minutes[clean$build == "A"], na.rm = TRUE), 1)`
minutes is an overestimate. Keeping the rows keeps their `quit_reason`, which is
complete and carries the finding. The log says what every summary of session
length is missing, so a reader can judge it.

**What would make it indefensible.** Dropping every incomplete row by reflex, which
silently takes *n* from `r nrow(clean)` to `r sum(complete.cases(clean))` and
removes the frustration quits the comparison is about. Or filling the blanks with 0
or with Build A's average, which invents session lengths and hides the logging
fault.
:::

**Write 3 sentences:** the blank session lengths are also a bug in Build A's
telemetry. Who should hear about it, and what goes in the bug report?

::: {.callout-important title="Before Part B"}
Part B uses `sessions` from the Setup block, not your `clean`. It is the same
sessions, exported again after the logging fault was fixed: the same rows, with the
gaps you found filled.

```{r}
#| label: part-a-bridge
identical(clean$session_id, sessions$session_id)
sum(is.na(sessions$session_minutes))
sum(is.na(sessions$deaths))
```

Everything Part B compares is therefore free of the bias you found in Exercise 05.
In a real playtest you would rarely get that second export. Your cleaning log is
what you would have instead.
:::

---

# Part B — Is the difference real, and does it matter?

Notes for this part: [From Sample to Population](../notes/07-sample-to-population.qmd)
and [Comparing Groups](../notes/08-comparing-groups.qmd). Report every comparison
by answering the six questions from Comparing Groups, in order: what happened, how
large, how uncertain, could it be chance, does it matter, what should the team do.

## Exercise 06 — Describe before you test `[Do]`

Produce one row per build with the number of sessions and the mean, median,
standard deviation and IQR of `wrong_turns`. Then draw side-by-side boxplots.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex06
sessions |>
  group_by(build) |>
  summarise(n      = n(),
            mean   = mean(wrong_turns),
            median = median(wrong_turns),
            sd     = sd(wrong_turns),
            iqr    = IQR(wrong_turns))
```

```{r}
#| label: ex06b
#| fig-alt: "Two boxplots of wrong turns per session. Build A's box sits around 8 to 10; Build B's is lower, around 3 to 7.5."
boxplot(wrong_turns ~ build, data = sessions, col = "grey85",
        main = "Wrong turns per session", ylab = "Wrong turns")
```

`n()` counts the rows in each group. Always report it: a difference means little
without the number of players behind it.
:::

**Write 1 sentence:** Build B's standard deviation is larger than Build A's. What
does that mean for a player?

## Exercise 07 — The independent-samples t-test `[Do]`

Run the t-test comparing `wrong_turns` between builds. Store the result, then pull
out the two means, the confidence interval and the p-value separately.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex07
result <- t.test(wrong_turns ~ build, data = sessions)
result

result$estimate
result$conf.int
result$p.value
```

Each player played one build only, so this is the independent-samples test.
Pulling out the parts with `$` is how you get the numbers into a sentence with
inline R rather than by copying.
:::

**Write 2 sentences:** is the confidence interval for Build A minus Build B, or
Build B minus Build A? How can you tell from the output?

## Exercise 08 — Could it be chance? Read the shuffle `[Do]`

If the signposting made no difference, the build label on each session would be
arbitrary. This code shuffles the labels 10,000 times, recomputes the gap in mean
wrong turns each time, and counts how many shuffled gaps are at least as large as
the real one. You do not need to be able to write it. Run it, and compare the
result with Exercise 07's p-value.

```{r}
#| label: ex08
#| fig-alt: "Bell-shaped histogram of shuffled gaps in mean wrong turns, centred on zero and running from about minus 2 to plus 2, with a red line at the observed gap of 3.4 far to the right of every bar."
gap <- function(y, g) mean(y[g == "A"]) - mean(y[g == "B"])

observed <- gap(sessions$wrong_turns, sessions$build)
set.seed(1)
shuffled <- replicate(10000, gap(sessions$wrong_turns, sample(sessions$build)))

observed
max(abs(shuffled))
mean(abs(shuffled) >= abs(observed))   # the shuffle's p-value

hist(shuffled, breaks = 40, col = "grey85", border = "white",
     xlim = range(c(shuffled, observed)),
     main = "Gaps from shuffled build labels",
     xlab = "Difference in mean wrong turns (A - B)")
abline(v = observed, col = "firebrick", lwd = 3)
```

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}
The real gap is `r round(observed, 1)` wrong turns; the largest of ten thousand
shuffled gaps is `r round(max(abs(shuffled)), 2)`. No shuffle came close, so the
shuffle's p-value is below 1 in 10,000, which agrees with the t-test's tiny
p-value. The shuffle shows what a p-value *is*: the share of a world without a
difference that looks at least as extreme as your data.
:::

**Write 2 sentences:** what does one shuffle represent, and why does the
proportion of extreme shuffles answer "could this be chance?"

## Exercise 09 — The same players, both builds `[Apply]`

`paired` records `r nrow(paired)` players who each played both builds. Run the
paired t-test on `wrong_turns_a` and `wrong_turns_b`, then run the unpaired test on
the same two columns. Compare the widths of the two confidence intervals.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex09
paired_test   <- t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)
unpaired_test <- t.test(paired$wrong_turns_a, paired$wrong_turns_b)

paired_test$conf.int
unpaired_test$conf.int

diff(paired_test$conf.int)
diff(unpaired_test$conf.int)
```

Same estimate, but the paired interval is
`r round(diff(paired_test$conf.int), 1)` wrong turns wide against
`r round(diff(unpaired_test$conf.int), 1)`. The paired test compares each player
with themselves, so differences in player skill cancel out. `diff()` of an
interval is its width.
:::

**Write 2 sentences:** the paired file does not record which build each player
played first. Why does that matter, and what would you change about the next
paired study?

## Exercise 10 — No difference, or cannot tell? `[Apply]`

Run the t-test on `enjoyment` by build and on `items_collected` by build. Neither
is significant. For which one can you say there is **no meaningful difference**,
and for which can you only say **we cannot tell**?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex10
enjoy <- t.test(enjoyment ~ build, data = sessions)
items <- t.test(items_collected ~ build, data = sessions)

enjoy$conf.int
items$conf.int

sessions |> summarise(mean_items = mean(items_collected))
```

Read each interval against the size of the variable. Enjoyment's interval runs
from `r round(enjoy$conf.int[1], 2)` to `r round(enjoy$conf.int[2], 2)` on a
seven-point scale: whatever the true difference is, it is too small to matter. That
is **no meaningful difference**. The items interval runs from
`r round(items$conf.int[1], 2)` to `r round(items$conf.int[2], 2)` items, when a
typical player collects about
`r round(mean(sessions$items_collected))`. Build B's players could be collecting
up to `r sprintf("%.1f", -items$conf.int[1])` more items each, or up to
`r sprintf("%.1f", items$conf.int[2])` fewer. That is **cannot tell**.

"Cannot tell" means this playtest was too small to see a difference of that size.
A bigger playtest, or a paired one, would narrow the interval enough to decide
([Testing in Depth](../going-further/testing-in-depth.qmd#two-ways-to-be-wrong)).
:::

**Write 2 sentences:** why is "not significant" a weaker statement than "the
confidence interval is narrow and sits around zero"?

## Exercise 11 — "Players play longer in Build B" `[Decide]`

```{r}
#| label: ex11-setup
#| include: false
tt_min  <- t.test(session_minutes ~ build, data = sessions)
not_fr  <- sessions[sessions$quit_reason != "Frustrated", ]
tt_nf   <- t.test(session_minutes ~ build, data = not_fr)
fr_med  <- tapply(sessions$session_minutes[sessions$quit_reason == "Frustrated"],
                  sessions$build[sessions$quit_reason == "Frustrated"], median)
stopifnot(tt_min$p.value < 0.05, diff(tt_min$estimate) > 0,
          tt_nf$p.value > 0.2, abs(diff(fr_med)) < 1)
```

`t.test(session_minutes ~ build)` gives `r if (tt_min$p.value < 0.001) "p < 0.001" else paste("p =", round(tt_min$p.value, 3))`,
and Build B's sessions are about `r round(unname(diff(tt_min$estimate)))` minutes
longer on average. The producer wants to put "players play
`r round(unname(diff(tt_min$estimate)))` minutes longer in Build B" in the pitch. Do you
let them? Test before you decide: compare session length between the builds for
players who did **not** quit in frustration.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex11
not_frustrated <- sessions |> filter(quit_reason != "Frustrated")
t.test(session_minutes ~ build, data = not_frustrated)

sessions |>
  filter(quit_reason == "Frustrated") |>
  group_by(build) |>
  summarise(n = n(), median_minutes = median(session_minutes))
```

**One defensible answer: no, because the claim describes the wrong thing.** Among
players who did not quit in frustration, Build A averaged
`r round(tt_nf$estimate[[1]], 1)` minutes and Build B
`r round(tt_nf$estimate[[2]], 1)` (p = `r round(tt_nf$p.value, 2)`). Frustration
quits end at a median of about `r round(mean(fr_med), 1)` minutes in both builds.
Build B does not make anyone play longer. It makes fewer people quit early, and
that pulls its average up. The honest pitch line is "fewer players quit in
frustration", which is also the stronger claim.

**What makes it defensible.** It checks what is driving the difference before
describing it, and it replaces the claim with one the data supports.

**What would make it indefensible.** Approving the line because the p-value is
below 0.05. The difference is real; the explanation attached to it is not.
:::

**Write 3 sentences:** a significant difference in a mean can come from every
player shifting a little or from a few players shifting a lot. How do you tell
which, and why does it change the design conclusion?

## Exercise 12 — Which number goes in the report? `[Decide]`

You have three estimates of how much the signposting reduced wrong turns: the
session-level test from Exercise 07, the paired study from Exercise 09, and a
player-level test that averages each player's `r unique(as.vector(table(sessions$player_id)))`
sessions first. Run the player-level test, then decide which result leads the
report.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex12
per_player <- sessions |>
  group_by(player_id, build) |>
  summarise(wrong_turns = mean(wrong_turns), .groups = "drop")
player_test <- t.test(wrong_turns ~ build, data = per_player)

player_test$conf.int
result$conf.int
paired_test$conf.int
```

**One defensible answer: lead with the player-level result, and cite the paired
study as support.** The session-level test treats
`r nrow(sessions)` sessions as independent when they come from
`r nrow(per_player)` players, so its interval is narrower than the data justifies.
The player-level interval, `r round(player_test$conf.int[1], 1)` to
`r round(player_test$conf.int[2], 1)` wrong turns, is honest about that. The paired
study, a different design with different players, points the same way, which is
worth more than any single p-value.

**What makes it defensible.** It picks the analysis whose assumptions match how
the data was collected, and it uses the other two as corroboration rather than
choosing between them.

**What would make it indefensible.** Picking whichever result has the smallest
p-value or the largest effect. That is choosing the answer, not the analysis.
:::

**Write 2 sentences:** why is agreement between two differently designed studies
stronger evidence than a very small p-value from one?

---

# Challenge — Build B was meant to be an improvement. Was it? `[Decide]`

The producer asks one question before approving Build B for the next milestone:
**was it an improvement?**

Build the evidence you would take into that meeting. Use more than one measure:
navigation, whether players stayed and finished, whether anything got worse, and
performance. `frames` is loaded in the Setup block. For each measure, keep the
evidence separate: what the plot shows, what the summaries say, what the test
says, and whether the difference is big enough to matter.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: challenge
# 1. Navigation: the question Build B was built to answer
t.test(wrong_turns ~ build, data = sessions)

# 2. Did players stay and finish? Rates per build, then a test for each
outcomes <- sessions |>
  group_by(build) |>
  summarise(n          = n(),
            frustrated = sum(quit_reason == "Frustrated"),
            completed  = sum(completed))
outcomes
prop.test(outcomes$frustrated, outcomes$n)
prop.test(outcomes$completed, outcomes$n)

# 3. Did anything get worse?
t.test(enjoyment ~ build, data = sessions)
t.test(score ~ build, data = sessions)

# 4. Performance, which nobody asked about
frames |>
  group_by(build) |>
  summarise(fps = 1000 / mean(ms), p99 = quantile(ms, 0.99))
```

```{r}
#| label: challenge-plot
#| fig-alt: "Boxplots of frame time for each build and zone. All six boxes sit near 17 ms; only Build B's Corridor has a column of outliers between about 34 and 55 ms."
boxplot(ms ~ build + zone, data = frames, col = "grey85", las = 2,
        main = "Frame times by build and zone", ylab = "Frame time (ms)", xlab = "")
```

```{r}
#| label: challenge-numbers
#| echo: false
wt   <- t.test(wrong_turns ~ build, data = sessions)
enj  <- t.test(enjoyment ~ build, data = sessions)
scr  <- t.test(score ~ build, data = sessions)
comp <- round(100 * tapply(sessions$completed, sessions$build, mean))
fr   <- table(sessions$build, sessions$quit_reason)[, "Frustrated"]
p99  <- round(tapply(frames$ms, frames$build, quantile, 0.99), 1)
fps  <- round(1000 / tapply(frames$ms, frames$build, mean))
pt_fr   <- prop.test(fr, table(sessions$build))
pt_comp <- prop.test(tapply(sessions$completed, sessions$build, sum), table(sessions$build))
slow <- frames[frames$ms > 1000 / 30, ]
p_fmt <- function(p) if (p < 0.001) "p < 0.001" else paste("p =", round(p, 2))

stopifnot(wt$p.value < 0.001, enj$p.value > 0.5, scr$p.value > 0.5,
          all(slow$build == "B"), all(slow$zone == "Corridor"),
          comp[["B"]] > comp[["A"]], fr[["B"]] < fr[["A"]],
          fps[["A"]] == fps[["B"]], p99[["B"]] > p99[["A"]],
          pt_fr$p.value < 0.01, pt_comp$p.value < 0.01)
```

| Measure | Descriptive | Statistical | Practical |
| :- | :- | :- | :- |
| Wrong turns | Mean `r round(wt$estimate[[1]], 1)` in A, `r round(wt$estimate[[2]], 1)` in B | 95% CI `r round(wt$conf.int[1], 1)` to `r round(wt$conf.int[2], 1)` fewer, `r p_fmt(wt$p.value)` | Large: players get lost far less |
| Frustration quits | `r fr[["A"]]` in A, `r fr[["B"]]` in B | `prop.test()`: `r p_fmt(pt_fr$p.value)` | Large: the main way Build A lost players |
| Completed the test | `r comp[["A"]]`% in A, `r comp[["B"]]`% in B | `prop.test()`: `r p_fmt(pt_comp$p.value)` | Large |
| Enjoyment | Means `r round(enj$estimate[[1]], 1)` and `r round(enj$estimate[[2]], 1)` | 95% CI `r round(enj$conf.int[1], 2)` to `r round(enj$conf.int[2], 2)`, `r p_fmt(enj$p.value)` | No cost to enjoyment |
| Score | Means `r round(scr$estimate[[1]])` and `r round(scr$estimate[[2]])` | `r p_fmt(scr$p.value)` | No change |
| Frame time | Both about `r fps[["A"]]` FPS; 99th percentile `r p99[["A"]]` ms in A, `r p99[["B"]]` ms in B | Not a comparison of means | Every one of the `r nrow(slow)` frames slower than 30 FPS is Build B, in the Corridor |

**The headline, in the six steps.**

1. **What happened?** Build B's players took fewer wrong turns, fewer quit in
   frustration, and more finished.
2. **How large?** `r round(wt$estimate[[1]] - wt$estimate[[2]], 1)` fewer wrong
   turns per session; `r fr[["A"]]` frustration quits against `r fr[["B"]]`.
3. **How uncertain?** The wrong-turns interval runs from
   `r round(wt$conf.int[1], 1)` to `r round(wt$conf.int[2], 1)` fewer.
4. **Could it be chance?** Not plausibly: `r p_fmt(wt$p.value)`, and the same for
   frustration and completion.
5. **Does it matter?** Yes. Even the smallest plausible reduction is noticeable,
   and none of it cost enjoyment or score. But the frame-time plot shows a hitch
   in Build B's Corridor that the average frame rate cannot see.
6. **What should the team do?** Keep the signposting; fix the Corridor hitch
   before Build B goes forward.

**What would make it indefensible.** Answering from wrong turns alone, because that
is the measure the build was designed to move. Or quoting average FPS as proof
that performance was unchanged. Both answers are supported by a real number, and
both miss what a player would actually experience.
:::

**Write 4 sentences**, using **What → Pattern → Meaning → Action**:

- **What** did you measure?
- **Pattern**: what does the evidence show, across more than one measure? Give
  the size and the interval for the main one.
- **Meaning**: what does that mean for the player?
- **Action**: what exactly do you recommend to the producer?

---

# Extension — Bring the survey in `[Apply]`

For anyone who finishes the Challenge with time to spare. It uses one new
function, `left_join()`, which you will meet properly in Stage 4.

```{r}
#| label: ext-setup
#| include: false
survey     <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)
joined_chk <- left_join(survey, select(sessions, session_id, build), by = "session_id")
dis_chk    <- tapply(joined_chk$objective_clear <= 2, joined_chk$build, sum)
pt_chk     <- prop.test(dis_chk, table(joined_chk$build))
stopifnot(nrow(joined_chk) == nrow(survey), !anyNA(joined_chk$build),
          dis_chk[["A"]] > dis_chk[["B"]], pt_chk$p.value < 0.01)
```

The survey does not record the build. Load `data/tidebreak_survey.csv`, then use
`left_join()` to add each session's `build` to it, matching on `session_id`. Then
compare the share of responses in each build that **disagreed** (1 or 2) with "I
always knew what my objective was", using `prop.test()`.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ext
survey <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)

joined <- survey |>
  left_join(select(sessions, session_id, build), by = "session_id")

nrow(joined)
sum(is.na(joined$build))

disagreed <- joined |>
  group_by(build) |>
  summarise(disagreed = sum(objective_clear <= 2), n = n())
disagreed
prop.test(disagreed$disagreed, disagreed$n)
```

`left_join()` keeps every survey row and adds `build` from the matching session;
checking that the row count is unchanged and that no `build` is `NA` confirms every
survey found its session. `r dis_chk[["A"]]` of `r table(joined_chk$build)[["A"]]`
Build A responses disagreed, against `r dis_chk[["B"]]` of
`r table(joined_chk$build)[["B"]]` in Build B
(`r if (pt_chk$p.value < 0.001) "p < 0.001" else paste("p =", round(pt_chk$p.value, 3))`).
:::

**Write 2 sentences:** the telemetry says Build B players took fewer wrong turns;
what does this survey result add, and what kind of evidence is it?

---

# Appendix — Reading `t.test()` output

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show the annotated output"}

```{r}
#| label: appendix-ttest
t.test(wrong_turns ~ build, data = sessions)
```

| Line | What it tells you |
| :- | :- |
| `Welch Two Sample t-test` | The independent-samples test, not assuming equal spreads. The right default |
| `t = ..., df = ...` | The test statistic and its degrees of freedom. Report them if asked; do not interpret them |
| `p-value = ...` | How surprising the data would be if the builds did not differ. `e-11` means "times 10 to the minus 11": tiny |
| `alternative hypothesis: ...` | Names the direction: the difference is *group A minus group B* |
| `95 percent confidence interval` | The plausible range for the difference, A minus B |
| `sample estimates` | The two group means |

:::

# Appendix — Common errors

::: {.callout-warning collapse="true" title="Show the error guide"}

| Message | What it usually means |
| :- | :- |
| `grouping factor must have exactly 2 levels` | `t.test(y ~ group)` needs exactly two groups. Filter to two first |
| `not enough 'x' observations` | A group is empty, usually a misspelled level such as `"a"` for `"A"` |
| `could not find function "filter"` (or `distinct`, `count`) | You have not run `library(dplyr)` |
| `'x' and 'n' must have the same length` | `prop.test()` needs one count and one total per group, in the same order |
| `object 'clean' not found` | You skipped Exercise 03, or ran Part A out of order |
| `unexpected '>'` on a line with `\|>` | Your R is older than 4.1. Update R |

:::
