---
title: "Lab 02 — Comparing Builds"
subtitle: "ggplot2 figures, t-tests, and whether a difference matters"
---

# Overview

Build B added lighting and signage to guide players through *Tidebreak*. This lab
asks the question the team built it to answer:

> **Does improved environmental guidance help players navigate the level?**

Part A rebuilds your Week 1 base R plots in `ggplot2`, then goes further than base
R comfortably can. Part B puts a size, an interval and a p-value on the
differences, and asks the question no function answers: does this difference
matter?

| Function | What it does |
| :- | :- |
| `ggplot()`, `aes()` | Start a figure: the data and the mapping |
| `geom_histogram()`, `geom_bar()`, `geom_boxplot()` | Draw it |
| `facet_wrap()` | One panel per group |
| `labs()` | Title and axis labels |
| `group_by()`, `summarise()` | One row of summaries per group |
| `t.test()` | Independent-samples and paired t-tests, with confidence intervals |
| `aggregate()` | Collapse several rows per player into one |

# Learning outcomes

By the end of this lab you can:

- build histograms, bar charts and boxplots in `ggplot2`, split by group with
  `fill` or facets, and label them for a reader;
- choose the figure that makes a comparison easiest to see, and defend the choice;
- run an independent-samples and a paired t-test, and pull the estimate, interval
  and p-value out of the result;
- tell "no meaningful difference" apart from "cannot tell" using a confidence
  interval;
- decide whether a statistically significant difference justifies a design change.

# Ground rules

- Draw before you test. Every Part B exercise assumes you have looked at the data
  first.
- Type the code. Open the solution **after** your attempt, even when your attempt
  worked.
- Every exercise ends with a short written answer. Write it. The Challenge is
  assessed on the writing, not the code.
- When you report a difference, give its size and its interval, not only the
  p-value.

# Setup (run this first)

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
paired   <- read_csv("data/tidebreak_paired.csv", show_col_types = FALSE)
frames   <- read_csv("data/tidebreak_frame_times.csv", show_col_types = FALSE)

nrow(sessions)
nrow(paired)
```

`sessions` has one row per player per session, with each player on one build only.
`paired` is a separate study in which every player played both builds.

---

# Part A — Figures with ggplot2

## Exercise 01 — A histogram, rebuilt `[Do]`

Redraw Lab 01's histogram of `score` in `ggplot2`, with bins 100 points wide.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex01
ggplot(sessions, aes(x = score)) +
  geom_histogram(binwidth = 100)
```

`binwidth` is in the units of the variable, so 100 means each bar covers 100
points of score.
:::

**Write 1 sentence:** point to the data, the mapping and the geometry in your code.

## Exercise 02 — A boxplot, rebuilt `[Do]`

Redraw Lab 01's boxplot of `score` by `archetype` in `ggplot2`.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex02
ggplot(sessions, aes(x = archetype, y = score)) +
  geom_boxplot()
```

The base R formula `score ~ archetype` becomes a mapping: the group on `x`, the
numeric variable on `y`.
:::

**Write 1 sentence:** what did base R make you type that `ggplot2` did not, or the
other way round?

## Exercise 03 — Quit reasons, by build, for a reader `[Apply]`

Draw the count of each `quit_reason`, with the two builds side by side in
different colours. Give it a title, drop the redundant x-axis label, and label the
y axis and the legend properly.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex03
ggplot(sessions, aes(x = quit_reason, fill = build)) +
  geom_bar(position = "dodge") +
  labs(title = "How sessions ended, by build",
       x = NULL, y = "Sessions", fill = "Build")
```

`fill = build` is **mapped**, so it changes with the data and earns a legend.
`position = "dodge"` stops the bars stacking, so each build's bar starts at zero
and the two can be compared directly.
:::

**Write 2 sentences:** remove `position = "dodge"` and look again. Which comparison
does the stacked version make harder, and why?

## Exercise 04 — Where Build A loses players `[Apply]`

Draw a histogram of `session_minutes` with one panel per build, stacked so the x
axes line up, with bins one minute wide. Where does Build A lose players that
Build B keeps?

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex04
#| fig-height: 5
ggplot(sessions, aes(x = session_minutes)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Session length, by build", x = "Minutes played", y = "Sessions")
```

`facet_wrap(~ build, ncol = 1)` is the line base R has no short equivalent for.
The fill is **set**, not mapped, because the panel labels already say which build
is which.
:::

**Write 2 sentences:** describe where Build A's extra early exits happen, and
connect them to one bar in Exercise 03.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: ex04-check
#| echo: false
early <- function(b) sum(sessions$build == b &
                         sessions$session_minutes > 11 & sessions$session_minutes <= 13)
```

Build A has a cluster of sessions ending between minutes 11 and 13:
`r early("A")` of them, against `r early("B")` in Build B. They line up with the
Frustrated bar in Exercise 03, where Build A has
`r sum(sessions$build == "A" & sessions$quit_reason == "Frustrated")` sessions to
Build B's `r sum(sessions$build == "B" & sessions$quit_reason == "Frustrated")`.
:::

## Exercise 05 — One figure for the level designer `[Decide]`

The level designer wants one figure that answers: *did the signposting reduce wrong
turns?* Draw at least two candidates for `wrong_turns` by build, pick one, and
justify it.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex05-facet
#| fig-height: 4
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```

```{r}
#| label: ex05-stacked
ggplot(sessions, aes(x = wrong_turns, fill = build)) +
  geom_histogram(binwidth = 1)
```

**One defensible answer: the faceted histogram.** The shared x axis makes the shift
obvious at a glance, and it shows something a boxplot flattens: Build B is wider as
well as lower, so the signposting helped most players and some hardly at all. A
side-by-side boxplot is also defensible for a reader who is used to them.

**What makes it defensible.** The figure makes the comparison the question asks
about the easiest thing on the page to see, and hides nothing the designer would
want to know.

**What would make it indefensible.** The stacked histogram: Build A's bars sit on
top of Build B's, so Build A's shape cannot be read. Or a bar chart of the two
means, which throws away the spread entirely.
:::

**Write 2 sentences:** what does your chosen figure show that the two means alone
would hide?

---

# Part B — Is the difference real, and does it matter?

## Exercise 06 — Describe before you test `[Do]`

Using `group_by()` and `summarise()`, produce one row per build with the number of
sessions and the mean, median, standard deviation and IQR of `wrong_turns`.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex06
sessions %>%
  group_by(build) %>%
  summarise(n      = n(),
            mean   = mean(wrong_turns),
            median = median(wrong_turns),
            sd     = sd(wrong_turns),
            iqr    = IQR(wrong_turns))
```

`n()` counts the rows in each group. Always report it: a difference means little
without the number of players behind it.
:::

**Write 1 sentence:** Build B's standard deviation is larger than Build A's. What
does that mean for a player?

## Exercise 07 — The independent-samples t-test `[Do]`

Run the t-test comparing `wrong_turns` between builds. Store the result, then pull
out the two means, the confidence interval and the p-value separately.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex07
result <- t.test(wrong_turns ~ build, data = sessions)
result

result$estimate
result$conf.int
result$p.value
```

Each player played one build only, so this is the independent-samples test.
Pulling out the parts with `$` is how you get the numbers into a sentence with
inline R rather than by copying.
:::

**Write 2 sentences:** is the confidence interval for Build A minus Build B, or
Build B minus Build A? How can you tell from the output?

## Exercise 08 — The same players, both builds `[Apply]`

`paired` records `r nrow(paired)` players who each played both builds. Run the
paired t-test on `wrong_turns_a` and `wrong_turns_b`, then run the unpaired test on
the same two columns. Compare the widths of the two confidence intervals.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex08
paired_test   <- t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)
unpaired_test <- t.test(paired$wrong_turns_a, paired$wrong_turns_b)

paired_test$conf.int
unpaired_test$conf.int

diff(paired_test$conf.int)
diff(unpaired_test$conf.int)
```

Same estimate, but the paired interval is
`r round(diff(paired_test$conf.int), 1)` wrong turns wide against
`r round(diff(unpaired_test$conf.int), 1)`. The paired test compares each player
with themselves, so differences in player skill cancel out.
:::

**Write 2 sentences:** the paired file does not record which build each player
played first. Why does that matter, and what would you change about the next
paired study?

## Exercise 09 — No difference, or cannot tell? `[Apply]`

Run the t-test on `enjoyment` by build and on `items_collected` by build. Neither
is significant. For which one can you say there is **no meaningful difference**,
and for which can you only say **we cannot tell**?

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex09
enjoy <- t.test(enjoyment ~ build, data = sessions)
items <- t.test(items_collected ~ build, data = sessions)

enjoy$conf.int
items$conf.int

tapply(sessions$items_collected, sessions$build, mean)
```

Read each interval against the size of the variable. Enjoyment's interval runs
from `r round(enjoy$conf.int[1], 2)` to `r round(enjoy$conf.int[2], 2)` on a
seven-point scale: whatever the true difference is, it is too small to matter. That
is **no meaningful difference**. The items interval runs from
`r round(items$conf.int[1], 2)` to `r round(items$conf.int[2], 2)` items, when a
typical player collects about
`r round(mean(sessions$items_collected))`. Build B's players could be collecting
up to `r sprintf("%.1f", -items$conf.int[1])` more items each, or up to
`r sprintf("%.1f", items$conf.int[2])` fewer. That is **cannot tell**.
:::

**Write 2 sentences:** why is "not significant" a weaker statement than "the
confidence interval is narrow and sits around zero"?

## Exercise 10 — "Players play longer in Build B" `[Decide]`

```{r}
#| label: ex10-setup
#| include: false
tt_min  <- t.test(session_minutes ~ build, data = sessions)
not_fr  <- sessions[sessions$quit_reason != "Frustrated", ]
tt_nf   <- t.test(session_minutes ~ build, data = not_fr)
fr_med  <- tapply(sessions$session_minutes[sessions$quit_reason == "Frustrated"],
                  sessions$build[sessions$quit_reason == "Frustrated"], median)
stopifnot(tt_min$p.value < 0.05, diff(tt_min$estimate) > 0,
          tt_nf$p.value > 0.2, abs(diff(fr_med)) < 1)
```

`t.test(session_minutes ~ build)` gives `r if (tt_min$p.value < 0.001) "p < 0.001" else paste("p =", round(tt_min$p.value, 3))`,
and Build B's sessions are about `r round(unname(diff(tt_min$estimate)))` minutes
longer on average. The producer wants to put "players play
`r round(unname(diff(tt_min$estimate)))` minutes longer in Build B" in the pitch. Do you
let them? Test before you decide: compare session length between the builds for
players who did **not** quit in frustration.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex10
not_frustrated <- sessions[sessions$quit_reason != "Frustrated", ]
t.test(session_minutes ~ build, data = not_frustrated)

frustrated <- sessions[sessions$quit_reason == "Frustrated", ]
tapply(frustrated$session_minutes, frustrated$build, median)
```

**One defensible answer: no, because the claim describes the wrong thing.** Among
players who did not quit in frustration, Build A averaged
`r round(tt_nf$estimate[[1]], 1)` minutes and Build B
`r round(tt_nf$estimate[[2]], 1)` (p = `r round(tt_nf$p.value, 2)`). Frustration
quits end at a median of about `r round(mean(fr_med), 1)` minutes in both builds.
Build B does not make anyone play longer. It makes fewer people quit early, and
that pulls its average up. The honest pitch line is "fewer players quit in
frustration", which is also the stronger claim.

**What makes it defensible.** It checks what is driving the difference before
describing it, and it replaces the claim with one the data supports.

**What would make it indefensible.** Approving the line because the p-value is
below 0.05. The difference is real; the explanation attached to it is not.
:::

**Write 3 sentences:** a significant difference in a mean can come from every
player shifting a little or from a few players shifting a lot. How do you tell
which, and why does it change the design conclusion?

## Exercise 11 — Which number goes in the report? `[Decide]`

You have three estimates of how much the signposting reduced wrong turns: the
session-level test from Exercise 07, the paired study from Exercise 08, and a
player-level test that averages each player's `r unique(as.vector(table(sessions$player_id)))`
sessions first. Run the player-level test, then decide which result leads the
report.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex11
per_player <- aggregate(wrong_turns ~ player_id + build, data = sessions, FUN = mean)
player_test <- t.test(wrong_turns ~ build, data = per_player)

player_test$conf.int
result$conf.int
paired_test$conf.int
```

**One defensible answer: lead with the player-level result, and cite the paired
study as support.** The session-level test treats
`r nrow(sessions)` sessions as independent when they come from
`r nrow(per_player)` players, so its interval is narrower than the data justifies.
The player-level interval, `r round(player_test$conf.int[1], 1)` to
`r round(player_test$conf.int[2], 1)` wrong turns, is honest about that. The paired
study, a different design with different players, points the same way, which is
worth more than any single p-value.

**What makes it defensible.** It picks the analysis whose assumptions match how
the data was collected, and it uses the other two as corroboration rather than
choosing between them.

**What would make it indefensible.** Picking whichever result has the smallest
p-value or the largest effect. That is choosing the answer, not the analysis.
:::

**Write 2 sentences:** why is agreement between two differently designed studies
stronger evidence than a very small p-value from one?

---

# Challenge — Build B was meant to be an improvement. Was it? `[Decide]`

The producer asks one question before approving Build B for the next milestone:
**was it an improvement?**

Build the evidence you would take into that meeting. Use more than one measure,
and keep four kinds of evidence separate: what the figures show, what the
descriptive statistics say, what the tests say, and whether each difference is big
enough to matter. `frames` is loaded in the Setup block.

::: {.callout-important collapse="true" title="One defensible answer"}

```{r}
#| label: challenge
# 1. Navigation: the question Build B was built to answer
t.test(wrong_turns ~ build, data = sessions)

# 2. Did players stay and finish?
table(sessions$build, sessions$quit_reason)
tapply(sessions$completed, sessions$build, mean)

# 3. Did anything get worse?
t.test(enjoyment ~ build, data = sessions)
t.test(score ~ build, data = sessions)

# 4. Performance, which nobody asked about
frames %>%
  group_by(build) %>%
  summarise(fps = 1000 / mean(ms), p99 = quantile(ms, 0.99))

ggplot(frames, aes(x = ms)) +
  geom_histogram(binwidth = 1, fill = "grey60") +
  facet_grid(build ~ zone) +
  labs(title = "Frame times by build and zone", x = "Frame time (ms)", y = "Frames")
```

```{r}
#| label: challenge-numbers
#| echo: false
wt   <- t.test(wrong_turns ~ build, data = sessions)
enj  <- t.test(enjoyment ~ build, data = sessions)
scr  <- t.test(score ~ build, data = sessions)
comp <- round(100 * tapply(sessions$completed, sessions$build, mean))
fr   <- table(sessions$build, sessions$quit_reason)[, "Frustrated"]
p99  <- round(tapply(frames$ms, frames$build, quantile, 0.99), 1)
fps  <- round(1000 / tapply(frames$ms, frames$build, mean))
slow <- frames[frames$ms > 1000 / 30, ]
p_fmt <- function(p) if (p < 0.001) "p < 0.001" else paste("p =", round(p, 2))

stopifnot(wt$p.value < 0.001, enj$p.value > 0.5, scr$p.value > 0.5,
          all(slow$build == "B"), all(slow$zone == "Corridor"),
          comp[["B"]] > comp[["A"]], fr[["B"]] < fr[["A"]],
          fps[["A"]] == fps[["B"]], p99[["B"]] > p99[["A"]])
```

| Measure | Visual | Descriptive | Statistical | Practical |
| :- | :- | :- | :- | :- |
| Wrong turns | Build B's histogram sits clearly to the left | Mean `r round(wt$estimate[[1]], 1)` in A, `r round(wt$estimate[[2]], 1)` in B | 95% CI `r round(wt$conf.int[1], 1)` to `r round(wt$conf.int[2], 1)` fewer, `r p_fmt(wt$p.value)` | Large: players get lost far less |
| Frustration quits | Build A's bar is far taller | `r fr[["A"]]` in A, `r fr[["B"]]` in B | Not tested this week | Large: the main way Build A lost players |
| Completed the test | — | `r comp[["A"]]`% in A, `r comp[["B"]]`% in B | Not tested this week | Large |
| Enjoyment | Panels look alike | Means `r round(enj$estimate[[1]], 1)` and `r round(enj$estimate[[2]], 1)` | 95% CI `r round(enj$conf.int[1], 2)` to `r round(enj$conf.int[2], 2)`, `r p_fmt(enj$p.value)` | No cost to enjoyment |
| Score | Boxes overlap | Means `r round(scr$estimate[[1]])` and `r round(scr$estimate[[2]])` | `r p_fmt(scr$p.value)` | No change |
| Frame time | A scatter of tiny bars far to the right in Build B's Corridor panel, and nowhere else. Easy to miss, even on the right figure | Both about `r fps[["A"]]` FPS; 99th percentile `r p99[["A"]]` ms in A, `r p99[["B"]]` ms in B | Not a comparison of means | Every one of the `r nrow(slow)` frames slower than 30 FPS is Build B, in the Corridor |

**So the defensible answer is: yes for navigation, and not yet for release.** The
signposting did what it was built to do. Players got lost less, fewer quit in
frustration, more finished, and none of it cost enjoyment or score. But Build B
also introduced a hitch in the Corridor that the frame-rate average cannot see.
Keep the signposting; fix the hitch before Build B goes forward.

**What would make it indefensible.** Answering from wrong turns alone, because that
is the measure the build was designed to move. Or quoting average FPS as proof
that performance was unchanged. Both answers are supported by a real number, and
both miss what a player would actually experience.
:::

**Write 4 sentences**, using **What → Pattern → Meaning → Action**:

- **What** did you measure?
- **Pattern** — what does the evidence show, across more than one measure?
- **Meaning** — what does that mean for the build?
- **Action** — what exactly do you recommend to the producer?

---

# Appendix — Reading `t.test()` output

::: {.callout-note collapse="true" title="Show the annotated output"}

```{r}
#| label: appendix-ttest
t.test(wrong_turns ~ build, data = sessions)
```

| Line | What it tells you |
| :- | :- |
| `Welch Two Sample t-test` | The independent-samples test, not assuming equal spreads. The right default |
| `t = ..., df = ...` | The test statistic and its degrees of freedom. Report them if asked; do not interpret them |
| `p-value = ...` | How surprising the data would be if the builds did not differ. `e-11` means "times 10 to the minus 11": tiny |
| `alternative hypothesis: ...` | Names the direction: the difference is *group A minus group B* |
| `95 percent confidence interval` | The plausible range for the difference, A minus B |
| `sample estimates` | The two group means |

:::

# Appendix — Common errors

::: {.callout-warning collapse="true" title="Show the error guide"}

| Message | What it usually means |
| :- | :- |
| `could not find function "ggplot"` | You have not run `library(ggplot2)` |
| ``Cannot use `+` with a single argument`` | A line starts with `+`. Move it to the end of the line above |
| `object 'build' not found` | A column name used outside `aes()` |
| ``stat_bin() requires a continuous x aesthetic`` | A histogram of a categorical column. Use `geom_bar()` |
| `grouping factor must have exactly 2 levels` | `t.test(y ~ group)` needs exactly two groups. Filter to two first |
| `not enough 'x' observations` | A group is empty, usually a misspelled level such as `"a"` for `"A"` |
| `could not find function "%>%"` | You have not run `library(dplyr)` |

:::
