---
title: "Lab 03 — Complete Playtest Analysis"
subtitle: "Figures, relationships, reshaping, and the report you will revise from"
---

# Overview

This lab ends the strand by answering one question from start to finish:

> **Does environmental guidance improve navigation without reducing enjoyment?**

Part A builds the figures and relationships the answer needs, in `ggplot2`, and
finds a correlation that means the opposite of what it seems to. Part B reshapes
the survey, then assembles everything into a nine-part report in Quarto.

**The report is the point.** It is the artefact you revise from for the Week 8
practical. Every exercise in Part B adds a piece of it, and you should leave the
lab with a `report.qmd` that renders.

| Function | What it does |
| :- | :- |
| `ggplot()`, `aes()`, `geom_*()` | Build a figure from data, mapping and geometry |
| `facet_wrap()`, `colour =` | Split a figure by group |
| `geom_smooth(method = "lm")` | Add a straight-line trend to a scatterplot |
| `labs()` | Title, axis and legend labels |
| `cor()` | Strength and direction of a straight-line relationship, −1 to +1 |
| `pivot_longer()` | Reshape wide columns into long rows |
| `table()`, `prop.table()` | Count, then turn counts into proportions |
| `left_join()` | Add columns from another table, matched on a key |
| `quarto render` | Turn a `.qmd` report into HTML |

# Learning outcomes

By the end of this lab you can:

- rebuild a base R plot in `ggplot2`, and build a figure base R cannot easily
  produce;
- measure a relationship with `cor()`, draw it with `geom_smooth()`, and check it
  within groups before trusting it;
- identify a confounding variable and say what it does to a conclusion;
- reshape a survey with `pivot_longer()` and summarise every item at once;
- assemble a nine-part playtest report in Quarto, with every number computed, and
  render it;
- write a finding as What → Pattern → Meaning → Action, including one that
  contradicts an earlier claim.

# Ground rules

- Choose, then justify. Several exercises have more than one sensible method; the
  written answer is where you say why you chose yours.
- Type the code. Open the solution **after** your attempt, even when your attempt
  worked.
- Keep every figure and number you produce. Part B puts them into the report.
- No number goes into the report by hand. If it is a number, it is inline R.
- If `library(tidyr)` or `library(ggplot2)` fails, run
  `install.packages(c("dplyr", "tidyr", "ggplot2"))` once in the Console.

# Setup (run this first)

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
survey   <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)

nrow(sessions)
dim(survey)
```

`sessions` is the cleaned session data from Lab 02: one row per player per session,
each player on one build. `survey` is the post-session questionnaire: one row per
session, eight agreement items rated 1 to 5.

---

# Part A — Visualise and relate

Notes for this part: [Visualising with ggplot2](../notes/visualising-with-ggplot2.qmd).

## Exercise 01 — A histogram, rebuilt `[Do]`

Redraw Lab 01's histogram of `score` in `ggplot2`, with bins 100 points wide and a
title and axis labels a reader could follow.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex01
ggplot(sessions, aes(x = score)) +
  geom_histogram(binwidth = 100, fill = "grey60", colour = "white") +
  labs(title = "Score per session", x = "Score", y = "Sessions")
```
:::

**Write 1 sentence:** point to the data, the mapping and the geometry in your code.

## Exercise 02 — A boxplot, rebuilt `[Do]`

Redraw Lab 01's boxplot of `score` by `archetype` in `ggplot2`, with labels.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex02
ggplot(sessions, aes(x = archetype, y = score)) +
  geom_boxplot() +
  labs(title = "Score by archetype", x = NULL, y = "Score")
```
:::

**Write 2 sentences:** the Sentinel box is the widest. Draw the Sentinel scores as a
histogram and say what the box was hiding.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: ex02-check
ggplot(filter(sessions, archetype == "Sentinel"), aes(x = score)) +
  geom_histogram(binwidth = 100, fill = "grey60", colour = "white") +
  labs(title = "Sentinel score", x = "Score", y = "Sessions")
```

Two humps, not one wide spread: Sentinels either anchor the round or collapse
early. A boxplot summarises one distribution and cannot show two.
:::

## Exercise 03 — The navigation figure `[Apply]`

Build the figure for part 4 of your report: the distribution of `wrong_turns` in
each build, stacked so the x axes line up, labelled for a level designer. This is
the figure base R cannot produce in one short call.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex03
#| fig-height: 5
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```

`facet_wrap(~ build, ncol = 1)` gives one panel per build on a shared x axis, so
the shift between builds is the first thing the reader sees. The fill is set, not
mapped, because the panel labels already name the builds.
:::

**Write 2 sentences:** what does this figure show that the two means from Lab 02
did not?

## Exercise 04 — What goes with score? `[Do]`

Use `cor()` to measure how strongly `score` is related to `awareness`, and to
`shots_fired`.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex04
cor(sessions$awareness, sessions$score)
cor(sessions$shots_fired, sessions$score)
```

`cor()` gives a number from −1 to +1. Near 0 means no straight-line relationship;
the sign gives the direction.
:::

**Write 2 sentences:** a designer wants to reward players who fire more shots,
"because they are the ones scoring". What do you tell them?

## Exercise 05 — The trade-off `[Apply]`

Draw `survivability` against `mobility` as a scatterplot with a straight-line
trend, and measure the relationship with `cor()`.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex05
ggplot(sessions, aes(x = mobility, y = survivability)) +
  geom_point(colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  labs(title = "Mobility against survivability",
       x = "Mobility rating", y = "Survivability rating")

cor(sessions$mobility, sessions$survivability)
```

`geom_smooth(method = "lm")` draws the straight line that best fits the points;
`se = FALSE` leaves out the shaded band. The line and `cor()` describe the same
relationship, one as a picture and one as a number.
:::

**Write 1 sentence:** describe the relationship in words a balance designer would
use.

## Exercise 06 — Faster players are more fragile? `[Decide]`

```{r}
#| label: ex06-setup
#| include: false
r_all  <- cor(sessions$mobility, sessions$survivability)
r_arch <- sapply(split(sessions, sessions$archetype),
                 function(d) cor(d$mobility, d$survivability))
stopifnot(r_all < -0.5, all(r_arch > 0))
```

The balance designer reads Exercise 05 and proposes a patch: *"Faster players are
more fragile, so when a Scout player gets faster, cut their health to match."*
Redraw the scatterplot with each archetype in its own colour and its own trend
line, then decide whether the patch follows from the data.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex06
ggplot(sessions, aes(x = mobility, y = survivability, colour = archetype)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Mobility against survivability, by archetype",
       x = "Mobility rating", y = "Survivability rating", colour = "Archetype")

sessions |>
  group_by(archetype) |>
  summarise(r = cor(mobility, survivability))
```

**One defensible answer: no, the patch reads the data backwards.** Across all
players the correlation is `r round(r_all, 2)`. Within every archetype it is
positive, from `r round(min(r_arch), 2)` to `r round(max(r_arch), 2)`. The
downward trend is the gap *between* archetypes: Scouts are built fast and fragile,
Sentinels slow and tough. Within any one archetype, better players are better at
both. **Archetype is the confounding variable**: it drives both measures, and
ignoring it reverses the direction of the relationship.

**What makes it defensible.** It checks the relationship inside the groups the
design is made of before drawing a conclusion about individuals, and it says what
the overall correlation *does* show: the archetypes are balanced against each other
as intended.

**What would make it indefensible.** Applying the patch because the correlation is
strong. A strong correlation across groups says nothing about what happens to one
player within a group.
:::

**Write 3 sentences:** name the confound, say what it does to the conclusion, and
say which part of the report this belongs in.

---

# Part B — Reshape and report

Notes for this part: [Interpreting and Reporting Evidence](../notes/reporting-evidence.qmd).

Every exercise in this part produces a piece of `report.qmd`. Create the file now,
in the project folder next to `data/`, and add to it as you go.

## Exercise 07 — Wide to long `[Do]`

Reshape `survey` so that it has one row per session per item, with columns
`session_id`, `item` and `rating`. Call it `long`, and check its size.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex07
long <- survey |>
  pivot_longer(-session_id, names_to = "item", values_to = "rating")

dim(long)
head(long, 3)
```

`-session_id` stacks every column except the key. `r nrow(survey)` sessions ×
8 items = `r nrow(long)` rows.
:::

**Write 1 sentence:** why can you not `group_by()` the items in the wide `survey`?

## Exercise 08 — Every item, both builds `[Apply]`

Use `table()` and `prop.table()` to show the distribution of ratings for every
item. Then join each session's `build` onto `long` and find the proportion of
players agreeing (4 or 5) with each item in each build. Which item moved most
between the builds?

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex08
round(prop.table(table(long$item, long$rating), margin = 1), 2)

agree_by_build <- long |>
  left_join(select(sessions, session_id, build), by = "session_id") |>
  group_by(item, build) |>
  summarise(agree = mean(rating >= 4), .groups = "drop") |>
  pivot_wider(names_from = build, values_from = agree) |>
  mutate(change = B - A) |>
  arrange(desc(abs(change)))

agree_by_build
```

`margin = 1` makes each row of the table sum to 1. `mean(rating >= 4)` is the
proportion agreeing, and `pivot_wider()`, the reverse of `pivot_longer()`, puts the
two builds side by side so the change is one subtraction.
:::

**Write 2 sentences:** `objective_clear` moved most. Which item would you have
*expected* to move most, given what Build B changed, and did it?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: ex08-check
#| echo: false
nav <- agree_by_build[agree_by_build$item == "level_navigable", ]
obj <- agree_by_build[agree_by_build$item == "objective_clear", ]
stopifnot(agree_by_build$item[1] == "objective_clear", nav$change < obj$change)
```

`level_navigable`, since Build B added guidance to help players navigate. It moved
from `r round(100 * nav$A)`% agreeing to `r round(100 * nav$B)`%, against
`objective_clear`'s `r round(100 * obj$A)`% to `r round(100 * obj$B)`%. Build B
cut wrong turns, but what players noticed was knowing where they were meant to go.
That is worth a sentence in the report.
:::

## Exercise 09 — "Without reducing enjoyment" `[Decide]`

The research question has two halves. Test the second: did Build B reduce
enjoyment? Use `enjoyment` from `sessions` (1–7) and `would_play_again` from the
survey, and decide what the report may claim.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex09
enjoy <- t.test(enjoyment ~ build, data = sessions)
enjoy$estimate
enjoy$conf.int

agree_by_build[agree_by_build$item == "would_play_again", ]
```

```{r}
#| label: ex09-check
#| include: false
wpa <- agree_by_build[agree_by_build$item == "would_play_again", ]
stopifnot(enjoy$p.value > 0.5, abs(wpa$change) < 0.1)
```

**One defensible answer: the report may claim no meaningful reduction.** Mean
enjoyment was `r round(enjoy$estimate[[1]], 2)` in Build A and
`r round(enjoy$estimate[[2]], 2)` in Build B. The 95% confidence interval for the
difference runs from `r round(enjoy$conf.int[1], 2)` to
`r round(enjoy$conf.int[2], 2)` on a seven-point scale, so any true difference is
small. Agreement with `would_play_again` was `r round(100 * wpa$A)`% and
`r round(100 * wpa$B)`%. Two different measures agree.

**What makes it defensible.** It rests on an interval, not on "not significant",
and on two measures rather than one. It claims *no meaningful reduction*, which the
evidence supports, rather than *no effect*, which no test can show.

**What would make it indefensible.** "The p-value is above 0.05, so Build B has no
effect on enjoyment." Or "Build B is just as enjoyable" from one measure with no
interval.
:::

**Write 2 sentences:** both measures were taken straight after one session. What
would you need to know before claiming Build B is as enjoyable over a week of play?

## Exercise 10 — Start the report, and render it `[Apply]`

In `report.qmd`, write:

- a YAML header with a title;
- the nine part headings from
  [Interpreting and Reporting Evidence](../notes/reporting-evidence.qmd);
- a setup chunk, hidden with `echo: false`, that loads the libraries and both CSVs;
- under part 1, the research question;
- under part 2, one sentence giving the number of sessions and players, with both
  numbers as inline R;
- under part 4, the figure from Exercise 03.

Then render it from the Terminal with `quarto render report.qmd` and open
`report.html`.

::: {.callout-tip collapse="true" title="Solution"}

````markdown
---
title: "Tidebreak Build B: Playtest Report"
---

```{{r}}
#| echo: false
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
survey   <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)
```

## 1. Research question

Does environmental guidance improve navigation without reducing enjoyment?

## 2. Experimental design

We analysed `r knitr::inline_expr("nrow(sessions)")` sessions from
`r knitr::inline_expr("length(unique(sessions$player_id))")` players, each of whom
played one build only.

## 3. Descriptive evidence

## 4. Visual evidence

```{{r}}
#| echo: false
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```

## 5. Statistical evidence

## 6. Practical significance

## 7. Qualitative evidence

## 8. Limitations

## 9. Design conclusion
````

The two inline expressions produce these values when the report renders:

```{r}
#| label: ex10
nrow(sessions)
length(unique(sessions$player_id))
```

`echo: false` hides the code and keeps its output, which is what a reader of a
report wants. Because the numbers are inline, the report corrects itself the next
time the data changes.
:::

**Write 2 sentences:** change one value in your copy of the CSV, render again, and
describe what changed in the report and what you did not have to do.

::: {.callout-note collapse="true" title="What you should see"}
The terminal ends with `Output created: report.html`. The page shows nine headings,
the sentence under part 2 reads "`r nrow(sessions)` sessions from
`r length(unique(sessions$player_id))` players", and the two-panel histogram sits
under part 4 with no code above it.

If the render stops with an error, the message names the line. The most common
causes are a data path written as `../data/...`, a chunk that uses a package it
did not load, and a missing closing ```` ``` ````.
:::

## Exercise 11 — Limitations and the conclusion `[Decide]`

Write parts 8 and 9 of the report. Part 8 lists what could make your conclusion
wrong. Part 9 is one finding in **What → Pattern → Meaning → Action** form that
answers the research question; a finding that does not name an action is not
finished ([rule 10](../notes/ten-rules.qmd)). Every number is inline R.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex11
wt <- t.test(wrong_turns ~ build, data = sessions)
wt$estimate
wt$conf.int

table(sessions$player_id)[1:4]
table(sessions$cohort, sessions$build)
```

```{r}
#| label: ex11-check
#| include: false
per_player <- unique(as.vector(table(sessions$player_id)))
n_players  <- length(unique(sessions$player_id))
stopifnot(length(per_player) == 1, wt$conf.int[1] > 0,
          round(enjoy$estimate[[1]], 2) == round(enjoy$estimate[[2]], 2))
```

**One defensible part 8.**

- Each of the `r n_players` players contributed `r per_player` sessions, so the
  `r nrow(sessions)` sessions are not independent, and the session-level test is
  more confident than the data justifies.
- Every measure was taken after one session. Nothing here shows whether the
  difference lasts.
- Cohorts were not perfectly balanced between the builds; the cohort table is in
  the report so the reader can judge.
- Comments record how a session felt, and do not agree with the wrong-turn counts
  (see the Challenge).
- Strong correlations in this data can reverse within archetypes (Exercise 06), so
  no relationship here is reported as a cause without checking within groups.

**One defensible part 9.**

> **What.** Wrong turns per session in `r nrow(sessions)` sessions, and enjoyment
> on a seven-point scale, compared between Build A and Build B.
>
> **Pattern.** Build B players averaged `r round(wt$estimate[[2]], 1)` wrong turns
> against Build A's `r round(wt$estimate[[1]], 1)`, a reduction of between
> `r round(wt$conf.int[1], 1)` and `r round(wt$conf.int[2], 1)` per session at 95%
> confidence. Enjoyment was unchanged, with a mean of
> `r round(enjoy$estimate[[1]], 2)` in both builds.
>
> **Meaning.** The guidance works, and costs nothing players noticed.
>
> **Action.** Keep Build B's guidance and carry it into the next build. Retest
> with an on-screen objective added, since objective clarity was the item that
> moved most.

**What makes it defensible.** Part 8 names limitations specific to this data, not
a generic "the sample was small". Part 9's Action follows from its Meaning, and its
Pattern gives a size and an interval.

**What would make it indefensible.** A part 8 that lists nothing the reader could
check, or a part 9 whose Action ("improve the level") no one could start on.
:::

**Write 3 sentences:** which limitation in your part 8 would most change your
Action if it turned out to matter, and how would you test it?

---

# Challenge — A finding that contradicts your own report `[Decide]`

A report that agrees with itself everywhere has usually stopped looking. Add a
finding to part 7 or part 8 of your report that contradicts a claim made earlier in
the same report, and write it so the reader is better informed, not confused.

::: {.callout-important collapse="true" title="One defensible answer"}

```{r}
#| label: challenge
themed <- sessions |>
  mutate(theme = case_when(
    grepl("lost|circles|meant to go|objective", comment) ~ "Lost or unclear",
    grepl("route|exit|signposting", comment)             ~ "Found the way",
    grepl("Died|hard", comment)                          ~ "Difficulty",
    TRUE                                                 ~ "General"
  ))

table(themed$theme, themed$build)
table(themed$theme, enjoyed = themed$enjoyment >= 5)
```

```{r}
#| label: challenge-check
#| include: false
lost <- table(themed$theme, themed$build)["Lost or unclear", ]
stopifnot(lost[["B"]] > lost[["A"]],
          all(themed$enjoyment[themed$theme == "Lost or unclear"] < 5))
```

**The contradiction.** Part 5 says Build B players took fewer wrong turns. Yet
`r lost[["B"]]` Build B players left a comment about being lost or unclear, against
`r lost[["A"]]` in Build A.

> **What.** One free-text comment per session, `r nrow(sessions)` in all,
> categorised into four themes.
>
> **Pattern.** "Lost or unclear" comments were more common in Build B
> (`r lost[["B"]]`) than in Build A (`r lost[["A"]]`), the opposite of the
> wrong-turn counts. Split by enjoyment instead of build, every one of those
> comments came from a player who rated enjoyment below 5.
>
> **Meaning.** The comments track how the session felt, not where the player went.
> They do not undermine the wrong-turn result, but they show that players who
> had a poor session describe it as getting lost whichever build they played.
>
> **Action.** Report the wrong-turn counts as the navigation evidence and the
> comments as evidence about mood. In the next playtest, ask a specific navigation
> question in the survey rather than relying on free text.

Other contradictions in this data that would serve: `level_navigable` barely moved
while wrong turns fell (Exercise 08); the mobility–survivability trade-off reverses
within archetypes (Exercise 06).

**What would make it indefensible.** Leaving the contradiction out because it
spoils the story, or including it without saying which evidence the reader should
trust and why.
:::

**Write 4 sentences**, using **What → Pattern → Meaning → Action**:

- **What** did you measure?
- **Pattern** — what does the evidence show, and what does it contradict?
- **Meaning** — which piece of evidence should the reader trust, and why?
- **Action** — what changes in the next playtest because of it?

---

# Output

Hand in two files: `report.qmd` and the `report.html` it renders to.

- All nine parts present, each with at least one sentence.
- Parts 3 to 7 each contain at least one computed number, table or figure from this
  lab or Lab 02.
- No number in the prose typed by hand.
- Part 9 written as What → Pattern → Meaning → Action, and the Challenge finding in
  part 7 or 8.
- The report renders with `quarto render report.qmd` and no errors.

This is the report you revise from for the Week 8 practical. Anything you could not
finish in the lab, finish before then: an unfinished part is easier to complete now
than to reconstruct later.

---

# Appendix — Common errors

::: {.callout-warning collapse="true" title="Show the error guide"}

| Message | What it usually means |
| :- | :- |
| `could not find function "ggplot"` | You have not run `library(ggplot2)` |
| `could not find function "pivot_longer"` | You have not run `library(tidyr)` |
| ``Cannot use `+` with a single argument`` | A line starts with `+`. Move it to the end of the line above |
| `object 'build' not found` | A column name used outside `aes()` |
| `'data/tidebreak_sessions.csv' does not exist` | `report.qmd` is not in the project folder next to `data/` |
| `object 'agree_by_build' not found` when rendering | The report uses an object you made in the Console. Put the code in a chunk in the report |

:::
