---
title: "Lab 02 — Comparing Builds"
subtitle: "Cleaning, transforming, t-tests, and whether a difference matters"
---

# Overview

Build B added lighting and signage to guide players through *Tidebreak*. This lab
asks the question the team built it to answer:

> **Does improved environmental guidance help players navigate the level?**

Part A takes the raw export the team received and gets it into a shape that can
answer that question, first in base R and then with the pipe, keeping a written
record of every decision. Part B puts a size, an interval and a p-value on the
differences, and asks the question no function answers: does this difference
matter?

| Function | What it does |
| :- | :- |
| `duplicated()` | TRUE for any row that repeats an earlier row |
| `is.na()`, `complete.cases()` | Find missing values, and rows with none |
| `trimws()`, `tolower()` | Make text labels consistent |
| `|>` | Pass the result on its left into the function on its right |
| `filter()`, `select()`, `mutate()`, `arrange()` | Keep rows, keep columns, change a column, sort |
| `group_by()`, `summarise()` | One row of summaries per group |
| `t.test()` | Independent-samples and paired t-tests, with confidence intervals |
| `aggregate()` | Collapse several rows per player into one |

# Learning outcomes

By the end of this lab you can:

- find duplicates, gaps and inconsistent labels in a raw export, and fix them in
  base R and again with the pipe;
- use `filter()`, `select()`, `mutate()` and `arrange()` to answer a question about
  the data in one readable pipeline;
- write a cleaning log that records every row dropped, why, and what it did to *n*;
- run an independent-samples and a paired t-test, and pull the estimate, interval
  and p-value out of the result;
- tell "no meaningful difference" apart from "cannot tell" using a confidence
  interval;
- decide whether a statistically significant difference justifies a design change.

# Ground rules

- Clean before you describe. Every change you make to the data goes in your
  cleaning log, including changes that remove no rows.
- Look before you test. Every Part B exercise assumes you have looked at the data
  first.
- Part A uses the pipe `|>`. Some Part B solutions use `%>%`, the older tidyverse
  pipe, which does the same job.
- Type the code. Open the solution **after** your attempt, even when your attempt
  worked.
- Every exercise ends with a short written answer. Write it. The Challenge is
  assessed on the writing, not the code.
- When you report a difference, give its size and its interval, not only the
  p-value.

# Setup (run this first)

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
paired   <- read_csv("data/tidebreak_paired.csv", show_col_types = FALSE)
frames   <- read_csv("data/tidebreak_frame_times.csv", show_col_types = FALSE)
raw      <- read_csv("data/tidebreak_sessions_raw.csv",
                     trim_ws = FALSE, show_col_types = FALSE)

nrow(raw)
nrow(sessions)
nrow(paired)
```

`raw` is the export as delivered, and Part A cleans it. `trim_ws = FALSE` stops
`read_csv()` quietly stripping spaces from text values, so you see the file the way
Excel or a database import would. `sessions` has one row per player per session,
with each player on one build only. `paired` is a separate study in which every
player played both builds.

---

# Part A — Clean and transform

Notes for this part: [Cleaning and Transforming](../notes/cleaning-and-transforming.qmd).

`raw` is the export the team actually received. Every problem in it is silent:
nothing throws an error, and every problem changes an answer. Part A finds them,
fixes them in base R, fixes them again with the pipe, and ends with the cleaning
log that goes in your report.

If you did the cleaning extension to Lab 01, Exercises 01–03 are revision. Do them
quickly and spend the time on Exercise 06.

## Exercise 01 — Rows exported twice `[Do]`

Count the rows of `raw` that are exact copies of an earlier row. Then count the
rows whose `session_id` has appeared before.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex01
sum(duplicated(raw))
sum(duplicated(raw$session_id))
```

The first counts rows that repeat every column; the second counts rows that repeat
only the key.
:::

**Write 2 sentences:** the two counts are equal. What would it mean if the second
were larger, and why could you not simply delete the extra rows?

## Exercise 02 — Where the gaps are `[Do]`

Count the missing values in every column of `raw` at once.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex02
colSums(is.na(raw))
```

`is.na()` on a data frame gives a TRUE/FALSE table the same shape as the data, and
`colSums()` counts the TRUEs in each column.
:::

**Write 1 sentence:** which column's gaps would you investigate first, and why
that one?

## Exercise 03 — Clean it in base R `[Apply]`

Make a copy of `raw` called `clean` with the duplicated rows removed. Then fix
`cohort` so that it has exactly three values, and check that it does. Note the
number of rows after each step.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex03
clean <- raw[!duplicated(raw), ]
nrow(clean)

table(clean$cohort)
clean$cohort <- tolower(trimws(clean$cohort))
table(clean$cohort)
```

`!duplicated()` keeps the first copy of every row. `trimws()` removes the stray
spaces and `tolower()` the inconsistent capitals, and `table()` before and after
is the only proof that the fix caught every variant.
:::

**Write 2 sentences:** the relabelling removed no rows. Why does it still belong in
the cleaning log?

## Exercise 04 — The same cleaning, with the pipe `[Apply]`

Rebuild `clean` as `clean_pipe` in one pipeline starting from `raw`: `filter()` out
the duplicated rows, then `mutate()` the `cohort` column. Confirm the two versions
are identical.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex04
clean_pipe <- raw |>
  filter(!duplicated(raw)) |>
  mutate(cohort = tolower(trimws(cohort)))

identical(clean, clean_pipe)
```

`filter()` keeps the rows where its condition is TRUE, which is what `[ , ]` did
in Exercise 03, and `mutate()` replaces the column in place. Each line is one step,
read top to bottom in the order it happens.
:::

**Write 2 sentences:** which version would you rather hand to a teammate who has to
re-run it on next week's export, and why?

## Exercise 05 — Who is missing? `[Apply]`

Using the pipe, list every row of `clean` with a blank `session_minutes` **or** a
blank `deaths`. Keep only `session_id`, `build`, `quit_reason`, `session_minutes`
and `deaths`, and sort by `build` then `quit_reason`.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex05
clean |>
  filter(is.na(session_minutes) | is.na(deaths)) |>
  select(session_id, build, quit_reason, session_minutes, deaths) |>
  arrange(build, quit_reason)
```

`|` means "or", so a row is kept if either value is missing. Sorting puts rows
with the same build and outcome next to each other, which is what makes the
pattern visible.
:::

**Write 2 sentences:** describe the pattern in the blank `session_minutes` rows,
and say how it differs from the pattern in the blank `deaths` rows.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: ex05-check
#| echo: false
blank_min <- clean[is.na(clean$session_minutes), ]
blank_dth <- clean[is.na(clean$deaths), ]
stopifnot(all(blank_min$build == "A"), all(blank_min$quit_reason == "Frustrated"),
          length(unique(blank_dth$quit_reason)) > 1)
```

All `r nrow(blank_min)` blank session lengths are Build A frustration quits. The
`r nrow(blank_dth)` blank death counts are spread across
`r length(unique(blank_dth$build))` builds and
`r length(unique(blank_dth$quit_reason))` different outcomes. One set looks like
something in the game failed to log; the other looks like noise.
:::

## Exercise 06 — Write the cleaning log `[Decide]`

Decide what to do with the rows that have gaps, then write the cleaning log for
`clean`: every step, the number of rows after it, and why. Your log is the record
a reader of the report would use to check your *n*
([rule 6](../notes/ten-rules.qmd)).

Before you decide, compare Build A's median `session_minutes` with the blanks left
out against the median of the Build A frustration quits that *were* recorded.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex06
tapply(clean$session_minutes, clean$build, median, na.rm = TRUE)

recorded_a_frustrated <- clean |>
  filter(build == "A", quit_reason == "Frustrated", !is.na(session_minutes))
median(recorded_a_frustrated$session_minutes)

nrow(raw)
nrow(clean)
nrow(clean[complete.cases(clean), ])
```

**One defensible answer: keep every row, and handle the gaps per analysis.**

| Step | Rows after | Why |
| :- | -: | :- |
| Raw export | `r nrow(raw)` | As delivered |
| Remove exact duplicates | `r nrow(clean)` | `r sum(duplicated(raw))` sessions exported twice; no key duplicates with different values |
| Standardise `cohort` | `r nrow(clean)` | `r sum(!raw$cohort[!duplicated(raw)] %in% c("Novice", "Regular", "Veteran"))` labels had stray spaces or inconsistent capitals |
| Keep blank `session_minutes` | `r nrow(clean)` | All `r nrow(blank_min)` are Build A frustration quits. Summaries of session length exclude them and say so |
| Keep blank `deaths` | `r nrow(clean)` | `r nrow(blank_dth)` rows, no pattern. Excluded from deaths summaries only |

**What makes it defensible.** The blank session lengths are not random. Recorded
Build A frustration quits have a median of
`r round(median(recorded_a_frustrated$session_minutes), 1)` minutes, so the
missing ones are almost certainly short, and Build A's median of
`r round(median(clean$session_minutes[clean$build == "A"], na.rm = TRUE), 1)`
minutes is an overestimate. Keeping the rows keeps their `quit_reason`, which is
complete and carries the finding. The log says what every summary of session
length is missing, so a reader can judge it.

**What would make it indefensible.** Running `complete.cases()` by reflex, which
silently takes *n* from `r nrow(clean)` to
`r nrow(clean[complete.cases(clean), ])` and removes the frustration quits the
comparison is about. Or filling the blanks with 0 or with Build A's average, which
invents session lengths and hides the logging fault.
:::

**Write 3 sentences:** the blank session lengths are also a bug in Build A's
telemetry. Who should hear about it, and what goes in the bug report?

::: {.callout-note title="Before Part B"}

Part B uses `sessions` from the Setup block, not your `clean`. It is the same
sessions, exported again after the logging fault was fixed: the same rows, with the
gaps you found filled.

```{r}
#| label: part-a-bridge
identical(clean$session_id, sessions$session_id)
sum(is.na(sessions$session_minutes))
sum(is.na(sessions$deaths))
```

Everything Part B compares is therefore free of the bias you found in Exercise 06.
In a real playtest you would rarely get that second export. Your cleaning log is
what you would have instead.
:::

---

# Part B — Is the difference real, and does it matter?

Notes for this part: [Comparing Playtest Conditions](../notes/comparing-conditions.qmd).

## Exercise 07 — Describe before you test `[Do]`

Using `group_by()` and `summarise()`, produce one row per build with the number of
sessions and the mean, median, standard deviation and IQR of `wrong_turns`.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex07
sessions %>%
  group_by(build) %>%
  summarise(n      = n(),
            mean   = mean(wrong_turns),
            median = median(wrong_turns),
            sd     = sd(wrong_turns),
            iqr    = IQR(wrong_turns))
```

`n()` counts the rows in each group. Always report it: a difference means little
without the number of players behind it.
:::

**Write 1 sentence:** Build B's standard deviation is larger than Build A's. What
does that mean for a player?

## Exercise 08 — The independent-samples t-test `[Do]`

Run the t-test comparing `wrong_turns` between builds. Store the result, then pull
out the two means, the confidence interval and the p-value separately.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex08
result <- t.test(wrong_turns ~ build, data = sessions)
result

result$estimate
result$conf.int
result$p.value
```

Each player played one build only, so this is the independent-samples test.
Pulling out the parts with `$` is how you get the numbers into a sentence with
inline R rather than by copying.
:::

**Write 2 sentences:** is the confidence interval for Build A minus Build B, or
Build B minus Build A? How can you tell from the output?

## Exercise 09 — The same players, both builds `[Apply]`

`paired` records `r nrow(paired)` players who each played both builds. Run the
paired t-test on `wrong_turns_a` and `wrong_turns_b`, then run the unpaired test on
the same two columns. Compare the widths of the two confidence intervals.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex09
paired_test   <- t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)
unpaired_test <- t.test(paired$wrong_turns_a, paired$wrong_turns_b)

paired_test$conf.int
unpaired_test$conf.int

diff(paired_test$conf.int)
diff(unpaired_test$conf.int)
```

Same estimate, but the paired interval is
`r round(diff(paired_test$conf.int), 1)` wrong turns wide against
`r round(diff(unpaired_test$conf.int), 1)`. The paired test compares each player
with themselves, so differences in player skill cancel out.
:::

**Write 2 sentences:** the paired file does not record which build each player
played first. Why does that matter, and what would you change about the next
paired study?

## Exercise 10 — No difference, or cannot tell? `[Apply]`

Run the t-test on `enjoyment` by build and on `items_collected` by build. Neither
is significant. For which one can you say there is **no meaningful difference**,
and for which can you only say **we cannot tell**?

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex10
enjoy <- t.test(enjoyment ~ build, data = sessions)
items <- t.test(items_collected ~ build, data = sessions)

enjoy$conf.int
items$conf.int

tapply(sessions$items_collected, sessions$build, mean)
```

Read each interval against the size of the variable. Enjoyment's interval runs
from `r round(enjoy$conf.int[1], 2)` to `r round(enjoy$conf.int[2], 2)` on a
seven-point scale: whatever the true difference is, it is too small to matter. That
is **no meaningful difference**. The items interval runs from
`r round(items$conf.int[1], 2)` to `r round(items$conf.int[2], 2)` items, when a
typical player collects about
`r round(mean(sessions$items_collected))`. Build B's players could be collecting
up to `r sprintf("%.1f", -items$conf.int[1])` more items each, or up to
`r sprintf("%.1f", items$conf.int[2])` fewer. That is **cannot tell**.
:::

**Write 2 sentences:** why is "not significant" a weaker statement than "the
confidence interval is narrow and sits around zero"?

## Exercise 11 — "Players play longer in Build B" `[Decide]`

```{r}
#| label: ex11-setup
#| include: false
tt_min  <- t.test(session_minutes ~ build, data = sessions)
not_fr  <- sessions[sessions$quit_reason != "Frustrated", ]
tt_nf   <- t.test(session_minutes ~ build, data = not_fr)
fr_med  <- tapply(sessions$session_minutes[sessions$quit_reason == "Frustrated"],
                  sessions$build[sessions$quit_reason == "Frustrated"], median)
stopifnot(tt_min$p.value < 0.05, diff(tt_min$estimate) > 0,
          tt_nf$p.value > 0.2, abs(diff(fr_med)) < 1)
```

`t.test(session_minutes ~ build)` gives `r if (tt_min$p.value < 0.001) "p < 0.001" else paste("p =", round(tt_min$p.value, 3))`,
and Build B's sessions are about `r round(unname(diff(tt_min$estimate)))` minutes
longer on average. The producer wants to put "players play
`r round(unname(diff(tt_min$estimate)))` minutes longer in Build B" in the pitch. Do you
let them? Test before you decide: compare session length between the builds for
players who did **not** quit in frustration.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex11
not_frustrated <- sessions[sessions$quit_reason != "Frustrated", ]
t.test(session_minutes ~ build, data = not_frustrated)

frustrated <- sessions[sessions$quit_reason == "Frustrated", ]
tapply(frustrated$session_minutes, frustrated$build, median)
```

**One defensible answer: no, because the claim describes the wrong thing.** Among
players who did not quit in frustration, Build A averaged
`r round(tt_nf$estimate[[1]], 1)` minutes and Build B
`r round(tt_nf$estimate[[2]], 1)` (p = `r round(tt_nf$p.value, 2)`). Frustration
quits end at a median of about `r round(mean(fr_med), 1)` minutes in both builds.
Build B does not make anyone play longer. It makes fewer people quit early, and
that pulls its average up. The honest pitch line is "fewer players quit in
frustration", which is also the stronger claim.

**What makes it defensible.** It checks what is driving the difference before
describing it, and it replaces the claim with one the data supports.

**What would make it indefensible.** Approving the line because the p-value is
below 0.05. The difference is real; the explanation attached to it is not.
:::

**Write 3 sentences:** a significant difference in a mean can come from every
player shifting a little or from a few players shifting a lot. How do you tell
which, and why does it change the design conclusion?

## Exercise 12 — Which number goes in the report? `[Decide]`

You have three estimates of how much the signposting reduced wrong turns: the
session-level test from Exercise 08, the paired study from Exercise 09, and a
player-level test that averages each player's `r unique(as.vector(table(sessions$player_id)))`
sessions first. Run the player-level test, then decide which result leads the
report.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex12
per_player <- aggregate(wrong_turns ~ player_id + build, data = sessions, FUN = mean)
player_test <- t.test(wrong_turns ~ build, data = per_player)

player_test$conf.int
result$conf.int
paired_test$conf.int
```

**One defensible answer: lead with the player-level result, and cite the paired
study as support.** The session-level test treats
`r nrow(sessions)` sessions as independent when they come from
`r nrow(per_player)` players, so its interval is narrower than the data justifies.
The player-level interval, `r round(player_test$conf.int[1], 1)` to
`r round(player_test$conf.int[2], 1)` wrong turns, is honest about that. The paired
study, a different design with different players, points the same way, which is
worth more than any single p-value.

**What makes it defensible.** It picks the analysis whose assumptions match how
the data was collected, and it uses the other two as corroboration rather than
choosing between them.

**What would make it indefensible.** Picking whichever result has the smallest
p-value or the largest effect. That is choosing the answer, not the analysis.
:::

**Write 2 sentences:** why is agreement between two differently designed studies
stronger evidence than a very small p-value from one?

---

# Challenge — Build B was meant to be an improvement. Was it? `[Decide]`

The producer asks one question before approving Build B for the next milestone:
**was it an improvement?**

Build the evidence you would take into that meeting. Use more than one measure,
and keep four kinds of evidence separate: what the figures show, what the
descriptive statistics say, what the tests say, and whether each difference is big
enough to matter. `frames` is loaded in the Setup block.

::: {.callout-important collapse="true" title="One defensible answer"}

```{r}
#| label: challenge
# 1. Navigation: the question Build B was built to answer
t.test(wrong_turns ~ build, data = sessions)

# 2. Did players stay and finish?
table(sessions$build, sessions$quit_reason)
tapply(sessions$completed, sessions$build, mean)

# 3. Did anything get worse?
t.test(enjoyment ~ build, data = sessions)
t.test(score ~ build, data = sessions)

# 4. Performance, which nobody asked about
frames %>%
  group_by(build) %>%
  summarise(fps = 1000 / mean(ms), p99 = quantile(ms, 0.99))

ggplot(frames, aes(x = ms)) +
  geom_histogram(binwidth = 1, fill = "grey60") +
  facet_grid(build ~ zone) +
  labs(title = "Frame times by build and zone", x = "Frame time (ms)", y = "Frames")
```

```{r}
#| label: challenge-numbers
#| echo: false
wt   <- t.test(wrong_turns ~ build, data = sessions)
enj  <- t.test(enjoyment ~ build, data = sessions)
scr  <- t.test(score ~ build, data = sessions)
comp <- round(100 * tapply(sessions$completed, sessions$build, mean))
fr   <- table(sessions$build, sessions$quit_reason)[, "Frustrated"]
p99  <- round(tapply(frames$ms, frames$build, quantile, 0.99), 1)
fps  <- round(1000 / tapply(frames$ms, frames$build, mean))
slow <- frames[frames$ms > 1000 / 30, ]
p_fmt <- function(p) if (p < 0.001) "p < 0.001" else paste("p =", round(p, 2))

stopifnot(wt$p.value < 0.001, enj$p.value > 0.5, scr$p.value > 0.5,
          all(slow$build == "B"), all(slow$zone == "Corridor"),
          comp[["B"]] > comp[["A"]], fr[["B"]] < fr[["A"]],
          fps[["A"]] == fps[["B"]], p99[["B"]] > p99[["A"]])
```

| Measure | Visual | Descriptive | Statistical | Practical |
| :- | :- | :- | :- | :- |
| Wrong turns | Build B's histogram sits clearly to the left | Mean `r round(wt$estimate[[1]], 1)` in A, `r round(wt$estimate[[2]], 1)` in B | 95% CI `r round(wt$conf.int[1], 1)` to `r round(wt$conf.int[2], 1)` fewer, `r p_fmt(wt$p.value)` | Large: players get lost far less |
| Frustration quits | Build A's bar is far taller | `r fr[["A"]]` in A, `r fr[["B"]]` in B | Not tested this week | Large: the main way Build A lost players |
| Completed the test | — | `r comp[["A"]]`% in A, `r comp[["B"]]`% in B | Not tested this week | Large |
| Enjoyment | Panels look alike | Means `r round(enj$estimate[[1]], 1)` and `r round(enj$estimate[[2]], 1)` | 95% CI `r round(enj$conf.int[1], 2)` to `r round(enj$conf.int[2], 2)`, `r p_fmt(enj$p.value)` | No cost to enjoyment |
| Score | Boxes overlap | Means `r round(scr$estimate[[1]])` and `r round(scr$estimate[[2]])` | `r p_fmt(scr$p.value)` | No change |
| Frame time | A scatter of tiny bars far to the right in Build B's Corridor panel, and nowhere else. Easy to miss, even on the right figure | Both about `r fps[["A"]]` FPS; 99th percentile `r p99[["A"]]` ms in A, `r p99[["B"]]` ms in B | Not a comparison of means | Every one of the `r nrow(slow)` frames slower than 30 FPS is Build B, in the Corridor |

**So the defensible answer is: yes for navigation, and not yet for release.** The
signposting did what it was built to do. Players got lost less, fewer quit in
frustration, more finished, and none of it cost enjoyment or score. But Build B
also introduced a hitch in the Corridor that the frame-rate average cannot see.
Keep the signposting; fix the hitch before Build B goes forward.

**What would make it indefensible.** Answering from wrong turns alone, because that
is the measure the build was designed to move. Or quoting average FPS as proof
that performance was unchanged. Both answers are supported by a real number, and
both miss what a player would actually experience.
:::

**Write 4 sentences**, using **What → Pattern → Meaning → Action**:

- **What** did you measure?
- **Pattern** — what does the evidence show, across more than one measure?
- **Meaning** — what does that mean for the build?
- **Action** — what exactly do you recommend to the producer?

---

# Appendix — Reading `t.test()` output

::: {.callout-note collapse="true" title="Show the annotated output"}

```{r}
#| label: appendix-ttest
t.test(wrong_turns ~ build, data = sessions)
```

| Line | What it tells you |
| :- | :- |
| `Welch Two Sample t-test` | The independent-samples test, not assuming equal spreads. The right default |
| `t = ..., df = ...` | The test statistic and its degrees of freedom. Report them if asked; do not interpret them |
| `p-value = ...` | How surprising the data would be if the builds did not differ. `e-11` means "times 10 to the minus 11": tiny |
| `alternative hypothesis: ...` | Names the direction: the difference is *group A minus group B* |
| `95 percent confidence interval` | The plausible range for the difference, A minus B |
| `sample estimates` | The two group means |

:::

# Appendix — Common errors

::: {.callout-warning collapse="true" title="Show the error guide"}

| Message | What it usually means |
| :- | :- |
| `could not find function "ggplot"` | You have not run `library(ggplot2)` |
| ``Cannot use `+` with a single argument`` | A line starts with `+`. Move it to the end of the line above |
| `object 'build' not found` | A column name used outside `aes()` |
| ``stat_bin() requires a continuous x aesthetic`` | A histogram of a categorical column. Use `geom_bar()` |
| `grouping factor must have exactly 2 levels` | `t.test(y ~ group)` needs exactly two groups. Filter to two first |
| `not enough 'x' observations` | A group is empty, usually a misspelled level such as `"a"` for `"A"` |
| `could not find function "%>%"` | You have not run `library(dplyr)` |
| `could not find function "filter"` | You have not run `library(dplyr)` |
| `object 'clean' not found` | You skipped Exercise 03, or ran Part A out of order |
| `unexpected '>'` on a line with `|>` | Your R is older than 4.1. Update R, or use `%>%` |

:::
