---
title: "Lab 05 — From Evidence to Report"
subtitle: "Stage 4 workshop: one complete finding, from Tidebreak and then from your own game"
---

::: {.callout-note appearance="simple" icon="false" title="Get the files"}
- **Data:** [tidebreak_sessions.csv](../data/tidebreak_sessions.csv){download="tidebreak_sessions.csv"}, [tidebreak_survey.csv](../data/tidebreak_survey.csv){download="tidebreak_survey.csv"}. Save them in the `data/` folder of your project.
:::

# Overview

This is the last lab, and it is a workshop. It brings the whole course together
into the thing your assessment needs:

> **One complete finding: research question → relevant data →
> descriptive, visual and statistical evidence → interpretation →
> evidence-backed recommendation, written under the report headings.**

Part A works through one finding on *Tidebreak*, quickly: does environmental
guidance improve navigation without reducing enjoyment? It adds the two things
Stage 4 is about, coding comments and triangulating evidence, and ends with a
`report.qmd` that renders. Part B is your project. Bring your team's data and your
draft report.

| Function | What it does |
| :- | :- |
| `pivot_longer()` | Reshape wide survey columns into long rows |
| `left_join()` | Add columns from another table, matched on a key |
| `case_when()`, `grepl()` | Give each comment a code |
| `` `r ` `` inline code | Put a computed number into a sentence |
| `quarto render` | Turn `report.qmd` into HTML |

# Learning outcomes

By the end of this lab you can:

- summarise every survey item by build, and decide what a result lets you claim;
- code free-text comments with a short codebook, and count the codes;
- triangulate telemetry, ratings and comments, and report a disagreement honestly;
- build a report in Quarto under the assessment headings, with every number
  computed, and render it;
- write a complete finding from your own game, from question to recommendation.

# Ground rules

- Part B is the point. Keep Part A moving; the solutions are there to be read if
  time is short.
- No number goes into the report by hand. If it is a number, it is inline R.
- Results reports; Discussion judges. Keep them apart.
- Every finding ends with an Action someone could start on tomorrow
  ([rule 10](../notes/ten-rules.qmd)).
- If `library(tidyr)` fails, run `install.packages("tidyr")` once in the Console.

# Setup (run this first)

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
survey   <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)

nrow(sessions)
dim(survey)
```

`sessions` is the cleaned session data: one row per player per session, each player
on one build. `survey` is the post-session questionnaire: one row per session,
eight agreement items rated 1 to 5.

---

# Part A — One finding, on Tidebreak

Notes for this part:
[Interpreting and Reporting Evidence](../notes/11-reporting-evidence.qmd) and
[Combining Evidence](../notes/12-combining-evidence.qmd).

## Exercise 01 — Every survey item, both builds `[Apply]`

Reshape `survey` long, join each session's `build` onto it, and find the proportion
of players agreeing (4 or 5) with each item in each build. Which item moved most
between the builds?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex01
long <- survey |>
  pivot_longer(-session_id, names_to = "item", values_to = "rating")

agree_by_build <- long |>
  left_join(select(sessions, session_id, build), by = "session_id") |>
  group_by(item, build) |>
  summarise(agree = mean(rating >= 4), .groups = "drop") |>
  pivot_wider(names_from = build, values_from = agree) |>
  mutate(change = B - A) |>
  arrange(desc(abs(change)))

agree_by_build
```

`pivot_longer()` makes `item` a column so one `group_by()` covers every item.
`pivot_wider()` puts the two builds side by side so the change is one subtraction
([Reshaping and Joining](../going-further/reshaping-and-joining.qmd)).
:::

**Write 2 sentences:** `objective_clear` moved most. Which item would you have
*expected* to move most, given what Build B changed, and did it?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answer"}

```{r}
#| label: ex01-check
#| echo: false
nav <- agree_by_build[agree_by_build$item == "level_navigable", ]
obj <- agree_by_build[agree_by_build$item == "objective_clear", ]
stopifnot(agree_by_build$item[1] == "objective_clear", nav$change < obj$change)
```

`level_navigable`, since Build B added guidance to help players navigate. It moved
from `r round(100 * nav$A)`% agreeing to `r round(100 * nav$B)`%, against
`objective_clear`'s `r round(100 * obj$A)`% to `r round(100 * obj$B)`%. Build B
cut wrong turns, but what players noticed was knowing where they were meant to go.
:::

## Exercise 02 — "Without reducing enjoyment" `[Decide]`

The research question has two halves. Test the second: did Build B reduce
enjoyment? Use `enjoyment` from `sessions` (1–7) and `would_play_again` from
Exercise 01, and decide what the report may claim.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex02
enjoy <- t.test(enjoyment ~ build, data = sessions)
enjoy$estimate
enjoy$conf.int

agree_by_build |> filter(item == "would_play_again")
```

```{r}
#| label: ex02-check
#| include: false
wpa <- agree_by_build[agree_by_build$item == "would_play_again", ]
stopifnot(enjoy$p.value > 0.5, abs(wpa$change) < 0.1)
```

**One defensible answer: the report may claim no meaningful reduction.** Mean
enjoyment was `r round(enjoy$estimate[[1]], 2)` in Build A and
`r round(enjoy$estimate[[2]], 2)` in Build B. The 95% confidence interval for the
difference runs from `r round(enjoy$conf.int[1], 2)` to
`r round(enjoy$conf.int[2], 2)` on a seven-point scale, so any true difference is
small. Agreement with `would_play_again` was `r round(100 * wpa$A)`% and
`r round(100 * wpa$B)`%. Two different measures agree.

**What makes it defensible.** It rests on an interval, not on "not significant",
and on two measures rather than one. It claims *no meaningful reduction*, which the
evidence supports, rather than *no effect*, which no test can show.

**What would make it indefensible.** "The p-value is above 0.05, so Build B has no
effect on enjoyment." Or "Build B is just as enjoyable" from one measure with no
interval.
:::

**Write 2 sentences:** both measures were taken straight after one session. What
would you need to know before claiming Build B is as enjoyable over a week of play?

## Exercise 03 — Code the comments `[Apply]`

Read every distinct comment with `count(comment)`. Then give each session one of
four codes, *Lost or unclear*, *Found the way*, *Difficulty* or *General*, using
`case_when()` and `grepl()`. Check every distinct comment against its code, then
count the codes by build.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex03
sessions |> count(comment, sort = TRUE)

coded <- sessions |>
  mutate(code = case_when(
    grepl("lost|circles|meant to go|objective", comment) ~ "Lost or unclear",
    grepl("route|exit|signposting", comment)             ~ "Found the way",
    grepl("Died|hard", comment)                          ~ "Difficulty",
    TRUE                                                 ~ "General"
  ))

coded |> distinct(comment, code)
coded |> count(code, build)
```

`case_when()` gives each comment the first code whose keywords it contains;
`TRUE ~ "General"` catches the rest. `distinct(comment, code)` is the check by
hand: every sentence next to the code it got.
:::

**Write 1 sentence:** write the codebook entry for *Lost or unclear*: what counts,
and one example.

## Exercise 04 — Do the sources agree? `[Decide]`

```{r}
#| label: ex04-setup
#| include: false
lost <- table(coded$code, coded$build)["Lost or unclear", ]
stopifnot(lost[["B"]] > lost[["A"]],
          all(coded$enjoyment[coded$code == "Lost or unclear"] < 5),
          all(coded$enjoyment[coded$code == "Found the way"] >= 5))
```

The telemetry says Build B players took far fewer wrong turns. Compare that with
the comment codes from Exercise 03, by build and then by whether the player
enjoyed the session (`enjoyment >= 5`). Decide what the report should say about
navigation, and which evidence it should trust for what.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex04
sessions |>
  group_by(build) |>
  summarise(mean_wrong_turns = mean(wrong_turns))

coded |> count(code, build)
coded |> count(code, enjoyed = enjoyment >= 5)
```

**The sources disagree.** Build B players took fewer wrong turns, yet
`r lost[["B"]]` Build B comments were coded *lost or unclear*, against
`r lost[["A"]]` in Build A. Split by enjoyment instead of build, the codes separate
cleanly: every *lost or unclear* comment came from a player who rated enjoyment
below 5, and every *found the way* comment from one who rated it 5 or above.

**One defensible answer.** Trust the telemetry for *what happened*: Build B reduced
wrong turns. Treat the comments as evidence about *how the session felt*: they
track enjoyment, not build. Report both, in Results, and explain the gap in the
Discussion. Recommend structured observation at the junctions in the next playtest,
because nothing in this data shows whether players notice when they are lost.

**What would make it indefensible.** Leaving the comments out because they spoil
the story, or reporting them as evidence that Build B made navigation worse.
:::

**Write 4 sentences**, using **What → Pattern → Meaning → Action**, for the
disagreement.

## Exercise 05 — Start the report, and render it `[Apply]`

Create `report.qmd` in the project folder, next to `data/`. Give it:

- a YAML header with a title;
- the eight assessment headings: Executive Summary, Introduction, Methodology,
  Results, Discussion, Conclusion, References, Appendices;
- a setup chunk, hidden with `echo: false`, that loads the libraries and both CSVs;
- under Introduction, the research question;
- under Methodology, one sentence giving the number of sessions and players, both
  as inline R;
- under Results, the navigation figure from [Lab 03](clinic-figures.qmd),
  Exercise 03, with a caption that states the takeaway.

Then render it from the Terminal with `quarto render report.qmd` and open
`report.html`.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

````markdown
---
title: "Tidebreak Build B: Playtest Report"
---

```{{r}}
#| echo: false
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
survey   <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)
```

## Executive Summary

## Introduction

Does environmental guidance improve navigation without reducing enjoyment?

## Methodology

We analysed `r knitr::inline_expr("nrow(sessions)")` sessions from
`r knitr::inline_expr("length(unique(sessions$player_id))")` players, each of whom
played one build only.

## Results

```{{r}}
#| label: fig-wrong-turns
#| fig-cap: "Players made fewer wrong turns in Build B."
#| echo: false
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```

## Discussion

## Conclusion

## References

## Appendices
````

The two inline expressions produce these values when the report renders:

```{r}
#| label: ex05
nrow(sessions)
length(unique(sessions$player_id))
```

`echo: false` hides the code and keeps its output, which is what a reader of a
report wants. Because the numbers are inline, the report corrects itself the next
time the data changes.
:::

**Write 2 sentences:** change one value in your copy of the CSV, render again, and
describe what changed in the report and what you did not have to do.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="What you should see"}
The terminal ends with `Output created: report.html`. The page shows eight
headings, the sentence under Methodology reads "`r nrow(sessions)` sessions from
`r length(unique(sessions$player_id))` players", and the two-panel histogram sits
under Results with its caption and no code above it.

If the render stops with an error, the message names the line. The most common
causes are a data path written as `../data/...`, a chunk that uses a package it
did not load, and a missing closing ```` ``` ````.
:::

## Exercise 06 — Discussion, Conclusion and Executive Summary `[Decide]`

Write the last three pieces of the *Tidebreak* report. Every number is inline R.

- **Discussion:** whether the reduction in wrong turns matters to the player; the
  disagreement from Exercise 04; and at least three limitations specific to this
  playtest.
- **Conclusion:** one finding in **What → Pattern → Meaning → Action** form that
  answers the research question.
- **Executive Summary:** the question, the finding and the action, in three
  sentences. Write it last.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex06
wt <- t.test(wrong_turns ~ build, data = sessions)
wt$estimate
wt$conf.int

sessions |> count(cohort, build)
```

```{r}
#| label: ex06-check
#| include: false
per_player <- unique(as.vector(table(sessions$player_id)))
n_players  <- length(unique(sessions$player_id))
stopifnot(length(per_player) == 1, wt$conf.int[1] > 0,
          round(enjoy$estimate[[1]], 2) == round(enjoy$estimate[[2]], 2))
```

**One defensible Discussion.**

> A reduction of `r round(wt$estimate[[1]] - wt$estimate[[2]], 1)` wrong turns per
> session is large enough for players to notice, and enjoyment did not fall. The
> comments disagree with the telemetry, but they track enjoyment rather than
> build, so we treat them as evidence about how sessions felt, not about
> navigation.
>
> Limitations:
>
> - Each of the `r n_players` players contributed `r per_player` sessions, so the
>   `r nrow(sessions)` sessions are not independent; the result holds when each
>   player counts once.
> - Every measure was taken after one session. Nothing here shows whether the
>   difference lasts.
> - Cohorts were not perfectly balanced between the builds; the cohort table is in
>   the appendices so the reader can judge.
> - No observation was recorded, so we cannot say whether players noticed when
>   they were lost.

**One defensible Conclusion.**

> **What.** Wrong turns per session in `r nrow(sessions)` sessions, and enjoyment
> on a seven-point scale, compared between Build A and Build B.
>
> **Pattern.** Build B players averaged `r round(wt$estimate[[2]], 1)` wrong turns
> against Build A's `r round(wt$estimate[[1]], 1)`, a reduction of between
> `r round(wt$conf.int[1], 1)` and `r round(wt$conf.int[2], 1)` per session at 95%
> confidence. Enjoyment was unchanged, with a mean of
> `r round(enjoy$estimate[[1]], 2)` in both builds.
>
> **Meaning.** The guidance works, and costs nothing players noticed.
>
> **Action.** Keep Build B's guidance and carry it into the next build. Retest
> with an on-screen objective added, since objective clarity was the survey item
> that moved most, and add an observation sheet at the first two junctions.

**One defensible Executive Summary.**

> Build B's environmental guidance cut wrong turns by about
> `r round(wt$estimate[[1]] - wt$estimate[[2]], 1)` per session without reducing
> enjoyment. Players' comments did not reflect the improvement, because they track
> how a session felt rather than where the player went. Keep the guidance, add an
> on-screen objective, and observe the junctions in the next playtest.

**What makes it defensible.** The limitations are specific to this data, and say
what was checked. The Conclusion's Action follows from its Meaning, and its
Pattern gives a size and an interval. The Executive Summary stands on its own.

**What would make it indefensible.** "The sample was small" as the only
limitation; a Conclusion whose Action ("improve the level") no one could start on;
or an Executive Summary that describes the method instead of the finding.
:::

**Write 3 sentences:** which limitation would most change your Action if it turned
out to matter, and how would you test it?

---

# Part B — One complete finding from your own game `[Decide]`

This is the part that matters. Allow at least half the workshop for it.

Work in your team's `report.qmd`. Pick **one** design question your playtest can
answer, and take it all the way through:

1. **Research question** (Introduction). One sentence a designer would recognise.
2. **Relevant data** (Methodology). Which measures, from which sources, on how many
   players and sessions. Point to your cleaning log.
3. **Evidence** (Results). Descriptive: a summary per group, with n. Visual: one
   figure with a caption that states the takeaway (Lab 03). Statistical: an
   interval, and a test if the question is a comparison (Lab 02) or a correlation
   checked within groups if it is a relationship (Lab 04). Qualitative: coded
   comments or observation, if you collected them.
4. **Interpretation** (Discussion). Does it matter to the player? Do the sources
   agree? What could make it wrong?
5. **Recommendation** (Conclusion). What → Pattern → Meaning → Action. An Action
   someone on your team could start on tomorrow.

Render the report after each step, not just at the end.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answer"}
There is no single right finding: it depends on your game. Check your finding
against the **Nine-Part Evidence Workflow** before you call it done.

- [ ] **Research question** is stated, in the Introduction.
- [ ] **Experimental design** is described, with n and the cleaning log, in the
  Methodology.
- [ ] **Descriptive evidence**: a summary per group, with n, in the Results.
- [ ] **Visual evidence**: one figure with a takeaway caption, in the Results.
- [ ] **Statistical evidence**: an interval, and a test or a within-group check,
  in the Results.
- [ ] **Practical significance**: whether it matters to the player, in the
  Discussion.
- [ ] **Qualitative evidence**: coded comments or observation, in the Results and
  Discussion, or a limitation saying you have none.
- [ ] **Limitations**: specific to your playtest, in the Discussion.
- [ ] **Design conclusion**: What → Pattern → Meaning → Action, in the Conclusion.
- [ ] No number typed by hand. The report renders.

Exercises 01 to 06 are a worked version of the whole process on *Tidebreak*.
:::

---

# What to leave with

- A `report.qmd` with the eight assessment headings that renders without errors.
- One complete finding from your own game, from question to recommendation.
- A list of what is still missing for the final report, from the checklist above.

---

# Appendix — Common errors

::: {.callout-warning collapse="true" title="Show the error guide"}

| Message | What it usually means |
| :- | :- |
| `could not find function "ggplot"` | You have not run `library(ggplot2)` |
| `could not find function "pivot_longer"` | You have not run `library(tidyr)` |
| ``Cannot use `+` with a single argument`` | A line starts with `+`. Move it to the end of the line above |
| `object 'build' not found` | A column name used outside `aes()` or outside a dplyr verb |
| `'data/tidebreak_sessions.csv' does not exist` | `report.qmd` is not in the project folder next to `data/` |
| `object 'agree_by_build' not found` when rendering | The report uses an object you made in the Console. Put the code in a chunk in the report |
| Every comment coded "General" | The keywords do not match: check capitals, since `grepl()` is case-sensitive |

:::
