---
title: "Lab 01 — Reading Data"
subtitle: "Stage 1: vectors, missing values, loading a playtest, and describing what happened"
---

::: {.callout-note appearance="simple" icon="false" title="Get the files"}
- **Data:** [tidebreak_sessions.csv](../data/tidebreak_sessions.csv){download="tidebreak_sessions.csv"}. Save them in the `data/` folder of your project.
:::

# Overview

This lab is your first hands-on session with R. You are not expected to have
written a line of it before, but read
[R from First Principles](../notes/02-r-first-principles.qmd) first: every idea in
Part A is introduced there.

The work is deliberately unglamorous. Before you can find anything interesting in
a playtest you have to be able to **load the data, look at it, and describe it**
without making a fool of yourself. That is what this lab trains. By the end you
will have answered the Stage 1 question, *what happened?*, for a real playtest, and
said what it means for the game.

| Function | What it does |
| :- | :- |
| `c()`, `length()`, `class()` | Build a vector; how many values, and what type |
| `is.na()`, `na.rm = TRUE` | Find missing values; leave them out of a calculation |
| `read_csv()`, `glimpse()` | Read a CSV; look at it without printing every row |
| `$` | Pull out one column as a vector |
| `count()` | Count each value of a categorical variable |
| `mean()`, `median()`, `sd()`, `IQR()`, `quantile()` | Describe a numeric variable |
| `group_by()`, `summarise()`, `n()`, `\|>` | The same summaries, once per group |
| `hist()`, `boxplot()` | See the shape of a variable |

# Learning outcomes

By the end of this lab you can:

- create and index vectors, and explain what `NA` does to a calculation
- load a CSV and inspect it without printing 120 rows
- classify any column as **nominal**, **ordinal**, **discrete** or **continuous**
- describe a numeric variable with an appropriate centre and spread, and draw it
- summarise a variable for each group with `group_by()` and `summarise()`
- read a retention curve, and say in one sentence when the mean is the wrong
  summary and why

# Ground rules

- Keep the CSV files in the `data/` folder of the project.
- Type the code. Do not copy and paste from the solutions until you have tried.
- Open the solution **after** your attempt, even when your attempt worked. The
  solution sometimes does it a better way.
- Every exercise ends with a short written answer. Write it. The writing is the
  part that matters; the code is how you get there.
- If a line of code fails, read the error before doing anything else. Most R
  errors name the exact problem.

# Setup (run this first)

```{r}
#| label: setup
library(readr)   # read_csv()
library(dplyr)   # count(), group_by(), summarise() and the rest

csv_file <- "data/tidebreak_sessions.csv"

if (!file.exists(csv_file))
  stop("CSV not found: ", csv_file,
       "\nCheck that the file sits in the data/ folder of the project.")

sessions <- read_csv(csv_file, show_col_types = FALSE)

nrow(sessions)
```

---

# Part A — Values, vectors and missing data

Notes for this part: [Planning a Playtest](../notes/01-planning-a-playtest.qmd),
[R from First Principles](../notes/02-r-first-principles.qmd) and
[Data, Variables and Types](../notes/03-data-and-types.qmd). Keep the
[R and Quarto Reference](../reference/r-quarto-reference.qmd) open while you work.

## Exercise 01 — Build a vector and look at it `[Do]`

Create a vector called `deaths` holding the death counts from six playtest
sessions: 4, 2, 7, 3, 0, 5.

Report how many values it holds and what type R thinks it is.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex01
deaths <- c(4, 2, 7, 3, 0, 5)

deaths
length(deaths)
class(deaths)
```

R stores plain numbers as numeric, whether or not they have a decimal point.
:::

**Write 1 sentence:** why does a vector have to hold values of a single type?

## Exercise 02 — Pull values out by position `[Do]`

Using `deaths` from Exercise 01, extract the first value, then the third and fifth
values together.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex02
deaths[1]
deaths[c(3, 5)]
```

Square brackets take a position, or a vector of positions built with `c()`.
:::

**Write 1 sentence:** R counts from 1, not 0. If you have written C# or C++, what
is the one mistake this will cause you?

## Exercise 03 — A missing value poisons the arithmetic `[Do]`

A seventh tester's death count was not recorded. Add `NA` to the end of `deaths`
and take the mean. Then take it again so that the result is a number.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex03
deaths <- c(deaths, NA)

mean(deaths)
mean(deaths, na.rm = TRUE)
```

`NA` is not zero and it is not "empty". It means *unknown*, and the mean of a set
containing an unknown value is itself unknown. R tells you that rather than
quietly guessing.
:::

**Write 2 sentences:** `na.rm = TRUE` makes the `NA` go away. Explain why that is
not the same thing as fixing the problem.

## Exercise 04 — Count with a condition `[Apply]`

Using the seven-value `deaths` vector, produce a TRUE/FALSE vector showing which
sessions had more than 3 deaths, use it to extract those death counts, then count
how many sessions that is and what proportion of the recorded sessions it is.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex04
deaths > 3
deaths[deaths > 3]
sum(deaths > 3, na.rm = TRUE)
mean(deaths > 3, na.rm = TRUE)
```

Two things worth noticing. `deaths[deaths > 3]` keeps the `NA`, because R does not
know whether the unknown value is above 3 or not. And `sum()` on a TRUE/FALSE
vector counts the TRUEs, because TRUE is 1 and FALSE is 0, so `mean()` gives the
proportion.
:::

**Write 1–2 sentences:** why is `sum(x > 3)` a count rather than a total?

## Exercise 05 — Build a small table of related vectors `[Apply]`

Create two more vectors the same length as `deaths`: `player` (P01 to P07) and
`build` (alternating `"A"` and `"B"`, starting with A). Combine all three into a
data frame called `mini`, then look at it with `glimpse()`.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex05
player <- c("P01", "P02", "P03", "P04", "P05", "P06", "P07")
build  <- c("A", "B", "A", "B", "A", "B", "A")

mini <- data.frame(player, build, deaths)

glimpse(mini)
```

Each vector becomes a column, and the column takes the vector's name. They must
all be the same length: one value per session.
:::

**Write 1 sentence:** what does a data frame give you that three separate vectors
do not?

---

# Part B — Loading and describing real data

Notes for this part: [Describing Variables](../notes/04-describing-variables.qmd)
and [Distributions and Shape](../notes/05-distributions-and-shape.qmd).

From here on you are working with `sessions`, loaded in the Setup block: 120
playtest sessions from the *Tidebreak* arena.

## Exercise 06 — Look at a data frame without drowning in it `[Do]`

Answer these without printing the whole table: how many rows and columns, what the
columns are called, and what type each one is.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex06
glimpse(sessions)
```

`glimpse()` gives you the size, every column's name, its type and its first few
values in one go. It is the first thing to run on any new data.
:::

**Write 1 sentence:** why is printing a whole data frame a bad habit, even when it
is small?

## Exercise 07 — Get at a single column `[Do]`

Pull out the `score` column with `$`. How many values does it hold, and what are
the lowest and highest scores?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex07
length(sessions$score)
min(sessions$score)
max(sessions$score)
```

`sessions$score` is an ordinary vector, so everything from Part A works on it. You
will also see `sessions[["score"]]` in other people's code; it does the same job.
:::

**Write 1 sentence:** the lowest score is far below the rest. What would you want
to know about that session before you trusted any average of `score`?

## Exercise 08 — Classify every column `[Apply]`

For each of these columns, state whether it is **nominal**, **ordinal**,
**discrete** or **continuous** (see
[Data, Variables and Types](../notes/03-data-and-types.qmd#types-of-data)):
`cohort`, `archetype`, `build`, `score`, `kills`, `deaths`, `session_minutes`,
`quit_reason`.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

| Column | Type | Why |
| :- | :- | :- |
| `cohort` | Ordinal | Novice → Regular → Veteran has a real order |
| `archetype` | Nominal | Scout is not more or less than Sentinel |
| `build` | Nominal | Two labels, no order |
| `score` | Discrete | Whole points, though large enough to treat as continuous |
| `kills` | Discrete | A count |
| `deaths` | Discrete | A count |
| `session_minutes` | Continuous | Measured time, one decimal place |
| `quit_reason` | Nominal | Four labels, no order |

The interesting one is `cohort`. R has loaded it as `chr`, plain text, and will
sort it alphabetically as Novice, Regular, Veteran, which happens to be the right
order by accident. Never rely on that.
:::

**Write 2 sentences:** pick one column above and explain what would go wrong if you
treated it as the wrong type.

## Exercise 09 — "Only one in five" `[Decide]`

```{r}
#| label: ex09-setup
#| include: false
fr_all <- sum(sessions$quit_reason == "Frustrated")
fr_a   <- sum(sessions$build == "A" & sessions$quit_reason == "Frustrated")
fr_b   <- sum(sessions$build == "B" & sessions$quit_reason == "Frustrated")
n_a    <- sum(sessions$build == "A")
n_b    <- sum(sessions$build == "B")
stopifnot(round(fr_all / nrow(sessions), 1) == 0.2, fr_a > 3 * fr_b)
```

The producer counts the quit reasons and says: *"Only one in five players quit in
frustration. Most players are fine, so frustration is not a priority."* Count
`quit_reason`, then find the share of frustration quits in each build, and decide
whether the producer's reading is defensible.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex09
sessions |> count(quit_reason)

sessions |>
  group_by(build) |>
  summarise(n = n(),
            frustrated = mean(quit_reason == "Frustrated"))
```

**One defensible answer: no, because "one in five" averages two different
games.** Overall, `r fr_all` of `r nrow(sessions)` sessions ended in frustration,
`r round(100 * fr_all / nrow(sessions))`%. Split by build, that is `r fr_a` of
`r n_a` in Build A (`r round(100 * fr_a / n_a)`%) and `r fr_b` of `r n_b` in
Build B (`r round(100 * fr_b / n_b)`%). One build loses a third of its players to
frustration; the other hardly any. The overall figure describes neither.

**What makes it defensible.** It breaks the count down by the one thing the
playtest was designed to compare before judging it, and it reads the numbers as
a difference between builds rather than a verdict on the game.

**What would make it indefensible.** Accepting "one in five" because it is smaller
than the number who completed the test, without asking where the one in five came
from. Or going the other way and calling the whole game frustrating, when Build B
barely is.
:::

**Write 2 sentences:** what would you say to the producer instead, and what would
you want to check next?

## Exercise 10 — Describe a numeric variable properly `[Apply]`

For `score` and for `session_minutes`, compute the mean, median, standard
deviation, quartiles and IQR.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex10
mean(sessions$score)
median(sessions$score)
sd(sessions$score)
quantile(sessions$score)
IQR(sessions$score)

sessions |>
  summarise(mean   = mean(session_minutes),
            median = median(session_minutes),
            sd     = sd(session_minutes),
            iqr    = IQR(session_minutes))
```

Both ways work. The first is quick for one variable; `summarise()` puts every
summary in one table, which is what you want once there is more than one.
`quantile()` with no second argument gives the five-number summary.
:::

**Write 2 sentences:** express the gap between mean and median as a percentage of
the median, for both variables. Which one is more skewed, and does its tail run
left or right?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answer"}

```{r}
#| label: ex10-check
#| echo: false
gap <- function(x) 100 * (mean(x) - median(x)) / median(x)
g_score <- gap(sessions$score)
g_mins  <- gap(sessions$session_minutes)
```

Score's mean sits `r sprintf("%.1f%%", abs(g_score))` below its median. Session
minutes sits `r sprintf("%.1f%%", abs(g_mins))` below. Both are **left-skewed**:
the mean falls below the median because a group of low values drags it down. But
session minutes is far more so.

Comparing the raw gaps (`r round(mean(sessions$score) - median(sessions$score))`
points against `r round(mean(sessions$session_minutes) - median(sessions$session_minutes), 1)`
minutes) would have told you the opposite, which is why the scale of the variable
always has to come into it.
:::

## Exercise 11 — See the shape `[Apply]`

Draw a histogram of `score` and a boxplot of `session_minutes`. Give both a title
and a sensible axis label.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex11
#| fig-alt: "Histogram titled Distribution of session score: most sessions between 1,200 and 1,800, with a separate small group around 500 to 700."
hist(sessions$score,
     breaks = 20,
     main   = "Distribution of session score",
     xlab   = "Score",
     col    = "grey85",
     border = "white")
```

```{r}
#| label: ex11b
#| fig-alt: "Horizontal boxplot titled How long testers played, running from about 8 to 32 minutes, with the median near 26 and a long lower whisker."
boxplot(sessions$session_minutes,
        main       = "How long testers played",
        xlab       = "Minutes",
        horizontal = TRUE,
        col        = "grey85")
```

:::

**Write 2 sentences:** set `breaks = 5` and then `breaks = 40`, and describe what
happens to the story the histogram tells. See [rule 4](../notes/ten-rules.qmd).

## Exercise 12 — Split a numeric variable by a group `[Apply]`

Is one archetype's score less predictable than the others? Summarise `score` for
each archetype (n, median, IQR and SD), then draw boxplots of `score` split by
`archetype`, in one plot.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex12
sessions |>
  group_by(archetype) |>
  summarise(n      = n(),
            median = median(score),
            iqr    = IQR(score),
            sd     = round(sd(score)))
```

```{r}
#| label: ex12b
#| fig-alt: "Four boxplots of score by archetype. Sentinel's box is much taller than the other three and stretches down to about 950."
boxplot(score ~ archetype,
        data = sessions,
        main = "Score by archetype",
        ylab = "Score",
        col  = "grey85")
```

`group_by()` splits the table by archetype and `summarise()` computes each summary
once per piece. In the boxplot, `score ~ archetype` reads as "score, split by
archetype".
:::

**Write 2 sentences:** one archetype has a much larger spread than the others.
Which one, and why is "larger spread" a more useful observation here than "lower
average"?

## Exercise 13 — Which build keeps players longer? `[Apply]`

Summarise `session_minutes` for each build: the number of sessions, the median, the
IQR and the share of sessions that lasted at least 20 minutes. Then draw the two
builds' session lengths as side-by-side boxplots.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: ex13
sessions |>
  group_by(build) |>
  summarise(n              = n(),
            median_minutes = median(session_minutes),
            iqr_minutes    = IQR(session_minutes),
            share_20_plus  = mean(session_minutes >= 20))
```

```{r}
#| label: ex13b
#| fig-alt: "Two boxplots of session length. Build A's box is wider and lower than Build B's."
boxplot(session_minutes ~ build, data = sessions,
        main = "Session length by build", ylab = "Minutes", col = "grey85")
```

`mean(session_minutes >= 20)` is the proportion trick from Part A, run once per
build inside `summarise()`.
:::

**Write 2 sentences:** what does the IQR tell you about Build A that the median
does not? Which number would you put in front of the design lead?

## Exercise 14 — When do we lose them? `[Decide]`

```{r}
#| label: ex14-setup
#| include: false
mins_a <- sessions$session_minutes[sessions$build == "A"]
mins_b <- sessions$session_minutes[sessions$build == "B"]
drop_a <- mean(mins_a >= 11) - mean(mins_a >= 13)
drop_b <- mean(mins_b >= 11) - mean(mins_b >= 13)
stopifnot(drop_a > 2 * drop_b)
```

The producer has the two mean session lengths, `r round(mean(mins_a), 1)` minutes
for Build A and `r round(mean(mins_b), 1)` for Build B, and says: *"Build A
sessions are about three minutes shorter. Players drift off a bit sooner, that's
all."*

A **retention curve** shows the share of sessions still going at every minute. Run
this code to draw one for each build; you do not need to be able to write it. Then
decide whether "drift off a bit sooner" is a fair description.

```{r}
#| label: ex14-curve
#| fig-alt: "Two step lines showing the share of sessions still playing at each minute. Build A's grey line drops sharply between minutes 11 and 13; Build B's red line stays higher until the late twenties."
mins_a <- sessions$session_minutes[sessions$build == "A"]
mins_b <- sessions$session_minutes[sessions$build == "B"]

minute  <- 0:32
still_a <- sapply(minute, function(m) mean(mins_a >= m))
still_b <- sapply(minute, function(m) mean(mins_b >= m))

plot(minute, still_a, type = "s", lwd = 2, col = "grey40", ylim = c(0, 1),
     main = "Share of sessions still playing",
     xlab = "Minute", ylab = "Share still playing")
lines(minute, still_b, type = "s", lwd = 2, col = "firebrick")
legend("bottomleft", legend = c("Build A", "Build B"), lwd = 2,
       col = c("grey40", "firebrick"), bty = "n")
```

`mean(mins_a >= m)` is the share of Build A sessions that lasted at least `m`
minutes, the proportion trick again; `sapply()` repeats it for every minute.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}
**One defensible answer: no. Build A does not drift; it drops.** Between minutes
11 and 13, Build A loses `r round(100 * drop_a)`% of its sessions; Build B loses
`r round(100 * drop_b)`% over the same two minutes. Something in the first dozen
minutes of Build A drives a block of players out at once, and Exercise 09 already
showed those quits are frustration. The investigation should start at minute 11 of
Build A: watch the replays, check where those players were.

**What makes it defensible.** It looks at *when* players left, not just the
average, and names a specific moment someone can investigate.

**What would make it indefensible.** Comparing the two means, agreeing that the
difference is "a bit", and moving on. The mean folds a sharp exit at minute 12 and
a long tail of full sessions into one number that describes neither
([rule 9](../notes/ten-rules.qmd)).
:::

**Write 2 sentences:** what would you ask the level designer to check first, and
what number would tell you, in the next playtest, that the fix worked?

---

# Challenge — Which cohort is the game actually built for? `[Decide]`

You have been asked a real question by the design lead: **is Tidebreak too hard
for novice players?**

Using only what is in this lab, produce the evidence you would take into that
conversation. There is no single correct answer. There is a defensible one and an
indefensible one.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show solution"}

```{r}
#| label: challenge
# One table: size, score, deaths, session length and frustration, per cohort
sessions |>
  group_by(cohort) |>
  summarise(n              = n(),
            mean_score     = round(mean(score)),
            mean_deaths    = round(mean(deaths), 2),
            median_minutes = median(session_minutes),
            frustrated     = round(mean(quit_reason == "Frustrated"), 2))
```

```{r}
#| label: challenge-plot
#| fig-alt: "Three boxplots of score by cohort, rising from Novice to Regular to Veteran, with a few low outliers in the Regular and Veteran groups."
# Show the shape, not just the numbers
boxplot(score ~ cohort, data = sessions,
        main = "Score by player cohort", ylab = "Score", col = "grey85")
```

**What the evidence shows.** Novices do score lower, and that is the finding
everyone expects. On its own it looks like a difficulty problem.

Every other measure contradicts it. Novices die at almost exactly the same rate as
everyone else, they play for *longer* than either other cohort, and they are less
likely to quit in frustration than regular players. A game that was too hard for
novices would show them dying more and leaving sooner. This one does not.

**So the defensible answer is: no.** The score gap looks like inexperience rather
than difficulty: novices are playing the game safely and slowly and scoring less
for it. The genuinely odd number in that table is the frustration rate among
*regular* players, which nobody asked about and which is the thing worth chasing
next.

**What would make it indefensible.** Comparing three mean scores, announcing that
novices are struggling, and recommending a difficulty change. The number is real
and the conclusion does not follow from it.
:::

**Write 3–4 sentences**, using **What → Pattern → Meaning → Action**:

- **What** did you measure?
- **Pattern**: what does the evidence show?
- **Meaning**: what might that mean for the design?
- **Action**: what is the one change or the one further test you would recommend?

**Going further:** Distributions and Shape says a normal distribution puts about
68% of values within 1 SD of the mean. If you want to check that against real
columns, try the exercise at the end of
[Is It Normal Enough?](../going-further/normality.qmd).

---

# Appendix — Mini dataset for Part A

If the CSV files are not available, use this in place of `sessions`.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show the mini dataset"}

```{r}
#| label: appendix-mini
#| eval: false
mini_sessions <- data.frame(
  player_id       = c("P01", "P02", "P03", "P04", "P05", "P06", "P07", "P08"),
  cohort          = c("Novice", "Veteran", "Regular", "Novice",
                      "Regular", "Veteran", "Regular", "Novice"),
  build           = c("A", "B", "A", "B", "A", "B", "A", "B"),
  score           = c(1370, 1795, 1690, 1544, 2019, 1614, 1652, 1331),
  kills           = c(7, 5, 5, 7, 12, 7, 6, 7),
  deaths          = c(4, 2, 2, 4, 3, 5, 4, NA),
  session_minutes = c(12.9, 24.6, 12.5, 18.8, 18.7, 29.4, 11.2, 27.0),
  quit_reason     = c("Frustrated", "Lost interest", "Frustrated",
                      "Completed test", "Completed test", "Completed test",
                      "Frustrated", "Technical fault")
)

glimpse(mini_sessions)
```

:::

# Appendix — Common first-week errors

::: {.callout-warning collapse="true" title="Show the error guide"}

| Message | What it usually means |
| :- | :- |
| `could not find function "read_csv"` | You forgot `library(readr)` |
| `could not find function "count"` (or `group_by`, `summarise`, `glimpse`) | You forgot `library(dplyr)` |
| `object 'sessions' not found` | You have not run the Setup block, or you restarted R |
| `object 'Score' not found` | Wrong capitalisation. R is case-sensitive |
| `$ operator is invalid for atomic vectors` | You used `$` on a vector instead of a data frame |
| `argument "x" is missing, with no default` | You called a function with no argument, usually a stray `()` |
| `unexpected symbol` / `unexpected ')'` | A typo: nearly always a missing comma or an unclosed bracket on the line above |
| The console shows `+` instead of `>` | R is waiting for you to finish an expression, often a pipe `\|>` at the end of the last line. Press **Esc** |

:::
