---
title: "Lab 01 — Reading Data"
subtitle: "Vectors, types, missing values, data frames, and describing one variable"
---

# Overview

This lab is your first contact with R. You are not expected to have written a line
of it before today.

The work is deliberately unglamorous. Before you can find anything interesting in
a playtest you have to be able to **load the data, look at it, and describe it**
without making a fool of yourself. That is what this sheet trains.

Everything here uses **base R** — the functions that ship with the language. You
will meet the tidyverse in Lab 02. Learning base R first means you will understand
what the tidyverse is doing for you rather than treating it as magic.

| Function | What it does |
| :- | :- |
| `c()` | Builds a vector from individual values |
| `length()`, `class()` | How many, and what type |
| `read_csv()` | Reads a CSV file into a data frame |
| `str()`, `head()`, `nrow()`, `names()` | Look at a data frame without printing all of it |
| `$`, `[ , ]` | Pull out a column, or a slice of rows and columns |
| `mean()`, `median()`, `sd()`, `range()` | Describe a numeric variable |
| `quantile()`, `IQR()` | Describe the spread and the tails |
| `table()` | Count a categorical variable |
| `hist()`, `boxplot()` | See the shape of a variable |

# Learning outcomes

By the end of this lab you can:

- create and index vectors, and explain what `NA` does to a calculation
- load a CSV into a data frame and inspect it without printing 120 rows
- classify any column as **nominal**, **ordinal**, **discrete** or **continuous**
- describe a numeric variable with an appropriate measure of centre and spread
- state, in one sentence, when the mean is the wrong summary and why

# Ground rules

- Keep the CSV files in the `data/` folder of the project.
- Type the code. Do not copy and paste from the solutions until you have tried.
- Open the solution **after** your attempt, even when your attempt worked. The
  solution sometimes does it a better way.
- Every exercise ends with a short written answer. Write it. The writing is the
  part that gets assessed later, not the code.
- If a line of code fails, read the error before doing anything else. Most R
  errors name the exact problem.

# Setup (run this first)

```{r}
#| label: setup
# readr gives us read_csv(), which is better behaved than base R's read.csv()
library(readr)

csv_file <- "data/tidebreak_sessions.csv"

if (!file.exists(csv_file))
  stop("CSV not found: ", csv_file,
       "\nCheck that the file sits in the data/ folder of the project.")

sessions <- read_csv(csv_file, show_col_types = FALSE)

nrow(sessions)
```

---

# Part A — Values, vectors and missing data

## Exercise 01 — Build a vector and look at it `[Do]`

Create a vector called `deaths` holding the death counts from six playtest
sessions: 4, 2, 7, 3, 0, 5.

Report how many values it holds and what type R thinks it is.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex01
deaths <- c(4, 2, 7, 3, 0, 5)

deaths
length(deaths)
class(deaths)
```

R stores plain numbers as doubles unless you force an integer with `4L`.
:::

**Write 1 sentence:** why does a vector have to hold values of a single type?

## Exercise 02 — Pull values out by position `[Do]`

Using `deaths` from Exercise 01, extract the first value, the third and fifth
values, and everything except the first value.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex02
deaths[1]
deaths[c(3, 5)]
deaths[-1]
```

Square brackets take a position, a vector of positions, or a negative position
meaning "all except this".
:::

**Write 1 sentence:** R counts from 1, not 0. If you have written C# or C++, what
is the one mistake this will cause you?

## Exercise 03 — A missing value poisons the arithmetic `[Do]`

A seventh tester's death count was not recorded. Add `NA` to the end of `deaths`
and take the mean. Then take it again so that the result is a number.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex03
deaths <- c(deaths, NA)

mean(deaths)
mean(deaths, na.rm = TRUE)
```

`NA` is not zero and it is not "empty". It means *unknown*, and the mean of a set
containing an unknown value is itself unknown. R tells you that rather than
quietly guessing.
:::

**Write 2 sentences:** `na.rm = TRUE` makes the error message go away. Explain why
that is not the same thing as fixing the problem.

## Exercise 04 — Filter a vector with a condition `[Apply]`

Using the seven-value `deaths` vector, produce a TRUE/FALSE vector showing which
sessions had more than 3 deaths, use it to extract those death counts, then count
how many sessions that is.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex04
deaths > 3
deaths[deaths > 3]
sum(deaths > 3, na.rm = TRUE)
```

Two things worth noticing. `deaths[deaths > 3]` keeps the `NA`, because R does not
know whether the unknown value is above 3 or not. And `sum()` on a TRUE/FALSE
vector counts the TRUEs, because TRUE is 1 and FALSE is 0.
:::

**Write 1–2 sentences:** why is `sum(x > 3)` a count rather than a total?

## Exercise 05 — Build a small table of related vectors `[Apply]`

Create two more vectors the same length as `deaths` — `player` (P01 to P07) and
`build` (alternating `"A"` and `"B"`, starting with A). Combine all three into a
data frame called `mini`, then show its structure and first three rows.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex05
player <- c("P01", "P02", "P03", "P04", "P05", "P06", "P07")
build  <- rep(c("A", "B"), length.out = 7)

mini <- data.frame(player, build, deaths)

str(mini)
head(mini, 3)
```

`rep()` with `length.out` repeats a pattern until it reaches the length you asked
for — useful for generating alternating conditions.
:::

**Write 1 sentence:** what does a data frame give you that three separate vectors
do not?

---

# Part B — Loading and describing real data

From here on you are working with `sessions`, loaded in the Setup block: 120 real
playtest sessions from the *Tidebreak* arena.

## Exercise 06 — Look at a data frame without drowning in it `[Do]`

Answer these without printing the whole table: how many rows and columns, what the
column names are, and what the first four rows look like.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex06
dim(sessions)
names(sessions)
head(sessions, 4)
```

`str(sessions)` gives you dimensions, names and types all at once, and is usually
the first thing to run on unfamiliar data.
:::

**Write 1 sentence:** why is printing a whole data frame a bad habit, even when it
is small?

## Exercise 07 — Get at a single column `[Do]`

Extract the `score` column three different ways: with `$`, with `[[ ]]`, and with
`[ , ]`. Confirm all three give the same thing.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex07
by_dollar  <- sessions$score
by_bracket <- sessions[["score"]]
by_comma   <- sessions[, "score", drop = TRUE]

identical(by_dollar, by_bracket)
identical(by_dollar, by_comma)
```

`drop = TRUE` matters: without it, `[ , ]` on a tibble hands you back a one-column
table rather than a plain vector, and the identical check fails.

Note the variable names. Never call a variable `c` — you would be shadowing `c()`,
the function you used to build every vector in Part A.
:::

**Write 1 sentence:** which of the three would you use in a script another person
has to read, and why?

## Exercise 08 — Classify every column `[Apply]`

For each of these columns, state whether it is **nominal**, **ordinal**,
**discrete** or **continuous**: `cohort`, `archetype`, `build`, `score`, `kills`,
`deaths`, `session_minutes`, `quit_reason`.

::: {.callout-tip collapse="true" title="Solution"}

| Column | Type | Why |
| :- | :- | :- |
| `cohort` | Ordinal | Novice → Regular → Veteran has a real order |
| `archetype` | Nominal | Scout is not more or less than Sentinel |
| `build` | Nominal | Two labels, no order |
| `score` | Discrete | Whole points, though large enough to treat as continuous |
| `kills` | Discrete | A count |
| `deaths` | Discrete | A count |
| `session_minutes` | Continuous | Measured time, one decimal place |
| `quit_reason` | Nominal | Four labels, no order |

The interesting one is `cohort`. R has loaded it as `chr` — plain text — and will
sort it alphabetically as Novice, Regular, Veteran, which happens to be the right
order by accident. Never rely on that.
:::

**Write 2 sentences:** pick one column above and explain what would go wrong if you
treated it as the wrong type.

## Exercise 09 — Count the categorical columns `[Do]`

Produce counts for `cohort`, `archetype` and `quit_reason`. Then produce a
cross-tabulation of `build` against `quit_reason`.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex09
table(sessions$cohort)
table(sessions$archetype)
table(sessions$quit_reason)

table(sessions$build, sessions$quit_reason)
```

The cross-tabulation is the one that earns its keep.
:::

**Write 2 sentences:** look at the cross-tabulation. What is the single most
noticeable difference between the two builds, and what would you want to check
next?

## Exercise 10 — Describe a numeric variable properly `[Apply]`

For `score` and for `session_minutes`, compute the mean, median, standard
deviation, range, quartiles and IQR.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex10
mean(sessions$score)
median(sessions$score)
sd(sessions$score)
range(sessions$score)
quantile(sessions$score)
IQR(sessions$score)

summary(sessions$session_minutes)
IQR(sessions$session_minutes)
```

`summary()` gives minimum, quartiles, median, mean and maximum in one call, which
is usually where to start. `quantile()` is there for when you need a specific
percentile rather than the standard five.
:::

**Write 2 sentences:** express the gap between mean and median as a percentage of
the median, for both variables. Which one is more skewed, and does its tail run
left or right?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: ex10-check
#| echo: false
gap <- function(x) 100 * (mean(x) - median(x)) / median(x)
g_score <- gap(sessions$score)
g_mins  <- gap(sessions$session_minutes)
```

Score sits `r sprintf("%.1f%%", abs(g_score))` below its median. Session minutes
sits `r sprintf("%.1f%%", abs(g_mins))` below. Both are **left-skewed** — the mean
falls below the median because a group of short sessions drags it down — but
session minutes is far more so.

Comparing the raw gaps (`r round(mean(sessions$score) - median(sessions$score))`
points against `r round(mean(sessions$session_minutes) - median(sessions$session_minutes), 1)`
minutes) would have told you the opposite, which is why the scale of the variable
always has to come into it.
:::

## Exercise 11 — See the shape `[Apply]`

Draw a histogram of `score` and a boxplot of `session_minutes`. Give both a title
and a sensible axis label.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex11
hist(sessions$score,
     breaks = 20,
     main   = "Distribution of session score",
     xlab   = "Score",
     col    = "grey85",
     border = "white")
```

```{r}
#| label: ex11b
boxplot(sessions$session_minutes,
        main       = "How long testers played",
        ylab       = "Minutes",
        horizontal = TRUE,
        col        = "grey85")
```

:::

**Write 2 sentences:** set `breaks = 5` and then `breaks = 40`, and describe what
happens to the story the histogram tells. See [rule 4](../notes/ten-rules.qmd).

## Exercise 12 — Split a numeric variable by a group `[Apply]`

Draw boxplots of `score` split by `archetype`, in one plot.

::: {.callout-tip collapse="true" title="Solution"}

```{r}
#| label: ex12
boxplot(score ~ archetype,
        data = sessions,
        main = "Score by archetype",
        ylab = "Score",
        col  = "grey85")
```

The `y ~ x` notation reads as "score **as explained by** archetype". You will meet
it again in every modelling function R has.
:::

**Write 2 sentences:** one archetype has a much larger spread than the others.
Which one, and why is "larger spread" a more useful observation here than "lower
average"?

---

# Challenge — Which cohort is the game actually built for? `[Decide]`

You have been asked a real question by the design lead: **is Tidebreak too hard
for novice players?**

Using only what is on this sheet, produce the evidence you would take into that
conversation. There is no single correct answer. There is a defensible one and an
indefensible one.

::: {.callout-important collapse="true" title="One defensible answer"}

```{r}
#| label: challenge
# 1. How many people are we talking about?
table(sessions$cohort)

# 2. Centre and spread of score, per cohort
tapply(sessions$score, sessions$cohort, summary)

# 3. Deaths, which is closer to "difficulty" than score is
tapply(sessions$deaths, sessions$cohort, mean)

# 4. And did they leave early?
table(sessions$cohort, sessions$quit_reason)

# 5. Show the shape, not just the numbers
boxplot(score ~ cohort,
        data = sessions,
        main = "Score by player cohort",
        ylab = "Score",
        col  = "grey85")
```

**What the evidence actually shows.**

```{r}
#| label: challenge-table
#| echo: false
cohorts <- c("Novice", "Regular", "Veteran")
frustrated <- tapply(sessions$quit_reason == "Frustrated", sessions$cohort, mean)

knitr::kable(
  data.frame(
    Measure = c("Mean score", "Mean deaths",
                "Median session length (min)", "Quit in frustration"),
    Novice  = c(round(tapply(sessions$score, sessions$cohort, mean)[["Novice"]]),
                round(tapply(sessions$deaths, sessions$cohort, mean)[["Novice"]], 2),
                round(tapply(sessions$session_minutes, sessions$cohort, median)[["Novice"]], 1),
                sprintf("%.0f%%", 100 * frustrated[["Novice"]])),
    Regular = c(round(tapply(sessions$score, sessions$cohort, mean)[["Regular"]]),
                round(tapply(sessions$deaths, sessions$cohort, mean)[["Regular"]], 2),
                round(tapply(sessions$session_minutes, sessions$cohort, median)[["Regular"]], 1),
                sprintf("%.0f%%", 100 * frustrated[["Regular"]])),
    Veteran = c(round(tapply(sessions$score, sessions$cohort, mean)[["Veteran"]]),
                round(tapply(sessions$deaths, sessions$cohort, mean)[["Veteran"]], 2),
                round(tapply(sessions$session_minutes, sessions$cohort, median)[["Veteran"]], 1),
                sprintf("%.0f%%", 100 * frustrated[["Veteran"]]))
  ),
  align = c("l", "r", "r", "r")
)
```

Novices do score lower, and that is the finding everyone expects. On its own it
looks like a difficulty problem.

Every other measure contradicts it. Novices die at almost exactly the same rate as
everyone else, they play for *longer* than either other cohort, and they are less
likely to quit in frustration than regular players. A game that was too hard for
novices would show them dying more and leaving sooner. This one does not.

**So the defensible answer is: no.** The score gap looks like inexperience rather
than difficulty — novices are playing the game safely and slowly and scoring less
for it. The genuinely odd number on that table is the frustration rate among
*regular* players, which nobody asked about and which is the thing worth chasing
next.

**What would make it indefensible.** Comparing three mean scores, announcing that
novices are struggling, and recommending a difficulty change. The number is real
and the conclusion does not follow from it.
:::

**Write 3–4 sentences**, using **What → Pattern → Meaning → Action**:

- **What** did you measure?
- **Pattern** — what does the evidence show?
- **Meaning** — what might that mean for the design?
- **Action** — what is the one change or the one further test you would recommend?

---

# Appendix — Mini dataset for Part A

If the CSV files are not available, use this in place of `sessions`.

::: {.callout-note collapse="true" title="Show the mini dataset"}

```{r}
#| label: appendix-mini
#| eval: false
mini_sessions <- data.frame(
  player_id       = c("P01", "P02", "P03", "P04", "P05", "P06", "P07", "P08"),
  cohort          = c("Novice", "Veteran", "Regular", "Novice",
                      "Regular", "Veteran", "Regular", "Novice"),
  build           = c("A", "B", "A", "B", "A", "B", "A", "B"),
  score           = c(1370, 1795, 1690, 1544, 2019, 1614, 1652, 1331),
  kills           = c(7, 5, 5, 7, 12, 7, 6, 7),
  deaths          = c(4, 2, 2, 4, 3, 5, 4, NA),
  session_minutes = c(12.9, 24.6, 12.5, 18.8, 18.7, 29.4, 11.2, 27.0),
  quit_reason     = c("Frustrated", "Lost interest", "Frustrated",
                      "Completed test", "Completed test", "Completed test",
                      "Frustrated", "Technical fault")
)

str(mini_sessions)
```

:::

# Appendix — Common first-week errors

::: {.callout-warning collapse="true" title="Show the error guide"}

| Message | What it usually means |
| :- | :- |
| `could not find function "read_csv"` | You forgot `library(readr)` |
| `object 'sessions' not found` | You have not run the Setup block, or you restarted R |
| `object 'Score' not found` | Wrong capitalisation. R is case-sensitive |
| `$ operator is invalid for atomic vectors` | You used `$` on a vector instead of a data frame |
| `argument "x" is missing, with no default` | You called a function with no argument, usually a stray `()` |
| `unexpected symbol` / `unexpected ')'` | A typo — nearly always a missing comma or an unclosed bracket on the line above |
| The console shows `+` instead of `>` | R is waiting for you to finish an expression. Press **Esc** |

:::
