---
title: "Planning a Playtest"
subtitle: "From a design question to data worth analysing"
body-classes: "wf-at-question wf-at-measure wf-at-collect"
aliases:
  - why-measure.html
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")

n_players  <- length(unique(sessions$player_id))
per_player <- unique(as.vector(table(sessions$player_id)))
builds_per_player <- tapply(sessions$build, sessions$player_id,
                            function(b) length(unique(b)))
stopifnot(length(per_player) == 1, all(builds_per_player == 1))
players_a <- length(unique(sessions$player_id[sessions$build == "A"]))
players_b <- length(unique(sessions$player_id[sessions$build == "B"]))

n_a    <- sum(sessions$build == "A")
n_b    <- sum(sessions$build == "B")
frus_a <- sum(sessions$build == "A" & sessions$quit_reason == "Frustrated")
frus_b <- sum(sessions$build == "B" & sessions$quit_reason == "Frustrated")

item_means <- sort(colMeans(survey[, -1]))
stopifnot(names(item_means)[1] == "objective_clear")
```

> Most of what goes wrong in a playtest analysis went wrong before anyone opened
> the data: a vague question, the wrong measure, the wrong players, or a design
> that cannot separate the build from the people playing it. This page is the
> plan you make before the first tester sits down. The playtest question running
> through it: *"Is Build A worse than Build B?"*

::: {.at-a-glance}
**Stage:** 1 · Core.
**Time:** 30 minutes in class; 15 more for the sections marked Useful when needed.
**Before this:** nothing. This is the first page.
**Key terms:** telemetry, observation, self-report, measure, between-subjects,
within-subjects, counterbalancing, random assignment, pilot, threshold.
**R you will write:** none. R starts in [R from First Principles](02-r-first-principles.qmd).
:::

## Learning Outcomes

By the end of these notes you can:

- name the three cases where a team's judgement of its own game reliably fails;
- distinguish telemetry, observation and self-report, and say what each can and
  cannot tell you;
- rewrite a vague design question as a measure you could record for a single
  player;
- choose between a between-subjects and a within-subjects design, and say how
  counterbalancing protects a within-subjects test;
- explain why a few players are enough to find problems but not to measure them;
- name the ethical obligations of a playtest that records player data;
- write down, before collecting anything, the result that would change your mind.

---

## Why This Matters for Playtesting

Every team makes decisions about their game constantly. Most of those decisions
are made from memory, taste and the loudest voice in the room, and most of the
time that is fine. Taste is why anyone plays your game rather than a spreadsheet.

Measurement earns its place in the cases where taste reliably fails. There are
three of them, and they recur in every project.

**You cannot see your own game any more.** You have played your level five hundred
times. You know where the key is. You cannot un-know it, and no amount of care
will let you experience your own tutorial as a stranger. This is the single most
common cause of a game that the team finds obvious and the player finds
impenetrable.

**You cannot feel a rare event.** A stutter that happens once every two seconds
occupies less than one per cent of frames. Your average frame rate will not show
it. Your players will talk about nothing else.

**You cannot remember twenty playtests.** After a session you remember the two
people who struggled most and the last person you watched. You do not remember the
distribution, and you will confidently reconstruct one that never existed.

A playtest fixes all three, but only if it is planned. The rest of this page is
the plan: seven decisions, made in this order, before anyone plays.

1. The design question.
2. The measures.
3. The design: who plays what.
4. How many players.
5. Who to recruit.
6. Ethics and consent.
7. The threshold that would change your mind, and a pilot.

---

## 1. Start with a Design Question

The question a designer asks is almost never answerable as asked.

| What you are asked | What you can actually measure |
| :- | :- |
| "Is the level too hard?" | Deaths per attempt; time to first completion; proportion who never finish; where attempts end |
| "Does the game feel good?" | Time to first successful interaction; retry rate after failure; rated responsiveness |
| "Is the tutorial working?" | Proportion who use the taught mechanic later, unprompted |
| "Do people like the art?" | Rated appeal; which screenshots they choose; where they stop to look |

Each right-hand cell names a **specific number you could write down after
watching one person play**. That is the test. If you cannot describe how you would
record it for a single player, you have not finished turning the question into a
measurement ([rule 2](ten-rules.qmd)).

The right-hand column is often several measures, not one, and they can disagree.
That is not a failure of the method. A tutorial that everybody completes and nobody
learns from is a real and common outcome, and only two measures together can find
it.

**For your project:** write down the one question your next playtest has to
answer, in the designer's words. Then rewrite it until you could record the answer
after watching a single player. If you cannot get there, the question is not ready
and neither is the playtest.

---

## 2. Choose Your Measures

Almost everything you analyse in a game project comes from one of three sources of
evidence. They answer different questions, so you need to know which one you are
holding.

| Source | What it tells you | Examples |
| :- | :- | :- |
| **Telemetry** | What happened in the game | deaths, completion time, wrong turns, attempts, objectives completed, frame times |
| **Observation** | How the player's behaviour unfolded | missed a doorway, backtracked, hesitated at an interaction, ignored a prompt |
| **Self-report** | What the player says they experienced | difficulty and enjoyment ratings, confidence, comments, questionnaires |

**Telemetry** is what the game records about itself while it runs. It is
objective, it arrives in large quantities, and it tells you *what* happened but
never *why*. Performance data (frame times, memory) is telemetry too, and it is the
one place where the usual summary, the average frame rate, actively misleads you.
[Distributions and Shape](05-distributions-and-shape.qmd) shows why.

**Observation** is what you see when you watch someone play: where they hesitate,
what they miss, where they go back on themselves. You record it as notes against a
time or a location, ideally against a short list of things to watch for, written
in advance. It catches what telemetry flattens: a player who took the right route
after thirty seconds of wandering looks fine in the log.

**Self-report** is surveys, interviews and think-aloud comments. It is the only
source that tells you how the game *felt*, and the least reliable. People
misremember, they are polite, and they answer the question they think you meant. A
rating tells you what a player believes about their experience, which is real and
useful, but it is not the same as what happened. How to ask well is covered in
[Data, Variables and Types](03-data-and-types.qmd#measuring-with-surveys).

**The sources disagree, and that is the point.** The most useful finding in a
playtest is often where they contradict each other: players say the level is fine
and the log shows nobody finds the side route; players say the game runs smoothly
and the frame log shows a hitch every two seconds. When they agree, the finding is
stronger than any one of them:

> Players made more wrong turns **(telemetry)**, observers saw repeated
> backtracking at the junction **(observation)**, and players rated the level as
> hard to navigate **(self-report)**.

Checking a finding against more than one source is called **triangulation**. A
team with one source never finds the contradictions, and cannot back up the
agreements.

**A measure is only usable if it is defined before you collect it.** For each one,
write down:

- **what exactly is recorded**, such as "a death event, with its cause";
- **when**, such as "on every death, timestamped from session start";
- **in what unit**, such as seconds, metres or a 1 to 5 rating;
- **with which identifiers**, so that it can be joined to everything else.

The last one is where telemetry most often fails. A death event with no session ID
cannot be matched to the player's survey; a survey with no session ID cannot be
matched to the match it describes. *Tidebreak* puts `session_id` and `player_id`
on every row of every file for exactly this reason.

---

## 3. Choose a Design: Who Plays What

The moment you compare two versions of anything, you have to decide who plays
which. There are two basic designs.

```{mermaid}
%%| label: fig-designs
%%| fig-cap: "Between-subjects compares two groups of players; within-subjects compares each player with themselves."
flowchart LR
  subgraph BETWEEN["Between-subjects"]
    direction TB
    G1["Group 1"] --> BA["Build A"]
    G2["Group 2"] --> BB["Build B"]
  end
  subgraph WITHIN["Within-subjects"]
    direction TB
    G3["Every player"] --> WA["Build A"]
    G3 --> WB["Build B"]
  end
  BETWEEN ~~~ WITHIN
```

**Between-subjects:** different players test each version. *Tidebreak*'s main
file is between-subjects: `r n_players` players each played `r per_player`
sessions, `r players_a` of them only on Build A and `r players_b` only on
Build B. Nobody saw both.

**Within-subjects:** every player tests every version. *Tidebreak*'s paired file is
within-subjects: `r nrow(paired)` players each played both builds, and their wrong
turns were recorded on each.

| | Between-subjects | Within-subjects |
| :- | :- | :- |
| Each player sees | one version | every version |
| Main strength | no carry-over from one version to the next | each player is compared with themselves, so differences in skill cancel out |
| Main weakness | differences between the groups can masquerade as differences between builds | order, learning and fatigue effects |
| Players needed | more | fewer |
| Use when | playing one version changes how you play the other, such as a puzzle you can only solve once | the versions can be played fresh, such as two control schemes |

**Why within-subjects is so powerful.** Players differ enormously before they
touch your build: some are better, faster or more patient than others. In a
between-subjects test, all of that variation sits on top of the difference you
are trying to see. In a within-subjects test, each player's own skill appears in
both measurements and cancels out. You will see exactly how much this matters in
the paired test of [Comparing Groups](08-comparing-groups.qmd).

### Random assignment

In a between-subjects test, **you decide who plays which version, by chance**.
Never let players choose, and never put the first ten through the door on Build A.
Players who choose, or who arrive early, differ from the rest in ways that will
show up in the results and be credited to the build. Random assignment does not
make the groups identical, but it makes any difference between them an accident
rather than a pattern.

### Order, learning and fatigue

A within-subjects test has the opposite problem. If every player plays A and then
B, then B gets players who have **learned** the controls and the map, and who are
more **tired** and more **bored**. Any of these can make B look better or worse
than it is. These are **order effects**.

The fix is **counterbalancing**: half the players play A first, half play B first,
decided at random.

```{mermaid}
%%| label: fig-counterbalance
%%| fig-cap: "Counterbalancing gives half the players each build first, so order effects fall equally on both builds instead of favouring one."
flowchart LR
  S["Players,<br/>split at random"] --> H1["Half 1"]
  S --> H2["Half 2"]
  H1 --> A1["Build A"] --> B1["Build B"]
  H2 --> B2["Build B"] --> A2["Build A"]
```

Learning and fatigue still happen, but now they happen equally to both builds, so
they no longer favour either one.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Going further: more than two versions"}
With three versions there are six possible orders (ABC, ACB, BAC, BCA, CAB, CBA).
Using all six equally is **full counterbalancing**. With four versions there are 24
orders, which quickly becomes impractical; a **Latin square** uses just four
orders, chosen so that each version appears in each position exactly once. The
idea is the same: no version is systematically first, or last.
:::

---

## 4. How Many Players? {.when-needed}

There is no single answer, because there are two different jobs a playtest can do.

**Finding problems needs few players.** If a door is hard to find, most players
will struggle with it, and you will see it in the first handful of sessions. A
well-known rule of thumb from usability research (Nielsen) is that around five
testers find most of the *obvious* problems in an interface. Each extra tester
mostly rediscovers what you already know.

**Measuring how often needs many.** "What proportion of players quit in
frustration?" is a different question. With five players, one person changes the
answer by 20 percentage points. The figure below takes *Tidebreak*'s
`r n_a` Build A sessions, in which `r pct_fmt(frus_a / n_a)` ended in frustration,
and draws small samples from them over and over to see what a smaller playtest
would have reported.

```{r}
#| label: fig-sample-size
#| echo: false
#| fig-cap: "The frustration rate a smaller playtest would have reported, from 1,000 random samples of Build A sessions at each size. The dashed line is the rate across all of them."
#| fig-alt: "Three horizontal strips of points, one per sample size. With 5 sessions the estimates are spread across most of the axis; with 15 they cluster more tightly; with 40 they sit close to the dashed line marking the true rate."
#| fig-height: 3.5
frus_flag <- sessions$quit_reason[sessions$build == "A"] == "Frustrated"
set.seed(2026)
sizes <- c(5, 15, 40)
est <- lapply(sizes, function(n) replicate(1000, mean(sample(frus_flag, n))))
names(est) <- paste(sizes, "sessions")
mid90 <- sapply(est, function(e) quantile(e, c(0.05, 0.95)))

par(mar = c(4, 7, 1, 1))
stripchart(est, method = "jitter", jitter = 0.25, pch = 16, cex = 0.5,
           col = adjustcolor(pal$line, alpha.f = 0.25), las = 1,
           xlab = "Reported frustration rate", xaxt = "n", xlim = c(0, 1))
axis(1, at = seq(0, 1, 0.2), labels = paste0(seq(0, 100, 20), "%"))
abline(v = mean(frus_flag), lty = 2, col = pal$highlight, lwd = 2)
```

With 5 sessions, the middle 90% of samples reported anything from
`r pct_fmt(mid90[1, 1])` to `r pct_fmt(mid90[2, 1])`. With 40, from
`r pct_fmt(mid90[1, 3])` to `r pct_fmt(mid90[2, 3])`. The true rate never changed;
only the number of sessions did.
[From Sample to Population](07-sample-to-population.qmd) explains why the spread
shrinks the way it does.

**Count players, not sessions.** *Tidebreak* recorded `r nrow(sessions)` sessions,
but only `r n_players` people. Four sessions from one unusually skilled player are
not four independent pieces of evidence. When you plan a sample size, plan the
number of people.

---

## 5. Who to Recruit

**The players you test should resemble the players you are making the game for.**
A playtest recruited from the studio's Discord samples its most committed fans. A
playtest of colleagues samples people who already know the controls. A playtest of
classmates samples people of one age with one set of habits. Each answers the
question "what do *these* people think?", which may not be the question you asked.

Decide in advance who your intended players are, and recruit to match. Record the
things about each player that could affect the results, such as experience with
the genre, so you can check them later. *Tidebreak* records each player's
`cohort`: Novice, Regular or Veteran.

No analysis afterwards corrects for who walked through the door. What you can do is
say who they were, in the report's limitations section.

---

## 6. Ethics and Consent

A playtest collects data from real people, and that comes with obligations. Your
institution or studio will have its own process; the principles behind it are the
same everywhere.

**Informed consent.** Before they play, tell testers what you are recording
(gameplay, screen, voice, survey answers), why, how long you will keep it, and who
will see it. Get their agreement, in writing.

**The right to stop.** Testers can stop at any time, without giving a reason, and
can ask for their data to be removed.

**Collect only what you need.** If the question is about navigation, you do not
need a tester's date of birth. Every extra field is a risk with no benefit.

**Separate identity from data.** Store names and contact details in one place and
gameplay data in another, linked only by a code such as `P01`. *Tidebreak*'s files
contain no names for this reason.

**Telemetry is personal data.** Under GDPR, data that can be linked back to an
identifiable person is personal data, and that includes gameplay logs tied to an
account or a player code you can decode. Coded data is still personal data while
the key exists; only data that can never be traced back to a person is anonymous.
Store it securely and delete it when you no longer need it.

**Take extra care with under-18s.** Testing with children needs a parent's or
guardian's consent as well as the child's own agreement.

---

## 7. Decide in Advance, then Pilot

**Write down the result that would change your mind**, before you collect anything:
"If more than a third of testers never find the second key, I will add a light
cue."

This costs nothing and protects you from the most comfortable mistake in the
discipline: running the test, getting an ambiguous result, and finding in it
whatever you already believed. If the threshold is written down beforehand, the
data gets to answer. If it is not, you will answer and attribute it to the data
([rule 1](ten-rules.qmd)).

**Then run a pilot**: the whole playtest, start to finish, with one or two people
who are not on the team. A pilot finds the problems that would otherwise ruin the
real test: a logger that drops the session ID, a survey question everyone
misreads, a session that runs twenty minutes over, a build that crashes on the lab
machines. Fix them, then begin.

**Quick check.** A team wants to know whether a new minimap helps players
navigate. They let testers choose whether to play with the minimap on or off. What
is wrong, and what should they do?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answer"}
Players who choose the minimap probably differ from those who do not: perhaps they
are less confident navigators. Any difference in wrong turns could be the players,
not the minimap. Assign the minimap at random, or better, have every player try
both, counterbalanced.
:::

---

## Worked Example

**The question as asked.** The *Tidebreak* team ran `r nrow(sessions)` sessions,
`r n_a` on Build A and `r n_b` on Build B. The designer asks "is Build A worse?",
which cannot be answered as asked: worse at what, for whom, by how much?

**The measures.** Two, from two different sources:

| Source | Measure | Recorded per session as |
| :- | :- | :- |
| Telemetry | Did the session end in a frustration quit? | `quit_reason` is `Frustrated` |
| Self-report | "I always knew what my objective was" | A rating from 1 to 5 |

**The design.** Between-subjects: each player was assigned to one build and played
it `r per_player` times. Different players on each build means differences between
the two groups of players are a possible explanation for any result, and the
report will have to say so.

**The threshold, written first.** Before anyone opens the file, the team writes:
*"If more than a quarter of the sessions on either build end in a frustration
quit, that build does not ship."*

**The result.** `r frus_a` of `r n_a` Build A sessions ended in a frustration quit
(`r pct_fmt(frus_a / n_a)`), against `r frus_b` of `r n_b` on Build B
(`r pct_fmt(frus_b / n_b)`). Build A crosses the line; Build B does not. Nobody has
to argue about it, because the argument happened before the data arrived.

**The second source.** Of the `r ncol(survey) - 1` survey items, the objective
item has the lowest mean, `r round(item_means[["objective_clear"]], 2)` on the 1 to
5 scale. Telemetry says *what* happened: players on Build A left in frustration.
The survey offers a candidate for *why*: players did not know what they were meant
to be doing. Neither source alone would have pointed a designer at the objective
marker.

---

## Interpretation Guidance

Every finding is read at three levels, and the discipline is keeping them apart.

**What the data shows.** `r frus_a` of `r n_a` Build A sessions ended in a
frustration quit, against `r frus_b` of `r n_b` on Build B. This is a count. It is
not in dispute.

**What the analysis suggests.** Something in Build A frustrates players that Build
B avoided, and the survey's weakest item points at objective clarity as a
candidate.

**What can legitimately be concluded.** Build A fails the threshold the team set
in advance. That is all. The survey figure is averaged across both builds, so on
its own it cannot say the unclear objective belongs to Build A. And because the
design is between-subjects, a skeptic can ask whether Build A's players were
simply different. Linking the objective to the frustration is a hypothesis for the
next test, not a finding.

---

## Common Mistakes

**Collecting first, asking later.** Logging everything and hoping a question
emerges produces large files and no decisions. Write the question down first
([rule 2](ten-rules.qmd)).

**Trusting one source.** A survey on its own tells you what players believe; a log
on its own tells you what they did; your notes tell you what you happened to see.
Treat any one alone as part of the evidence.

**Letting players choose their condition.** The people who pick an option differ
from those who do not, and the difference gets credited to the option.

**Running every player in the same order.** Without counterbalancing, the second
build always gets more practised, more tired players.

**Counting sessions as players.** Four sessions from one player are one person's
evidence, four times.

**Testing only people you know.** Friends and colleagues are polite, practised and
unrepresentative.

**Logging without identifiers.** A death event with no session or player ID cannot
be joined to anything.

**Skipping the pilot.** The first real session should not be the one where you
discover the logger is broken.

**Setting the threshold after seeing the result.** A threshold chosen afterwards
is always met by exactly the right margin. If it was not written down beforehand,
it is not a threshold ([rule 1](ten-rules.qmd)).

---

## Practice

Each takes two or three minutes.

1. **Make it measurable.** Rewrite "is the corridor too dangerous?" as one
   telemetry measure, one thing an observer would watch for, and one self-report
   measure, each recordable for a single player.
2. **Choose a design.** You want to compare two control schemes for a racing game,
   and you can recruit 12 players for an afternoon. Between or within? What must
   you do to make it fair?
3. **Critique.** A student team tests their puzzle game on 30 classmates in one
   afternoon. Everyone plays the old version, then the new one. The new version
   takes 40% less time to solve. List three reasons the new version might not
   actually be better.
4. **Write the threshold.** Count the sessions that ended in a technical fault,
   then write the threshold that would make you delay a release over it.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
1. Telemetry: deaths in the corridor per session. Observation: does the player
   stop, retreat or look around before going in? Self-report: agreement with "I
   understood why I died". Players who die there often *and* do not understand why
   describe a different problem from players who die there often and accept it.
2. Within-subjects. Control schemes can be played fresh, twelve players is too few
   to split into two groups, and comparing each player with themselves cancels out
   the large differences in driving skill. Counterbalance: six players, chosen at
   random, try scheme A first; the other six try B first.
3. Any three of: players had already solved the puzzles once, so the second
   attempt was always going to be faster (learning, and no counterbalancing; and
   puzzles are a case for between-subjects); classmates are not the target
   audience; the team knew the testers, who may have tried harder on the version
   they knew was new; there was no threshold written in advance; and nobody
   piloted the test, so problems in the first version's build or logging may have
   slowed it down.
4. `r sum(sessions$quit_reason == "Technical fault")` of `r nrow(sessions)`
   sessions ended in a technical fault. Any threshold is defensible if it was
   written first and names a number: "if more than one session in twenty ends in a
   technical fault, we do not release". "If it seems like a lot" is not a
   threshold.
:::

---

## Summary

- Measurement earns its place where taste fails: you cannot see your own game, feel
  a rare event, or remember twenty playtests.
- A question is measurable when you can say how you would record it for one
  player.
- Telemetry says what happened, observation says how it unfolded, and self-report
  says what players experienced. Where they agree the finding is strong; where
  they disagree there is something to find.
- Between-subjects designs avoid carry-over; within-subjects designs cancel out
  differences between players. Assign at random, and counterbalance the order.
- A few players find problems. Measuring how often needs many more. Count people,
  not sessions.
- Recruit players like your audience, get informed consent, collect only what you
  need, and treat telemetry as personal data.
- Write the threshold that would change your mind before you look, then pilot.

| Key term | Meaning |
| :- | :- |
| **Telemetry** | What the game records about itself while it runs |
| **Observation** | What you record by watching someone play |
| **Self-report** | What players tell you about their experience |
| **Triangulation** | Checking a finding against more than one source of evidence |
| **Measure** | A specific value you could record for one player |
| **Between-subjects** | Each player tests one version |
| **Within-subjects** | Every player tests every version |
| **Random assignment** | Deciding by chance who plays which version |
| **Order effect** | A difference caused by what a player did first, not by the version |
| **Counterbalancing** | Varying the order so order effects fall equally on every version |
| **Pilot** | A full rehearsal of the playtest with one or two people |
| **Threshold** | The result, written in advance, that would change your decision |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 1 (decide what would change your mind),
  rule 2 (a number without a question is decoration) and rule 8 (small samples lie
  confidently)
- [Statistics Reference: sample and population](../reference/reference-statistics.qmd#describing-data)
- [Lab 00: Setting Up](../labs/setting-up.qmd) and
  [Lab 01: Reading Data](../labs/reading-data.qmd)
- Next: [R from First Principles](02-r-first-principles.qmd)

---

## Reflective Questions

- Pick a mechanic in your own project. Write the designer's question, then three
  different measures that would bear on it, then say which one you would trust
  most and why.
- Name something about your own game that you can no longer judge, because you
  have played it too often.
- Your next playtest compares two versions of a boss fight. Would you run it
  between or within subjects? What would a skeptic say about your choice?
- Your team wants to log every input a player makes "in case it is useful". What
  would you say, as the person responsible for the testers' data?
