---
title: "R from First Principles"
subtitle: "What R is, why we use it, and how to read and write it"
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")

# Draw a vector as a row of boxes, with positions above and values inside.
# `keep` highlights the elements an index selects.
draw_vec <- function(values, y, keep = rep(TRUE, length(values)), label = "") {
  n <- length(values)
  for (i in seq_len(n)) {
    rect(i - 0.45, y - 0.3, i + 0.45, y + 0.3,
         col = if (keep[i]) pal$highlight else "white",
         border = if (keep[i]) pal$highlight else pal$guide)
    text(i, y, values[i], col = if (keep[i]) "white" else pal$line, font = 2)
    text(i, y + 0.45, i, cex = 0.7, col = pal$guide)
  }
  mtext(label, side = 2, at = y, las = 1, line = 0.5, family = "mono", cex = 0.85)
}
```

> R is the language every analysis in this module is written in. This page
> introduces it from nothing: what it is, why a games team would use it, where you
> type it, and the dozen ideas that everything else is built from. It is written
> for programmers, so it says where R behaves like C# and, more usefully, where it
> does not. Read it before Lab 01; come back to it whenever R surprises you.

::: {.at-a-glance}
**Time:** pre-reading, about 45 minutes with R open beside you.
**Before this:** R and an editor installed ([Lab 00](../labs/setting-up.qmd)).
**Key terms:** console, script, Quarto document, project, object, assignment,
vector, index, function, argument, `NA`, data frame, package.
**R introduced:** `<-`, `c()`, `class()`, `length()`, `[ ]`, `seq()`, `rep()`,
`sum()`, `mean()`, `is.na()`, `data.frame()`, `$`, `str()`, `head()`,
`summary()`, `library()`, `read_csv()`, `args()`, `?`.
:::

## Learning Outcomes

By the end of these notes you can:

- explain why a games team would analyse playtest data in R rather than a
  spreadsheet, and why this module teaches base R before the tidyverse;
- say what the console, a script, a Quarto document and a project are for;
- create objects, check their type, and predict what R does when types mix;
- build, combine and index vectors, including with a logical condition;
- call a function with positional and named arguments, and find its help page;
- handle missing values in a calculation;
- build a data frame, inspect it, and pull out columns and rows;
- install and load a package, read a CSV, and read an error message for what it
  is telling you.

---

## Why This Matters for Playtesting

A playtest produces a few thousand rows of telemetry, a survey export and a frame
log, and next week it produces them again for the new build. The question is not
whether you can analyse them once. It is whether you can analyse them **again**,
identically, in a minute, when the build changes.

That is the case for writing your analysis as code. A spreadsheet records the
answer; a script records **how you got it**, so it can be checked, repeated on
next week's data, and handed to a teammate who gets exactly the same result.

---

## Why R?

**R is a language built for data.** Where most languages treat a table of numbers
as something you have to build, R starts there. Taking the mean of a column,
counting categories or drawing a histogram is one line:

```{r}
#| label: fig-one-line
#| fig-cap: "One line of R: the distribution of every Tidebreak session length."
#| fig-alt: "Histogram of session length in minutes, with most sessions between 20 and 32 minutes and a smaller group under 15."
#| fig-height: 3.5
hist(sessions$session_minutes, col = pal$fill, border = "white",
     main = "", xlab = "Session length (minutes)")
```

**Against a spreadsheet.** Spreadsheets are excellent for looking at a small table
and poor at everything this module needs: they hide the steps, so nobody can check
how a number was made; they invite editing the data by hand, so the original is
lost; and they have to be redone, click by click, when new data arrives. A
widely cited 2010 economics paper by Reinhart and Rogoff was later found to
contain a spreadsheet formula that left five countries out of an average, an error
nobody spotted for three years because the steps were hidden in cells. Code does
not prevent mistakes, but it leaves them where someone can find them.

**Against Python.** Python is the other obvious choice and a perfectly good one;
its data libraries can do everything in these notes. R is used here because its
statistics and plotting are built in rather than added, because Quarto and R work
together with no setup, and because most statistics teaching material is written
in it. Everything you learn transfers: a data frame, a vector and a grouped
summary are the same ideas in both.

**R is free, and it is reproducible.** Anyone can install it, every analysis in
these notes runs on any machine, and a Quarto document re-runs its own code every
time it is rendered, so its numbers cannot drift away from its data.

### Base R first, then the tidyverse

R comes with a large set of built-in functions, called **base R**. On top of it,
most modern R uses the **tidyverse**: a family of packages (`readr`, `dplyr`,
`ggplot2` and others) with a consistent, readable style.

This module teaches base R first and the tidyverse afterwards: `tapply()` before
`group_by()`, `hist()` before `ggplot()`. That is deliberate. The tidyverse is
easier to read and harder to understand; base R shows you what is actually
happening. If you later read *R for Data Science*, you will find it teaches the
other way round, starting with `ggplot2`. Both orders work; this one means the
tidyverse never feels like magic.

---

## Where You Write R

There are four places R code lives, and each has a job.

```{mermaid}
%%| label: fig-where-code-lives
%%| fig-cap: "The console is for trying things. Scripts and Quarto documents are for keeping them. A project holds it all together."
flowchart LR
  subgraph PROJECT["Project folder"]
    direction TB
    S["Script<br/>analysis.R"]
    Q["Quarto document<br/>report.qmd"]
    D[("data/<br/>CSV files")]
  end
  C["Console<br/>type, press Enter,<br/>see the answer"]
  S -->|"run a line<br/>Ctrl+Enter"| C
  Q -->|"run a chunk"| C
  Q -->|"Render"| H["Web page<br/>report.html"]
  D -.->|"read_csv()"| S
  D -.->|"read_csv()"| Q
  classDef keep fill:#f8f9fa,stroke:#6c757d,color:#212529
  classDef out fill:#fbeaea,stroke:#b22222,color:#212529
  class S,Q,D keep
  class C,H out
```

**The console** is where R runs. Type a line, press Enter, see the result. It is
for trying things out, and nothing typed there is kept.

**A script** (`.R` file) is a text file of R code. You write lines in it and send
them to the console with `Ctrl+Enter` (`Cmd+Enter` on macOS) in both VS Code and
RStudio. A script is your analysis: if it is not in the script, it did not happen.

**A Quarto document** (`.qmd` file) mixes prose and R code. Rendering it runs
every chunk in order and produces a web page with the text, code, numbers and
plots together. Your playtest report is one of these.

**A project** is the folder that holds everything: scripts, documents and a
`data/` folder. Working inside the project means R looks for files relative to
that folder, which is why every path in these notes starts `data/`.

---

## Values and Objects

At its simplest, R is a calculator.

```{r}
#| label: calc
2 + 3
1370 / 4
2^10
```

The `[1]` in front of each answer is not part of it. It is R telling you this line
of output starts with the first value; you will see why it matters when an answer
has many values.

**To keep a value, give it a name** with the assignment arrow `<-`, read aloud as
"gets":

```{r}
#| label: assign
score <- 1370
score
score * 2
```

`score` is now an **object**: a named value that lives in your R session until you
close it or overwrite it. Names are case-sensitive (`Score` is a different object),
cannot start with a number, and should say what the value is. `session_minutes`
is a good name; `x2` is not.

::: {.callout-note collapse="true" title="Going further: `<-` or `=`?"}
`score = 1370` also works, and R programmers argue about which is better. These
notes use `<-` for assignment and keep `=` for naming arguments inside a function
call, such as `mean(x, na.rm = TRUE)`. That way, `=` always means "this argument
gets this value" and `<-` always means "this object gets this value". In RStudio,
`Alt+-` types the arrow; VS Code can be given the same shortcut in its keyboard
settings.
:::

---

## Types

Every value has a **type**. `class()` tells you which.

```{r}
#| label: types
class(1370)
class("Scout")
class(TRUE)
```

| Type | What it holds | Examples |
| :- | :- | :- |
| **numeric** | numbers, whole or decimal | `1370`, `24.6` |
| **character** | text, always in quotes | `"Scout"`, `"S001"` |
| **logical** | `TRUE` or `FALSE` | the answer to any comparison |

**Comparisons produce logicals.** `==` tests equality (a single `=` is not a
comparison), `!=` tests inequality, and `<`, `>`, `<=`, `>=` work as you would
expect:

```{r}
#| label: compare
score > 1500
score == 1370
"Scout" == "scout"
```

The last one is `FALSE` because text comparison is case-sensitive, a fact that
matters a great deal when you come to clean real data.

**R does not make you declare types.** If you come from C#, this is the first big
difference: there is no `int score = 1370;`. R works the type out from the value,
and it will convert between types without asking. That is convenient until it is
not:

```{r}
#| label: coerce
#| warning: true
as.numeric("24.6")
as.numeric("twenty")
```

The second line returns `NA`, R's marker for a missing value, with a warning. It
does not stop. A column of numbers with one typo in it can quietly become a column
of text, and every calculation on it then fails or misleads. Checking types is the
first thing you do with any new data.

::: {.callout-note collapse="true" title="Going further: integers"}
R has a separate **integer** type for whole numbers, written with an `L`: `4L`.
You will rarely need to create one, but you will see it: base R's `read.csv()`
stores whole-number columns as integers, while `readr`'s `read_csv()` stores all
numbers as numeric (printed `<dbl>`, for double precision). For analysis the
difference almost never matters.
:::

---

## Vectors

In C#, a single number is the basic unit and a collection is something you build.
In R it is the other way round. **The basic unit is the vector**: an ordered
collection of values, all of the same type. A single number is just a vector of
length one, which is what the `[1]` was telling you.

Build a vector with `c()`, for *combine*:

```{r}
#| label: vector
deaths <- c(4, 2, 7, 3, 0, 5)
deaths
length(deaths)
```

```{r}
#| label: fig-vector
#| echo: false
#| fig-cap: "A vector: values in order, each with a position. R counts positions from 1."
#| fig-alt: "Six boxes in a row containing 4, 2, 7, 3, 0 and 5, with positions 1 to 6 written above them."
#| fig-height: 1.3
par(mar = c(0, 6, 0, 1))
plot.new(); plot.window(xlim = c(0.4, 6.6), ylim = c(0.6, 1.6))
draw_vec(deaths, 1, keep = rep(FALSE, 6), label = "deaths")
```

**Arithmetic works on every element at once.** This is called **vectorisation**,
and it replaces most of the loops you would write in C#:

```{r}
#| label: vectorised
deaths + 1
deaths * 10
```

**Summary functions collapse a vector to one value:**

```{r}
#| label: summaries
sum(deaths)
mean(deaths)
max(deaths)
```

**A comparison on a vector gives one logical per element**, and because R counts
`TRUE` as 1 and `FALSE` as 0, summing or averaging that vector counts it:

```{r}
#| label: logical-count
deaths > 2
sum(deaths > 2)
mean(deaths > 2)
```

`sum(deaths > 2)` is "how many sessions had more than two deaths";
`mean(deaths > 2)` is "what proportion did". This one idea produces a large share
of the numbers in a playtest report.

**Every element has the same type.** Mix them and R converts everything to the
most general type, usually character:

```{r}
#| label: mixed
c(1, "two", 3)
```

The numbers are now text, in quotes. This is the same silent conversion as before,
and it is why a vector, and every column of a table, holds one type only.

**Handy vector builders:**

```{r}
#| label: builders
1:6
seq(0, 30, by = 10)
rep("A", 3)
```

---

## Indexing

Square brackets pull elements out of a vector. If you write C# or C++, the one
mistake you will make is this: **R counts from 1, not 0.**

```{r}
#| label: index
deaths[1]
deaths[c(3, 5)]
deaths[-1]
deaths[deaths > 2]
```

```{r}
#| label: fig-index
#| echo: false
#| fig-cap: "Four ways to index the same vector. Highlighted boxes are the elements each one returns."
#| fig-alt: "Four rows of the deaths vector. deaths[1] highlights the first box; deaths[c(3, 5)] highlights the third and fifth; deaths[-1] highlights all but the first; deaths[deaths > 2] highlights the boxes containing 4, 7, 3 and 5."
#| fig-height: 3.6
par(mar = c(0, 11, 0, 1))
plot.new(); plot.window(xlim = c(0.4, 6.6), ylim = c(0.5, 4.7))
draw_vec(deaths, 4, keep = seq_along(deaths) == 1,        label = "deaths[1]")
draw_vec(deaths, 3, keep = seq_along(deaths) %in% c(3, 5), label = "deaths[c(3, 5)]")
draw_vec(deaths, 2, keep = seq_along(deaths) != 1,         label = "deaths[-1]")
draw_vec(deaths, 1, keep = deaths > 2,                     label = "deaths[deaths > 2]")
```

| Index | Means | Returns |
| :- | :- | :- |
| `[1]` | the element at position 1 | one value |
| `[c(3, 5)]` | the elements at positions 3 and 5 | two values |
| `[-1]` | every element **except** position 1 | five values |
| `[deaths > 2]` | the elements where the condition is `TRUE` | the values above 2 |

**The last one is the most important.** A **logical index** keeps the elements
where a condition holds, and the condition can come from a different vector of the
same length. If `archetype` holds each session's archetype, then
`deaths[archetype == "Scout"]` is the deaths for Scout sessions only. That is how
you filter data in base R.

---

## Functions and Arguments

A **function** takes inputs, called **arguments**, and returns a result. You have
already used several: `c()`, `class()`, `sum()`, `mean()`.

**Arguments have names and an order.** `args()` shows both:

```{r}
#| label: args
args(round)
```

`round()` takes `x`, the number, and `digits`, which defaults to 0. You can supply
arguments **by position**, in the order the function declares them, or **by
name**, in any order:

```{r}
#| label: named
round(3.14159, 2)
round(x = 3.14159, digits = 2)
round(digits = 2, x = 3.14159)
```

All three give the same answer. Without names, position is everything, and
swapping the values changes the meaning:

```{r}
#| label: swapped
round(2, 3.14159)
```

That rounds 2 to three decimal places. **The usual style is to pass the main input
by position and everything else by name**: `mean(deaths, na.rm = TRUE)`,
`seq(0, 30, by = 10)`. It reads well, and it lets you skip arguments you are happy
to leave at their defaults.

**Every function has a help page.** Type `?round` in the console. The page lists
every argument, its default, what the function returns, and examples at the
bottom, which are usually the quickest way to understand it.

::: {.callout-note collapse="true" title="Going further: writing your own function"}
You can define functions with `function()`. This one turns a logical vector into a
percentage, with a default of one decimal place:

```{r}
#| label: own-function
pct <- function(x, digits = 1) {
  round(100 * mean(x), digits)
}

pct(deaths > 2)
pct(deaths > 2, digits = 0)
```

The last expression in the body is the return value; there is no `return`
keyword needed. You will not need to write functions for the labs, but it is how
R programmers stop repeating themselves.
:::

---

## Missing Values

`NA` means a value should be there but was not recorded: a session length the
logger lost, a survey question the player skipped. It is not zero and not empty.
It means **unknown**.

```{r}
#| label: na
minutes <- c(12.9, 24.6, NA, 18.8)
mean(minutes)
```

**Any calculation involving an unknown is unknown.** R will not guess. Tell it to
leave the missing values out with `na.rm = TRUE`, and find them with `is.na()`:

```{r}
#| label: na-rm
mean(minutes, na.rm = TRUE)
is.na(minutes)
sum(is.na(minutes))
```

`sum(is.na(x))` is the count of missing values: the same "sum a logical to count
it" trick as before. Before you drop missing values, ask why they are missing.
Sometimes the pattern of what is missing is itself a finding, which is the subject
of [Cleaning and Transforming](06-cleaning-and-transforming.qmd).

---

## Data Frames

A playtest table is a **data frame**: a set of vectors of equal length, one per
column, side by side. Each column has one type; each row is one observation.

```{mermaid}
%%| label: fig-dataframe
%%| fig-cap: "A data frame is a set of equal-length vectors, one per column."
flowchart LR
  V1["session<br/>S001 S002 S003"] --> DF[("data frame<br/>3 rows × 3 columns")]
  V2["archetype<br/>Scout Tank Scout"] --> DF
  V3["deaths<br/>4 2 7"] --> DF
  classDef v fill:#f8f9fa,stroke:#6c757d,color:#212529
  classDef d fill:#fbeaea,stroke:#b22222,color:#212529
  class V1,V2,V3 v
  class DF d
```

You can build one by hand from vectors:

```{r}
#| label: df-build
mini <- data.frame(
  session   = c("S001", "S002", "S003", "S004"),
  archetype = c("Scout", "Tank", "Scout", "Support"),
  deaths    = c(4, 2, 7, 3)
)
mini
```

**Look at a data frame without printing it all.** A real playtest file has
hundreds of rows. These four functions are the first things you run on any table:

```{r}
#| label: df-inspect
nrow(mini)
names(mini)
str(mini)
```

`str()`, for *structure*, is the most useful: it gives the size, every column's
name, its type and its first few values in one go. `head()` shows the first six
rows, and `summary()` gives a quick numeric summary of every column.

**`$` pulls out one column**, as a vector:

```{r}
#| label: df-dollar
mini$deaths
mean(mini$deaths)
```

**`[rows, columns]` slices.** Leave a side empty to mean "all of them". The comma
is not optional: it is what separates rows from columns.

```{r}
#| label: df-slice
mini[2, ]
mini[, c("session", "deaths")]
mini[mini$deaths > 3, ]
```

The last line is a logical index on rows: keep the rows where `deaths` is above 3.
It is exactly the vector idea from before, applied to a table.

**Add a column by assigning to it:**

```{r}
#| label: df-add
mini$high_deaths <- mini$deaths > 3
mini
```

---

## Packages and Reading Data

**A package** is a bundle of extra functions someone else has written. R's
strength is the thousands of them available. Using one is two steps:

```{r}
#| label: packages
#| eval: false
install.packages("readr")   # once per computer: downloads and installs it
library(readr)              # every session: makes its functions available
```

Installing is like adding a library to a project; loading is like the `using`
statement at the top of a C# file. Forget the `library()` call and R reports
`could not find function`, which is the most common error in the first week.

**Reading a CSV** is one line with `readr`'s `read_csv()`:

```{r}
#| label: read
library(readr)
sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
dim(sessions)
```

The path is relative to the project folder. `dim()` gives rows and columns:
`r nrow(sessions)` sessions, `r ncol(sessions)` variables. `read_csv()` returns a
**tibble**, the tidyverse's version of a data frame. Everything above works on it
unchanged; it just prints more tidily, showing the first ten rows and each column's
type:

```{r}
#| label: tibble-print
head(sessions[, c("session_id", "cohort", "build", "deaths", "session_minutes")])
```

`<chr>` is character and `<dbl>` is numeric. Whether those types match what each
column *means* is a separate question, and the subject of
[Data, Variables and Types](03-data-and-types.qmd).

::: {.callout-note collapse="true" title="Going further: `::` and `glimpse()`"}
`readr::read_csv()` calls `read_csv()` from the `readr` package without loading
the whole package first. You will see this notation in documentation and in these
notes, where it makes clear which package a function belongs to.

`dplyr::glimpse(sessions)` is a tidyverse alternative to `str()`: one line per
column, with its type and first few values, laid out to fit the screen.
:::

---

## Reading Error Messages

R's error messages are terse, but they nearly always name the problem. Read the
error before changing anything. Here are the ones you will meet most, produced
live:

```{r}
#| label: err-object
#| error: true
mean(sesions$score)
```

**`object not found`**: a typo in a name (`sesions`), or an object you have not
created yet in this session. Check spelling and capitals, and check you ran the
line that creates it.

```{r}
#| label: err-function
#| error: true
summarise_scores(sessions)
```

**`could not find function`**: a typo in a function name, or a package that is not
loaded. Add the `library()` call.

```{r}
#| label: err-binary
#| error: true
"24" + 1
```

**`non-numeric argument to binary operator`**: arithmetic on text. A number has
been read or typed as character. Check with `class()`.

```r
read_csv("tidebreak_sessions.csv", show_col_types = FALSE)
```

```
Error: 'tidebreak_sessions.csv' does not exist in current working directory:
'C:/Users/you/L8-3DGD-Playtest-Analytics'.
```

**`does not exist`** or **`cannot open file`**: the path is wrong. Here the
`data/` folder is missing from it. Paths are relative to the project folder.

**`unexpected symbol`**, **`unexpected string constant`** or
**`unexpected ')'`**: a syntax error, almost always a missing comma, bracket or
quote just before the point R names.

**A warning is not an error.** An error stops the code. A **warning** lets it
finish but tells you something may be wrong, as `as.numeric("twenty")` did above.
Never ignore a warning without reading it.

---

## Quarto in One Page

A Quarto document is a text file with three kinds of content:

````markdown
---
title: "Build B playtest report"
---

## Session length

Build B sessions were longer. The median was `r knitr::inline_expr("median(b$session_minutes)")` minutes.

```{{r}}
#| label: session-hist
hist(b$session_minutes)
```
````

**The header** between the `---` lines sets the title and options. **Prose** is
written in Markdown: `##` for a heading, `**bold**`, `*italic*`, `-` for a list.
**Code chunks** start with ```` ```{r} ```` and end with ```` ``` ````; lines
starting `#|` are chunk options, such as a label or whether to show the code.

**Inline code** puts a computed number straight into a sentence: the backtick-r
expression above becomes the actual median when the document is rendered. A number
written this way can never go out of date. Every number in these notes is produced
like that.

**Rendering** runs every chunk from the top, in order, in a fresh session, and
writes the web page: the Render button in either editor, or `quarto render` in a
terminal. Because it starts fresh, a document that renders proves the analysis
works from the data alone, not from something you happened to type in the console
earlier.

---

## Worked Example

**How many Build A sessions ended in frustration?** This is the first number in
[Planning a Playtest](01-planning-a-playtest.qmd). Here it is built from nothing,
one idea at a time.

**Load and look.** Every analysis starts the same way:

```{r}
#| label: worked-load
sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
str(sessions[, c("build", "quit_reason")])
```

**One condition gives a logical vector**, one value per session:

```{r}
#| label: worked-condition
is_a <- sessions$build == "A"
head(is_a)
sum(is_a)
```

Summing it counts the Build A sessions: `r sum(sessions$build == "A")`.

**Two conditions combine with `&`**, "and": `TRUE` only where both are true.

```{r}
#| label: worked-combine
frustrated_a <- sessions$build == "A" & sessions$quit_reason == "Frustrated"
sum(frustrated_a)
mean(frustrated_a[is_a])
```

`r sum(frustrated_a)` sessions. The last line takes the proportion among Build A
sessions only, using the logical index from earlier:
`r pct_fmt(mean(frustrated_a[is_a]))`.

**Check it a second way.** `table()` with two vectors counts every combination at
once:

```{r}
#| label: worked-table
table(sessions$build, sessions$quit_reason)
```

The `A` row of the `Frustrated` column says `r sum(frustrated_a)`, the same answer.
Getting a number two different ways is the cheapest check there is.

Nothing in this example is more than the ideas on this page: an object, a
comparison, a logical vector, `sum()`, `mean()`, an index and a function call.

---

## Reading What R Tells You

**`[1]` and friends are positions, not output.** When a long vector wraps across
lines, each line starts with the position of its first value: `[1]`, then `[19]`
and so on.

**`NA` in a result means an unknown went in.** Find it with `is.na()` before
reaching for `na.rm = TRUE`.

**A tibble prints its size and types first.** `# A tibble: 120 × 26` and the
`<chr>`/`<dbl>` row under the names are the first check on any table.

**Silence is success.** Assignment prints nothing. If you want to see an object,
type its name.

**Red text is not always an error.** Messages and warnings also print in red in
most editors. Read the first word: `Error` stops the code, `Warning` does not.

---

## Common Mistakes

**Using `=` to compare.** `if (x = 3)` is not a test. Comparison is `==`.

**Counting from 0.** `deaths[0]` returns an empty vector, not the first value.
R starts at 1.

**Getting the case wrong.** `Sessions`, `sessions` and `SESSIONS` are three
different objects, and `"Scout"` does not equal `"scout"`.

**Forgetting the quotes, or adding them.** `"build"` is the text *build*; `build`
is an object called build. `sessions$"build"` happens to work, but
`table("build")` counts the word, not the column.

**Forgetting `library()`.** Installed is not loaded. Every new session, and every
Quarto document, needs its own `library()` calls.

**Running lines out of order.** The console remembers everything you have ever run
in the session; a rendered document remembers only what is in it, from the top.
If it works in the console and fails when rendered, something is missing from the
document.

**Naming an object after a function.** `mean <- 5` or `data <- read_csv(...)`
works, but makes later code confusing. Pick names that describe the contents.

**Leaving out `na.rm = TRUE`, or adding it without thinking.** The first gives
`NA`; the second silently drops data you may need to explain.

---

## Practice

Each takes two or three minutes. Try each in the console before opening the
answer.

**1. Predict.** What does each line return? `c(1, 2, 3) * 2`, `sum(c(TRUE, FALSE,
TRUE))`, `c(5, 10, 15)[-2]`, `length(c("a", "b", NA))`.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-predict
c(1, 2, 3) * 2
sum(c(TRUE, FALSE, TRUE))
c(5, 10, 15)[-2]
length(c("a", "b", NA))
```

`NA` still occupies a position, so the last vector has length 3.
:::

**2. Fix it.** Each line has one error. Find it without running the line first.
`mean(session$deaths)`, `sessions[sessions$build = "A", ]`,
`Mean(sessions$deaths)`.

::: {.callout-note collapse="true" title="Check your answer"}
`session` should be `sessions` (object not found). `=` should be `==` inside the
condition (an error about an unexpected `=`). `Mean` should be `mean`: function
names are case-sensitive too (could not find function).
:::

**3. Compute.** How many *Tidebreak* sessions lasted longer than 30 minutes, and
what proportion is that?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-long
sum(sessions$session_minutes > 30)
mean(sessions$session_minutes > 30)
```

`r sum(sessions$session_minutes > 30)` sessions, or
`r pct_fmt(mean(sessions$session_minutes > 30))`. A comparison, then `sum()` to
count and `mean()` for the proportion.
:::

**4. Reorder.** `seq(from = 0, to = 100, by = 25)` produces 0, 25, 50, 75, 100.
Write it with the arguments in the order `by`, `to`, `from`. Then write it with no
names at all.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-seq
seq(by = 25, to = 100, from = 0)
seq(0, 100, 25)
```

With names, any order works. Without them, the values must follow the function's
own order, which `args(seq)` shows.
:::

---

## Summary

- R is a language built for data. Code records how you got an answer, so it can be
  checked and repeated; a spreadsheet records only the answer.
- Try things in the console; keep them in a script or Quarto document; keep
  everything in a project.
- `<-` creates an object. Every value has a type; R converts types silently, so
  check them.
- The vector is R's basic unit. Arithmetic is vectorised, and summing a logical
  vector counts it.
- R counts from 1. Index by position, by exclusion, or by a logical condition.
- Arguments go by position or by name. Pass the main input by position and options
  by name. `?name` opens the help.
- `NA` means unknown and spreads through calculations. `na.rm = TRUE` leaves it
  out; `is.na()` finds it.
- A data frame is a set of equal-length vectors. `str()` first, then `$` and
  `[rows, columns]`.
- `install.packages()` once, `library()` every session. Read the error message
  before changing anything.

| Key term | Meaning |
| :- | :- |
| **Console** | Where R runs code and prints results; nothing is kept |
| **Script** | A `.R` file of code you keep and re-run |
| **Quarto document** | A `.qmd` file of prose and code that renders to a web page |
| **Project** | The folder holding an analysis; paths are relative to it |
| **Object** | A named value, created with `<-` |
| **Vector** | An ordered set of values of one type |
| **Index** | A position, exclusion or condition in `[ ]` that selects elements |
| **Function** | Code that takes arguments and returns a result |
| **Argument** | An input to a function, given by position or by name |
| **`NA`** | A missing value: unknown, not zero |
| **Data frame** | A table of equal-length column vectors; a tibble is one kind |
| **Package** | A bundle of extra functions, installed once and loaded with `library()` |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 3 (know what kind of number you are holding)
  and rule 6 (cleaning is analysis)
- [R and Quarto Reference](../reference/r-quarto-reference.qmd): every function on
  this page, in one table
- [Lab 00: Setting Up](../labs/setting-up.qmd) and
  [Lab 01: Reading Data](../labs/reading-data.qmd), Part A
- Next: [Data, Variables and Types](03-data-and-types.qmd)

---

## Reflective Questions

- Think of an analysis you have done in a spreadsheet. What would a teammate need
  to repeat it exactly next week, and how much of that would be written down?
- Vectorised arithmetic replaces most loops. Write, in C#, what `sum(deaths > 2)`
  does, and count the lines.
- R converts types silently where C# would refuse to compile. Name one way that
  helps an analyst and one way it hurts.
- A teammate's Quarto report renders on their machine but not on yours. List three
  things that could cause it, from this page.
