---
title: "R from First Principles"
subtitle: "What R is, why we use it, and how to read and write it"
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")

# Draw a vector as a row of boxes, with positions above and values inside.
# `keep` highlights the elements an index selects.
draw_vec <- function(values, y, keep = rep(TRUE, length(values)), label = "") {
  n <- length(values)
  for (i in seq_len(n)) {
    rect(i - 0.45, y - 0.3, i + 0.45, y + 0.3,
         col = if (keep[i]) pal$highlight else "white",
         border = if (keep[i]) pal$highlight else pal$guide)
    text(i, y, values[i], col = if (keep[i]) "white" else pal$line, font = 2)
    text(i, y + 0.45, i, cex = 0.7, col = pal$guide)
  }
  mtext(label, side = 2, at = y, las = 1, line = 0.5, family = "mono", cex = 0.85)
}
```

> R is the language every analysis in this course is written in. This page
> introduces it from nothing: what it is, why a games team would use it, where you
> type it, and the handful of ideas everything else is built from. It is written
> for programmers, so it says where R behaves like C# and, more usefully, where it
> does not. Read it before Stage 1, and come back whenever R surprises you.

::: {.at-a-glance}
**Stage:** before Stage 1 · Core.
**Time:** pre-reading, about 45 minutes with R open beside you.
**Before this:** R and an editor installed ([Lab 00](../labs/setting-up.qmd)).
**Key terms:** console, script, Quarto document, project, object, assignment,
vector, index, function, argument, `NA`, data frame, package, pipe.
**R you will write:** `<-`, `c()`, `class()`, `length()`, `sum()`, `mean()`,
`is.na()`, `$`, `library()`, `read_csv()`, `glimpse()`, `|>`, `count()`, `?`.
:::

## Learning Outcomes

By the end of these notes you can:

- explain why a games team would analyse playtest data in code rather than a
  spreadsheet;
- say what the console, a script, a Quarto document and a project are for;
- create objects, check their type, and predict what R does when types mix;
- build vectors, do arithmetic on them, and count with a logical condition;
- call a function with positional and named arguments, and find its help page;
- handle missing values in a calculation;
- load a package, read a CSV, look at the result with `glimpse()`, and pull out a
  column;
- read an error message for what it is telling you.

---

## Why This Matters for Playtesting

A playtest produces a few thousand rows of telemetry, a survey export and a frame
log, and next week it produces them again for the new build. The question is not
whether you can analyse them once. It is whether you can analyse them **again**,
identically, in a minute, when the build changes.

That is the case for writing your analysis as code. A spreadsheet records the
answer; a script records **how you got it**, so it can be checked, repeated on
next week's data, and handed to a teammate who gets exactly the same result.

---

## Why R?

**R is a language built for data.** Where most languages treat a table of numbers
as something you have to build, R starts there. Taking the mean of a column,
counting categories or drawing a histogram is one line:

```{r}
#| label: fig-one-line
#| fig-cap: "One line of R draws the distribution of every Tidebreak session length."
#| fig-alt: "Histogram of session length in minutes, with most sessions between 20 and 32 minutes and a smaller group under 15."
#| fig-height: 3.5
hist(sessions$session_minutes, col = pal$fill, border = "white",
     main = "", xlab = "Session length (minutes)")
```

**Against a spreadsheet.** Spreadsheets are fine for looking at a small table and
poor at everything this course needs. They hide the steps, so nobody can check how
a number was made. They invite editing the data by hand, so the original is lost.
And they have to be redone, click by click, when new data arrives. A widely cited
2010 economics paper by Reinhart and Rogoff was later found to contain a
spreadsheet formula that left five countries out of an average, an error nobody
spotted for three years because the steps were hidden in cells. Code does not
prevent mistakes, but it leaves them where someone can find them.

**Against Python.** Python is the other obvious choice and a perfectly good one.
R is used here because its statistics and plotting are built in rather than added,
and because Quarto and R work together with no setup. Everything you learn
transfers: a data frame, a vector and a grouped summary are the same ideas in both.

### Base R and the tidyverse

R comes with a large set of built-in functions, called **base R**. On top of it,
most modern R uses the **tidyverse**: a family of packages (`readr`, `dplyr`,
`ggplot2` and others) with a consistent, readable style.

This course uses both, with one way to do each job:

- **Handling data:** the tidyverse. `read_csv()` to load it, then verbs such as
  `count()`, `filter()` and `summarise()`, chained with the pipe `|>`.
- **Summary statistics:** base R's `mean()`, `median()`, `sd()` and friends, which
  the tidyverse uses too.
- **Plots:** base R in Stages 1 and 2 (`hist()`, `boxplot()`, `plot()`), which are
  quick and good for looking while you think. `ggplot2` arrives in Stage 3, for
  figures other people will read.

**You do not need to memorise this:** code you find online will often do the same
jobs in other ways. Where that matters, these notes show you what the other way
looks like, so you can read it, and say that you do not need to write it.

---

## Where You Write R

There are four places R code lives, and each has a job.

```{mermaid}
%%| label: fig-where-code-lives
%%| fig-cap: "Try things in the console; keep them in a script or Quarto document, inside one project folder."
flowchart LR
  subgraph PROJECT["Project folder"]
    direction TB
    S["Script<br/>analysis.R"]
    Q["Quarto document<br/>report.qmd"]
    D[("data/<br/>CSV files")]
  end
  C["Console"]
  S -->|"run a line<br/>Ctrl+Enter"| C
  Q -->|"run a chunk"| C
  Q -->|"Render"| H["Web page<br/>report.html"]
  D -.->|"read_csv()"| S
  D -.->|"read_csv()"| Q
  class C,H accent
```

**The console** is where R runs. Type a line, press Enter, see the result. It is
for trying things out, and nothing typed there is kept.

**A script** (`.R` file) is a text file of R code. You write lines in it and send
them to the console with `Ctrl+Enter` (`Cmd+Enter` on macOS) in both VS Code and
RStudio. A script is your analysis: if it is not in the script, it did not happen.

**A Quarto document** (`.qmd` file) mixes prose and R code. Rendering it runs
every chunk in order and produces a web page with the text, code, numbers and
plots together. Your playtest report will be one of these; the full story is in
Stage 4, [Interpreting and Reporting Evidence](11-reporting-evidence.qmd#quarto-the-report-is-the-analysis).

**A project** is the folder that holds everything: scripts, documents and a
`data/` folder. Working inside the project means R looks for files relative to
that folder, which is why every path in these notes starts `data/`.

**For your project:** make one project folder for your team's playtest analysis
now, with a `data/` folder inside it. Every export from every playtest goes there,
unedited. You will thank yourself in March.

---

## Values and Objects

At its simplest, R is a calculator.

```{r}
#| label: calc
2 + 3
1370 / 4
2^10
```

The `[1]` in front of each answer is not part of it. It is R telling you this line
of output starts with the first value; you will see why it matters when an answer
has many values.

**To keep a value, give it a name** with the assignment arrow `<-`, read aloud as
"gets":

```{r}
#| label: assign
score <- 1370
score
score * 2
```

`score` is now an **object**: a named value that lives in your R session until you
close it or overwrite it. Names are case-sensitive (`Score` is a different object),
cannot start with a number, and should say what the value is. `session_minutes`
is a good name; `x2` is not.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Going further: `<-` or `=`?"}
`score = 1370` also works, and R programmers argue about which is better. These
notes use `<-` for assignment and keep `=` for naming arguments inside a function
call, such as `mean(x, na.rm = TRUE)`. That way, `=` always means "this argument
gets this value" and `<-` always means "this object gets this value". In RStudio,
`Alt+-` types the arrow; VS Code can be given the same shortcut in its keyboard
settings.
:::

---

## Types

Every value has a **type**. `class()` tells you which.

```{r}
#| label: types
class(1370)
class("Scout")
class(TRUE)
```

| Type | What it holds | Examples |
| :- | :- | :- |
| **numeric** | numbers, whole or decimal | `1370`, `24.6` |
| **character** | text, always in quotes | `"Scout"`, `"S001"` |
| **logical** | `TRUE` or `FALSE` | the answer to any comparison |

**Comparisons produce logicals.** `==` tests equality (a single `=` is not a
comparison), `!=` tests inequality, and `<`, `>`, `<=`, `>=` work as you would
expect:

```{r}
#| label: compare
score > 1500
score == 1370
"Scout" == "scout"
```

The last one is `FALSE` because text comparison is case-sensitive, which matters a
great deal when you come to clean real data.

**R does not make you declare types.** If you come from C#, this is the first big
difference: there is no `int score = 1370;`. R works the type out from the value,
and it will convert between types without asking. That is convenient until it is
not:

```{r}
#| label: coerce
#| warning: true
as.numeric("24.6")
as.numeric("twenty")
```

The second line returns `NA`, R's marker for a missing value, with a warning. It
does not stop. A column of numbers with one typo in it can quietly become a column
of text, and every calculation on it then fails or misleads. Checking types is the
first thing you do with any new data.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Going further: integers"}
R has a separate **integer** type for whole numbers, written with an `L`: `4L`.
You will rarely need to create one, but you will see it: base R's `read.csv()`
stores whole-number columns as integers, while `readr`'s `read_csv()` stores all
numbers as numeric (printed `<dbl>`, for double precision). For analysis the
difference almost never matters.
:::

---

## Vectors

In C#, a single number is the basic unit and a collection is something you build.
In R it is the other way round. **The basic unit is the vector**: an ordered
collection of values, all of the same type. A single number is just a vector of
length one, which is what the `[1]` was telling you.

Build a vector with `c()`, for *combine*:

```{r}
#| label: vector
deaths <- c(4, 2, 7, 3, 0, 5)
deaths
length(deaths)
```

```{r}
#| label: fig-vector
#| echo: false
#| fig-cap: "A vector holds values in order, each with a position. R counts positions from 1."
#| fig-alt: "Six boxes in a row containing 4, 2, 7, 3, 0 and 5, with positions 1 to 6 written above them."
#| fig-height: 1.3
par(mar = c(0, 6, 0, 1))
plot.new(); plot.window(xlim = c(0.4, 6.6), ylim = c(0.6, 1.6))
draw_vec(deaths, 1, keep = rep(FALSE, 6), label = "deaths")
```

**Arithmetic works on every element at once.** This is called **vectorisation**,
and it replaces most of the loops you would write in C#:

```{r}
#| label: vectorised
deaths + 1
deaths * 10
```

**Summary functions collapse a vector to one value:**

```{r}
#| label: summaries
sum(deaths)
mean(deaths)
max(deaths)
```

**A comparison on a vector gives one logical per element**, and because R counts
`TRUE` as 1 and `FALSE` as 0, summing or averaging that vector counts it:

```{r}
#| label: logical-count
deaths > 2
sum(deaths > 2)
mean(deaths > 2)
```

`sum(deaths > 2)` is "how many sessions had more than two deaths";
`mean(deaths > 2)` is "what proportion did". **You need to know** this one: it
produces a large share of the numbers in a playtest report.

**Every element has the same type.** Mix them and R converts everything to the
most general type, usually character:

```{r}
#| label: mixed
c(1, "two", 3)
```

The numbers are now text, in quotes. This is the same silent conversion as before,
and it is why a vector, and every column of a table, holds one type only.

---

## Indexing

Square brackets pull elements out of a vector. If you write C# or C++, the one
mistake you will make is this: **R counts from 1, not 0.**

```{r}
#| label: index
deaths[1]
deaths[c(3, 5)]
deaths[deaths > 2]
```

```{r}
#| label: fig-index
#| echo: false
#| fig-cap: "The same vector indexed three ways. A condition inside the brackets keeps the elements where it is TRUE."
#| fig-alt: "Three rows of the deaths vector. deaths[1] highlights the first box; deaths[c(3, 5)] highlights the third and fifth; deaths[deaths > 2] highlights the boxes containing 4, 7, 3 and 5."
#| fig-height: 2.8
par(mar = c(0, 11, 0, 1))
plot.new(); plot.window(xlim = c(0.4, 6.6), ylim = c(0.5, 3.7))
draw_vec(deaths, 3, keep = seq_along(deaths) == 1,        label = "deaths[1]")
draw_vec(deaths, 2, keep = seq_along(deaths) %in% c(3, 5), label = "deaths[c(3, 5)]")
draw_vec(deaths, 1, keep = deaths > 2,                     label = "deaths[deaths > 2]")
```

**The last one is the important one.** A **logical index** keeps the elements
where a condition holds. It is the same idea as filtering rows of a table, which
you will do with `filter()` in Stage 2.

**For now,** that is all the indexing you need to write. You will see square
brackets used on whole tables in code online; the box below shows what they mean.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Going further: indexing a table with [rows, columns]"}
Base R slices a data frame with `[rows, columns]`. Leave a side empty to mean "all
of them"; the comma separates rows from columns.

```{r}
#| label: df-slice
sessions[2, c("session_id", "deaths")]
nrow(sessions[sessions$deaths > 8, ])
```

The second line keeps the rows where `deaths` is above 8 and counts them. You only
need to recognise this. In these notes the same jobs are done with `filter()` and
`select()`, which read more like the question you are asking.
:::

---

## Functions and Arguments

A **function** takes inputs, called **arguments**, and returns a result. You have
already used several: `c()`, `class()`, `sum()`, `mean()`.

**Arguments have names and an order.** `args()` shows both:

```{r}
#| label: args
args(round)
```

`round()` takes `x`, the number, and `digits`, which defaults to 0. You can supply
arguments **by position**, in the order the function declares them, or **by
name**, in any order:

```{r}
#| label: named
round(3.14159, 2)
round(x = 3.14159, digits = 2)
round(digits = 2, x = 3.14159)
```

All three give the same answer. Without names, position is everything, and
swapping the values changes the meaning:

```{r}
#| label: swapped
round(2, 3.14159)
```

That rounds 2 to three decimal places. **Pass the main input by position and
everything else by name**: `mean(deaths, na.rm = TRUE)`. It reads well, and it lets
you skip arguments you are happy to leave at their defaults.

**Every function has a help page.** Type `?round` in the console. The page lists
every argument, its default, what the function returns, and examples at the
bottom, which are usually the quickest way to understand it.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Going further: writing your own function"}
You can define functions with `function()`. This one turns a logical vector into a
percentage, with a default of one decimal place:

```{r}
#| label: own-function
pct <- function(x, digits = 1) {
  round(100 * mean(x), digits)
}

pct(deaths > 2)
pct(deaths > 2, digits = 0)
```

The last expression in the body is the return value; there is no `return`
keyword needed. You will not need to write functions for the labs, but it is how
R programmers stop repeating themselves.
:::

---

## Missing Values

`NA` means a value should be there but was not recorded: a session length the
logger lost, a survey question the player skipped. It is not zero and not empty.
It means **unknown**.

```{r}
#| label: na
minutes <- c(12.9, 24.6, NA, 18.8)
mean(minutes)
```

**Any calculation involving an unknown is unknown.** R will not guess. Tell it to
leave the missing values out with `na.rm = TRUE`, and find them with `is.na()`:

```{r}
#| label: na-rm
mean(minutes, na.rm = TRUE)
is.na(minutes)
sum(is.na(minutes))
```

`sum(is.na(x))` is the count of missing values: the same "sum a logical to count
it" trick as before. Before you drop missing values, ask why they are missing.
Sometimes the pattern of what is missing is itself a finding, which is the subject
of [Cleaning and Transforming](06-cleaning-and-transforming.qmd).

---

## Data Frames

A playtest table is a **data frame**: a set of vectors of equal length, one per
column, side by side. Each column has one type; each row is one observation.

```{mermaid}
%%| label: fig-dataframe
%%| fig-cap: "A data frame is a set of equal-length vectors, one per column."
flowchart LR
  V1["session<br/>S001 S002 S003"] --> DF[("data frame<br/>3 rows × 3 columns")]
  V2["archetype<br/>Scout Tank Scout"] --> DF
  V3["deaths<br/>4 2 7"] --> DF
  class DF accent
```

You can build a small one by hand from vectors, which is handy for trying an idea
out:

```{r}
#| label: df-build
mini <- data.frame(
  session   = c("S001", "S002", "S003", "S004"),
  archetype = c("Scout", "Tank", "Scout", "Support"),
  deaths    = c(4, 2, 7, 3)
)
mini
```

**`$` pulls out one column**, as a vector, and everything from the Vectors section
works on it:

```{r}
#| label: df-dollar
mini$deaths
mean(mini$deaths)
sum(mini$deaths > 3)
```

In practice you rarely type a data frame in. You read one from a file.

---

## Packages and Reading Data

**A package** is a bundle of extra functions someone else has written. Using one is
two steps:

```{r}
#| label: packages
#| eval: false
install.packages("tidyverse")   # once per computer: downloads and installs it
library(readr)                  # every session: makes its functions available
```

Installing is like adding a library to a project; loading is like the `using`
statement at the top of a C# file. Forget the `library()` call and R reports
`could not find function`, which is the most common error in the first week.

**Reading a CSV** is one line with `readr`'s `read_csv()`:

```{r}
#| label: read
library(readr)
library(dplyr)
sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
```

The path is relative to the project folder. `read_csv()` returns a **tibble**, the
tidyverse's version of a data frame. Everything above works on it unchanged.

**Look before you do anything else.** `glimpse()`, from `dplyr`, shows the size of
the table and one line per column: its name, its type and its first few values.

```{r}
#| label: glimpse
glimpse(sessions)
```

`r nrow(sessions)` rows, `r ncol(sessions)` columns. `<chr>` is character and
`<dbl>` is numeric. Whether those types match what each column *means* is a
separate question, and the subject of [Data, Variables and Types](03-data-and-types.qmd).

**You will also see** `str()`, `head()` and `summary()` doing similar jobs in other
people's code. You do not need to write them.

---

## The Pipe

Most analysis is a short chain of steps: take the data, do this, then that. The
**pipe**, `|>`, writes the chain in the order it happens. Read it aloud as "and
then":

```{r}
#| label: pipe-count
sessions |> count(build)
```

"Take `sessions`, and then count the rows for each build." The pipe passes whatever
is on its left in as the first argument of the function on its right, so this is
exactly `count(sessions, build)`. With one step there is no difference; with four
steps, the piped version still reads top to bottom.

```{r}
#| label: pipe-count2
sessions |> count(build, quit_reason)
```

Two columns in `count()` counts every combination: how each build's sessions
ended. You will add more verbs to the chain as the course goes on:
`group_by()` and `summarise()` in [Describing Variables](04-describing-variables.qmd),
and `filter()`, `select()` and `mutate()` in Stage 2.

**You will also see** `%>%` in code online. It is an older pipe that does the same
job. Read it as `|>`.

---

## Reading Error Messages

R's error messages are terse, but they nearly always name the problem. Read the
error before changing anything. Here are the ones you will meet most, produced
live:

```{r}
#| label: err-object
#| error: true
mean(sesions$score)
```

**`object not found`**: a typo in a name (`sesions`), or an object you have not
created yet in this session. Check spelling and capitals, and check you ran the
line that creates it.

```{r}
#| label: err-function
#| error: true
summarise_scores(sessions)
```

**`could not find function`**: a typo in a function name, or a package that is not
loaded. Add the `library()` call.

```{r}
#| label: err-binary
#| error: true
"24" + 1
```

**`non-numeric argument to binary operator`**: arithmetic on text. A number has
been read or typed as character. Check with `class()`.

```r
read_csv("tidebreak_sessions.csv", show_col_types = FALSE)
```

```
Error: 'tidebreak_sessions.csv' does not exist in current working directory:
'C:/Users/you/L8-3DGD-Playtest-Analytics'.
```

**`does not exist`** or **`cannot open file`**: the path is wrong. Here the
`data/` folder is missing from it. Paths are relative to the project folder.

**`unexpected symbol`**, **`unexpected string constant`** or
**`unexpected ')'`**: a syntax error, almost always a missing comma, bracket or
quote just before the point R names.

**A warning is not an error.** An error stops the code. A **warning** lets it
finish but tells you something may be wrong, as `as.numeric("twenty")` did above.
Never ignore a warning without reading it.

---

## Worked Example

**How many Build A sessions ended in frustration?** This is the first number in
[Planning a Playtest](01-planning-a-playtest.qmd). Here it is built from nothing,
one idea at a time.

**Load and look.** Every analysis starts the same way:

```{r}
#| label: worked-load
sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
glimpse(sessions |> select(build, quit_reason))
```

**One condition gives a logical vector**, one value per session:

```{r}
#| label: worked-condition
is_a <- sessions$build == "A"
sum(is_a)
```

Summing it counts the Build A sessions: `r sum(sessions$build == "A")`.

**Two conditions combine with `&`**, "and": `TRUE` only where both are true.

```{r}
#| label: worked-combine
frustrated_a <- sessions$build == "A" & sessions$quit_reason == "Frustrated"
sum(frustrated_a)
```

`r sum(frustrated_a)` sessions, out of `r sum(is_a)`.

**Check it a second way.** `count()` with two columns counts every combination at
once:

```{r}
#| label: worked-count
sessions |> count(build, quit_reason)
```

The `A`, `Frustrated` row says `r sum(frustrated_a)`, the same answer. Getting a
number two different ways is the cheapest check there is.

Nothing in this example is more than the ideas on this page: an object, a
comparison, a logical vector, `sum()`, `$`, and one pipe.

---

## Reading What R Tells You

**`[1]` and friends are positions, not output.** When a long vector wraps across
lines, each line starts with the position of its first value: `[1]`, then `[19]`
and so on.

**`NA` in a result means an unknown went in.** Find it with `is.na()` before
reaching for `na.rm = TRUE`.

**A tibble prints its size and types first.** `# A tibble: 120 × 26` and the
`<chr>`/`<dbl>` row under the names are the first check on any table.

**Silence is success.** Assignment prints nothing. If you want to see an object,
type its name.

**Red text is not always an error.** Messages and warnings also print in red in
most editors. Read the first word: `Error` stops the code, `Warning` does not.

---

## Common Mistakes

**Using `=` to compare.** `sessions$build = "A"` is not a test. Comparison is `==`.

**Counting from 0.** `deaths[0]` returns an empty vector, not the first value.
R starts at 1.

**Getting the case wrong.** `Sessions`, `sessions` and `SESSIONS` are three
different objects, and `"Scout"` does not equal `"scout"`.

**Forgetting the quotes, or adding them.** `"build"` is the text *build*; `build`
is an object called build. `sum("deaths")` is an error; `sum(sessions$deaths)` is
a total.

**Forgetting `library()`.** Installed is not loaded. Every new session, and every
Quarto document, needs its own `library()` calls.

**Running lines out of order.** The console remembers everything you have run in
the session; a rendered document remembers only what is in it, from the top. If it
works in the console and fails when rendered, something is missing from the
document.

**Naming an object after a function.** `mean <- 5` or `data <- read_csv(...)`
works, but makes later code confusing. Pick names that describe the contents.

**Leaving out `na.rm = TRUE`, or adding it without thinking.** The first gives
`NA`; the second silently drops data you may need to explain.

---

## Practice

Each takes two or three minutes. Try each in the console before opening the
answers.

1. **Predict.** What does each line return? `c(1, 2, 3) * 2`,
   `sum(c(TRUE, FALSE, TRUE))`, `c(5, 10, 15)[2]`, `length(c("a", "b", NA))`.
2. **Fix it.** Each line has one error. Find it without running the line first.
   `mean(session$deaths)`, `sum(sessions$build = "A")`, `Mean(sessions$deaths)`.
3. **Compute.** How many *Tidebreak* sessions lasted longer than 30 minutes, and
   what proportion is that?
4. **Count.** How many sessions came from each cohort?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
**1.** Run them to check:

```{r}
#| label: practice-predict
c(1, 2, 3) * 2
sum(c(TRUE, FALSE, TRUE))
c(5, 10, 15)[2]
length(c("a", "b", NA))
```

`NA` still occupies a position, so the last vector has length 3.

**2.** `session` should be `sessions` (object not found). `=` should be `==` (an
error about an unexpected `=`). `Mean` should be `mean`: function names are
case-sensitive too (could not find function).

**3.** A comparison, then `sum()` to count and `mean()` for the proportion:

```{r}
#| label: practice-long
sum(sessions$session_minutes > 30)
mean(sessions$session_minutes > 30)
```

`r sum(sessions$session_minutes > 30)` sessions, or
`r pct_fmt(mean(sessions$session_minutes > 30))`.

**4.** One pipe:

```{r}
#| label: practice-cohort
sessions |> count(cohort)
```
:::

---

## Summary

- R is a language built for data. Code records how you got an answer, so it can be
  checked and repeated; a spreadsheet records only the answer.
- Try things in the console; keep them in a script or Quarto document; keep
  everything in a project.
- `<-` creates an object. Every value has a type; R converts types silently, so
  check them.
- The vector is R's basic unit. Arithmetic is vectorised, and summing a logical
  vector counts it.
- R counts from 1. A condition in square brackets keeps the elements where it is
  `TRUE`.
- Arguments go by position or by name. Pass the main input by position and options
  by name. `?name` opens the help.
- `NA` means unknown and spreads through calculations. `na.rm = TRUE` leaves it
  out; `is.na()` finds it.
- Read data with `read_csv()`, look at it with `glimpse()`, pull out a column with
  `$`.
- The pipe `|>` reads "and then". `sessions |> count(build)` counts sessions per
  build.
- `install.packages()` once, `library()` every session. Read the error message
  before changing anything.

| Key term | Meaning |
| :- | :- |
| **Console** | Where R runs code and prints results; nothing is kept |
| **Script** | A `.R` file of code you keep and re-run |
| **Quarto document** | A `.qmd` file of prose and code that renders to a web page |
| **Project** | The folder holding an analysis; paths are relative to it |
| **Object** | A named value, created with `<-` |
| **Vector** | An ordered set of values of one type |
| **Index** | A position or condition in `[ ]` that selects elements |
| **Function** | Code that takes arguments and returns a result |
| **Argument** | An input to a function, given by position or by name |
| **`NA`** | A missing value: unknown, not zero |
| **Data frame** | A table of equal-length column vectors; a tibble is one kind |
| **Package** | A bundle of extra functions, installed once and loaded with `library()` |
| **Pipe** | `\|>`: pass the result on the left into the function on the right |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 3 (know what kind of number you are holding)
  and rule 6 (cleaning is analysis)
- [R and Quarto Reference](../reference/r-quarto-reference.qmd): every function on
  this page, in one table
- [Lab 00: Setting Up](../labs/setting-up.qmd) and
  [Lab 01: Reading Data](../labs/reading-data.qmd), Part A
- Next: [Data, Variables and Types](03-data-and-types.qmd)

---

## Reflective Questions

- Think of an analysis you have done in a spreadsheet. What would a teammate need
  to repeat it exactly next week, and how much of that would be written down?
- Vectorised arithmetic replaces most loops. Write, in C#, what `sum(deaths > 2)`
  does, and count the lines.
- R converts types silently where C# would refuse to compile. Name one way that
  helps an analyst and one way it hurts.
- A teammate's Quarto report renders on their machine but not on yours. List three
  things that could cause it, from this page.
