---
title: "Data, Variables and Types"
subtitle: "What kind of number are you holding?"
body-classes: "wf-at-clean"
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")

n_sess    <- nrow(sessions)
n_vars    <- ncol(sessions)
n_players <- length(unique(sessions$player_id))
per_player <- unique(as.vector(table(sessions$player_id)))
stopifnot(length(per_player) == 1)

items      <- survey[, -1]
n_items    <- ncol(items)
obj        <- survey$objective_clear
obj_agree  <- mean(obj >= 4)
obj_dis    <- mean(obj <= 2)
item_agree <- sort(sapply(items, function(x) mean(x >= 4)))
item_dis   <- sapply(items, function(x) mean(x <= 2))
neg_items  <- names(item_dis)[item_dis > sapply(items, function(x) mean(x >= 4))]
stopifnot(names(which.max(item_dis)) == "objective_clear",
          names(which.max(item_agree)) == "visual_style")
code_list  <- function(x) paste0("`", x, "`", collapse = ", ")
```

> Before you can summarise a column you have to know what kind of thing is in it.
> This page sorts every variable you will meet into one of four types, shows what
> each type lets you do, and explains why R cannot make that decision for you.
> The playtest question running through it: *"The objective item averages 2.6 out
> of 5. Is that a finding?"*

::: {.at-a-glance}
**Time:** 30 minutes in class; 15 more for the Going further sections.
**Before this:** [R from First Principles](02-r-first-principles.qmd).
**Key terms:** observation, variable, tidy data, nominal, ordinal, discrete,
continuous, binary, Likert item, sample, population.
**R introduced:** `factor()` (preview). Uses `class()` and `table()` from
[R from First Principles](02-r-first-principles.qmd).
:::

## Learning Outcomes

By the end of these notes you can:

- say what one row and one column of a playtest table represent, and name the
  unit of observation;
- recognise a tidy table, and explain why a survey export usually is not one;
- classify any variable as **nominal**, **ordinal**, **discrete** or
  **continuous**, and say which summaries each type allows;
- explain the difference between the type a variable *is* and the type R
  *stores* it as, and name the common traps;
- choose a defensible summary for a rating-scale item, and say what assumption a
  mean of ratings makes;
- distinguish a sample from a population, and say why recruitment matters.

---

## Why This Matters for Playtesting

R will compute the mean of anything that is stored as a number. Player IDs,
difficulty settings, survey ratings, the build number: if it is a number in the
file, `mean()` returns an answer, and the answer looks exactly as authoritative
whether it means something or not.

**The type of a variable decides which questions you can ask of it.** You can count
quit reasons but not average them. You can put difficulty settings in order but
not say that Hard is twice Normal. You can say one session had twice as many
deaths as another. None of these rules is enforced by R. They are enforced by you,
and getting them wrong produces numbers that survive all the way into a design
meeting ([rule 3](ten-rules.qmd)).

---

## Observations and Variables

A data table has two directions, and each has a name.

An **observation** is one unit you measured: one row. A **variable** is one thing
you recorded about every observation: one column. A **value** is a single cell.

```{r}
#| label: obs-var-table
#| echo: false
knitr::kable(
  head(sessions[, c("session_id", "player_id", "cohort", "build",
                    "deaths", "session_minutes", "quit_reason")], 4),
  caption = "The first four rows of tidebreak_sessions.csv, seven of its columns."
)
```

In `tidebreak_sessions.csv` there are `r n_sess` observations and `r n_vars`
variables. Each observation is **one player in one session**. That is the file's
**unit of observation**, and it is worth stating out loud, because it is not the
same as "one player": the `r n_sess` sessions come from `r n_players` players,
`r per_player` sessions each.

**Every file has its own unit.** The other *Tidebreak* files record different
things, and mixing them up is how analyses end up counting the wrong thing:

| File | One row is | Rows |
| :- | :- | :- |
| `tidebreak_sessions.csv` | one player in one session | `r nrow(sessions)` |
| `tidebreak_survey.csv` | one player's survey after one session | `r nrow(survey)` |
| `tidebreak_paired.csv` | one player who played both builds | `r nrow(paired)` |
| `tidebreak_frame_times.csv` | one rendered frame | `r format(nrow(frames), big.mark = ",")` |

**In a report you would write:** "`r n_sess` sessions from `r n_players` players",
never just "`r n_sess` players".

---

## Tidy Data

Data arrives in whatever shape the tool that produced it happened to use. Almost
every analysis tool, R included, assumes one particular shape, and it has a name.

A table is **tidy** when:

1. each row is one observation;
2. each column is one variable;
3. each cell holds one value.

The sessions file is tidy. The survey file is not, quite, and the way it fails is
the most common failure there is. Here are its first three rows:

```{r}
#| label: tidy-wide
#| echo: false
knitr::kable(head(survey, 3),
             caption = "Wide: one row per session, one column per question.")
```

The eight question columns look like eight variables, but they are really values
of **one** variable, "which question", spread across the column headings. That
is convenient for reading and awkward for analysis: to ask "which question was
rated lowest?" you have to handle eight columns separately.

The tidy version has one row per session **per question**:

```{r}
#| label: tidy-long
#| echo: false
wide3 <- as.data.frame(head(survey, 3))
long3 <- data.frame(
  session_id = rep(wide3$session_id, each = n_items),
  item       = rep(names(items), times = 3),
  rating     = as.vector(t(wide3[, -1]))
)
knitr::kable(head(long3, 10),
             caption = paste0("Long: the same three sessions as ", nrow(long3),
                              " rows. The first ten are shown."))
```

Now "item" is a column you can group by, filter on and plot by, and the whole
survey can be summarised in one operation. Turning wide into long is called
**reshaping**, and you will do it with `pivot_longer()` in
[Interpreting and Reporting Evidence](11-reporting-evidence.qmd) and Lab 03.

::: {.callout-tip collapse="true" title="Quick check"}
A telemetry export has columns `player_id`, `deaths_level1`, `deaths_level2`,
`deaths_level3`. Is it tidy? What would the tidy version's columns be?

**Answer.** No. "Level" is a variable whose values are hiding in the column names.
The tidy version has `player_id`, `level` and `deaths`, with three rows per player.
:::

---

## Types of Data

Every variable is one of two kinds, and each kind splits in two.

**Categorical** variables have values that are **labels**. **Numerical** variables
have values that are **quantities**, so arithmetic on them means something.

```{mermaid}
%%| label: fig-types-mindmap
%%| fig-cap: "The four types of variable, with Tidebreak examples of each."
mindmap
  root((Variable))
    Categorical: values are labels
      Nominal: no order
        quit_reason
        archetype
        build
        completed, TRUE or FALSE
      Ordinal: ordered, unknown gaps
        cohort
        survey ratings 1 to 5
        difficulty rating 1 to 7
    Numerical: values are quantities
      Discrete: counts
        deaths
        kills
        wrong_turns
      Continuous: measurements
        session_minutes
        distance_m
        frame time in ms
```

**Nominal** variables are labels with no order. `quit_reason`, `archetype` and
`build` are nominal: Scout is not more or less than Sentinel, just different. You
can count each category and say which is most common. You cannot average them.

**Ordinal** variables are labels with an order but unknown gaps. `cohort` runs
Novice, Regular, Veteran, and that order means something, but nobody can say that
Regular is "halfway" between the other two. You can sort them and find the middle
one. Adding or averaging them assumes the gaps are equal, which nothing guarantees.

**Discrete** variables are counts of whole things. `deaths`, `kills` and
`wrong_turns` can be 0, 1, 2 and so on, never 2.4 in a single session. Arithmetic
works: one session can have twice the deaths of another, and the *average* number
of deaths across sessions can perfectly well be 2.4.

**Continuous** variables are measurements that can fall anywhere in a range:
`session_minutes`, `distance_m`, frame time. The only limit on how finely you can
record them is the precision of the instrument.

**Binary** variables are a special nominal case with exactly two values.
`completed` is `TRUE` or `FALSE`. They are worth singling out because of a useful
trick: R counts `TRUE` as 1 and `FALSE` as 0, so `mean(sessions$completed)` is the
**proportion** of sessions completed, `r pct_fmt(mean(sessions$completed))`.

When you are unsure, ask two questions in order:

```{mermaid}
%%| label: fig-types-decision
%%| fig-cap: "Two questions classify any variable."
flowchart TD
  A{"Are the values labels<br/>or quantities?"} -->|Labels| B{"Do the labels have<br/>a natural order?"}
  B -->|No| N["Nominal<br/>quit_reason, build"]
  B -->|Yes| O["Ordinal<br/>cohort, ratings"]
  A -->|Quantities| C{"Are they counts<br/>of whole things?"}
  C -->|Yes| D["Discrete<br/>deaths, wrong_turns"]
  C -->|"No, measured"| E["Continuous<br/>session_minutes"]
  classDef out fill:#f8f9fa,stroke:#b22222,color:#212529
  class N,O,D,E out
```

### What each type lets you do

Broadly, the types form a ladder: each allows what the one before it allows, plus
something new. The one wrinkle is counting: a continuous variable has almost no
repeated values, so you count it in ranges, or **bins**, instead.

| You want to... | Nominal | Ordinal | Discrete | Continuous |
| :- | :-: | :-: | :-: | :-: |
| count each value, find the most common (the **mode**) | yes | yes | yes | in bins |
| put values in order, find the middle (the **median**) | no | yes | yes | yes |
| add, subtract, find the **mean** | no | assumed | yes | yes |
| say "twice as much" | no | no | yes | yes |
| draw a... | bar chart | bar chart | bar chart or histogram | histogram |

The "assumed" for ordinal is the subject of a long argument, and it gets its own
section below.

::: {.callout-note collapse="true" title="Going further: interval and ratio"}
Statistics textbooks often use a second scheme, from the psychologist Stanley
Stevens: **nominal, ordinal, interval, ratio**. The first two are the same as
here. The difference between the last two is whether **zero means "none"**.

Deaths and session length are **ratio** variables: zero deaths means no deaths, so
"twice as many deaths" makes sense. Temperature in Celsius is **interval**: zero
is an arbitrary point, so 20°C is not "twice as hot" as 10°C. A rating scale, even
treated as numeric, is at best interval: nobody thinks a 6 out of 7 for enjoyment
means twice the enjoyment of a 3.

Discrete against continuous is a different question: whether values come in whole
steps. A variable can be discrete and ratio (deaths) or continuous and ratio
(session length).
:::

---

## The Rating-Scale Question

```{r}
#| label: fig-obj-bar
#| fig-cap: "Responses to “I always knew what my objective was”, one bar per point on the scale."
#| fig-alt: "Bar chart of the five response options for the objective item. The bars for 1, 2 and 3 are the tallest; very few players chose 5."
#| fig-height: 3.5
par(mar = c(4, 4, 1, 1))
barplot(table(factor(obj, levels = 1:5)),
        names.arg = c("1 Strongly disagree", "2", "3", "4", "5 Strongly agree"),
        col = pal$fill, border = "white", ylab = "Responses", cex.names = 0.8)
```

The survey asks players to rate statements from 1 (strongly disagree) to 5
(strongly agree). Is a rating a number?

**Strictly, it is ordinal.** You know 4 is above 3. You do not know that the step
from 2 to 3 is the same size as the step from 4 to 5, and a player choosing 3 may
mean "neutral", "don't know" or "didn't think about it".

**In practice, people often average ratings anyway**, especially when a scale has
five or more points and the responses are spread out. That convention is
widespread and defensible. It is still an assumption, and you should know when you
are making it.

**When a finding rests on the mean, show the distribution instead.** The objective
item averages `r round(mean(obj), 2)`. That number is hard to act on: is 2.6 bad?
The distribution in @fig-obj-bar is not: `r pct_fmt(obj_dis)` of responses
disagreed that the player always knew their objective, and only
`r pct_fmt(obj_agree)` agreed. A designer can act on "almost half the time,
players did not know what they were meant to be doing".

**In a report you would write:** "`r pct_fmt(obj_dis)` of `r length(obj)`
responses disagreed (1 or 2 on a 5-point scale) that they always knew their
objective; `r pct_fmt(obj_agree)` agreed (4 or 5)."

---

## What R Thinks the Type Is

Statistical type and R type are different things. R's type, introduced in
[R from First Principles](02-r-first-principles.qmd#types), describes **how the
value is stored**. The statistical type describes **what the value means**. R
cannot work out the second from the first.

```{r}
#| label: r-types
class(sessions$quit_reason)
class(sessions$deaths)
class(sessions$session_minutes)
class(sessions$completed)
```

`read_csv()` gives you three storage types for this file: **character** for text,
**numeric** (printed as `<dbl>`) for every number, and **logical** for
`TRUE`/`FALSE`. Here is how those line up with the four statistical types:

| Statistical type | How R stores it after `read_csv()` | What R does not know |
| :- | :- | :- |
| Nominal | character, e.g. `quit_reason` | nothing missing: `mean()` of text returns `NA` with a warning |
| Binary | logical, e.g. `completed` | nothing missing: `mean()` gives a proportion |
| Ordinal | character (`cohort`) **or** numeric (survey ratings) | the order of text labels; that numeric ratings are not measurements |
| Discrete | numeric, e.g. `deaths` | that the values are whole counts |
| Continuous | numeric, e.g. `session_minutes` | nothing: this is what numeric storage is for |

**Four traps follow from that table.**

**Ordered labels sort alphabetically.** `table(sessions$cohort)` lists Novice,
Regular, Veteran, which happens to be right only because the alphabet agrees. For
an order the alphabet does not know, such as Easy, Normal, Hard, you tell R with a
**factor**:

```{r}
#| label: factor-preview
cohort_f <- factor(sessions$cohort, levels = c("Novice", "Regular", "Veteran"))
table(cohort_f)
```

You will use `factor()` properly in [Cleaning and Transforming](06-cleaning-and-transforming.qmd).

**Ratings arrive as plain numbers.** Every survey item is numeric, so `mean()` and
`sd()` work on them without complaint. Whether they *should* is the decision in the
previous section, and R cannot make it for you.

**Identifiers can look like numbers.** *Tidebreak* uses text IDs (`P01`,
`S001`), which is good practice for exactly this reason. A file that numbers its
players 1 to 30 lets R compute an "average player", which is meaningless. If you
ever see a mean ID, a summary has run over the wrong columns.

**Whole numbers do not tell you discrete from continuous.** `time_alive_s` is
recorded in whole seconds, so every value looks like a count. It is a measurement
of time, rounded by the logger: continuous. The type is decided by what the number
measures, not by whether it has a decimal point. (Base R's `read.csv()` makes this
trap visible: it stores `time_alive_s` as **integer**, a storage type for whole
numbers, which says nothing about whether the variable is a count.)

---

## Measuring with Surveys

The survey is the only source that tells you what players thought, so it deserves
the same care as the telemetry.

**A Likert item** is a statement with an ordered set of responses, usually five or
seven points from strongly disagree to strongly agree. *Tidebreak*'s survey uses
five points; its `difficulty` and `enjoyment` ratings use seven. More points give
finer distinctions; fewer are quicker to answer and easier to label clearly.

**Good items share a few habits:**

- **One idea per item.** "The controls were responsive and intuitive" asks two
  things; a player who found them responsive but confusing cannot answer it.
- **Neutral wording.** "How much did you enjoy the improved lighting?" tells the
  player what you want to hear.
- **Labelled anchors.** Say what 1 and 5 mean, and keep the direction the same
  across items.
- **Asked straight after play**, before memory rewrites the session.

**Prefer an established questionnaire to an invented one.** Questionnaires such
as the **System Usability Scale** (SUS, usability), the **Player Experience
Inventory** (PXI) and the **Game Experience Questionnaire** (GEQ) have been tested
on many players. That means their items are known to measure something
consistently, and your results can be compared with other studies. An item you
wrote last night has none of that. Check the published evidence for any
questionnaire before adopting it; some widely used ones have been criticised.

**Summarise ordinal items as distributions.** For each item, report the count or
percentage at each point, or the percentage who agree (4 or 5). Show the whole
survey at once with a stacked bar per item:

```{r}
#| label: fig-survey-stacked
#| fig-cap: "Every survey item as a stacked bar, from 1 (strongly disagree, red) to 5 (strongly agree, blue), ordered by the share who agreed."
#| fig-alt: "Horizontal stacked bars for the eight survey items, ordered by share agreeing. Visual style is almost entirely agreement; objective clear, death understood and difficulty fair have more red disagreement than blue agreement."
#| fig-height: 4.5
props <- sapply(items[names(item_agree)],
                function(x) prop.table(table(factor(x, levels = 1:5))))
likert_cols <- c("#b22222", "#e8a3a3", "grey85", "#9bb0c9", "#3a5f8a")
op <- par(mar = c(4, 11, 3, 1))
barplot(props, horiz = TRUE, las = 1, border = "white", col = likert_cols,
        xlab = "Share of responses", cex.names = 0.8)
legend("top", inset = c(0, -0.13), xpd = TRUE, horiz = TRUE, bty = "n",
       legend = 1:5, fill = likert_cols, cex = 0.8)
par(op)
```

@fig-survey-stacked tells you at a glance what eight means would not.
`r length(neg_items)` items have more players disagreeing than agreeing
(`r code_list(neg_items)`), and the objective item has the largest share
disagreeing, `r pct_fmt(max(item_dis))`. At the other end, `visual_style` is close
to unanimous: `r pct_fmt(max(item_agree))` agreed. The large neutral share on most
items, the grey in the middle, is also information: a mean would fold it away.

---

## Sample and Population

The `r n_players` *Tidebreak* players are a **sample**: the people who happened to
come to this playtest. The **population** is everyone the game is being made for,
and you never get to measure it.

**Every number you compute describes the sample.** The question you care about is
always about the population: not "did these players struggle?" but "will players
struggle?". Run the test again with different players and every figure on this
page moves. How far it can move is the subject of
[From Sample to Population](07-sample-to-population.qmd). Until then, report how many players a number came from every time you quote it
([rule 8](ten-rules.qmd)).

**Who is in the sample matters as much as how many.** A playtest recruited from the
studio's Discord samples the game's most committed fans, not its audience. A
playtest of colleagues samples people who already know the controls. No analysis
afterwards corrects for who walked through the door, which is why recruitment is
part of [planning the playtest](01-planning-a-playtest.qmd).

---

## Worked Example

**Is a mean of 2.6 a finding?** The *Tidebreak* team summarises its survey with
`colMeans()` and gets eight numbers. The objective item is lowest at
`r round(mean(obj), 2)`. The draft report says: "Objective clarity scored 2.6 out
of 5."

**Step 1: what type is it?** A 1 to 5 agreement rating: ordinal. The mean treats
the gap from 1 to 2 as equal to the gap from 4 to 5. Possibly defensible, but an
assumption.

**Step 2: what does the distribution say?** `r sum(obj <= 2)` of
`r length(obj)` responses were 1 or 2; `r sum(obj >= 4)` were 4 or 5;
`r sum(obj == 3)` sat at 3.

**Step 3: rewrite the finding.** "`r pct_fmt(obj_dis)` of players disagreed that
they always knew their objective; `r pct_fmt(obj_agree)` agreed." The mean said the
item was low. The distribution says *how many players* had the problem, which is
what a designer needs to decide whether to fix it.

**Step 4: check the unit.** The survey has one row per session, and each player
completed `r per_player` sessions. "`r pct_fmt(obj_dis)` of players" is really
"`r pct_fmt(obj_dis)` of responses". The honest version says so.

---

## Interpretation Guidance

**What the data shows.** The type of each variable, read from what it measures.
`quit_reason` is nominal; `objective_clear` is ordinal; `deaths` is discrete.

**What the analysis suggests.** A summary chosen to fit the type: counts and
proportions for categories, distributions for ratings, means and medians for
quantities.

**What can legitimately be concluded.** Only what the type allows. "Quit reasons
averaged 2.3" is not a conclusion; it is an error. "Players rated the objective 2.6"
is a conclusion with a hidden assumption; state it, or report the distribution.

---

## Common Mistakes

**Averaging a label.** Code the cohorts as 1, 2 and 3 and R will average them
happily. The answer means nothing ([rule 3](ten-rules.qmd)).

**Trusting R's type.** `<dbl>` says how a value is stored, not what it means. A
rating, a count and a measurement can all print as `<dbl>`.

**Reporting a mean of ratings without the distribution.** The mean hides whether
everyone was lukewarm or half loved it and half hated it.

**Counting sessions as players.** `r n_sess` rows is `r n_players` people. Say both.

**Spreading one variable across columns.** `deaths_level1`, `deaths_level2` and
`deaths_level3` is one variable, `deaths`, and one more, `level`, wearing a
disguise.

**Inventing survey items.** A home-made item has no evidence that it measures what
you think it does. Use an established questionnaire where one exists.

---

## Practice

Each takes two or three minutes.

**1. Classify.** Nominal, ordinal, discrete or continuous: `kills`, `quit_reason`,
`cohort`, `distance_m`, `visual_style`, `completed`, `difficulty`,
`time_alive_s`, `session_id`.

::: {.callout-note collapse="true" title="Check your answer"}
`kills` discrete; `quit_reason` nominal; `cohort` ordinal; `distance_m`
continuous; `visual_style` ordinal (a 1 to 5 rating); `completed` nominal, and
binary; `difficulty` ordinal (a 1 to 7 rating); `time_alive_s` continuous,
although it is recorded in whole seconds; `session_id` is not a variable you
analyse at all. It is an **identifier**: a label for the row.
:::

**2. Predict.** What does R print for `class(sessions$cohort)` and
`class(survey$visual_style)`? Which of the two is R closer to being right about?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-class
class(sessions$cohort)
class(survey$visual_style)
```

Character and numeric. R is closer with `cohort`: `mean()` of text returns `NA`
with a warning instead of a number. It will average `visual_style` without a murmur, and whether it
should is your decision.
:::

**3. Critique.** A teammate's report says: "Average cohort was 1.9, so our testers
were mostly Regulars." What is wrong, and what should it say instead?

::: {.callout-note collapse="true" title="Check your answer"}
Cohort is ordinal; the 1.9 comes from coding Novice, Regular, Veteran as 1, 2, 3
and assuming equal gaps. Report the counts instead:

```{r}
#| label: practice-cohort
table(cohort_f)
```

"`r sum(sessions$cohort == "Regular")` of `r n_sess` sessions were played by
Regulars" says the same thing without the invented arithmetic.
:::

**4. Compute.** What proportion of sessions were completed? Do it with one
function.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-completed
mean(sessions$completed)
```

`completed` is logical, so `mean()` treats `TRUE` as 1 and `FALSE` as 0 and
returns the proportion: `r pct_fmt(mean(sessions$completed))`.
:::

---

## Summary

- One row is one observation; one column is one variable. Name the unit: here,
  one player in one session.
- Tidy data has one observation per row, one variable per column, one value per
  cell. Survey exports usually are not tidy.
- Variables are nominal, ordinal, discrete or continuous. The type decides which
  summaries are legal.
- A rating is ordinal. Averaging it is an assumption; when the finding depends on
  it, report the distribution.
- R's storage type is not the statistical type. Check every column yourself.
- Use established questionnaires, and summarise ordinal items as distributions.
- Every number describes the sample. Report how many players it came from.

| Key term | Meaning |
| :- | :- |
| **Observation** | One unit measured: one row |
| **Variable** | One thing recorded about every observation: one column |
| **Unit of observation** | What one row represents, e.g. one player in one session |
| **Tidy data** | One observation per row, one variable per column, one value per cell |
| **Nominal** | Labels with no order |
| **Ordinal** | Labels with an order but unknown gaps |
| **Discrete** | Counts of whole things |
| **Continuous** | Measurements anywhere in a range |
| **Binary** | A nominal variable with two values, such as TRUE/FALSE |
| **Likert item** | A statement rated on an ordered agree–disagree scale |
| **Sample** | The players you actually measured |
| **Population** | Everyone the conclusion is meant to apply to |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 3 (know what kind of number you are holding)
  and rule 8 (small samples lie confidently)
- [Statistics Reference: describing data](../reference/reference-statistics.qmd#describing-data)
- [Lab 01: Reading Data](../labs/reading-data.qmd), Part B
- Next: [Describing Variables](04-describing-variables.qmd)

---

## Reflective Questions

- Take a variable from your own project's telemetry. Which of the four types is
  it, and what would break if you treated it as the type next door?
- Your survey has a 1 to 7 enjoyment rating. A teammate wants to report the mean;
  you want to report the distribution. Make the best case for each.
- Your game logs one row per death, with a timestamp and a player ID. What is the
  unit of observation, and what would you have to do to answer "how many deaths
  per session?"
- Which established questionnaire would you use to evaluate your own project's
  controls, and what would you lose by writing your own items instead?
