---
title: "Describing Variables"
subtitle: "Centre, spread, and the pictures that keep them honest"
body-classes: "wf-at-describe"
aliases:
  - describing-what-you-have.html
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")

# Worked example: the Sentinel
arch_mean   <- tapply(sessions$score, sessions$archetype, mean)
arch_median <- tapply(sessions$score, sessions$archetype, median)
arch_sd     <- tapply(sessions$score, sessions$archetype, sd)
stopifnot(names(which.min(arch_mean)) == "Sentinel",
          names(which.max(arch_sd))   == "Sentinel")
rank_word  <- c("highest", "second-highest", "third-highest", "lowest")
sen_median <- rank_word[rank(-arch_median)[["Sentinel"]]]
sentinel   <- sessions$score[sessions$archetype == "Sentinel"]
others_sd  <- arch_sd[names(arch_sd) != "Sentinel"]
low_cut    <- 700
stopifnot(sum(sentinel < low_cut) > 0, all(sentinel[sentinel < low_cut] < 700))

# SD by hand
d    <- c(4, 2, 7, 3, 0, 5)
dev  <- d - mean(d)

# Boxplot anatomy
score_stats <- boxplot.stats(sessions$score)
score_out   <- score_stats$out
stopifnot(all(sessions$archetype[sessions$score %in% score_out] == "Sentinel"))

# time_alive_s, for the four-part description
ta      <- sessions$time_alive_s
ta_cap  <- max(ta)
n_cap   <- sum(ta == ta_cap)

deaths_mode <- as.numeric(names(which.max(table(sessions$deaths))))
stopifnot(deaths_mode != median(sessions$deaths))

n_frus <- sum(sessions$quit_reason == "Frustrated")
```

> Before you compare anything to anything, you have to be able to say what one
> variable looks like. This page gives you the vocabulary for that, centre and
> spread, and the three pictures that check it: the bar chart, the histogram and
> the boxplot. It ends with describing a variable group by group, which is where
> most playtest questions live.
> The playtest question running through it: *"The Sentinel has the lowest average
> score. Is it underpowered?"*

::: {.at-a-glance}
**Stage:** 1 · Core.
**Time:** 30 minutes in class; 20 more for the sections marked Useful when needed.
**Before this:** [Data, Variables and Types](03-data-and-types.qmd).
**Key terms:** frequency table, proportion, mode, mean, median, range, standard
deviation, quartile, IQR, percentile, histogram, bin, boxplot, outlier.
**R you will write:** `count()`, `mutate()`, `group_by()`, `summarise()`, `n()`,
`mean()`, `median()`, `sd()`, `IQR()`, `quantile()`, `barplot()`, `hist()`,
`boxplot()`.
:::

## Learning Outcomes

By the end of these notes you can:

- describe a categorical variable with counts, proportions and its mode, and draw
  it as a bar chart;
- choose between the mean and the median, and read the gap between them;
- describe spread with the range, IQR and standard deviation, say in words what a
  standard deviation measures, and say why the range is rarely enough;
- read percentiles and a five-number summary;
- draw and read a histogram and a boxplot, and say what each shows that a summary
  hides;
- summarise a variable for each group with `group_by()` and `summarise()`, and
  compare the groups with side-by-side boxplots;
- describe a variable in one sentence that gives its centre, spread, shape and size.

---

## Why This Matters for Playtesting

A playtest file is a few thousand numbers. Nobody can read them; everybody needs
to know what they say. A **summary** is a number that stands in for a whole
column, and choosing it well is the difference between a report that helps a
designer and one that misleads them.

Every summary throws information away. That is what a summary is for. The risk is
that it throws away the finding, which is why every summary on this page comes with
a picture that checks it ([rule 4](ten-rules.qmd)).

---

## Describing a Categorical Variable

A categorical variable has only one kind of summary: **how many of each**.

A **frequency table** counts each category. `count()` builds one, and
`sort = TRUE` puts the biggest first. A **proportion** is each count divided by the
total; `mutate()` adds it as a new column:

```{r}
#| label: cat-table
library(dplyr)
quit_counts <- sessions |>
  count(quit_reason, sort = TRUE) |>
  mutate(prop = round(n / sum(n), 2))
quit_counts
```

Read the pipe aloud: take the sessions, count them by quit reason, and then add a
column that divides each count by the total. The most common category is the
**mode**: here, `r quit_counts$quit_reason[1]`.

**One proportion on its own** is quicker with the trick from
[R from First Principles](02-r-first-principles.qmd#vectors): the mean of a
condition is the proportion of rows where it is true.

```{r}
#| label: cat-one-prop
mean(sessions$quit_reason == "Frustrated")
```

The picture is a **bar chart**: one bar per category, height proportional to the
count. Sort the bars by height unless the categories have their own order.

```{r}
#| label: fig-cat-bar
#| fig-cap: "Most Tidebreak sessions ended with the test completed; frustration was the most common way to leave early."
#| fig-alt: "Bar chart of quit reasons. Completed test is by far the tallest bar, followed by Frustrated, Lost interest, and a short Technical fault bar."
#| fig-height: 3.5
barplot(quit_counts$n, names.arg = quit_counts$quit_reason,
        col = pal$fill, border = "white", ylab = "Sessions")
```

**In a report you would write:** "`r n_frus` of `r nrow(sessions)` sessions
(`r pct_fmt(n_frus / nrow(sessions))`) ended in a frustration quit."

**You will also see** `table()` and `prop.table()` doing the same job in older
code. You do not need to write them.

---

## Centre: the Typical Value

Three summaries claim to describe "the typical value", and they can disagree.

The **mean** adds every value and divides by how many there are. It uses every
value, which is its strength, and it is pulled by extreme values, which is its
weakness.

The **median** is the middle value when the data is sorted: half the values are
below it, half above. It ignores how extreme the extremes are, which is its
strength and its weakness.

The **mode** is the most common value. It is the only centre a nominal variable
has, and for a count it is often worth knowing: the most common number of deaths
in a session is `r deaths_mode`, while the median is
`r median(sessions$deaths)`.

```{r}
#| label: fig-mean-median
#| echo: false
#| fig-cap: "Nine players' completion times. One slow player pulls the mean up; the median does not move."
#| fig-alt: "Dot plot of nine completion times. Eight cluster between 4 and 7 minutes and one sits alone at 19. A dashed line marks the median at 5 minutes; a solid red line marks the mean at about 6.8 minutes, near the right edge of the cluster."
#| fig-height: 2.4
times <- c(4, 4, 5, 5, 5, 6, 6, 7, 19)
par(mar = c(4, 1, 1.5, 1))
stripchart(times, method = "stack", pch = 19, cex = 1.6, col = pal$line,
           offset = 0.6, xlim = c(0, 20), ylim = c(0.9, 1.6),
           xlab = "Completion time (minutes)")
abline(v = median(times), lty = 2, lwd = 2, col = pal$guide)
abline(v = mean(times), lwd = 2, col = pal$highlight)
text(median(times), 1.55, "median", pos = 2, cex = 0.8, col = pal$line)
text(mean(times), 1.55, "mean", pos = 4, cex = 0.8, col = pal$highlight)
label_illustration("Illustration: invented times")
```

In the figure, eight players finish in 4 to 7 minutes and one takes 19. The median,
`r median(times)` minutes, is a fair description of a typical player. The mean,
`r round(mean(times), 1)` minutes, describes nobody: it is slower than
`r sum(times < mean(times))` of the `r length(times)` players, pulled there by one.

**When the mean and median disagree, the data is lopsided, and the gap tells you
which way.** If the mean sits above the median, a few large values are pulling it
up. If it sits below, a few small values are pulling it down. That lopsidedness is
called **skew**, and it is the main subject of
[Distributions and Shape](05-distributions-and-shape.qmd).

**Which to use.** The median when the data is lopsided or has extreme values; the
mean when it is roughly symmetric, or when the total matters (total damage dealt
is the mean times the number of players, and the median cannot tell you that). The
mode for categories, and for counts where "most common" is the question.

---

## Spread: How Much Players Differ

The centre alone is nearly useless. Two levels with the same average completion
time, one where everybody takes four minutes and one where half take one minute
and half take seven, are not the same level. **Spread** measures how much the
values differ from each other.

**The range** is the maximum minus the minimum. Easy to quote and nearly always
misleading, because a single unusual player sets it on their own.

**The standard deviation** (SD) is the typical distance of a value from the mean.
Small means players cluster tightly around the average; large means the average is
a poor guide to any individual. It is in the same units as the data.

You do not need to calculate it by hand, but seeing it once makes the number
make sense. Here it is for the six death counts from
[R from First Principles](02-r-first-principles.qmd):

| Step | Values |
| :- | :- |
| The data | `r paste(d, collapse = ", ")` |
| 1. Find the mean | `r mean(d)` |
| 2. Distance of each value from the mean | `r paste(dev, collapse = ", ")` |
| 3. Square each distance, so negatives do not cancel positives | `r paste(dev^2, collapse = ", ")` |
| 4. Average the squares (dividing by n − 1 = `r length(d) - 1`): the **variance** | `r round(var(d), 2)` |
| 5. Square root, back to the original units: the **SD** | `r round(sd(d), 2)` |

```{r}
#| label: fig-sd-distance
#| echo: false
#| fig-cap: "The standard deviation summarises these distances: how far, typically, each value sits from the mean."
#| fig-alt: "Six points at heights 4, 2, 7, 3, 0 and 5, with a horizontal red line at the mean of 3.5 and a vertical grey segment from each point to the line."
#| fig-height: 3
par(mar = c(4, 4, 1, 1))
plot(seq_along(d), d, pch = 19, cex = 1.4, col = pal$line, ylim = c(0, 7.5),
     xlab = "Session", ylab = "Deaths", xaxt = "n")
axis(1, at = seq_along(d))
abline(h = mean(d), col = pal$highlight, lwd = 2)
segments(seq_along(d), d, seq_along(d), mean(d), col = pal$guide, lwd = 2)
text(4.5, mean(d), paste("mean =", mean(d)), pos = 3, cex = 0.8, col = pal$highlight)
```

In R it is one function:

```{r}
#| label: sd-r
sd(d)
```

::: {.callout-note collapse="true" appearance="simple" icon="false" title="The maths, if you want it"}
For values $x_1, \dots, x_n$ with mean $\bar{x}$:

$$\text{variance} = \frac{\sum (x_i - \bar{x})^2}{n - 1} \qquad \text{SD} = \sqrt{\text{variance}}$$

Read aloud: subtract the mean from every value, square, add them up, divide by one
fewer than the number of values, take the square root. Dividing by $n - 1$ rather
than $n$ is a small correction that makes the sample SD a better estimate of the
population's; with a few dozen players the difference is negligible.
:::

**The SD belongs to the mean.** It measures distance from the mean, so it inherits
the mean's weakness: a few extreme values inflate it. For lopsided data, use the
next measure instead.

**Quartiles** split the sorted data into four equal parts. A quarter of the values
lie below the **first quartile** (Q1), half below the median, and three quarters
below the **third quartile** (Q3). The **interquartile range** (IQR) is Q3 − Q1:
the width of the middle half of your players. It ignores the extremes entirely,
which makes it the most robust measure of spread you have.

**Percentiles** generalise this: the 90th percentile has 90% of values below it.
`quantile()` returns any you ask for, and with no second argument it gives the
**five-number summary**: minimum, Q1, median, Q3, maximum.

```{r}
#| label: spread-r
quantile(sessions$time_alive_s)
IQR(sessions$time_alive_s)
quantile(sessions$time_alive_s, c(0.1, 0.9))
```

The middle half of sessions lasted between `r quantile(ta, 0.25)` and
`r quantile(ta, 0.75)` seconds before the player died or the match ended: an IQR
of `r IQR(ta)` seconds.

**Pair the centre with its spread.** Mean with SD; median with IQR. Mixing them
("median 133, SD 33") pairs a summary that ignores extremes with one that is
inflated by them.

---

## Seeing One Variable

### The histogram

A **histogram** divides the range of a numeric variable into equal-width intervals,
called **bins**, and draws a bar for how many values fall in each. It shows the
whole distribution: where the bulk sits, whether it is lopsided, whether there is
a second hump, whether anything sits apart from the rest.

```{r}
#| label: fig-hist-distance
#| fig-cap: "Distance travelled per session. Roughly symmetric: the bulk sits in the middle and falls away evenly on both sides."
#| fig-alt: "Histogram of distance travelled per session, from about 100 to 650 metres. The bars are tallest around 350 to 450 metres and fall away on both sides fairly evenly."
#| fig-height: 3.5
hist(sessions$distance_m, breaks = 15, col = pal$fill, border = "white",
     main = "", xlab = "Distance travelled (m)", ylab = "Sessions")
```

**The number of bins changes what you see.** Too few and a second hump merges into
the first; too many and random noise looks like structure. `breaks` sets roughly
how many bars R draws. Always try more than one; the worked example below shows
what one setting can hide.

### The boxplot

A **boxplot** compresses the distribution into five numbers and a rule for
outliers. It is less detailed than a histogram for one variable, and much better
for comparing several groups side by side.

```{r}
#| label: fig-box-anatomy
#| echo: false
#| fig-cap: "The anatomy of a boxplot, drawn for every Tidebreak session's score."
#| fig-alt: "A horizontal boxplot of score with labels. The box runs from Q1 to Q3 with a thick line at the median. Whiskers extend to the most extreme values within 1.5 box-lengths. Seven points sit separately at the far left, labelled outliers."
#| fig-height: 3.2
q <- quantile(sessions$score, c(0.25, 0.5, 0.75))
par(mar = c(4, 1, 2.5, 1))
boxplot(sessions$score, horizontal = TRUE, col = pal$fill, ylim = c(450, 2100),
        xlab = "Score", pch = 19, outcol = pal$highlight, frame = FALSE)
lab <- function(x, txt, y = 1.33) {
  segments(x, 1.22, x, y - 0.04, col = pal$guide)
  text(x, y, txt, cex = 0.75, pos = 3, xpd = NA)
}
lab(q[1], "Q1", 1.3)
lab(q[2], "median", 1.42)
lab(q[3], "Q3", 1.3)
lab(score_stats$stats[1], "whisker end", 1.3)
lab(score_stats$stats[5], "whisker end", 1.3)
text(mean(range(score_out)), 0.78, "outliers", cex = 0.75, col = pal$highlight)
arrows(q[1], 0.7, q[3], 0.7, code = 3, length = 0.05, col = pal$guide)
text(mean(q[c(1, 3)]), 0.62, "IQR: the middle half", cex = 0.75)
```

| Part | What it shows |
| :- | :- |
| Thick line | The median |
| Box | Q1 to Q3: the middle half of the values |
| Whiskers | The most extreme values within 1.5 box-lengths of the box |
| Separate points | **Outliers**: values beyond the whiskers, drawn individually |

The seven points at the left are scores more than 1.5 IQRs below Q1. An outlier
rule is a prompt to look, not a verdict; what to do about outliers is in
[Distributions and Shape](05-distributions-and-shape.qmd#outliers).

### The strip chart {.when-needed}

With a small group, a **strip chart** shows every value as a point, so nobody
mistakes five players for a distribution:

```{r}
#| label: fig-strip
#| fig-cap: "Deaths per session for each cohort, one point per session. Points are jittered sideways so that equal values do not hide each other."
#| fig-alt: "Three vertical strips of points for Novice, Regular and Veteran, showing deaths from 0 to 7. All three cohorts spread across a similar range."
#| fig-height: 3.5
stripchart(deaths ~ cohort, data = sessions, vertical = TRUE,
           method = "jitter", jitter = 0.2, pch = 19,
           col = adjustcolor(pal$line, alpha.f = 0.5),
           ylab = "Deaths", xlab = "Cohort")
```

---

## Describing by Group

Most playtest questions compare groups: builds, cohorts, archetypes. Two verbs do
the work. `group_by()` splits the table into one piece per group, and `summarise()`
calculates the same summaries for each piece:

```{r}
#| label: by-group
sessions |>
  group_by(cohort) |>
  summarise(n             = n(),
            median_alive  = median(time_alive_s),
            iqr_alive     = IQR(time_alive_s))
```

One row per cohort, one column per summary. `n()` counts the rows in each group:
always include it, because a summary from five players and one from fifty are not
the same evidence ([rule 8](ten-rules.qmd)).

A boxplot with the formula `y ~ group` draws the same split. Read `~` as "split
by":

```{r}
#| label: fig-by-group
#| fig-cap: "Veterans survive longest on average, but the boxes overlap heavily: plenty of Novices outlasted plenty of Veterans."
#| fig-alt: "Three boxplots of time alive in seconds for Novice, Regular and Veteran. The medians rise from Novice to Veteran, and all three boxes overlap substantially."
#| fig-height: 3.5
boxplot(time_alive_s ~ cohort, data = sessions, col = pal$fill,
        ylab = "Time alive (s)", xlab = "Cohort")
```

Read the boxes, not only the medians: Veterans have the highest median, but the
three boxes overlap heavily.

**A proportion per group** is a mean of a condition inside `summarise()`. How
often did each build's sessions end in frustration?

```{r}
#| label: prop-by-group
sessions |>
  group_by(build) |>
  summarise(n = n(),
            frustrated = mean(quit_reason == "Frustrated"))
```

Comparing proportions, not counts, is what makes groups of different sizes
comparable.

**You will also see** `tapply()` and `aggregate()` doing this job in older code.
You do not need to write them.

**For your project:** almost every question your team asks will be "is it
different for this group?": this build, this level, these players. Write the
`group_by()` and `summarise()` for it before you write anything else.

---

## Describing Two Variables Together

The right picture and summary depend on the types of the two variables:

| The two variables | Picture | Summary | Where |
| :- | :- | :- | :- |
| Numeric and categorical | Boxplots side by side: `boxplot(y ~ group)` | Median, IQR and n per group | this page |
| Categorical and categorical | Bar chart per group | Counts, then a proportion per group | this page |
| Numeric and numeric | Scatterplot | Correlation | Stage 3, [Relationships](10-relationships-and-correlation.qmd) |

You have already done the first two in the section above. **Two categorical
variables** start with `count()` on both, which counts every combination:

```{r}
#| label: crosstab
sessions |> count(build, quit_reason)
```

Then turn the one you care about into a proportion per group, as above:
`r pct_fmt(mean(sessions$quit_reason[sessions$build == "A"] == "Frustrated"))` of
Build A sessions ended in frustration, against
`r pct_fmt(mean(sessions$quit_reason[sessions$build == "B"] == "Frustrated"))` of
Build B.

**For now,** leave two numeric variables alone. Whether deaths go with difficulty,
or awareness with score, is the subject of Stage 3, once you have your own data to
ask it of.

---

## The Four-Part Description

A description of a numeric variable has four parts, and one sentence can hold them
all: **centre, spread, shape and size**.

> Median time alive was `r median(ta)` seconds (IQR `r IQR(ta)`,
> n = `r length(ta)`). The distribution has a tail of short sessions on the left and
> a pile-up on the right: `r n_cap` sessions sit at exactly `r ta_cap` seconds, the
> full match, because those players never died.

**For your project:** pick the one number your playtest question depends on and
write its four-part sentence. If you cannot fill in all four parts, you have not
looked at it properly yet.

Notice what that sentence does *not* say. The mean, `r round(mean(ta), 1)`, is only
`r round(abs(mean(ta) - median(ta)), 1)` seconds from the median, which might
suggest a symmetric distribution. It is not symmetric: the left tail pulls the mean
down, and the ceiling at `r ta_cap` seconds stops the right side from pulling it
back up. The two effects happen to balance. **A small gap between mean and median
does not prove symmetry**; only the histogram can show you that.

```{r}
#| label: fig-time-alive
#| fig-cap: "Time alive has a tail of short sessions on the left and a tall final bar where the match timer cut off sessions in which the player never died."
#| fig-alt: "Histogram of time alive from 60 to 180 seconds. Short bars on the left rise to a broad bulk between 100 and 170 seconds, with the tallest bar at the right edge, 180 seconds."
#| fig-height: 3.5
hist(ta, breaks = 12, col = pal$fill, border = "white",
     main = "", xlab = "Time alive (s)", ylab = "Sessions")
```

---

## Worked Example

**Is the Sentinel underpowered?** Sort the four archetypes by mean score and the
Sentinel comes last, at `r round(arch_mean[["Sentinel"]])` points. A balance pass
built on that number buffs the Sentinel.

**Step 1: compare centres.** Sort by median instead and the Sentinel is the
`r sen_median`, at `r round(arch_median[["Sentinel"]])`. The mean and median
disagree by `r round(arch_median[["Sentinel"]] - arch_mean[["Sentinel"]])` points,
with the mean below: something is dragging the mean down without moving the middle.

**Step 2: compare spreads.** The Sentinel's SD is
`r round(arch_sd[["Sentinel"]])` points, against `r round(min(others_sd))` to
`r round(max(others_sd))` for the other three archetypes.

```{r}
#| label: worked-by-archetype
sessions |>
  group_by(archetype) |>
  summarise(n      = n(),
            mean   = round(mean(score)),
            median = round(median(score)),
            sd     = round(sd(score)))
```

**Step 3: look.** The histogram answers it, but only at the right bin setting:

```{r}
#| label: fig-sentinel-bins
#| code-fold: true
#| code-summary: "Show the code"
#| fig-cap: "The same 30 Sentinel scores at three bin settings. With three bins the low cluster merges into the first bar; with more, it stands apart."
#| fig-alt: "Three histograms of Sentinel score side by side. With breaks = 3 there are three wide bars. With breaks = 5 and breaks = 12, a separate group of low scores is visible at the left, with an empty gap before the main body of scores."
#| fig-height: 3
op <- par(mfrow = c(1, 3), mar = c(4, 3, 2, 0.5))
for (k in c(3, 5, 12)) {
  hist(sentinel, breaks = k, col = pal$fill, border = "white",
       main = paste("breaks =", k), xlab = "Score", ylab = "", cex.main = 0.9)
}
par(op)
```

`r sum(sentinel < low_cut)` of the `r length(sentinel)` Sentinel sessions scored
under `r low_cut` and sit in a clump of their own,
`r min(sentinel[sentinel >= low_cut]) - max(sentinel[sentinel < low_cut])` points
below the next-lowest Sentinel game. They are the seven outliers in the boxplot
above. With `breaks = 3`, the clump disappears into the first bar.

**Step 4: rewrite the finding.** The Sentinel either anchors the round or collapses
early. It is not underpowered; it is inconsistent ([rule 4](ten-rules.qmd)).

**What should the team do?** Not a flat buff: that raises the scores of the
players already doing well. Watch the collapsing games and find what they have in
common, then design for that.

---

## Interpretation Guidance

**What the data shows.** The Sentinel's mean score is the lowest of the four
archetypes; its median is not; its scores fall into two groups.

**What the analysis suggests.** The low mean comes from a minority of collapsing
games, not from a generally weak character.

**What can legitimately be concluded.** A flat buff is not supported by this data.
Why some Sentinel games collapse is a new question; "some players do not know how
to play the role" is one candidate, and it needs a different kind of evidence
before it becomes a conclusion.

**Pick the summary to fit the shape, and say why.** Median and IQR for anything
lopsided or with outliers; mean and SD only when the histogram is roughly
symmetric. Whichever you use, report n.

---

## Common Mistakes

**Quoting a mean without looking.** The Sentinel's mean is a correct calculation
that describes no Sentinel game. Draw the histogram first ([rule 4](ten-rules.qmd)).

**Quoting the range as the spread.** One extreme player sets the range on their
own. Quote the IQR.

**Mixing pairs.** Median with SD, or mean with IQR. Pair each centre with its own
spread.

**Trusting one bin setting.** A histogram drawn once is one argument among several
the data could make. Redraw it with fewer and more bars.

**Reading a small mean–median gap as symmetry.** A tail and a ceiling can cancel
out. Look at the shape.

**Comparing groups of different sizes by count.** 20 frustration quits out of 60 is
not the same as 20 out of 100. Use row proportions.

**Leaving out n.** "Players rated it 2.6" means nothing without "out of how many"
([rule 8](ten-rules.qmd)).

---

## Practice

Each takes two or three minutes.

1. **Predict.** For frame times, where most frames are quick and a few are very
   slow, will the mean be above or below the median? Check with `mean(frames$ms)`
   and `median(frames$ms)`.
2. **Compute.** Describe `distance_m` in one sentence with centre, spread and size.
   Does the mean sit above or below the median, and does it matter?
3. **Critique.** A teammate writes: "Average deaths per session was 3.0, with a
   range of 0 to 7." What would you add or change?
4. **Compute.** Use `count()` to find the most common number of deaths. Why is
   counting sensible for `deaths` but not for `distance_m`?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
**1.** Above. A few very slow frames pull the mean up; the median ignores how slow
they are.

```{r}
#| label: practice-predict
mean(frames$ms)
median(frames$ms)
```

The gap is small here only because the slow frames are rare, which is the subject
of the worked example in [Distributions and Shape](05-distributions-and-shape.qmd).

**2.**

```{r}
#| label: practice-distance
sessions |>
  summarise(n = n(),
            median = median(distance_m),
            iqr = IQR(distance_m),
            mean = mean(distance_m))
```

Median distance was `r median(sessions$distance_m)` m (IQR
`r IQR(sessions$distance_m)`, n = `r nrow(sessions)`). The mean sits
`r if (mean(sessions$distance_m) > median(sessions$distance_m)) "above" else "below"`
the median by
`r round(abs(mean(sessions$distance_m) - median(sessions$distance_m)), 1)` m, which
is tiny against a spread of that size. The histogram earlier on this page confirms
the distribution is roughly symmetric, so the mean and SD would be fine too.

**3.** Add n (`r nrow(sessions)` sessions from
`r length(unique(sessions$player_id))` players). Replace the range with a spread
that is not set by single sessions: SD `r round(sd(sessions$deaths), 1)` if the
distribution is roughly symmetric, which the strip chart suggests it is, or the
IQR. The most common count, `r deaths_mode`, is also worth a mention, because it is
not the same as the average.

**4.**

```{r}
#| label: practice-deaths
sessions |> count(deaths)
```

The mode is `r deaths_mode` deaths. Deaths is discrete with only a handful of
possible values, so counting each one is informative.
`r length(unique(sessions$distance_m))` of the `r nrow(sessions)` `distance_m`
values are different, so counting it gives a list of ones; a continuous variable
needs a histogram instead.
:::

---

## Summary

- Describe a categorical variable with counts, proportions and its mode; draw a bar
  chart.
- The mean uses every value and is pulled by extremes; the median is the middle
  value and is not. When they disagree, the data is lopsided.
- The SD is the typical distance from the mean; the IQR is the width of the middle
  half. Pair mean with SD and median with IQR. The range is rarely enough.
- A histogram shows the shape; try several bin settings. A boxplot compares groups;
  its separate points are outliers to investigate, not to delete.
- `group_by()` then `summarise()` gives one row of summaries per group. Always
  include `n()`. Compare groups with proportions, not counts.
- A full description gives centre, spread, shape and n. A small mean–median gap
  does not prove symmetry.

| Key term | Meaning |
| :- | :- |
| **Frequency table** | Count of each category |
| **Proportion** | A count divided by the total |
| **Mode** | The most common value |
| **Mean** | The sum divided by the count |
| **Median** | The middle value when sorted |
| **Range** | Maximum minus minimum |
| **Variance** | The average squared distance from the mean |
| **Standard deviation** | The square root of the variance: the typical distance from the mean |
| **Quartiles** | The values that cut the sorted data into four equal parts |
| **IQR** | Q3 − Q1: the width of the middle half |
| **Percentile** | The value with that percentage of the data below it |
| **Five-number summary** | Minimum, Q1, median, Q3, maximum |
| **Histogram** | Bars counting values in equal-width bins |
| **Boxplot** | Median, quartiles, whiskers and outliers in one picture |
| **Cross-tabulation** | Counts of every combination of two categorical variables |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 4 (look at the shape before you quote the
  average), rule 5 (the mean is the wrong summary more often than you think) and
  rule 8 (small samples lie confidently)
- [Statistics Reference: centre and spread](../reference/reference-statistics.qmd#centre-and-spread)
- [R and Quarto Reference: describing one variable](../reference/r-quarto-reference.qmd#describing-one-variable),
  [describing by group](../reference/r-quarto-reference.qmd#describing-by-group)
  and [plots](../reference/r-quarto-reference.qmd#plots)
- [Lab 01: Reading Data](../labs/reading-data.qmd), Part B
- Next: [Distributions and Shape](05-distributions-and-shape.qmd)

---

## Reflective Questions

- Why is the median usually the right summary for completion time, but the mean
  the right one for total damage dealt across a team?
- You draw a histogram of level completion times and see two clear humps. List
  three things that could have caused that, and say how you would tell them apart.
- A teammate reports "average session length is 22 minutes". What two further
  numbers would you ask for before letting that into a report?
- The Sentinel's outliers are seven real games. Under what circumstances, if any,
  would you leave them out of a balance analysis?
