---
title: "Distributions and Shape"
subtitle: "Why the average is never the whole story"
body-classes: "wf-at-visualise wf-at-describe"
aliases:
  - distributions-and-variation.html
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")

# Frame times
a <- frames$ms[frames$build == "A"]
b <- frames$ms[frames$build == "B"]
fps <- function(x) round(1000 / mean(x), 1)
pct <- function(x, q) round(unname(quantile(x, q)), 1)
slow   <- 1000 / 30
slow_b <- frames[frames$build == "B" & frames$ms > slow, ]
gap    <- unique(diff(sort(slow_b$frame)))
stopifnot(mean(b) < mean(a), median(b) < median(a),
          quantile(b, 0.95) < quantile(a, 0.95),
          sum(a > slow) == 0,
          all(slow_b$zone == "Corridor"),
          length(gap) == 1)
corridor_a <- frames$ms[frames$build == "A" & frames$zone == "Corridor"]
corridor_b <- frames$ms[frames$build == "B" & frames$zone == "Corridor"]

# Normal comparison
within_sd <- function(x, k) mean(abs(x - mean(x)) < k * sd(x))
aw <- sessions$awareness
sm <- sessions$session_minutes

# Session length by build, and retention
mins_a <- sessions$session_minutes[sessions$build == "A"]
mins_b <- sessions$session_minutes[sessions$build == "B"]
still  <- function(x, m) mean(x >= m)
stopifnot(round(still(mins_a, 11), 2) == 0.88, round(still(mins_a, 13), 2) == 0.68)
early_a <- sum(mins_a < 15); early_b <- sum(mins_b < 15)

# Outliers
score_out <- boxplot.stats(sessions$score)$out
stopifnot(all(sessions$archetype[sessions$score %in% score_out] == "Sentinel"))
frame_out_a <- length(boxplot.stats(a)$out)
frame_out_b <- length(boxplot.stats(b)$out)
```

> A summary is a single number standing in for a whole column. This page is about
> what it stands in for: the **distribution**, the full pattern of values. It names
> the shapes you will meet, introduces the normal distribution that later notes
> rely on, deals with outliers, and shows why the most important number in a
> performance capture is one that hardly anyone quotes. The playtest question
> running through it: *"Both builds run at 59 FPS. Which one runs better?"*

::: {.at-a-glance}
**Time:** 30 minutes in class; 20 more for the Going further sections.
**Before this:** [Describing Variables](04-describing-variables.qmd).
**Key terms:** distribution, symmetric, skew, bimodal, floor, ceiling, normal
distribution, z-score, tail, outlier, retention curve.
**R introduced:** `dnorm()`, `curve()`, `boxplot.stats()`, `abline()`,
`lines()`, plus `hist()`, `boxplot()`, `quantile()` and `tapply()` from
[Describing Variables](04-describing-variables.qmd).
:::

## Learning Outcomes

By the end of these notes you can:

- describe a distribution by its shape: symmetric, skewed left or right, bimodal,
  bounded by a floor or ceiling;
- say which way skew pulls the mean, and why game data is rarely symmetric;
- describe the normal distribution, use the 68–95–99.7 rule, and compute a z-score;
- check a variable against the normal shape and say whether the match is close
  enough to matter;
- say whether an outlier is an error, an extreme player or the finding itself, and
  what to do in each case;
- compare two groups' distributions, including with a retention curve;
- read the mean, median, SD and high percentiles together and say what each adds.

---

## Why This Matters for Playtesting

Every design question is really a question about players, plural, and players
differ. A level that takes eight minutes on average can be a level everyone
finishes in eight minutes, or a level half the testers finish in four and the other
half abandon. Those are different games with different fixes, and the average
cannot tell them apart.

**The players you design for are rarely the average player.** The ones who quit,
the ones who get stuck, the ones whose machine stutters: they live in the edges of
the distribution, and the edges are exactly what a mean averages away.

**Variation is information, not noise.** When two builds differ in how spread out
their players are, that difference is often the finding. A build that tightens the
spread has made the experience more consistent, even if the average has not moved.

---

## What a Distribution Is

A **distribution** is the full pattern of a variable's values: which values occur,
and how often. A histogram draws it. Every summary statistic is an attempt to
describe a distribution in one number, and every one of them fails for some
distribution.

Describing a distribution means saying four things about its picture:

1. **Centre**: where the bulk sits.
2. **Spread**: how wide the bulk is.
3. **Shape**: symmetric or lopsided, one hump or two, cut off at an edge or not.
4. **Unusual values**: anything sitting apart from the rest.

The first two were [Describing Variables](04-describing-variables.qmd). This page
is about the last two.

---

## The Shapes You Will Meet

```{r}
#| label: fig-shapes
#| echo: false
#| fig-cap: "Four shapes. Top row: idealised illustrations. Bottom row: the same shape in Tidebreak data. The solid red line is the mean; the dashed line is the median. *The frame-time panel's vertical axis is cut off so that the rare slow frames are visible; the spike near 17 ms runs far off the top."
#| fig-alt: "Eight histograms in two rows. Top row, simulated: a symmetric bell, a right-skewed shape with a long right tail, a left-skewed shape with a long left tail, and a two-humped shape. Bottom row, Tidebreak: awareness, roughly symmetric with mean and median together; Build B frame times, a spike cut off at the top with a long thin right tail of slow frames; session length, piled up against 32 minutes with a tail to the left; Sentinel score, a small group of low scores and a larger group of high scores. In each skewed panel the mean sits further towards the tail than the median."
#| fig-height: 5.5
#| fig-width: 9
set.seed(42)
sim <- list(
  "Symmetric"  = rnorm(2000, 50, 10),
  "Right skew" = rgamma(2000, shape = 2, scale = 8),
  "Left skew"  = 60 - rgamma(2000, shape = 2, scale = 8),
  "Bimodal"    = c(rnorm(700, 30, 6), rnorm(1300, 65, 8))
)
real <- list(
  "Symmetric: awareness"        = aw,
  "Right skew: Build B frame ms*" = b,
  "Left skew: session minutes"  = sm,
  "Bimodal: Sentinel score"     = sessions$score[sessions$archetype == "Sentinel"]
)
panel <- function(x, title, breaks = 20, ylim = NULL) {
  hist(x, breaks = breaks, col = pal$fill, border = "white", main = title,
       xlab = "", ylab = "", yaxt = "n", cex.main = 0.85, ylim = ylim)
  abline(v = mean(x), col = pal$highlight, lwd = 2)
  abline(v = median(x), col = pal$line, lwd = 2, lty = 2)
}
op <- par(mfrow = c(2, 4), mar = c(2.5, 0.8, 2.2, 0.8), oma = c(0, 0, 1.2, 0))
for (n in names(sim)) panel(sim[[n]], n)
mtext("Illustration: simulated data", side = 3, outer = TRUE, adj = 1,
      cex = 0.7, col = pal$guide)
for (n in names(real)) {
  if (grepl("frame", n)) panel(real[[n]], n, breaks = 60, ylim = c(0, 40))
  else panel(real[[n]], n, breaks = 12)
}
par(op)
```

**Symmetric.** The two sides mirror each other. The mean and median sit together,
and either describes the centre. *Tidebreak*'s awareness rating is close to
symmetric: mean `r round(mean(aw), 1)`, median `r median(aw)`.

**Right skew.** A long tail of large values. Most frames are quick and a few are
very slow; most players finish in a few minutes and a few take far longer. The
mean is pulled out towards the tail, above the median.

**Left skew.** A long tail of small values. Most *Tidebreak* sessions run close to
the 32-minute limit and a minority end early: mean `r round(mean(sm), 1)` minutes,
median `r round(median(sm), 1)`. The mean is pulled down, below the median.
Look closely and session length also has a small second hump of early quits; where
it comes from is the subject of [Comparing Distributions](#comparing-distributions).

**Bimodal.** Two humps. Almost always, two groups of players have been mixed
together: those who found the key and those who did not, sessions that completed
and sessions that crashed. The mean of a bimodal distribution falls in the valley
between the humps, where the fewest players are. The Sentinel's scores are the
example from [Describing Variables](04-describing-variables.qmd#worked-example).

**The mean is always pulled towards the tail; the median is not.** That is why the
gap between them tells you which way the data leans, and why a large gap is a
reason to report the median.

### Floors and ceilings

**Game data is rarely symmetric, because it is usually bounded.** Times have a
**floor**: nobody finishes faster than the fastest possible route, so time data
often has a long right tail. Many measures also have a **ceiling**: a match timer,
a maximum score, the top of a rating scale. Values pile up against it and the tail
runs the other way.

*Tidebreak* has both kinds of bounded time. Frame times have a floor and a long
right tail. Session length is capped at `r max(sm)` minutes and time alive at
`r max(sessions$time_alive_s)` seconds, so both pile up on the right with a tail to
the left. **Expect skew in any bounded measure, and check which way it runs**,
rather than assuming the bell shape from school statistics.

::: {.callout-tip collapse="true" title="Quick check"}
Players rate a game's difficulty from 1 to 7, and most choose 6 or 7. Which way is
the distribution skewed, and is the mean above or below the median?

**Answer.** Left skew: values pile up against the ceiling at 7, with a tail of
lower ratings to the left. The tail pulls the mean down, below the median.
:::

---

## The Normal Distribution

One shape has a name of its own, because so much of statistics is built on it.

The **normal distribution** is the symmetric bell curve. It is completely
described by two numbers: its **mean**, which sets where the centre is, and its
**SD**, which sets how wide it is. Every normal distribution has the same shape;
only the centre and width change.

```{r}
#| label: fig-normal
#| echo: false
#| fig-cap: "The 68–95–99.7 rule. In a normal distribution, about 68% of values lie within 1 SD of the mean, 95% within 2 SDs and 99.7% within 3."
#| fig-alt: "A bell curve with the x axis marked in standard deviations from −3 to +3. The region within 1 SD of the mean is shaded darkest and labelled 68%; the region within 2 SDs is labelled 95%; within 3 SDs, 99.7%."
#| fig-height: 3.8
par(mar = c(4, 1, 1.5, 1))
xs <- seq(-3.6, 3.6, length.out = 400)
plot(xs, dnorm(xs), type = "n", axes = FALSE, xlab = "Standard deviations from the mean",
     ylab = "", ylim = c(0, 0.47))
shade <- function(k, col) {
  x <- seq(-k, k, length.out = 200)
  polygon(c(-k, x, k), c(0, dnorm(x), 0), col = col, border = NA)
}
shade(3, "grey92"); shade(2, "grey80"); shade(1, "#e8a3a3")
lines(xs, dnorm(xs), lwd = 2)
axis(1, at = -3:3, labels = c("−3", "−2", "−1", "mean", "+1", "+2", "+3"))
arrows(-1, 0.43, 1, 0.43, code = 3, length = 0.05); text(0, 0.45, "68%", cex = 0.85)
arrows(-2, 0.09, 2, 0.09, code = 3, length = 0.05); text(0, 0.11, "95%", cex = 0.85)
arrows(-3, 0.02, 3, 0.02, code = 3, length = 0.05); text(2.9, 0.045, "99.7%", cex = 0.8)
label_illustration("Illustration: the theoretical curve")
```

**The 68–95–99.7 rule.** In any normal distribution, about 68% of values lie within
1 SD of the mean, about 95% within 2 SDs, and nearly all, 99.7%, within 3. This is
what gives the SD its practical meaning: for normal-shaped data, "the mean plus or
minus 2 SDs" covers almost everyone.

**A z-score** says how many SDs a value sits from the mean:

$$z = \frac{\text{value} - \text{mean}}{\text{SD}}$$

A player with an awareness rating of 90 has
$z = (90 - `r round(mean(aw), 1)`) / `r round(sd(aw), 1)` = `r round((90 - mean(aw)) / sd(aw), 1)`$:
nearly 2 SDs above average, which the rule says puts them in roughly the top few
per cent. z-scores let you compare values measured on different scales: "two SDs
above average" means the same thing for a score, a time or a rating.

### Is it normal enough?

Real data is never exactly normal. The useful question is whether it is close
enough for the rule to work. Draw the histogram on a density scale and lay the
normal curve with the same mean and SD over it:

```{r}
#| label: fig-normal-check
#| fig-cap: "The normal curve with each variable's own mean and SD, laid over its histogram. Awareness follows it closely; session length does not."
#| fig-alt: "Two histograms. Left: awareness, with a bell curve that follows the bars closely. Right: session length, where the bell curve overestimates the middle and misses the tall bars piled up near 30 minutes."
#| fig-height: 3.5
op <- par(mfrow = c(1, 2), mar = c(4, 3, 2, 1))
hist(aw, breaks = 12, freq = FALSE, col = pal$fill, border = "white",
     main = "Awareness", xlab = "Rating", ylab = "")
curve(dnorm(x, mean(aw), sd(aw)), add = TRUE, lwd = 2, col = pal$highlight)
hist(sm, breaks = 12, freq = FALSE, col = pal$fill, border = "white",
     main = "Session length", xlab = "Minutes", ylab = "")
curve(dnorm(x, mean(sm), sd(sm)), add = TRUE, lwd = 2, col = pal$highlight)
par(op)
```

The rule makes a checkable prediction, so check it:

| Variable | Within 1 SD | Within 2 SDs | Verdict |
| :- | :-: | :-: | :- |
| Normal distribution | 68% | 95% | |
| `awareness` | `r pct_fmt(within_sd(aw, 1))` | `r pct_fmt(within_sd(aw, 2))` | close: the rule works |
| `distance_m` | `r pct_fmt(within_sd(sessions$distance_m, 1))` | `r pct_fmt(within_sd(sessions$distance_m, 2))` | symmetric but flatter than a bell |
| `session_minutes` | `r pct_fmt(within_sd(sm, 1))` | `r pct_fmt(within_sd(sm, 2))` | not normal: skewed and capped |
| Build B frame ms | `r pct_fmt(within_sd(b, 1))` | `r pct_fmt(within_sd(b, 2))` | not normal: a spike and a long tail |

For session length, "mean ± 1 SD" covers only `r pct_fmt(within_sd(sm, 1))` of
sessions, so quoting the mean and SD as if they described a bell would mislead.
For frame times, the SD is inflated by a handful of very slow frames, and the
mean ± 2 SDs covers almost every frame while saying nothing about the ones that
matter.

**Why it matters later.** Even when individual values are far from normal, the
*averages* of samples tend to be close to normal. That fact, the central limit
theorem, is what makes confidence intervals and t-tests work, and it is the
subject of [From Sample to Population](07-sample-to-population.qmd#the-central-limit-theorem).
When later notes ask whether data is
"approximately normal", this is the shape they mean.

::: {.callout-note collapse="true" title="Going further: Q-Q plots"}
A **Q-Q plot** is a sharper check than a histogram. It plots each value against
where it would fall if the data were perfectly normal; normal data lies along a
straight line, and skew bends the ends away from it. In base R:

```{r}
#| label: qq
#| fig-height: 3.2
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
qqnorm(aw, main = "Awareness", pch = 19, cex = 0.6); qqline(aw, col = pal$highlight)
qqnorm(sm, main = "Session length", pch = 19, cex = 0.6); qqline(sm, col = pal$highlight)
par(op)
```

Awareness hugs the line. Session length flattens out at the top, where values pile
up against the 32-minute cap. You will meet Q-Q plots again when checking the
assumptions of a t-test.
:::

---

## Outliers

An **outlier** is a value far from the rest. A boxplot flags any value more than
1.5 IQRs beyond the box, and draws it as a separate point. **That rule is a prompt
to look, not a verdict.** With large samples it flags plenty of perfectly ordinary
values: of Build A's `r length(a)` frames, `r frame_out_a` are boxplot outliers,
and Build A has no performance problem at all.

An outlier is one of three things, and the response to each is different:

```{mermaid}
%%| label: fig-outlier-tree
%%| fig-cap: "What to do about an outlier depends on what it is."
flowchart TD
  O["A value sits apart<br/>from the rest"] --> Q1{"Is it possible?"}
  Q1 -->|"No: a session of −3 minutes,<br/>a rating of 7 on a 1–5 scale"| E["An error<br/>Fix or remove it,<br/>and record that you did"]
  Q1 -->|Yes| Q2{"Is it one player<br/>or a pattern?"}
  Q2 -->|"One unusual player"| X["An extreme player<br/>Keep it; report the median<br/>rather than the mean"]
  Q2 -->|"A cluster, or a<br/>repeating pattern"| F["The finding<br/>Keep it and investigate:<br/>it is why you ran the test"]
  classDef err fill:#f8f9fa,stroke:#6c757d,color:#212529
  classDef find fill:#fbeaea,stroke:#b22222,color:#212529
  class E,X err
  class F find
```

| The outlier is | Tidebreak example | What you do |
| :- | :- | :- |
| An error | None in the cleaned file; [Cleaning and Transforming](06-cleaning-and-transforming.qmd) finds some in the raw export | Fix it or remove it, and record that you did |
| An extreme but ordinary player | The `r sum(sessions$time_alive_s == max(sessions$time_alive_s))` sessions where the player never died | Keep it; report the median rather than the mean |
| The finding | The `r length(score_out)` score outliers, every one a Sentinel game | Keep it and investigate; it is why you ran the test |

**Never delete an outlier because it is inconvenient.** Every row you drop is a
decision with a consequence ([rule 6](ten-rules.qmd)), and the value most tempting
to drop is often the one the playtest was run to find ([rule 9](ten-rules.qmd)).

---

## Comparing Distributions

Most design questions compare two distributions: Build A against Build B, Novices
against Veterans. **Compare the whole distributions, not just their centres**
([rule 7](ten-rules.qmd)).

The boxplot is the standard tool, and usually a good one. But it has a blind spot:

```{r}
#| label: fig-box-vs-hist
#| fig-cap: "Session length by build. The boxplots (left) show that Build A's sessions are more spread out. Only the histograms (right) show why: Build A has two separate groups of sessions."
#| fig-alt: "Left: two boxplots of session length. Build A's box is wider and lower than Build B's. Right: two stacked histograms. Build A's has two clear humps, one around 10 to 14 minutes and one around 26 to 32. Build B's rises steadily to a single peak near 30 minutes."
#| fig-height: 4.2
layout(matrix(c(1, 1, 2, 3), nrow = 2), widths = c(1, 1.6))
par(mar = c(4, 4, 2, 1))
boxplot(session_minutes ~ build, data = sessions, col = pal$fill,
        ylab = "Session length (minutes)", xlab = "Build", main = "Boxplots",
        cex.main = 0.9)
par(mar = c(2.5, 3, 2, 1))
hist(mins_a, breaks = seq(6, 32, 2), col = pal$build_a, border = "white",
     main = "Build A", xlab = "", ylab = "", cex.main = 0.9, ylim = c(0, 14))
hist(mins_b, breaks = seq(6, 32, 2), col = pal$build_b, border = "white",
     main = "Build B", xlab = "", ylab = "", cex.main = 0.9, ylim = c(0, 14))
layout(1)
```

The boxplots say Build A's sessions are shorter and more spread out: an IQR of
`r round(IQR(mins_a), 1)` minutes against `r round(IQR(mins_b), 1)`. True, but it
misses the point. Build A's histogram has **two humps**: a group of sessions that
ended between about 10 and 14 minutes, and a group that ran to the end. A boxplot
compresses every distribution into one box, so it cannot show two humps. Its wide
box is the two groups averaged into one.
`r early_a` of `r length(mins_a)` Build A sessions ended before minute 15, against
`r early_b` of `r length(mins_b)` in Build B.

### The retention curve

When the variable is "how long did players keep playing", there is a picture made
for the question. A **retention curve** plots, for every minute, the proportion of
sessions still going:

```{r}
#| label: fig-retention
#| fig-cap: "Retention: the share of sessions still going at each minute. Build A loses a fifth of its sessions between minutes 11 and 13."
#| fig-alt: "Two step lines starting at 100%. Build B's stays above 90% until about minute 14 and then declines. Build A's drops steeply between minutes 11 and 13, from 88% to 68%, then declines slowly. Both fall sharply near 32 minutes, the end of the session."
#| fig-height: 3.8
minute <- 0:32
ret_a <- sapply(minute, function(m) mean(mins_a >= m))
ret_b <- sapply(minute, function(m) mean(mins_b >= m))

par(mar = c(4, 4, 1, 1))
plot(minute, ret_a, type = "s", lwd = 2, col = pal$build_a, ylim = c(0, 1),
     xlab = "Minute", ylab = "Share of sessions still playing", yaxt = "n")
axis(2, at = seq(0, 1, 0.25), labels = paste0(seq(0, 100, 25), "%"), las = 1)
lines(minute, ret_b, type = "s", lwd = 2, col = pal$build_b)
abline(v = c(11, 13), lty = 3, col = pal$guide)
legend("bottomleft", legend = c("Build A", "Build B"), lwd = 2,
       col = c(pal$build_a, pal$build_b), bty = "n")
```

Each point is a single `mean()` of a comparison, the "proportion that pass a test"
idea from [R from First Principles](02-r-first-principles.qmd#vectors), repeated
for every minute. Build A drops from `r pct_fmt(still(mins_a, 11))` still playing
at minute 11 to `r pct_fmt(still(mins_a, 13))` at minute 13. Build B, over the same
two minutes, goes from `r pct_fmt(still(mins_b, 11))` to
`r pct_fmt(still(mins_b, 13))`. Something about the first dozen minutes of Build A
drives players away, and the retention curve says exactly when.

---

## Reading Several Summaries Together

No single statistic describes a distribution, but a small set of them, read
together, usually does.

| Summary | What it tells you | What it misses |
| :- | :- | :- |
| Mean | The total spread evenly across players | Skew, clusters, tails |
| Median | The middle player | How extreme the extremes are |
| Standard deviation | Typical distance from the mean | Which direction the spread goes |
| IQR | Width of the middle half | Everything outside the middle half |
| 95th or 99th percentile | How bad the worst few per cent get | The typical case |

**Where the summaries disagree, look at the picture.** Mean and median apart means
skew. A large SD with a small IQR means a few extreme values. A normal median with
a bad 99th percentile means a rare problem that the average will never show. Each
disagreement points at a specific shape, and the histogram confirms which one
([rule 4](ten-rules.qmd)).

---

## Worked Example

**Which build runs better?** The *Tidebreak* team captured `r length(a)` frames
from each build, split across `r length(unique(frames$zone))` level zones. The
frame-rate counter reports Build A at `r fps(a)` FPS and Build B at `r fps(b)` FPS.
By the number everyone quotes, Build B is marginally the faster build.

**Step 1: more centres.** The median agrees: `r round(median(a), 2)` ms per frame
for Build A, `r round(median(b), 2)` ms for Build B. So does the 95th percentile:
`r pct(a, 0.95)` ms against `r pct(b, 0.95)` ms. Three summaries, one verdict.

**Step 2: the tail.** The 99th percentile disagrees. Build A's is `r pct(a, 0.99)`
ms; Build B's is `r pct(b, 0.99)` ms,
`r sprintf("%.1f", quantile(b, 0.99) / quantile(a, 0.99))` times as long. The SD
says the same thing more quietly: `r round(sd(a), 2)` ms for Build A,
`r round(sd(b), 2)` ms for Build B.

```{r}
#| label: worked-summary
frame_summary <- function(x) {
  round(c(mean = mean(x), median = median(x), sd = sd(x),
          p95 = unname(quantile(x, 0.95)), p99 = unname(quantile(x, 0.99)),
          fps = 1000 / mean(x)), 2)
}
rbind(A = frame_summary(a), B = frame_summary(b))
```

**Step 3: look.**

```{r}
#| label: fig-frames-hist
#| fig-cap: "Build B's frame times, with the 95th and 99th percentiles marked. The vertical axis is cut off at 30 frames so the rare slow frames show. The 95th percentile sits at the edge of the main spike; only the 99th reaches the slow frames out on the right."
#| fig-alt: "Histogram of Build B frame times. A narrow spike between about 13 and 20 milliseconds runs off the top of the chart. A scatter of short bars lies far to the right, between 30 and 55 milliseconds. A dashed line marks the 95th percentile inside the spike; a red line marks the 99th percentile out among the slow frames."
#| fig-height: 3.6
par(mar = c(4, 4, 1, 1))
hist(b, breaks = 80, col = pal$fill, border = pal$fill, main = "",
     xlab = "Frame time (ms)", ylab = "Frames", ylim = c(0, 30))
text(20.5, 26, "<- main spike continues
    to over 800 frames", pos = 4, cex = 0.75,
     col = pal$line)
abline(v = quantile(b, 0.95), lty = 2, lwd = 2, col = pal$line)
abline(v = quantile(b, 0.99), lwd = 2, col = pal$highlight)
legend("topright", bty = "n", lwd = 2, lty = c(2, 1),
       col = c(pal$line, pal$highlight),
       legend = c("95th percentile", "99th percentile"))
```

**Step 4: find it.** Count the frames slower than 30 FPS (`r round(slow, 1)` ms).
Build A has `r sum(a > slow)`. Build B has `r nrow(slow_b)`, every one of them in
the Corridor zone, and exactly `r gap` frames apart: at 60 FPS, a hitch
`r round(60 / gap, 1)` times a second for as long as the player is in the corridor.

```{r}
#| label: fig-frame-trace
#| fig-cap: "Frame times in the Corridor, first 200 frames. Build B hitches at a fixed interval; Build A does not."
#| fig-alt: "Two line traces of frame time over 200 frames. Build A's grey line stays flat around 17 milliseconds. Build B's red line also sits around 17 milliseconds but spikes to over 40 at regular intervals."
#| fig-height: 3.6
par(mar = c(4, 4, 1, 1))
plot(corridor_b[1:200], type = "l", col = pal$build_b, ylim = c(10, 65),
     xlab = "Frame", ylab = "Frame time (ms)")
lines(corridor_a[1:200], col = pal$build_a)
abline(h = 1000 / 60, lty = 2, col = pal$guide)
legend("top", legend = c("Build A", "Build B"), horiz = TRUE,
       col = c(pal$build_a, pal$build_b), lty = 1, bty = "n")
```

`plot()` with `type = "l"` joins the points with a line instead of drawing them
separately: the right choice when the order of the values matters, as it does for
frames in time. The dashed line is the 16.7 ms budget for 60 FPS.

**Why the mean cannot see it.** The slow frames are
`r round(100 * nrow(slow_b) / length(b), 1)`% of Build B's total. Spread across
every frame of the capture, they barely move the mean, and Build B's faster frames
elsewhere more than make up for them. The 95th percentile sits below the problem
because the problem is rarer than one frame in twenty. Only a summary aimed at the
worst one per cent can see it, which is why the 99th percentile is the standard
figure in performance capture. The mean is not approximately wrong here. It is
structurally incapable of showing the thing players experience.

---

## Interpretation Guidance

**What the data shows.** Build A and Build B have near-identical mean frame times.
Build B has `r nrow(slow_b)` frames slower than 30 FPS, all in the Corridor, and
Build A has `r sum(a > slow)`.

**What the analysis suggests.** Something in Build B's Corridor costs a frame at a
fixed interval. The regular spacing points at a periodic job, such as streaming,
garbage collection or a timed effect, rather than at scene complexity.

**What can legitimately be concluded.** Build B introduced a hitch confined to the
Corridor, and a report that quotes average FPS alone would have shipped it. The
cause is not established by frame times; that needs a profiler.

**Quote the tail with the centre.** For anything where the worst case is what
players remember, such as frame time, load time or matchmaking wait, report the
median and a high percentile together. "Median `r round(median(b), 1)` ms, 99th
percentile `r round(quantile(b, 0.99))` ms" is a performance report.
"`r round(1000 / mean(b))` FPS" is a slogan.

---

## Common Mistakes

**Reporting average FPS as the performance result.** It is the number the engine
shows and the number least able to show a hitch.

**Comparing groups by their means alone.** Two builds with the same mean can be
different games ([rule 7](ten-rules.qmd)). Draw the distributions side by side.

**Trusting a boxplot to show shape.** It cannot show two humps. Draw the histogram
as well.

**Assuming the bell shape.** Bounded measures, which is most game measures, are
usually skewed. Check which way.

**Quoting mean ± SD for skewed data.** It assumes a normal shape the data does not
have. Use the median and IQR, or percentiles.

**Deleting outliers to clean the chart.** Find out which of the three kinds each one
is first. The Sentinel's low scores are the finding, not the noise.

**Stopping at the 95th percentile.** A problem affecting one frame in a hundred is
invisible at the 95th. Match the percentile to how rare the problem could be.

---

## Practice

Each takes two or three minutes.

**1. Predict.** Before running anything: will `items_collected` be skewed? Which
way? Then check with a histogram, the mean and the median.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-items
#| fig-height: 3
hist(sessions$items_collected, breaks = 0:9 - 0.5, col = pal$fill,
     border = "white", main = "", xlab = "Items collected")
mean(sessions$items_collected)
median(sessions$items_collected)
```

Close to symmetric: the mean, `r round(mean(sessions$items_collected), 2)`, sits
almost on the median, `r median(sessions$items_collected)`. A count with a floor at
zero *can* be right-skewed, but only when the typical value is near the floor; here
it is well away from it.
:::

**2. Compute.** Find the 99th percentile of frame time for each zone in Build B.
Which zone is responsible?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-zone
frames_b <- frames[frames$build == "B", ]
tapply(frames_b$ms, frames_b$zone, quantile, 0.99)
```

The Corridor, at `r round(quantile(corridor_b, 0.99), 1)` ms. The other two zones
are at most
`r round(max(tapply(frames_b$ms, frames_b$zone, quantile, 0.99)[c("Atrium", "Gantry")]), 1)` ms.
Splitting a summary by where it happened turns "Build B stutters" into "Build B
stutters in one place", which is a ticket someone can fix.
:::

**3. z-score.** A session scored 2,000. Using the mean and SD of all scores, what
is its z-score? Is the 68–95–99.7 rule a safe way to say how unusual it is?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-z
(2000 - mean(sessions$score)) / sd(sessions$score)
```

About `r round((2000 - mean(sessions$score)) / sd(sessions$score), 1)` SDs above
the mean. But score is not normal: the Sentinel collapses give it a left tail,
which inflates the SD and drags the mean down. The rule would be a rough guide at
best. The percentile is more honest:
`r pct_fmt(mean(sessions$score < 2000))` of sessions scored below 2,000.
:::

**4. Critique.** "Build A's sessions had a wider spread than Build B's, so Build A
is less consistent." Is that the best description of what happened?

::: {.callout-note collapse="true" title="Check your answer"}
It is true but it misses the finding. Build A is not uniformly inconsistent; it
has two distinct groups, one quitting around minutes 10 to 14 and one playing to
the end. "A fifth of Build A sessions end between minutes 11 and 13, mostly in
frustration" names a moment a designer can investigate. The wider spread is a
symptom of that, and only the histogram or the retention curve shows it.
:::

---

## Summary

- A distribution is the full pattern of values. Describe it by centre, spread,
  shape and unusual values.
- Skew pulls the mean towards the tail. Bounded game measures are usually skewed:
  a floor gives a right tail, a ceiling a left one.
- Two humps usually mean two groups mixed together. A boxplot cannot show them.
- The normal distribution is the symmetric bell; 68% of values lie within 1 SD,
  95% within 2. Check before assuming it, by overlaying the curve.
- A z-score is the distance from the mean in SDs.
- An outlier is an error, an extreme player or the finding. Work out which before
  doing anything to it.
- Compare whole distributions. For "how long did players stay", use a retention
  curve.
- For performance data, quote a high percentile. Both builds run at about
  `r round(1000 / mean(frames$ms))` FPS, and only one of them stutters.

| Key term | Meaning |
| :- | :- |
| **Distribution** | The full pattern of a variable's values |
| **Skew** | A long tail on one side; the mean is pulled towards it |
| **Bimodal** | Two humps, usually two groups mixed together |
| **Floor, ceiling** | A lower or upper limit values cannot pass |
| **Normal distribution** | The symmetric bell curve, defined by its mean and SD |
| **68–95–99.7 rule** | The share of normal values within 1, 2 and 3 SDs of the mean |
| **z-score** | How many SDs a value sits from the mean |
| **Tail** | The extreme values at either end of a distribution |
| **Outlier** | A value far from the rest; an error, an extreme player or the finding |
| **Retention curve** | The share of sessions still going at each point in time |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 4 (look at the shape before you quote the
  average), rule 7 (compare distributions, not averages) and rule 9 (sometimes the
  rare value is the whole story)
- [Statistics Reference: shape](../reference/reference-statistics.qmd#shape)
- [R and Quarto Reference: plots](../reference/r-quarto-reference.qmd#plots)
- [Lab 01: Reading Data](../labs/reading-data.qmd), Part B
- Next: [Cleaning and Transforming](06-cleaning-and-transforming.qmd)

---

## Reflective Questions

- A producer asks for "one number" to show the build's performance has improved.
  Which number do you give, and what do you say about the numbers you left out?
- Build A's session lengths are more spread out than Build B's. Is a narrower spread
  always better for a game? Give a case where it would not be.
- You find three sessions with a score of zero. What would you check before
  deciding they are errors, and what would change your mind?
- The Corridor hitch affects
  `r round(100 * nrow(slow_b) / length(b), 1)`% of Build B's frames. At what point
  does a rare problem stop being worth fixing, and who should decide?
