---
title: "Cleaning and Transforming"
subtitle: "Every cleaning choice is part of the analysis"
body-classes: "wf-at-clean"
aliases:
  - cleaning-and-transforming.html
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")
library(dplyr)

original <- sessions
raw      <- read_csv("data/tidebreak_sessions_raw.csv",
                     trim_ws = FALSE, show_col_types = FALSE)

deduped <- raw[!duplicated(raw), ]
fixed   <- tolower(trimws(deduped$cohort))
mislabelled <- sum(!deduped$cohort %in% c("Novice", "Regular", "Veteran"))
blank   <- deduped[is.na(deduped$session_minutes), ]
kept    <- deduped[complete.cases(deduped), ]

frus  <- function(d, b) sum(d$build == b & d$quit_reason == "Frustrated")
med_a <- function(d) median(d$session_minutes[d$build == "A"], na.rm = TRUE)

stopifnot(sum(duplicated(raw)) == sum(duplicated(raw$session_id)),
          nrow(deduped) == nrow(original),
          all(blank$build == "A"),
          all(blank$quit_reason == "Frustrated"),
          length(unique(fixed)) == 3,
          med_a(deduped) > med_a(original),
          sum(deduped$session_minutes <= 0, na.rm = TRUE) == 0,
          all(deduped$enjoyment %in% 1:7),
          all(head(deduped$archetype[order(deduped$score)], 5) == "Sentinel"))
```

> No playtest export arrives ready to analyse. This page has two halves. The first
> is **cleaning**: finding missing values, duplicates, impossible values and
> broken labels, and dealing with each without quietly changing the answer. The
> second is **transforming**: the pipe and the six `dplyr` verbs that reshape a
> clean table into the one your question needs. The playtest question running
> through it: *"Which build keeps players longer?"*, and how a single line of
> cleaning nearly hides the answer.

::: {.at-a-glance}
**Stage:** 2 · Core.
**Time:** 30 minutes in class; 20 more to work through the R.
**Before this:** [Distributions and Shape](05-distributions-and-shape.qmd).
**Key terms:** missing value, duplicate, key, impossible value, factor, cleaning
log, pipe, verb, split–apply–combine.
**R you will write:** `is.na()`, `distinct()`, `trimws()`, `tolower()`,
`factor()`, `filter()`, `select()`, `mutate()`, `arrange()`, `group_by()`,
`summarise()`, `count()`, `n()`.
:::

## Learning Outcomes

By the end of these notes you can:

- say what an `NA` means in a playtest export, and check whether the pattern of
  missing values is itself a finding;
- find exact duplicates and duplicates on the key, and remove them with
  `distinct()`;
- tell an impossible value from an unusual one, and treat each correctly;
- repair inconsistent category labels and convert text to numbers without losing
  data silently;
- keep a cleaning log that records every row dropped and what it did to *n*;
- chain `filter()`, `select()`, `mutate()`, `arrange()`, `group_by()` and
  `summarise()` with the pipe to answer a playtest question.

---

## Why This Matters for Playtesting

Telemetry drops events when a build crashes. Spreadsheets get pasted into twice.
Someone types "veteran" where the form said "Veteran". Every one of these has to be
dealt with before any summary means anything, and every way of dealing with it is a
choice.

**Cleaning choices change conclusions.** Dropping the rows with a blank cell is one
line of R. If the blanks are not scattered at random, that line removes a
particular kind of player, and the summary you compute afterwards describes the
players who were left. The worked example below does exactly this to the strongest
finding in the *Tidebreak* data.

**That is why cleaning goes in the report.** A reader who cannot see what you
removed cannot judge what you found. Cleaning is analysis
([rule 6](ten-rules.qmd)): record every choice where the reader can see it, not
hidden in a script.

---

# Part 1: Cleaning

You already know `count()`, `mutate()`, `group_by()` and `summarise()` from
Stage 1. This part also uses `filter()`, which keeps the rows where a condition is
true: `raw |> filter(build == "A")` is the Build A rows. Part 2 covers it properly.

The raw *Tidebreak* export is `tidebreak_sessions_raw.csv`. Load it so you see the
file as delivered:

```{r}
#| label: load-raw
library(dplyr)
raw <- read_csv("data/tidebreak_sessions_raw.csv",
                trim_ws = FALSE, show_col_types = FALSE)
nrow(raw)
```

`read_csv()` normally strips spaces from the edges of text values without telling
you. `trim_ws = FALSE` switches that off, so you see what a spreadsheet or database
import would see.

Here is the whole cleaning process for this file, with the number of rows after
each step. Every step is explained below.

```{r}
#| label: cleaning-pipeline-gen
#| echo: false
#| output: asis
# Writes a Mermaid block so the row counts are computed, not typed
cat("```{mermaid}\n",
    "%%| label: fig-cleaning-pipeline\n",
    "%%| fig-cap: \"The cleaning pipeline for the raw Tidebreak export, with the number of rows after each step.\"\n",
    "flowchart LR\n",
    sprintf('  R["Raw export<br/>%d rows"] --> D["Remove exact<br/>duplicates<br/>%d rows"]\n',
            nrow(raw), nrow(deduped)),
    sprintf('  D --> L["Fix cohort<br/>labels<br/>%d rows"]\n', nrow(deduped)),
    sprintf('  L --> I["Check impossible<br/>values<br/>%d rows"]\n', nrow(deduped)),
    sprintf('  I --> M["Keep blank<br/>session lengths<br/>%d rows"]\n', nrow(deduped)),
    sprintf('  M --> C(["Clean data<br/>%d rows"])\n', nrow(deduped)),
    "  class C action\n",
    "```\n", sep = "")
```

Only one step removes rows. The others change values or decide to leave something
alone, and they belong in the record just as much.

---

## What `NA` Actually Means

`NA` is R's marker for a **missing value**: a cell with nothing in it. It does not
mean zero, and it does not mean the event did not happen. It means the export does
not know. A blank `deaths` cell is not a player who never died.

**`NA` is contagious.** Any calculation that touches an `NA` returns `NA`, because
R will not guess what the missing value was. Adding `na.rm = TRUE` tells R to leave
the blanks out, which is itself a decision: it assumes the missing players look like
the ones you can see.

```{r}
#| label: demo-na
colSums(is.na(raw))

mean(raw$deaths)
mean(raw$deaths, na.rm = TRUE)
```

`is.na()` returns `TRUE` for each blank cell, and `colSums()` adds them up column
by column: a count of blanks in every column at once. The first `mean()` shows the
contagion; the second is a decision.

Three questions, in this order, before you do anything to a missing value:

1. **How much is missing?** A handful of blanks in `r nrow(raw)` rows is different
   from half a column.
2. **Where is it missing?** Split the count by build, cohort and outcome. If the
   blanks cluster, something produced them.
3. **Is the pattern the finding?** A build whose quit path never writes the session
   length has a telemetry bug, and the players on that path are the ones the
   playtest was run to find ([rule 9](ten-rules.qmd)).

**Common mistake:** dropping every row with a blank, "to be safe". You will see
it done with `na.omit()`, `drop_na()` or `complete.cases()`. Count what it would
remove before you ever do it:

```{r}
#| label: demo-complete
sum(complete.cases(raw))   # rows with no blank anywhere
```

The worked example below shows what that one line would have cost.

---

## Duplicates

A **duplicate** is a row that appears more than once. It inflates *n*, and it
counts one player's experience twice.

An **exact duplicate** repeats every column. It is almost always an export or
copy-paste fault, and removing it is safe. A **duplicate on the key** repeats the
identifier, here `session_id`, but may differ elsewhere. That is two records
claiming to be the same session, and you have to work out which one is right before
you remove either.

```{r}
#| label: demo-duplicated
# Exact duplicates: how many rows disappear when repeats are removed?
nrow(raw) - nrow(distinct(raw))

# Duplicates on the key: which session IDs appear more than once?
raw |> count(session_id) |> filter(n > 1)

clean <- raw |> distinct()
nrow(clean)
```

`distinct()` keeps the first copy of every row and drops the repeats. **Check both
counts.** Here `r sum(duplicated(raw))` rows are exact copies and
`r sum(duplicated(raw$session_id))` session IDs are repeated: the same number, so
every key duplicate is an exact copy and removing them is safe. If they did not
match, you would look at the differing rows first.

---

## Impossible Versus Unusual

An **impossible value** cannot happen in the game: a negative session length, a
rating of 9 on a 1–7 scale, more objectives captured than the map has. It is an
error. Fix it if you can trace the cause, otherwise set it to `NA`, and log it
either way.

An **unusual value** is extreme but possible: a player who never died, a Sentinel
who scored almost nothing. It is data. Removing it because it spoils a chart is
deleting the players you most need to hear from.

**The test is the game's rules, not the chart.** A boxplot flags points by distance
from the rest. Only you know whether the game allows that value. Write the rule
down, such as `session_minutes > 0` or `enjoyment` in 1–7, then check the data
against it:

```{r}
#| label: demo-impossible
clean |> filter(session_minutes <= 0) |> nrow()     # sessions of zero or less
clean |> filter(!enjoyment %in% 1:7) |> nrow()      # ratings off the 1-7 scale

clean |> arrange(score) |> select(session_id, archetype, score) |> head(5)
```

`%in%` asks "is this value one of these?", and `!` turns it round.

No impossible lengths or ratings. The five lowest scores are all Sentinels: unusual,
not impossible, and they are the bimodal Sentinel finding from
[Describing Variables](04-describing-variables.qmd#worked-example). They stay. The
[outlier decision tree](05-distributions-and-shape.qmd#outliers) is the same
reasoning in diagram form.

---

## Labels and Types

**Inconsistent labels.** R compares text exactly. `"Veteran"`, `"veteran"` and
`"Veteran "` with a trailing space are three different cohorts to `count()`, and a
grouped summary will report three groups where the design had one. `trimws()`
strips leading and trailing spaces and `tolower()` makes the case consistent. Use
them together inside `mutate()`, and always count before and after:

```{r}
#| label: demo-labels
clean |> count(cohort)

clean <- clean |> mutate(cohort = tolower(trimws(cohort)))
clean |> count(cohort)
```

`r length(unique(deduped$cohort))` spellings become three cohorts.

**Factors.** A **factor** is a categorical variable that knows its possible values
and their order. Setting the levels puts the cohorts in order of experience rather
than the alphabet, and it makes a misspelled label visible: anything not in
`levels` becomes `NA` rather than a fourth group ([rule 3](ten-rules.qmd)).

```{r}
#| label: demo-factor
clean <- clean |>
  mutate(cohort = factor(cohort, levels = c("novice", "regular", "veteran")))
clean |> count(cohort)
```

**Numbers stored as text.** When one cell in a numeric column contains `"n/a"` or a
comma decimal, `read_csv()` reads the whole column as text. `as.numeric()` converts
it back, and turns every value it cannot read into `NA` with only a warning:

```{r}
#| label: demo-as-numeric
#| warning: true
minutes_text <- c("12.9", "24.6", "n/a", "8,6")
as.numeric(minutes_text)
```

A constructed example, since this export's numbers all arrived as numbers. Two of
four values became `NA`, and the only sign was a warning. Count the `NA`s before and
after converting: any increase is data you just lost.

---

## Keeping a Cleaning Log

A **cleaning log** is a list of every change made to the data, in order, with the
number of rows before and after. It is short, and it goes in the report.

| Step | Rows after | Why |
| :- | -: | :- |
| Raw export | `r nrow(raw)` | As delivered |
| Remove exact duplicates | `r nrow(deduped)` | `r sum(duplicated(raw))` sessions exported twice |
| Standardise `cohort` labels | `r nrow(deduped)` | `r mislabelled` values re-spaced or re-cased; no rows removed |
| Check impossible values | `r nrow(deduped)` | None found against the game's rules |
| Keep blank `session_minutes` | `r nrow(deduped)` | All `r nrow(blank)` are Build A frustration quits; see the worked example |

**A step that removes no rows still goes in the log.** Relabelling changes which
group a player is counted in, and the reader needs to know that happened.

**For your project:** start the cleaning log the day the first export arrives,
in the same Quarto document as the analysis. It goes in the Methodology section
of your report, and a reader should be able to reproduce your *n* from it.

---

## Worked Example: the Blanks Are the Finding

**Which build keeps players longer?** After removing the
`r sum(duplicated(raw))` duplicates, `r nrow(blank)` sessions have a blank
`session_minutes`. Every one of them is Build A. Every one of them ended in a
frustration quit.

```{r}
#| label: worked-where-missing
clean |> count(build, quit_reason, missing = is.na(session_minutes))
```

`count()` can count a condition as well as a column: `missing = is.na(...)` makes
a TRUE/FALSE column and counts each combination.

That pattern is not random, so both obvious ways of handling it are wrong.

**Choice 1: drop incomplete rows.** `complete.cases()` keeps only rows with no
blank anywhere. It keeps `r nrow(kept)` of `r nrow(deduped)` sessions. Before,
Build A had `r frus(deduped, "A")` frustration quits against Build B's
`r frus(deduped, "B")`. After, it has `r frus(kept, "A")` against
`r frus(kept, "B")`. One line of cleaning removed
`r round(100 * (frus(deduped, "A") - frus(kept, "A")) / frus(deduped, "A"))`% of
the evidence for Build A's biggest problem, and nothing warned you.

**Choice 2: leave them out of the median.** With `na.rm = TRUE`, Build A's median
session is `r round(med_a(deduped), 1)` minutes. But the Build A frustration quits
that *were* recorded have a median of
`r round(median(deduped$session_minutes[deduped$build == "A" & deduped$quit_reason == "Frustrated"], na.rm = TRUE), 1)`
minutes. The missing sessions are almost certainly short ones, so leaving them out
makes Build A look better than it is.

**What the complete data says.** For this exercise the unbroken export exists, so
the damage can be measured. Build A's true median is
`r round(med_a(original), 1)` minutes, not `r round(med_a(deduped), 1)`. The blank
cells were never noise. They were the shortest, angriest sessions in the playtest,
lost by a telemetry bug on exactly the path the playtest was meant to examine.

**The defensible handling** keeps the rows, counts them from `quit_reason`, which
survived, reports the median with a note that `r nrow(blank)` short sessions are
missing from it, and files a bug against Build A's session logging.

**What should the team do?** Fix the logger first, because the next playtest
will have the same hole. Then report Build A's frustration quits from the column
that is complete, and say plainly which number is an overestimate and why.

---

# Part 2: Transforming

A clean table is rarely the table your question needs. You need some of the rows,
some of the columns, a new column computed from others, a different order, or one
row per group. The `dplyr` package, part of the tidyverse, does each of these with
one **verb**, and the pipe joins the verbs together.

## The Pipe

You met the pipe in Stage 1. As a reminder, `|>` passes the result on its left in
as the **first argument** of the function on its right:

| With the pipe | Means |
| :- | :- |
| `x \|> f()` | `f(x)` |
| `x \|> f(y)` | `f(x, y)` |
| `x \|> f() \|> g()` | `g(f(x))` |

Its value is in chaining. "The median session length of Build A's frustration
quits, to one decimal place":

```{r}
#| label: pipe-chain
clean |>
  filter(build == "A", quit_reason == "Frustrated") |>
  pull(session_minutes) |>
  median(na.rm = TRUE) |>
  round(1)
```

```{mermaid}
%%| label: fig-pipe
%%| fig-cap: "A pipeline reads in the order things happen: each step takes the result of the one before it."
flowchart LR
  C[("clean")] --> F["filter()<br/>Build A quits"]
  F --> P["pull()<br/>session_minutes"]
  P --> M["median()"]
  M --> R["round(1)"]
  R --> A(["answer"])
  class A action
```

It reads as a sentence: *take the clean data, keep Build A's frustration quits,
pull out the session lengths, take the median, round it.* Each line is one step,
and deleting a line deletes one step. `pull()` extracts one column as a vector,
the pipe-friendly version of `$`.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Going further: the same question without the pipe"}
You will see code written the other two ways. Both give the same answer.

```{r}
#| label: pipe-other-ways
# Nested: read from the inside out
round(median(clean$session_minutes[clean$build == "A" &
                                     clean$quit_reason == "Frustrated"],
             na.rm = TRUE), 1)

# Intermediate objects: clear, but leaves a trail of names
a_frus <- clean[clean$build == "A" & clean$quit_reason == "Frustrated", ]
round(median(a_frus$session_minutes, na.rm = TRUE), 1)
```

You will also see `%>%`, the older tidyverse pipe. It does the same job here. In
RStudio, `Ctrl+Shift+M` types a pipe; *Tools → Global Options → Code* chooses which
one.
:::

---

## The Six Verbs

Each verb takes a data frame first and returns a data frame, which is what makes
them chain. To see what each one does, here is a small table of six sessions that
every example below starts from:

```{r}
#| label: verbs-few
few <- clean |>
  select(session_id, build, cohort, deaths, kills, session_minutes) |>
  head(6)
few
```

| Verb | What it does | The question it answers |
| :- | :- | :- |
| `select()` | Keep some columns | "I only need these measures" |
| `filter()` | Keep some rows | "Only Build A", "only the frustration quits" |
| `mutate()` | Add or change a column | "Kills per death", "fix the labels" |
| `arrange()` | Sort the rows | "Shortest sessions first" |
| `group_by()` | Mark the groups | "For each build..." |
| `summarise()` | One row of results per group | "...what is the median?" |

Inside every verb you refer to columns by their bare names, `deaths` rather than
`few$deaths`: the verb already knows which table it is working on.

### `select()`: keep columns

```{r}
#| label: verb-select
few |> select(session_id, build, deaths)
```

Six rows, three columns. `select(-cohort)` would keep everything *except*
`cohort`.

### `filter()`: keep rows

```{r}
#| label: verb-filter
few |> filter(build == "A", deaths >= 3)
```

Only the rows where every condition is `TRUE`. A comma between conditions means
"and"; use `|` for "or".

### `mutate()`: add a column

```{r}
#| label: verb-mutate
few |> mutate(kd = round(kills / pmax(deaths, 1), 2))
```

A new column, computed row by row from the others. `pmax(deaths, 1)` replaces a
zero with 1 so that a player who never died does not divide by zero.

### `arrange()`: sort

```{r}
#| label: verb-arrange
few |> arrange(desc(session_minutes))
```

Longest session first. Without `desc()` it sorts smallest first. `arrange()` puts
`NA` last whichever way you sort.

### `group_by()` and `summarise()`: one row per group

```{r}
#| label: verb-summarise
few |>
  group_by(build) |>
  summarise(sessions = n(), mean_deaths = mean(deaths))
```

Six rows in, one row per build out. `group_by()` changes nothing you can see; it
marks the groups so that `summarise()` computes once for each. `n()` counts the
rows in each group.

### Chaining them

The verbs earn their keep together:

```{r}
#| label: demo-verbs
clean |>
  filter(build == "A", quit_reason == "Frustrated") |>
  mutate(logged = !is.na(session_minutes)) |>
  select(session_id, cohort, session_minutes, logged) |>
  arrange(session_minutes) |>
  head(4)
```

Read it as a sentence: take the clean sessions, keep Build A's frustration quits,
add a column saying whether the length was logged, keep four columns, sort by
length, show the first four. The shortest *recorded* quits come first, because the
blank ones are sorted last.

---

## Split–Apply–Combine

`group_by()` with `summarise()` does three things, which together are called
**split–apply–combine**:

```{mermaid}
%%| label: fig-split-apply-combine
%%| fig-cap: "Split the column by group, apply a function to each piece, combine the results into one table."
flowchart LR
  D[("wrong_turns")] -->|split by build| A["Build A<br/>sessions"]
  D -->|split by build| B["Build B<br/>sessions"]
  A -->|"apply mean()"| MA["mean for A"]
  B -->|"apply mean()"| MB["mean for B"]
  MA -->|combine| T(["one row<br/>per build"])
  MB -->|combine| T
  class T action
```

*What is the average number of wrong turns in each build?*

```{r}
#| label: demo-group-one
clean |>
  group_by(build) |>
  summarise(mean_wrong = mean(wrong_turns))
```

`group_by()` does the split, `summarise()` does the apply, and the combine is a
table with one row per group.

The point of it shows when the question grows. *How many sessions does each build
have, how many are missing a length, and what is the median length of the rest?*

```{r}
#| label: demo-group-many
clean |>
  group_by(build) |>
  summarise(
    n       = n(),
    missing = sum(is.na(session_minutes)),
    median  = median(session_minutes, na.rm = TRUE)
  )
```

One `group_by()`, as many summaries as you like. The answer is the worked example
in one table: all `r nrow(blank)` missing lengths are Build A's, and the median
beside them is computed without them.

**You will also see** `tapply()` and `aggregate()` doing split–apply–combine in
older code. You do not need to write them.

---

## Interpretation Guidance

**What the data shows.** The raw export has `r sum(duplicated(raw))` duplicated
sessions, `r mislabelled` mislabelled cohorts and `r nrow(blank)` blank session
lengths, all in Build A frustration quits.

**What the analysis suggests.** Build A's quit path does not write the session
length. The blanks are a telemetry fault concentrated in the sessions that matter
most.

**What can legitimately be concluded.** Build A's frustration-quit count is
reliable, because it comes from `quit_reason`, which is complete. Its median
session length is an overestimate by an amount you cannot know from this export.
Say both.

**Report what you did, not only what you found.** The cleaning log goes in the
report's Methodology section. A reader should be able to reproduce your *n* from
it.

---

## Common Mistakes

**Dropping incomplete rows by reflex.** `na.omit()` and `complete.cases()` are one
line each and can remove the finding. Check where the blanks are first.

**Treating `NA` as zero.** A blank deaths cell is unknown, not a player who never
died. Replacing it with 0 invents a very good player.

**Using `na.rm = TRUE` without saying so.** It is the right tool, and it changes
what the number describes. Report how many values it left out.

**Fixing labels without checking.** Run `count()` before and after. A label you did
not know about survives `tolower()`.

**Deleting unusual values as if they were impossible.** The game's rules decide
what is impossible. The chart does not.

**Forgetting to keep the result.** `clean |> filter(build == "A")` prints the
filtered rows and changes nothing. Assign it, `a_only <- clean |> filter(...)`,
to keep it.

**Using `=` inside `filter()`.** `filter(build = "A")` is an error; the comparison
is `==`.

**Cleaning in a spreadsheet.** Edits made by hand cannot be seen, repeated or
reviewed. Every change belongs in code, where it runs again on the next export.

---

## Practice

Each takes two or three minutes. Start from `clean`, the de-duplicated data with
fixed labels from Part 1.

1. **Predict.** Before running it: how many rows will
   `clean |> filter(build == "B", quit_reason == "Frustrated")` return? Use what
   you know from [Planning a Playtest](01-planning-a-playtest.qmd#worked-example).
2. **Compute.** Which rows have a blank `session_minutes` or a blank `deaths`?
   Count them by build and quit reason.
3. **Rewrite.** You find `tapply(clean$score, clean$archetype, median)` in a
   teammate's script. Rewrite it as a pipeline with `group_by()` and
   `summarise()`, and add the number of sessions per archetype.
4. **Critique.** A teammate's script starts `clean <- na.omit(raw)` and then
   reports Build A's frustration rate. What is wrong, and what should the first
   lines of the script do instead?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
**1.** Piping into `nrow()` counts the rows the filter kept:

```{r}
#| label: practice-predict
clean |> filter(build == "B", quit_reason == "Frustrated") |> nrow()
```

`r frus(deduped, "B")`, the Build B frustration count from the worked example in
Planning a Playtest.

**2.** `|` means "or":

```{r}
#| label: practice-incomplete
clean |>
  filter(is.na(session_minutes) | is.na(deaths)) |>
  count(build, quit_reason)
```

The `r nrow(blank)` Build A frustration quits are the blank lengths; the blank
`deaths` cells are scattered across other outcomes. Those two groups need
different handling, and dropping every incomplete row treats them the same.

**3.**

```{r}
#| label: practice-archetype
clean |>
  group_by(archetype) |>
  summarise(n = n(), median_score = median(score))
```

Adding `n = n()` costs one argument. Always report *n* next to a group summary
([rule 8](ten-rules.qmd)).

**4.** `na.omit()` drops every row with any blank, which removes all
`r nrow(blank)` of Build A's unlogged frustration quits, so the reported
frustration rate is too low. The script should remove duplicates, fix labels, then
**count** the missing values by build and outcome before deciding anything, and log
each step.
:::

---

## Summary

- `NA` means unknown. Count it, find where it is, and ask whether the pattern is
  the finding before doing anything to it.
- Check duplicates both exactly and on the key.
- Impossible values are errors; unusual values are data. The game's rules decide
  which is which.
- `trimws()` and `tolower()` fix labels; `factor()` fixes their order; `count()`
  checks the result.
- Keep a cleaning log with *n* at every step. It goes in the report.
- The pipe passes its left side in as the first argument on its right, so a
  pipeline reads in the order things happen.
- The six verbs select columns, filter rows, mutate columns, arrange rows, and
  group and summarise. `group_by()` + `summarise()` is split–apply–combine.
- In *Tidebreak*, the `r nrow(blank)` blank cells were the frustration quits.
  Dropping them would have hidden Build A's worst problem.

| Key term | Meaning |
| :- | :- |
| **Missing value (`NA`)** | A value that should exist but was not recorded |
| **Exact duplicate** | A row repeated in every column |
| **Key** | The column that identifies a row, such as `session_id` |
| **Impossible value** | A value the game's rules do not allow: an error |
| **Unusual value** | An extreme but possible value: data |
| **Factor** | A categorical variable with fixed, ordered levels |
| **Cleaning log** | Every change to the data, in order, with *n* after each |
| **Pipe (`\|>`)** | Passes the result on its left as the first argument on its right |
| **Verb** | A `dplyr` function that takes a data frame and returns one |
| **Split–apply–combine** | Split by group, apply a function to each piece, combine the results |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 3 (know what kind of number you are holding),
  rule 6 (cleaning is analysis), rule 8 (small samples lie confidently) and rule 9
  (sometimes the rare value is the whole story)
- [R and Quarto Reference: cleaning](../reference/r-quarto-reference.qmd#cleaning)
  and [transforming with dplyr](../reference/r-quarto-reference.qmd#transforming-with-dplyr)
- [Lab 02: Comparing Builds](../labs/comparing-builds.qmd#part-a-clean-the-export),
  Part A
- [Stage 2 demo](../demos/stage-2.qmd)
- Next: [From Sample to Population](07-sample-to-population.qmd)

---

## Reflective Questions

- A colleague removes every row with a blank cell "to be safe". What would you need
  to know about those rows before you agreed with them?
- Build A's median session length is wrong in a direction you know but by an amount
  you do not. Should the report quote it at all? What would you write instead?
- Two rows share a `session_id` but have different scores. What evidence would you
  use to decide which one to keep, and what do you do if there is none?
- Your own game's telemetry will have blanks. Which of your measures would you
  expect a crash or a rage-quit to leave blank, and what would that do to your
  results?
