---
title: "From Sample to Population"
subtitle: "Why the next playtest would give different numbers"
body-classes: "wf-at-uncertainty"
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")

# Four playtests of 30, in file order
batch       <- ceiling(seq_len(nrow(sessions)) / 30)
batch_means <- round(tapply(sessions$score, batch, mean))

# An imaginary population: the 120 session lengths copied 1,000 times
pop      <- rep(sessions$session_minutes, times = 1000)
pop_mean <- mean(pop)
pop_sd   <- sd(pop)

set.seed(7)
draw_means <- function(n, reps = 1000) replicate(reps, mean(sample(pop, n)))
means_10  <- draw_means(10)
means_30  <- draw_means(30)
means_100 <- draw_means(100)

se_30 <- pop_sd / sqrt(30)

# Twenty playtests of 30, each with a 95% confidence interval
set.seed(1)
cis <- t(replicate(20, {
  x <- sample(pop, 30)
  t.test(x)$conf.int
}))
missed <- sum(cis[, 1] > pop_mean | cis[, 2] < pop_mean)
stopifnot(missed >= 1, missed <= 3)

# Build B session length: one real interval
mins_b <- sessions$session_minutes[sessions$build == "B"]
n_b    <- length(mins_b)
se_b   <- sd(mins_b) / sqrt(n_b)
ci_b   <- t.test(mins_b)$conf.int
```

> Every number in a playtest report describes the players who happened to turn
> up. Bring in a different group and every number moves. This page is about how
> far: what **sampling variability** is, how the **standard error** measures it,
> and how a **confidence interval** turns one playtest's number into an honest
> range for the population you care about. The playtest question running through
> it: *"Build B sessions averaged 24 minutes. How sure can we be?"*

::: {.at-a-glance}
**Time:** 15 minutes in class (the simulation figures); the rest is reading.
**Before this:** [Distributions and Shape](05-distributions-and-shape.qmd),
especially the normal distribution.
**Key terms:** sampling variability, sampling distribution, standard error,
central limit theorem, confidence interval, margin of error.
**R introduced:** `sample()`, `replicate()`, `sqrt()`, `qt()`, `t.test()` for one
group.
:::

## Learning Outcomes

By the end of these notes you can:

- explain why a second playtest with different players would give different
  numbers, even if nothing about the game changed;
- describe a sampling distribution, and say what the standard error measures;
- explain how sample size affects the standard error, and why quadrupling the
  players only halves it;
- state the central limit theorem in plain English, and say why it matters for
  data that is not normal;
- compute and interpret a 95% confidence interval for a mean, and say what the
  "95%" does and does not mean.

---

## Why This Matters for Playtesting

[Planning a Playtest](01-planning-a-playtest.qmd#how-many-players) showed that a
five-session playtest can report a frustration rate anywhere from 0% to 60% when
the true rate is a third. That spread is not a flaw in the method. It is what
happens whenever you measure some players and draw conclusions about all of them.

**Every comparison in Part 4 rests on this page.** Build B's sessions were longer
than Build A's. Is that because Build B is better, or because Build B happened to
get the more patient players? You cannot answer that until you know how much a
number would move by chance alone. That amount has a name, the standard error, and
this page shows where it comes from.

---

## Sampling Variability

Split the `r nrow(sessions)` *Tidebreak* sessions into four batches of 30, in the
order they were recorded, and treat each as a separate playtest. Same game, same
build mix, same procedure. The mean score of each batch:

```{r}
#| label: batches
batch <- ceiling(seq_len(nrow(sessions)) / 30)
round(tapply(sessions$score, batch, mean))
```

Four playtests, four different answers, from `r min(batch_means)` to
`r max(batch_means)`. Nothing changed except which players were in each batch.

**Sampling variability** is this: a different sample gives a different number,
even from the same population. It is not a mistake and it cannot be avoided. The
only question is how big it is, and that depends on two things: **how much the
players differ** (the SD) and **how many of them you measured** (n).

---

## The Sampling Distribution

To see sampling variability properly, you need many playtests, not four. Real
studios cannot afford that, but a simulation can.

Imagine the population of everyone who will ever play *Tidebreak*, and suppose
their session lengths look exactly like the `r nrow(sessions)` you recorded. Build
that population by copying the sessions a thousand times over, then run a thousand
imaginary playtests of 30 players each, and record each playtest's mean:

```{r}
#| label: simulate
pop <- rep(sessions$session_minutes, times = 1000)

set.seed(7)
means_30 <- replicate(1000, mean(sample(pop, 30)))
```

`sample(pop, 30)` draws 30 players at random; `mean()` is that playtest's result;
`replicate(1000, ...)` runs it a thousand times. The thousand means form the
**sampling distribution**: the distribution of a summary across every playtest you
might have run.

```{r}
#| label: fig-sampling-dist
#| fig-cap: "Left: session length across the imaginary population, lopsided and two-humped. Right: the means of 1,000 playtests of 30 players drawn from it, which form a bell."
#| fig-alt: "Two histograms. Left: the population of session lengths, piled up near 30 minutes with a second smaller hump around 12 minutes. Right: the distribution of 1,000 sample means, a narrow symmetric bell centred near 23 minutes, with a red normal curve laid over it that fits closely."
#| fig-height: 3.6
op <- par(mfrow = c(1, 2), mar = c(4, 3, 2, 1))
hist(pop, breaks = 26, col = pal$fill, border = "white", main = "Population",
     xlab = "Session length (minutes)", ylab = "", yaxt = "n", cex.main = 0.9)
abline(v = pop_mean, col = pal$highlight, lwd = 2)
hist(means_30, breaks = 30, freq = FALSE, col = pal$fill, border = "white",
     main = "1,000 playtest means, n = 30", xlim = range(pop),
     xlab = "Mean session length (minutes)", ylab = "", yaxt = "n", cex.main = 0.9)
curve(dnorm(x, mean(means_30), sd(means_30)), add = TRUE, lwd = 2,
      col = pal$highlight)
par(op)
```

Two things stand out.

**The means are centred on the truth.** The population mean is
`r round(pop_mean, 1)` minutes; the average of the thousand playtest means is
`r round(mean(means_30), 1)`. Any one playtest misses, but not systematically in
one direction.

**The means are far less spread out than the players.** Individual sessions run
from `r min(pop)` to `r max(pop)` minutes. Playtest means of 30 players fall almost
all between `r round(quantile(means_30, 0.025), 1)` and
`r round(quantile(means_30, 0.975), 1)`. Averaging cancels out most of the
differences between players.

---

## The Standard Error

The SD of the sampling distribution has its own name: the **standard error** (SE).
It is the typical distance between one playtest's mean and the population's mean:
**how far your number is likely to be off**.

You never get to run a thousand playtests, but you do not need to. The standard
error can be estimated from a single sample:

$$\text{SE} = \frac{\text{SD}}{\sqrt{n}}$$

Check it against the simulation. The population SD is `r round(pop_sd, 2)`, so for
n = 30 the formula gives `r round(pop_sd, 2)` / √30 = `r round(se_30, 2)`. The SD
of the thousand simulated means is `r round(sd(means_30), 2)`. The formula works.

**Bigger samples give smaller standard errors, but slowly.** Because n is under a
square root, you need **four times** the players to **halve** the standard error:

```{r}
#| label: fig-sample-size-se
#| fig-cap: "Means of 1,000 simulated playtests at three sizes. Each quadrupling of n roughly halves the spread."
#| fig-alt: "Three horizontal strips of points. With n = 10 the playtest means spread widely around the dashed true mean; with n = 30 they are tighter; with n = 100 they cluster closely around it."
#| fig-height: 3.4
par(mar = c(4, 6, 1, 1))
stripchart(list("n = 10" = means_10, "n = 30" = means_30, "n = 100" = means_100),
           method = "jitter", jitter = 0.25, pch = 16, cex = 0.4, las = 1,
           col = adjustcolor(pal$line, alpha.f = 0.3),
           xlab = "Playtest mean session length (minutes)")
abline(v = pop_mean, lty = 2, lwd = 2, col = pal$highlight)
```

| Players per playtest | SE from the formula | SD of 1,000 simulated means |
| :-: | :-: | :-: |
| 10 | `r round(pop_sd / sqrt(10), 2)` | `r round(sd(means_10), 2)` |
| 30 | `r round(pop_sd / sqrt(30), 2)` | `r round(sd(means_30), 2)` |
| 100 | `r round(pop_sd / sqrt(100), 2)` | `r round(sd(means_100), 2)` |

**SD and SE answer different questions.** The SD describes the players: how much
one session differs from another. The SE describes your *estimate*: how much your
mean would move if you ran the playtest again. More players does not make the
players less varied, so the SD stays put; it makes your estimate steadier, so the
SE shrinks.

---

## The Central Limit Theorem

Look again at the left panel of @fig-sampling-dist. Session length is lopsided and
two-humped, nothing like a bell. Yet the means on the right form a bell almost
perfectly.

That is the **central limit theorem**: **whatever shape the individual values
have, the means of reasonably large samples are approximately normal.** The
individual players can be skewed, bimodal or capped; their averages still pile up
in a bell around the true mean.

It is why the normal distribution from
[Distributions and Shape](05-distributions-and-shape.qmd#the-normal-distribution)
matters even for data that is not normal. The 68–95–99.7 rule does not describe
the session lengths, but it does describe the sampling distribution of their mean.
So about 95% of playtest means fall within 2 standard errors of the true mean. That
single fact is what confidence intervals and t-tests are built on.

**How large is "reasonably large"?** For mildly skewed data, 30 is usually plenty;
the simulation above uses 30 and the bell is clear. For heavily skewed data, such
as frame times with rare hitches, it takes more.

---

## Confidence Intervals

If 95% of playtest means fall within about 2 SEs of the true mean, then turning it
round, the true mean is within about 2 SEs of your playtest's mean 95% of the time.
That range is a **95% confidence interval**:

$$\text{mean} \pm \text{about } 2 \times \text{SE}$$

The "about 2" is 1.96 for large samples; for smaller ones R uses a slightly larger
number from the **t distribution**, which allows for the SD itself being estimated
from the sample. The part after the ± is the **margin of error**.

**For real:** Build B's `r n_b` sessions.

```{r}
#| label: ci-by-hand
mins_b <- sessions$session_minutes[sessions$build == "B"]

m  <- mean(mins_b)
se <- sd(mins_b) / sqrt(length(mins_b))
multiplier <- qt(0.975, df = length(mins_b) - 1)

c(mean = m, se = se, multiplier = multiplier)
m + c(-1, 1) * multiplier * se
```

And the same in one line, which is how you will usually do it:

```{r}
#| label: ci-t-test
t.test(mins_b)$conf.int
```

**In a report you would write:** "Build B sessions averaged
`r round(mean(mins_b), 1)` minutes (95% CI `r ci_fmt(ci_b)`, n = `r n_b`)." The
interval says the data is consistent with a population average anywhere from
`r round(ci_b[1], 1)` to `r round(ci_b[2], 1)` minutes. That honesty is the point:
a reader now knows how much weight the `r round(mean(mins_b), 1)` can bear.

::: {.callout-note collapse="true" title="Going further: these sessions are not quite independent"}
The interval treats Build B's `r n_b` sessions as `r n_b` independent
observations. They are not: they come from `r length(unique(sessions$player_id[sessions$build == "B"]))`
players who each played `r unique(as.vector(table(sessions$player_id)))` times,
and one player's sessions resemble each other. The honest effective sample size
is somewhere between the number of players and the number of sessions, so the true
interval is somewhat wider than this one. Handling repeated measurements properly
needs methods beyond this module; the practical lesson is to report both counts
([rule 8](ten-rules.qmd)).
:::

### What "95%" means

This is the most misread number in statistics. **The 95% describes the method,
not this particular interval.** If you ran the playtest many times and computed an
interval each time, about 95% of those intervals would contain the true population
mean. Any one interval either contains it or does not; you never know which.

```{r}
#| label: fig-twenty-cis
#| fig-cap: "Twenty simulated playtests of 30 players, each with its 95% confidence interval. The dashed line is the true population mean. Red intervals miss it."
#| fig-alt: "Twenty horizontal line segments stacked vertically, each with a dot at its centre. A vertical dashed line marks the true mean. Most segments cross the dashed line; the ones that do not are drawn in red."
#| fig-height: 4.6
miss <- cis[, 1] > pop_mean | cis[, 2] < pop_mean
par(mar = c(4, 4, 1, 1))
plot(NA, xlim = range(cis), ylim = c(1, 20), yaxt = "n",
     xlab = "Mean session length (minutes)", ylab = "Playtest")
axis(2, at = c(1, 5, 10, 15, 20), las = 1)
segments(cis[, 1], 1:20, cis[, 2], 1:20, lwd = 2,
         col = ifelse(miss, pal$highlight, pal$line))
points(rowMeans(cis), 1:20, pch = 19, cex = 0.7,
       col = ifelse(miss, pal$highlight, pal$line))
abline(v = pop_mean, lty = 2, lwd = 2, col = pal$guide)
```

In this run, `r missed` of the 20 intervals missed the true mean, close to the one
in twenty the method promises. A playtest that misses has done nothing wrong; it
drew an unusual set of players, which is exactly what the 5% allows for.

**So do not say** "there is a 95% chance the true mean is between
`r round(ci_b[1], 1)` and `r round(ci_b[2], 1)`". Say "the method that produced
this interval captures the true mean 95% of the time", or, in a report, simply
give the interval and let it stand for "the plausible range".

**Wider means less certain.** An interval narrows with more players (a smaller SE)
and widens with more varied players (a larger SD). A 99% interval is wider than a
95% one: more confidence costs precision.

::: {.callout-note collapse="true" title="Going further: the bootstrap"}
The formula relies on the central limit theorem. The **bootstrap** gets an interval
without it, by treating your sample as a stand-in for the population and resampling
it. Draw `r n_b` sessions *with replacement* from Build B's `r n_b`, take the mean,
and repeat a thousand times; the middle 95% of those means is the interval.

```{r}
#| label: bootstrap
set.seed(3)
boot_means <- replicate(1000, mean(sample(mins_b, replace = TRUE)))
quantile(boot_means, c(0.025, 0.975))
```

It agrees closely with `t.test()`'s `r ci_fmt(ci_b)`. The bootstrap earns its keep
for summaries with no simple formula, such as a median or a 99th percentile:
replace `mean` with `median` in the line above and it still works.
:::

---

## Worked Example

**Build B sessions averaged 24 minutes. How sure can we be?** The draft report
says: "Build B sessions last 24 minutes on average."

**Step 1: what is the sample?** `r n_b` sessions from
`r length(unique(sessions$player_id[sessions$build == "B"]))` players, one
playtest. A different group would give a different mean.

**Step 2: how much would it move?** The SD of Build B's session lengths is
`r round(sd(mins_b), 2)` minutes. The standard error is that divided by
√`r n_b`: `r round(se_b, 2)` minutes. A second playtest's mean would typically land
within about `r round(se_b, 1)` minutes of this one.

**Step 3: give the range.** The 95% confidence interval is `r ci_fmt(ci_b)`
minutes.

**Step 4: rewrite.** "Build B sessions averaged `r round(mean(mins_b), 1)` minutes
(95% CI `r ci_fmt(ci_b)`; `r n_b` sessions from
`r length(unique(sessions$player_id[sessions$build == "B"]))` players)." The claim
is now as strong as the data, and no stronger.

**Step 5: notice what is still missing.** This is one build on its own. Whether
Build B's sessions are longer than Build A's is a different question, because both
numbers wobble. Comparing two groups, and asking whether the gap could be chance, is
[Comparing Groups](08-comparing-groups.qmd).

---

## Interpretation Guidance

**What the data shows.** One playtest's mean: `r round(mean(mins_b), 1)` minutes
for Build B.

**What the analysis suggests.** The population mean is plausibly anywhere from
`r round(ci_b[1], 1)` to `r round(ci_b[2], 1)` minutes, with values near the centre
of that range more plausible than values at its ends.

**What can legitimately be concluded.** A range, not a point. A finding that
depends on the exact value of a mean, such as "sessions must average 25 minutes",
cannot be settled by a playtest whose interval spans both sides of 25.

---

## Common Mistakes

**Treating one playtest's number as a fact about the game.** It is a fact about
those players. Give the interval.

**Confusing the SD and the SE.** The SD describes the players; the SE describes
your estimate. Reporting an SE as if it were the spread of the players makes the
players look far more alike than they are.

**Reading "95%" as the chance this interval is right.** It describes how often the
method works, not this interval.

**Expecting a bigger sample to fix everything.** Four times the players halves the
SE. Doubling the playtest buys less precision than it seems.

**Forgetting that sessions are not players.** Repeated sessions from the same
player are not independent, so an interval built on sessions is somewhat too
narrow. Report both counts.

**Assuming the central limit theorem rescues tiny samples.** With five players and
skewed data, the sample mean is not reliably normal, and the interval is not
reliable either.

---

## Practice

Each takes two or three minutes.

**1. Predict.** A playtest of 25 players gives an SE of 2 minutes. Roughly what SE
would 100 players give? And 400?

::: {.callout-note collapse="true" title="Check your answer"}
About 1 minute with 100 players (four times the players, half the SE) and about
0.5 with 400 (sixteen times the players, a quarter of the SE). Each halving costs
four times as many players.
:::

**2. Compute.** Find the 95% confidence interval for the mean session length of
**Build A**, and compare its width with Build B's.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-ci-a
mins_a <- sessions$session_minutes[sessions$build == "A"]
ci_a <- t.test(mins_a)$conf.int
ci_a
diff(ci_a)
diff(t.test(mins_b)$conf.int)
```

Build A's interval is `r ci_fmt(t.test(sessions$session_minutes[sessions$build == "A"])$conf.int)`
minutes, and it is wider: the same number of sessions, but Build A's session lengths
are more spread out (SD
`r round(sd(sessions$session_minutes[sessions$build == "A"]), 1)` against
`r round(sd(mins_b), 1)`), because of its two groups of players.
:::

**3. Critique.** A report says: "Mean enjoyment was 4.1 (SE 0.08), so players
agreed closely about how much they enjoyed it." What is wrong?

::: {.callout-note collapse="true" title="Check your answer"}
The SE says how precisely the *mean* is known, not how much players agreed. With
`r nrow(sessions)` sessions, the SE is small even if players disagreed widely. To
describe agreement, report the SD or the distribution. (Enjoyment is also a 1 to 7
rating, so the distribution is the better summary anyway; see
[Data, Variables and Types](03-data-and-types.qmd#the-rating-scale-question).)
:::

**4. Interpret.** In @fig-twenty-cis, one of the red intervals came from a
playtest. Did that playtest do anything wrong?

::: {.callout-note collapse="true" title="Check your answer"}
No. It drew an unusual set of players by chance. A 95% method is designed to miss
about one time in twenty, and nothing in the data warns you when it is your turn.
That is why a single surprising result is a reason to run another playtest, not to
redesign the game.
:::

---

## Summary

- A different sample gives a different number. That is sampling variability, and
  it is unavoidable.
- The sampling distribution is the distribution of a summary across every playtest
  you might have run. Its SD is the standard error: how far your estimate is
  typically off.
- SE = SD / √n. Four times the players halves the SE.
- The SD describes the players; the SE describes your estimate.
- The central limit theorem: means of reasonably large samples are approximately
  normal, whatever the shape of the data.
- A 95% confidence interval is about mean ± 2 SE. The 95% describes the method: it
  captures the true value in about 95% of playtests.
- Report a mean with its interval and n.

| Key term | Meaning |
| :- | :- |
| **Sampling variability** | Different samples from the same population give different results |
| **Sampling distribution** | The distribution of a summary across many possible samples |
| **Standard error** | The SD of the sampling distribution: how far an estimate is typically off |
| **Central limit theorem** | Means of reasonably large samples are approximately normal |
| **Confidence interval** | A range of plausible values for the population, given the sample |
| **Margin of error** | The part after the ± in a confidence interval |
| **Bootstrap** | Estimating a sampling distribution by resampling your own sample |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 8 (small samples lie confidently)
- [Statistics Reference: from sample to population](../reference/reference-statistics.qmd#from-sample-to-population)
- [Lab 02: Comparing Builds](../labs/comparing-builds.qmd), Part B
- [Week 2 demo](../demos/week-2.qmd)
- Next: [Comparing Groups](08-comparing-groups.qmd)

---

## Reflective Questions

- Your playtest found a mean session length of 24 minutes, and your producer wants
  "the number". Do you give them 24, or the interval? What do you say to justify
  your choice?
- Doubling the playtest from 30 to 60 players narrows the interval by less than a
  third. When is that worth the cost, and when is it not?
- The central limit theorem rescues means from skewed data. Does it rescue the
  99th percentile of frame time? What would you do instead?
- Your interval for Build B is built on sessions, but players played four sessions
  each. Is the true interval wider or narrower, and why does it matter for a
  design decision?
