---
title: "Comparing Groups"
subtitle: "Is the difference real, and does it matter?"
body-classes: "wf-at-compare wf-at-interpret"
aliases:
  - comparing-conditions.html
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")
library(dplyr)

wt_a <- sessions$wrong_turns[sessions$build == "A"]
wt_b <- sessions$wrong_turns[sessions$build == "B"]

tt_wt  <- t.test(wrong_turns ~ build, data = sessions)
gap_wt <- unname(tt_wt$estimate[1] - tt_wt$estimate[2])

# Permutation tests
gap <- function(y, g) mean(y[g == "A"]) - mean(y[g == "B"])
set.seed(1)
shuffled_wt <- replicate(10000, gap(sessions$wrong_turns, sample(sessions$build)))
set.seed(2)
shuffled_df <- replicate(10000, gap(sessions$difficulty, sample(sessions$build)))
obs_df  <- gap(sessions$difficulty, sessions$build)
perm_p_df <- mean(abs(shuffled_df) >= abs(obs_df))

tt_diff <- t.test(difficulty ~ build, data = sessions)
gap_dif <- unname(tt_diff$estimate[2] - tt_diff$estimate[1])
ci_dif  <- -rev(tt_diff$conf.int)
tt_dth  <- t.test(deaths ~ build, data = sessions)
tt_enj  <- t.test(enjoyment ~ build, data = sessions)

tt_pair <- t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)
tt_unp  <- t.test(paired$wrong_turns_a, paired$wrong_turns_b)
d_pair  <- paired$wrong_turns_a - paired$wrong_turns_b

atrium  <- frames[frames$zone == "Atrium", ]
tt_atr  <- t.test(ms ~ build, data = atrium)
gap_atr <- unname(tt_atr$estimate[1] - tt_atr$estimate[2])

per_player <- aggregate(wrong_turns ~ player_id + build, data = sessions, FUN = mean)
tt_player  <- t.test(wrong_turns ~ build, data = per_player)

n_players <- length(unique(sessions$player_id))
n_per     <- unique(as.vector(table(sessions$player_id)))

wt_out <- tapply(sessions$wrong_turns, sessions$build,
                 function(x) length(boxplot.stats(x)$out))

# Proportions
frus_n  <- tapply(sessions$quit_reason == "Frustrated", sessions$build, sum)
build_n <- table(sessions$build)
pt_frus <- prop.test(frus_n, build_n)

# Effect size
cohen_d <- (mean(wt_a) - mean(wt_b)) / sqrt((var(wt_a) + var(wt_b)) / 2)

# Power at 5 sessions per build, for the Practice answer
set.seed(5)
power_5 <- mean(replicate(2000, t.test(sample(wt_a, 5, TRUE),
                                       sample(wt_b, 5, TRUE))$p.value < 0.05))

stopifnot(tt_wt$p.value < 0.001, tt_wt$conf.int[1] > 0,
          max(abs(shuffled_wt)) < abs(gap_wt),
          perm_p_df < 0.05, perm_p_df > 0.01,
          tt_diff$p.value < 0.05, gap_dif > 0, gap_dif < 0.5, ci_dif[1] < 0.1,
          tt_dth$p.value > 0.5,
          tt_enj$p.value > 0.5, max(abs(tt_enj$conf.int)) < 0.5,
          tt_pair$p.value < 0.001,
          diff(tt_pair$conf.int) < diff(tt_unp$conf.int),
          tt_atr$p.value < 0.001, abs(gap_atr) < 0.5,
          tt_player$p.value < 0.001,
          isTRUE(all.equal(unname(diff(rev(tt_player$estimate))), gap_wt)),
          length(n_per) == 1,
          pt_frus$p.value < 0.01,
          power_5 < 0.5,
          quantile(wt_a, 0.25) > quantile(wt_b, 0.75))
```

> Every build comparison ends with a difference, because two groups of people
> never produce identical numbers. This page is about the two questions that
> follow: **could the difference be chance?** and **is it big enough to matter?**
> It builds the idea of a p-value from a simple shuffle, then introduces the
> t-test, paired designs, and how to write the result as a recommendation. The
> playtest question running through it: *"Did the new signposting reduce wrong
> turns?"*

::: {.at-a-glance}
**Stage:** 2 · Core.
**Time:** 30 minutes in class; 15 more for the sections marked Useful when needed.
**Before this:** [From Sample to Population](07-sample-to-population.qmd).
**Key terms:** observed difference, null hypothesis, p-value, significance level,
t-test, paired design, effect size, practical significance.
**R you will write:** `t.test(y ~ group)`, `t.test(x, y, paired = TRUE)`, and
`$estimate`, `$conf.int`, `$p.value` to pull out the parts.
:::

## Learning Outcomes

By the end of these notes you can:

- report a comparison in six steps: what happened, how large, how uncertain,
  could it be chance, does it matter, and what the team should do;
- explain a p-value through a shuffle, and say what it is and is not;
- run and read an independent-samples t-test and a paired t-test, and choose
  between them from the design;
- tell statistical significance from practical significance;
- write an evidence-backed design recommendation from a comparison.

---

## Why This Matters for Playtesting

Build B's players took fewer wrong turns, or scored more, or quit less. The
question is never "is there a difference?" but **"is this difference telling me
something about the build, or only about the players who happened to turn up?"**

**A playtest is a sample.** [From Sample to Population](07-sample-to-population.qmd)
showed that every mean wobbles from one playtest to the next. When you compare two
groups, both means wobble, and so does the gap between them. Before you rebuild a
level on the strength of a difference, you need to know whether that difference
would survive a second playtest.

**And a real difference can still be worthless.** With enough data, a change too
small for any player to notice becomes statistically detectable. A result can be
certain and irrelevant at the same time. Deciding whether it matters is a design
judgement, and no test makes it for you.

---

## Six Questions, in Order

Every comparison in this course is reported by answering six questions, in this
order. It is the **What → Pattern → Meaning → Action** frame, expanded for a
comparison:

| | Question | What you use | Frame |
| :- | :- | :- | :- |
| 1 | **What happened?** | Side-by-side plots, a summary per group | Pattern |
| 2 | **How large was the difference?** | The difference, in the measure's own units | Pattern |
| 3 | **How uncertain is the estimate?** | The 95% confidence interval for the difference | Pattern |
| 4 | **Could random sampling plausibly explain it?** | The p-value | Pattern |
| 5 | **Does the difference matter to the player experience?** | Design judgement | Meaning |
| 6 | **What should the team do?** | A recommendation | Action |

**Look at the plot before running the test.** The test is question 4, not
question 1. A p-value with no picture and no size attached cannot be interpreted.
If the boxplots show a gap nobody would care about, no p-value changes that
([rule 7](ten-rules.qmd)).

The rest of this page works through the six questions on one result: did Build B's
signposting reduce wrong turns?

---

## What Happened, and How Large?

The *Tidebreak* team added signposting to Build B. Did players get lost less?

```{r}
#| label: fig-wt-look
#| fig-cap: "Build B's players took fewer wrong turns. The boxes do not overlap, but the individual sessions do."
#| fig-alt: "Two boxplots of wrong turns per session, Build A higher than Build B, with every session drawn as a jittered point on top. Build A's points cluster around 8 to 10; Build B's around 4 to 7, with overlap between the two."
#| fig-height: 3.8
boxplot(wrong_turns ~ build, data = sessions, col = pal$fill, outline = FALSE,
        ylab = "Wrong turns per session", xlab = "Build")
stripchart(wrong_turns ~ build, data = sessions, vertical = TRUE, add = TRUE,
           method = "jitter", jitter = 0.15, pch = 19,
           col = adjustcolor(c(pal$build_a, pal$build_b), alpha.f = 0.5))
```

```{r}
#| label: wt-describe
sessions |>
  group_by(build) |>
  summarise(n = n(), mean = mean(wrong_turns), sd = sd(wrong_turns))
```

Build A's players averaged `r round(mean(wt_a), 1)` wrong turns per session and
Build B's `r round(mean(wt_b), 1)`. The **observed difference** is
`r round(gap_wt, 1)` wrong turns: your best single estimate of the true difference,
and almost certainly not exactly right. The two boxes do not overlap, but the
points do: plenty of individual Build A sessions had fewer wrong turns than plenty
of Build B sessions.

**Give the size in the measure's own units.** "`r round(gap_wt, 1)` fewer wrong
turns per session" is something a designer can picture. That is the **effect
size** in its simplest form.

---

## How Uncertain Is It?

A **confidence interval for the difference** works the same way as the interval
for one mean in [From Sample to Population](07-sample-to-population.qmd#confidence-intervals).
Here it runs from `r ci_fmt(tt_wt$conf.int)` wrong turns. You will get it from
`t.test()` in a moment.

**Read it as a range of plausible sizes**: the signposting saved somewhere between
`r round(tt_wt$conf.int[1], 1)` and `r round(tt_wt$conf.int[2], 1)` wrong turns per
session. **If an interval includes zero, "no difference" is still plausible**; this
one is nowhere near zero.

---

## Could It Be Chance? The Shuffle

Suppose the signposting did nothing. Then the build label on each session is
arbitrary: each player would have taken the same number of wrong turns whichever
build they had been given. So you could shuffle the labels, give every session a
random build, and the gap between the "builds" would be just as large as the real
one.

That is a test you can run. **Shuffle the build labels, recompute the gap, and
repeat ten thousand times.** The shuffled gaps show what differences chance alone
produces. If the real gap looks like them, chance is a fine explanation. If it is
far outside them, chance is not. This is called a **permutation test**.

**You do not need to write this code.** The idea is what matters: it is what every
p-value means.

```{r}
#| label: permutation
#| code-fold: true
#| code-summary: "Show the code"
gap <- function(y, g) mean(y[g == "A"]) - mean(y[g == "B"])

observed <- gap(sessions$wrong_turns, sessions$build)
set.seed(1)
shuffled <- replicate(10000, gap(sessions$wrong_turns, sample(sessions$build)))

observed
max(abs(shuffled))
mean(abs(shuffled) >= abs(observed))
```

```{r}
#| label: fig-permutation
#| code-fold: true
#| code-summary: "Show the code"
#| fig-cap: "If the build made no difference, shuffling the labels would produce gaps like the real one. In 10,000 shuffles, chance never came close to the real gap."
#| fig-alt: "A bell-shaped histogram of shuffled differences centred on zero, running from about −2 to +2. A red vertical line marks the observed difference of 3.4, far to the right of every bar."
#| fig-height: 3.4
par(mar = c(4, 4, 1, 1))
hist(shuffled_wt, breaks = 40, col = pal$fill, border = "white", main = "",
     xlim = c(-3.6, 3.6), xlab = "Difference in mean wrong turns (A − B)",
     ylab = "Shuffles")
abline(v = gap_wt, col = pal$highlight, lwd = 3)
text(gap_wt, par("usr")[4] * 0.9, "observed", pos = 2, col = pal$highlight, cex = 0.85)
```

The real gap is `r round(gap_wt, 1)`. The largest gap in ten thousand shuffles is
`r round(max(abs(shuffled_wt)), 2)`.

**The proportion of shuffles at least as extreme as the real gap is the
p-value.** Here not one of ten thousand was, so p is below 1 in 10,000. If the
signposting did nothing, a gap this large would essentially never happen.

**Most results are less clear-cut.** The same shuffle on the 1–7 difficulty rating
gives a gap of `r round(abs(obs_df), 2)` points (Build B rated harder), and
`r pct_fmt(perm_p_df, 1)` of shuffles produce a gap at least that large in either
direction:

```{r}
#| label: fig-pvalue-tail
#| code-fold: true
#| code-summary: "Show the code"
#| fig-cap: "The p-value is a tail area: the share of shuffles (shaded) at least as extreme as the observed difficulty gap, in either direction."
#| fig-alt: "A bell-shaped histogram of shuffled differences in difficulty rating centred on zero. The bars beyond the observed gap on both sides are shaded red; they are a small fraction of the total."
#| fig-height: 3.4
lim <- ceiling(max(abs(shuffled_df)) * 20) / 20 + 0.05
br  <- seq(-lim, lim, by = 0.05)
h  <- hist(shuffled_df, breaks = br, plot = FALSE)
par(mar = c(4, 4, 1, 1))
plot(h, col = ifelse(abs(h$mids) >= abs(obs_df), pal$highlight, pal$fill),
     border = "white", main = "", xlab = "Difference in mean difficulty (A − B)",
     ylab = "Shuffles")
abline(v = c(obs_df, -obs_df), lty = 2, col = pal$line)
```

That `r pct_fmt(perm_p_df, 1)` is the difficulty p-value. Chance alone produces a
gap that size in roughly one shuffle in `r round(1 / perm_p_df)`: unusual, but not
unheard of.

---

## The Language of Testing

The shuffle has a formal vocabulary.

The **null hypothesis** is the claim of no difference: here, that the build makes
no difference to wrong turns. The shuffle is a simulation of the world where the
null is true. The **alternative hypothesis** is that the build does make a
difference.

The **p-value** is the probability of seeing a difference at least as large as the
one observed, *if the null hypothesis were true*. A small p-value means the data
would be surprising in a world where the builds did not differ.

::: {.callout-important}
**A p-value is not the probability that the null hypothesis is true.**

It is calculated by *assuming* the null is true, so it cannot also measure how
likely the null is. p = 0.03 does not mean "a 3% chance the builds are the same".
It means "if the builds were the same, a gap this large would turn up about 3% of
the time".
:::

**The significance level**, written α (alpha), is the threshold you choose
**before** looking, conventionally 0.05. If p falls below it, the result is called
**statistically significant**: "unlikely to be chance alone", and nothing more.
Choosing it in advance is the same discipline as deciding what would change your
mind before you look at the data ([rule 1](ten-rules.qmd)).

**Not significant does not mean no difference.** A large p-value means this sample
could not rule out chance. That can be because there is no difference, or because
there were too few players to see one. The confidence interval tells you which.

**Going further:** the two ways a test can be wrong, statistical power, and why
testing many things at once produces false alarms are on
[Testing in Depth](../going-further/testing-in-depth.qmd).

---

## The t-Test

Shuffling ten thousand times works, but for comparing two means there is a formula
that gives almost the same answer instantly: the **t-test**. It is what you will
use.

```{r}
#| label: ttest
result <- t.test(wrong_turns ~ build, data = sessions)
result
```

`t.test(y ~ group)` is the **independent-samples t-test**, for when different
players tested each build. Read `~` as "split by", as in a boxplot. R runs the
Welch version by default, which does not assume the two builds have the same
spread; that is the right default. Read three parts of the output:

- the **sample estimates**: the two means (questions 1 and 2);
- the **95 percent confidence interval**: for Build A minus Build B (question 3);
- the **p-value** (question 4).

`$` pulls single parts out of the result, which is how you get numbers into a
sentence with inline R instead of copying them by hand:

```{r}
#| label: ttest-parts
result$estimate
result$conf.int
result$p.value
```

**The t-test and the shuffle agree.** For wrong turns, both say p is tiny. For
difficulty, the shuffle gave p = `r round(perm_p_df, 3)` and the t-test gives
`r p_fmt(tt_diff$p.value)`. The shuffle shows what the p-value *means*; the t-test
is how you will compute it.

---

## Does It Matter? Practical Significance

**Practical significance** is whether a difference is large enough to change the
player's experience, and so a design decision. Statistical significance cannot
tell you this, because it depends on how much data you have as well as on how big
the difference is.

**Lead with the size.** "Build B's players took `r round(gap_wt, 1)` fewer wrong
turns per session (95% CI `r ci_fmt(tt_wt$conf.int)`)." A designer can judge
whether 3 wrong turns matters; nobody can judge a p-value. For your game, the
important question is whether the *smallest* plausible difference, the low end of
the interval, is still large enough to matter.

**The Atrium shows why.** The frame-time capture has `r nrow(atrium)` frames from
the Atrium zone. There, Build B's frames are `r sprintf("%.2f", abs(gap_atr))` ms
faster than Build A's, `r p_fmt(tt_atr$p.value)`. That is about as statistically
significant as a result gets, and no player can perceive a difference of less than
half a millisecond per frame. Nobody should schedule work around it.

| | Practically small | Practically large |
| :- | :- | :- |
| **Statistically significant** | Real but not worth acting on | Act on it |
| **Not significant** | Nothing to see, if the interval is narrow | Test more players: the sample was too small to tell |

**Always report the size of the difference.** "Significant" alone hides which cell
of that table you are in.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Going further: Cohen's d"}
When measures are on different scales, a **standardised** effect size lets you
compare them. **Cohen's d** divides the difference in means by the typical SD of
the two groups:

$$d = \frac{\text{mean}_A - \text{mean}_B}{\text{pooled SD}}$$

For wrong turns, d = `r round(cohen_d, 2)`: the groups are well over one SD apart.
A common rough guide calls 0.2 small, 0.5 medium and 0.8 large, but those labels
come from psychology; in a playtest, "large" means large enough to change a design
decision, which only the units can tell you.
:::

---

## Two Designs: Independent and Paired

How the playtest was run decides which test you use
([Planning a Playtest](01-planning-a-playtest.qmd#choose-a-design-who-plays-what)).

**Independent samples.** Different players test each build. That is how the
*Tidebreak* sessions file was collected: `r n_players` players, each of whom played
only one build. Use the independent-samples t-test, as above.

**Paired.** The same players test both builds. `tidebreak_paired.csv` records a
second study in which `r nrow(paired)` players each played both. Each player is
compared with themselves, so their skill cancels out. Use the **paired t-test**:

```{r}
#| label: fig-paired
#| code-fold: true
#| code-summary: "Show the code"
#| fig-cap: "Each line is one player in the paired study. Most lines slope down, and players who took many wrong turns in Build A tend to take many in Build B too."
#| fig-alt: "Thirty grey lines joining each player's wrong turns in Build A on the left to Build B on the right. Most slope downward; players who took many wrong turns in Build A tend to take many in Build B too. A thick red line joins the two means."
#| fig-height: 4
par(mar = c(4, 4, 1, 1))
matplot(rbind(paired$wrong_turns_a, paired$wrong_turns_b), type = "l", lty = 1,
        col = adjustcolor(pal$line, alpha.f = 0.4), xaxt = "n", xlim = c(0.8, 2.2),
        xlab = "", ylab = "Wrong turns")
axis(1, at = 1:2, labels = c("Build A", "Build B"))
lines(1:2, colMeans(paired[, c("wrong_turns_a", "wrong_turns_b")]),
      col = pal$highlight, lwd = 4)
```

```{r}
#| label: paired-test
t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)
```

`paired = TRUE` compares each player's two values. The estimate is the **mean
difference**: on average, a player took `r round(tt_pair$estimate, 1)` fewer wrong
turns in Build B. `r sum(d_pair > 0)` of `r length(d_pair)` players improved,
`r sum(d_pair == 0)` stayed the same and `r sum(d_pair < 0)` got worse.

**Ignoring the pairing throws away what the design bought you.** The same numbers
tested as if the players were different people give the same estimate, but the
interval is `r round(diff(tt_unp$conf.int), 1)` wrong turns wide against
`r round(diff(tt_pair$conf.int), 1)` for the paired test. The paired test removes
the player-to-player variation; the unpaired test has to allow for it.

**Paired designs bring their own problems:** learning, order and fatigue effects.
If everyone plays Build A first, some of Build B's improvement is just practice.
Counterbalancing fixes it, but `tidebreak_paired.csv` does not record which build
each player played first, so this study cannot rule it out.

**For your project:** if your team tests a change on the same few players before
and after, that is a paired design. Use `paired = TRUE`, and counterbalance the
order if you possibly can.

---

## Comparing Proportions {.when-needed}

Not every outcome is a number. "Did the session end in a frustration quit?" is yes
or no, and the comparison is between two **proportions**:
`r frus_n[["A"]]` of `r build_n[["A"]]` Build A sessions
(`r pct_fmt(frus_n[["A"]] / build_n[["A"]])`) against `r frus_n[["B"]]` of
`r build_n[["B"]]` in Build B (`r pct_fmt(frus_n[["B"]] / build_n[["B"]])`).

`prop.test()` takes the counts of "yes" and the group sizes:

```{r}
#| label: prop-test
frus <- sessions |>
  group_by(build) |>
  summarise(frustrated = sum(quit_reason == "Frustrated"), n = n())
frus

prop.test(frus$frustrated, frus$n)
```

The difference is `r round(100 * (frus_n[["A"]] / build_n[["A"]] - frus_n[["B"]] / build_n[["B"]]))`
percentage points, with a 95% interval from
`r round(100 * pt_frus$conf.int[1])` to `r round(100 * pt_frus$conf.int[2])` points,
`r p_fmt(pt_frus$p.value)`. Build A's frustration rate is higher, and even the low
end of the interval is a difference a designer would care about. The same six
questions, for a yes/no outcome.

---

## Checking the Assumptions {.when-needed}

A t-test is reliable when these hold closely enough. None needs a formal test; each
needs a look.

| Assumption | What it means | How to check | *Tidebreak* |
| :- | :- | :- | :- |
| **Independence** | One observation does not influence another | Think about how the data was collected | Each player contributed `r n_per` sessions, so the `r nrow(sessions)` sessions are really `r n_players` players |
| **No strong outliers** | No few values dominate the means | Boxplots | `r sum(wt_out)` flagged wrong-turn counts across `r nrow(sessions)` sessions |
| **Roughly normal** | One hump, no long tail; matters less with a few dozen per group | A histogram | Fine for wrong turns; not for frame times |
| **Appropriate design** | Independent test for separate groups, paired test for the same people | How the playtest was run | Sessions: independent. Paired file: paired |

**A t-test compares means**, so it is the wrong tool when the mean is the wrong
summary. Frame times are the example: the problem is in the tail, as
[Distributions and Shape](05-distributions-and-shape.qmd#worked-example) showed.

**The independence problem has a simple check.** Average each player's sessions
into one value, then compare players rather than sessions:

```{r}
#| label: per-player
per_player <- sessions |>
  group_by(player_id, build) |>
  summarise(wrong_turns = mean(wrong_turns), .groups = "drop")
nrow(per_player)
t.test(wrong_turns ~ build, data = per_player)$conf.int
```

That leaves `r nrow(per_player)` players instead of `r nrow(sessions)` sessions.
`.groups = "drop"` just tidies the result. The difference is the same and the
interval is a little wider, `r ci_fmt(tt_player$conf.int)`, because there are fewer
independent observations. The conclusion survives.

---

## Which Test?

```{mermaid}
%%| label: fig-which-test
%%| fig-cap: "Choose the test from the outcome and the design, after you have looked at the data. A heavy tail means the mean is the wrong thing to compare."
flowchart TD
  S["Plot both groups"] --> Q1{"Yes/no or<br/>a number?"}
  Q1 -->|Yes/no| P(["prop.test()"])
  Q1 -->|A number| Q2{"Same players<br/>in both?"}
  Q2 -->|Yes| PT(["Paired t-test"])
  Q2 -->|No| Q3{"Mean the right<br/>summary?"}
  Q3 -->|Yes| IT(["Independent t-test"])
  Q3 -->|"No: heavy tail"| PM(["Compare the tail<br/>by permutation"])
  class S accent
```

The last branch is rare in playtests, and the [Testing in
Depth](../going-further/testing-in-depth.qmd#a-shuffle-for-any-summary) page shows
how to do it.

---

## Worked Example

**Did the signposting reduce wrong turns, without costing enjoyment?** The six
questions, in order.

**1. What happened?** @fig-wt-look: Build B's box sits lower. Build A's median is
`r median(wt_a)` wrong turns and Build B's is `r median(wt_b)`. The boxplot flags
`r wt_out[["A"]]` outlying sessions in Build A and `r wt_out[["B"]]` in Build B, too
few to dominate either mean.

**2. How large?** `r round(mean(wt_a), 1)` against `r round(mean(wt_b), 1)`: a
difference of `r round(gap_wt, 1)` wrong turns per session.

**3. How uncertain?** The 95% confidence interval is `r ci_fmt(tt_wt$conf.int)`
wrong turns. Nowhere near zero.

**4. Could it be chance?** `r p_fmt(tt_wt$p.value)`, and no shuffle in ten thousand
came close. It also survives counting each player once, and the separate paired
study points the same way.

**5. Does it matter?** The smallest plausible reduction,
`r round(tt_wt$conf.int[1], 1)` wrong turns per session, is still large enough for
a player to feel. And it did not cost enjoyment: the enjoyment difference is
`r p_fmt(tt_enj$p.value)`, with an interval from `r ci_fmt(tt_enj$conf.int, 2)`
points on a 7-point scale, so any change is under half a point either way.

**6. What should the team do?** Keep the signposting.

**In the report you would write:**

> Players using Build B made fewer wrong turns than players using Build A: on
> average `r round(gap_wt, 1)` fewer per session (95% CI
> `r ci_fmt(tt_wt$conf.int)`, `r p_fmt(tt_wt$p.value)`). A difference this large
> is very unlikely to be sampling variation alone. More importantly, even the
> smallest plausible reduction is large enough to improve navigation, and
> enjoyment did not change (95% CI for the difference `r ci_fmt(tt_enj$conf.int, 2)`
> points). The team should keep the revised signposting.

**Was Build B harder?** The same steps on the 1–7 difficulty rating give a second
significant result. Build B was rated `r round(gap_dif, 2)` points harder,
`r p_fmt(tt_diff$p.value)`, with a confidence interval from `r ci_fmt(ci_dif, 2)`
points.

Three things say not to act on it. The size is small: under half a point on a
seven-point scale. The interval's lower end, `r round(ci_dif[1], 2)`, is a
difference no player would report. And the play data does not back it up: deaths
did not differ between the builds (`r p_fmt(tt_dth$p.value)`). Statistically
significant, practically marginal, and unsupported by what players actually did.
**What should the team do?** Note it, and check it in the next playtest.

---

## Interpretation Guidance

**What the data shows.** In this playtest, Build B's players averaged
`r round(gap_wt, 1)` fewer wrong turns per session than Build A's (95% CI
`r ci_fmt(tt_wt$conf.int)`, `r p_fmt(tt_wt$p.value)`). In a separate study where
`r nrow(paired)` players played both builds, they took
`r round(tt_pair$estimate, 1)` fewer in Build B (95% CI
`r ci_fmt(tt_pair$conf.int)`).

**What the analysis suggests.** The signposting helps players find their way, and
by enough to matter. Two different designs, with different players, point the same
way.

**What can legitimately be concluded.** The reduction is very unlikely to be
chance, and even its smallest plausible size is a noticeable improvement. The
paired study cannot separate the signposting from practice, because play order was
not recorded; the independent study does not have that problem, which is why the
two together are stronger than either alone.

"The result was significant" is not a finding. It names no size and no action
([rule 10](ten-rules.qmd)).

---

## Common Mistakes

**Reading the p-value as the chance the null is true.** It is the chance of data
this extreme *if* the null is true.

**Reading "not significant" as "no difference".** It means this sample could not
tell. Check the confidence interval: narrow and around zero is evidence of no
meaningful difference; wide is a sample too small to say.

**Reading "significant" as "important".** The Atrium frame times are among the
most significant results in the whole playtest and among the least important.

**Testing before looking.** A t-test on data you have not drawn can hide outliers,
skew or two groups mixed together. Plot first, every time.

**Using the independent test on paired data.** It gives an interval far wider than
the data supports. Match the test to the design.

**Testing every measure and reporting what came up.** Decide which comparisons
answer your question before you run any of them. Test enough things and one will
come up by luck.

**Running too small a playtest and concluding "no effect".** A handful of players
per build will often miss even a large effect.

**Counting sessions as players.** Check that the result survives when each player
counts once.

---

## Practice

Each takes two or three minutes.

1. **Predict.** If you shuffled the build labels and recomputed the gap in mean
   *enjoyment*, would the real gap stand out from the shuffles? Then run the t-test
   and use the confidence interval to answer: did the signposting cost enjoyment?
2. **Compute.** Run the paired data through both the paired and the unpaired
   t-test. Why is the paired interval narrower?
3. **Decide.** Build B's sessions were longer on average. Run the test on
   `session_minutes`. Is the difference statistically significant? Is it
   practically significant?
4. **Critique.** A team tested 12 players, 6 per build, found p = 0.4 for wrong
   turns, and concluded: "The new signposting has no effect." What would you tell
   them?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
**1.**

```{r}
#| label: practice-enjoyment
t.test(enjoyment ~ build, data = sessions)
```

No: the p-value is `r round(tt_enj$p.value, 2)`, so the real gap looks exactly like
the shuffles. The interval says more: it runs from `r ci_fmt(tt_enj$conf.int, 2)`
points on a seven-point scale, so any true difference is smaller than half a point
either way. That is evidence the signposting did not cost enjoyment, a stronger
statement than "not significant".

**2.**

```{r}
#| label: practice-paired
t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)$conf.int
t.test(paired$wrong_turns_a, paired$wrong_turns_b)$conf.int
```

Players who take many wrong turns in one build take many in the other. The paired
test compares each player with themselves, so that player-to-player variation drops
out. The unpaired test leaves it in and has to allow for it.

**3.**

```{r}
#| label: practice-minutes
t.test(session_minutes ~ build, data = sessions)
```

```{r}
#| label: practice-minutes-numbers
#| include: false
tt_min <- t.test(session_minutes ~ build, data = sessions)
ci_min <- -rev(tt_min$conf.int)
```

Statistically, yes: `r p_fmt(tt_min$p.value)`. Practically, it is hard to say. The
interval says Build B's sessions were `r ci_fmt(ci_min)` minutes longer: anywhere
from `r round(ci_min[1] * 60)` seconds to `r round(ci_min[2], 1)` minutes. The
direction is clear and the size is not, which is a case for a larger playtest
rather than a decision. (Session length is also two-humped in Build A, so a
difference in means hides *how* the builds differ; the retention curve in
[Distributions and Shape](05-distributions-and-shape.qmd#the-retention-curve) shows
it.)

**4.** With 5 sessions per build, even an effect as large as *Tidebreak*'s is
detected only about `r pct_fmt(power_5)` of the time, and 6 is not much better. So
p = 0.4 is what you would often see whether or not the signposting works. "No
effect" is not supported. They should report the difference and its confidence
interval, which will be wide, and run a larger or paired playtest. [Testing in
Depth](../going-further/testing-in-depth.qmd#two-ways-to-be-wrong) shows where that
number comes from.
:::

---

## Summary

- Report a comparison in six steps: what happened, how large, how uncertain, could
  it be chance, does it matter, what should the team do. Plot before you test.
- A shuffle (permutation test) shows what differences chance produces. The p-value
  is the share of shuffles at least as extreme as the real result.
- A p-value is how surprising the data would be if there were no difference. It is
  not the probability that there is no difference.
- The t-test is how you compute it for two means: `t.test(y ~ group)` for
  different players, `paired = TRUE` for the same players.
- Lead with the size of the difference and its interval. Statistically significant
  means unlikely to be chance; practically significant means big enough to matter
  to the player.
- End with what the team should do.

| Key term | Meaning |
| :- | :- |
| **Observed difference** | The gap between the groups in your sample |
| **Permutation test** | Shuffling group labels to see what differences chance produces |
| **Null hypothesis** | The claim of no difference |
| **p-value** | The chance of a difference at least this large if the null were true |
| **Significance level (α)** | The p-value threshold chosen in advance, usually 0.05 |
| **Effect size** | How big the difference is, in the measure's units or standardised |
| **t-test** | The standard test for a difference in means |
| **Paired design** | The same players measured under both conditions |
| **Practical significance** | Whether a difference is big enough to change the player's experience |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 1 (decide what would change your mind before you
  look), rule 7 (compare distributions, not averages), rule 8 (small samples lie
  confidently) and rule 10 (a finding that does not name an action is not
  finished)
- [Testing in Depth](../going-further/testing-in-depth.qmd): errors and power,
  multiple comparisons, Q-Q plots, a shuffle for any summary
- [Statistics Reference: testing a difference](../reference/reference-statistics.qmd#testing-a-difference)
- [R and Quarto Reference: relationships and tests](../reference/r-quarto-reference.qmd#relationships-and-tests)
- [Lab 02: Comparing Builds](../labs/comparing-builds.qmd#part-b-is-the-difference-real-and-does-it-matter),
  Part B
- [Stage 2 demo](../demos/stage-2.qmd)
- Next, in Stage 3: [Visualising with ggplot2](09-visualising-with-ggplot2.qmd)

---

## Reflective Questions

- Build B's difficulty rating was significantly higher, but deaths did not change.
  What would you need to see in the next playtest to take the difficulty result
  seriously?
- A colleague says the paired study proves the signposting works because its
  p-value is smaller. What is wrong with that argument, and what does the paired
  study add that the sessions data does not?
- You can afford 30 testers for the next build comparison. Would you run them as
  two independent groups of 15 or as one paired group of 30? What would you have to
  control in the paired design?
- Pick a change your team made to your own game this term. Write the six questions
  you would answer to decide whether it worked, and the measure you would use for
  each.
