---
title: "Visualising with ggplot2"
subtitle: "From a plot you look at to a figure someone else reads"
body-classes: "wf-at-describe"
aliases:
  - visualising-with-ggplot2.html
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")
```

> A finding leaves your hands as a figure. This page introduces `ggplot2`, the
> tool for figures somebody else will read, now that you already know what a
> histogram, a boxplot and a scatterplot are for. It teaches enough to make one
> report-quality figure from your own playtest data. The playtest question running
> through it: *"Did the new signposting reduce wrong turns, and how do we show
> it?"*

::: {.at-a-glance}
**Stage:** 3A · Core. Taught in a clinic, November or December.
**Time:** 30 minutes, then the clinic lab on your own data.
**Before this:** [Describing Variables](04-describing-variables.qmd) and
[Distributions and Shape](05-distributions-and-shape.qmd): the base R versions of
every plot here.
**Key terms:** aesthetic mapping, geometry, mapping vs setting, fill, colour,
facet.
**R you will write:** `ggplot()`, `aes()`, `geom_histogram()`, `geom_bar()`,
`geom_boxplot()`, `geom_point()`, `facet_wrap()`, `labs()`.
:::

## Learning Outcomes

By the end of these notes you can:

- name the three parts every `ggplot2` figure is built from, and point to each in
  a line of code;
- tell mapping an aesthetic apart from setting one, and fix the bug that comes from
  confusing them;
- choose between `geom_histogram()`, `geom_bar()`, `geom_boxplot()` and
  `geom_point()` from the kind of variable you are drawing;
- use `fill`, `colour` and faceting to split a figure by group, and say which
  comparison each one makes easy;
- give a figure a title, axis labels and a caption a reader outside the team can
  follow;
- produce one figure from your team's own playtest data that is fit for the final
  report.

---

## Why This Matters for Playtesting

A finding leaves your hands as a figure. The producer, the level designer and the
person approving the next sprint will not read your console output. They will look
at one chart for a few seconds and remember what it seemed to say.

**The chart is the claim.** If the figure puts the wrong thing side by side, the
reader draws the wrong conclusion, however careful the analysis behind it was.
Choosing what to compare, and making that comparison the easiest thing on the page
to see, is part of the analysis rather than decoration added after it.

**The plots in Stages 1 and 2 were for you.** `hist()` and `boxplot()` answer a question in one
line while you are thinking. They stay useful. What they are not built for is the
figure that has to survive being pasted into a report, split by three groups and
read by someone who was not in the room.

---

## The Grammar: Data, Mapping, Geometry

`ggplot2` describes every figure with the same three parts:

> **data → aesthetic mapping → geometry**

The **data** is a data frame, one row per observation. The **aesthetic mapping**
says which column drives which visual property: position on the x axis, position
on the y axis, colour, fill. The **geometry** is the kind of mark drawn: bars,
boxes, points.

```r
ggplot(sessions, aes(x = score)) +
  geom_histogram()
```

Read it as: *take `sessions`, put `score` on the x axis, draw it as a histogram.*
`aes()` is where the mapping lives. Each further piece of the figure is added with
`+`, which is why a `ggplot2` figure is written as a stack of short lines rather
than one long function call.

**The grammar is what makes it a system.** Swap `geom_histogram()` for
`geom_boxplot()` and the same mapping draws a different chart. Add `fill = build`
to the mapping and every geometry splits by build in the way that suits it. You
learn three ideas once instead of learning every chart type separately.

```{mermaid}
%%| label: fig-grammar
%%| fig-cap: "A ggplot2 figure is built in layers, each added with +. The first three are required; the rest refine it."
flowchart LR
  D[("Data<br/>ggplot(sessions)")] --> M["Aesthetic mapping<br/>aes(x = score)"]
  M --> G["Geometry<br/>+ geom_histogram()"]
  G --> F["Facets<br/>+ facet_wrap(~ build)"]
  F --> L["Labels<br/>+ labs(title = ...)"]
  L --> P(["Figure"])
  class D,M,G accent
```

---

## Mapping and Setting

An aesthetic can be **mapped** or **set**, and the difference is where you write
it.

| | Written | Means | Result |
| :- | :- | :- | :- |
| Mapped | Inside `aes()`: `aes(fill = build)` | Colour depends on the data | One colour per build, and a legend |
| Set | Outside `aes()`: `geom_bar(fill = "grey70")` | Colour is fixed | Every bar the same colour, no legend |

**Mapping carries information; setting is styling.** If the colour tells the reader
something about the data, map it. If it is only there to look tidy, set it. The
most common `ggplot2` bug is putting a fixed colour inside `aes()`, covered in
Common Mistakes below.

---

## Choosing the Geometry

The geometry follows from the type of the variables you are drawing
([Data, Variables and Types](03-data-and-types.qmd#types-of-data),
[rule 3](ten-rules.qmd)), exactly as in Stage 1.

| You want to show | Variable types | Geometry | Base R you used before |
| :- | :- | :- | :- |
| The shape of one numeric variable | Discrete or continuous | `geom_histogram()` | `hist()` |
| Counts of one categorical variable | Nominal or ordinal | `geom_bar()` | `barplot()` |
| A numeric variable compared across groups | Categorical and numeric | `geom_boxplot()` | `boxplot(y ~ x)` |
| Two numeric variables together | Numeric and numeric | `geom_point()` | `plot(x, y)` |
| A value changing over time or order | Numeric against time | `geom_line()` | `plot(x, y, type = "l")` |

The first three are this page. Scatterplots are the starting point of
[Relationships and Correlation](10-relationships-and-correlation.qmd), the second
Stage 3 clinic.

**`geom_bar()` counts for you.** Give it the raw column and it tallies the
categories itself, so you do not need `count()` first.

---

## Splitting by Group: Fill, Colour and Facets

Most playtest questions compare groups: two builds, three cohorts, four
archetypes. `ggplot2` gives you three ways to split a figure, and each makes a
different comparison easy.

**`fill`** colours the inside of areas: bars, boxes, histogram bins. **`colour`**
colours lines and points, and the outline of areas. Map a group to `fill` for
bars and boxes, and to `colour` for points.

**Faceting** draws one small panel per group, all on the same axes:

```r
+ facet_wrap(~ build)
```

Read `~ build` as "split by build". Panels that share an axis let the eye compare
position directly, which is what makes faceting the strongest of the three when
the question is *how do these distributions differ?* ([rule 7](ten-rules.qmd)).

**Colour compares within a panel; facets compare between panels.** Two groups
overlaid in one panel invite a comparison at each x value. Two groups faceted
invite a comparison of whole shapes. Decide which question the reader is meant to
answer, then pick the split that answers it.

---

## Base R or ggplot2?

Both are correct tools, and neither replaces the other.

| | Base R | `ggplot2` |
| :- | :- | :- |
| Built for | Quick inspection while you think | A figure somebody else will read |
| Speed to a first picture | One line | Two or three lines |
| Splitting by group | Awkward beyond `y ~ group` | One more line: `fill`, `colour` or a facet |
| Legends and labels | Added by hand | Generated from the mapping |
| Changing the chart type | Rewrite the call | Swap the geometry |

**Use base R to find the finding and `ggplot2` to show it.** In practice you will
run `hist(sessions$wrong_turns)` a dozen times while exploring and build one
`ggplot2` figure for the report. That is the right division of labour, not a
failure to commit to one tool.

**That is why Stages 1 and 2 used base R.** You were inspecting: describing a
variable, checking a clean, looking at two builds before testing them. Every plot
had one reader, you, and one line was all it needed to earn. Now the reader
changes. Your final report is for someone who was not in the room, and building
that figure is the job `ggplot2` is designed for. Base R has not been replaced: you
will still reach for `hist()` first.

---

## Worked Examples

```{r}
#| label: worked-data
#| include: false
library(readr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)

wt_a <- sessions$wrong_turns[sessions$build == "A"]
wt_b <- sessions$wrong_turns[sessions$build == "B"]

stopifnot(median(wt_b) < median(wt_a),
          IQR(wt_b) > IQR(wt_a),
          max(wt_b) >= quantile(wt_a, 0.75))
```

**Did the new signposting reduce wrong turns?** Build B added lighting and signage
to guide players through the level. Both builds logged how many wrong turns each
tester took, `r nrow(sessions)` sessions in all. The question is a comparison of
two distributions, and the three drafts below draw the same data three ways.

**Draft 1: everything at once.**

```{r}
#| label: worked-draft-1
#| fig-alt: "Default ggplot2 histogram of wrong turns for all sessions: dark grey bars, no title, axes labelled only count and wrong_turns. The bars are broad at low values and peak around 9."
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1)
```

This is a correct histogram of the wrong question. It pools both builds, so it
describes the playtest as a whole and says nothing about whether Build B helped.
`binwidth = 1` gives one bar per whole number of wrong turns, which suits a count.

**Draft 2: fill by build.**

```{r}
#| label: worked-draft-2
#| fig-alt: "The same histogram with bars filled by build, Build A in red stacked on Build B in teal. Build B's bars sit mostly at lower values and Build A's at higher ones, but the stacking makes Build B's shape hard to read."
ggplot(sessions, aes(x = wrong_turns, fill = build)) +
  geom_histogram(binwidth = 1)
```

Mapping `fill = build` splits the bars and adds a legend without being asked. But
`geom_histogram()` stacks groups by default: Build A's bars sit on top of Build
B's, so Build A's bars start at a different height in every column. You can read
Build B's shape. You cannot read Build A's, and you cannot compare the two.

**Draft 3: facet by build.**

```{r}
#| label: worked-draft-3
#| fig-cap: "Build B's sessions sit further left, and are also more spread out: the signposting helped most players, not all of them."
#| fig-alt: "Two faceted histograms of wrong turns on the same axis, Build A above and Build B below, titled Wrong turns per session, by build. Build A peaks around 9; Build B sits further left, around 3 to 7."
#| fig-height: 5
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```

One panel per build, stacked vertically so the x axes line up. Now the comparison
is immediate. Build A's sessions bunch between `r quantile(wt_a, 0.25)` and
`r quantile(wt_a, 0.75)` wrong turns, around a median of `r median(wt_a)`. Build
B's centre has moved left to a median of `r median(wt_b)`.

The picture shows a second thing that the medians do not. Build B is **wider**,
not just further left: its interquartile range is `r IQR(wt_b)` wrong turns
against Build A's `r IQR(wt_a)`, and its worst session took `r max(wt_b)`. The
signposting helped most players a lot and some players hardly at all. That is a
different finding from "Build B reduced wrong turns", and it points at a different
next step: find out who the signposting did not reach.

Note that `fill = "grey60"` is **set**, outside `aes()`. The panels already say
which build is which, so colour would repeat information the reader has. That is
the kind of decision that separates a figure from a plot.

**What should the team do?** Keep the signposting, and find out who it did not
reach: check whether the high-wrong-turn sessions in Build B share a cohort, an
archetype or a route.

---

## Rebuilding the Stage 1 Plots

Each example rebuilds a base R plot from Stage 1 in `ggplot2`, so you can see what
changes and what does not.

```{r}
#| label: demo-load
library(readr)
library(dplyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
```

```{r}
#| label: demo-histogram
#| fig-alt: "Histogram titled Distribution of session score. Most sessions score between 1,200 and 1,800, with a small separate group around 500 to 700 on the left."
ggplot(sessions, aes(x = score)) +
  geom_histogram(binwidth = 100, fill = "grey60", colour = "white") +
  labs(title = "Distribution of session score", x = "Score", y = "Sessions")
```

`geom_histogram()` is `hist()`. Set `binwidth` in the units of the variable rather
than guessing a number of bins; the shape you see depends on it
([rule 4](ten-rules.qmd)). `colour = "white"` outlines each bar so neighbouring
bars stay distinct.

```{r}
#| label: demo-bar
#| fig-alt: "Grouped bar chart titled How sessions ended, by build. Build B has more completed tests and more lost-interest endings; Build A has four times as many frustrated endings."
ggplot(sessions, aes(x = quit_reason, fill = build)) +
  geom_bar(position = "dodge") +
  labs(title = "How sessions ended, by build",
       x = NULL, y = "Sessions", fill = "Build")
```

`geom_bar()` counts each quit reason itself. `position = "dodge"` puts the builds
side by side instead of stacking them, for the reason Draft 2 showed. The
Frustrated pair is the one to look at:
`r sum(sessions$build == "A" & sessions$quit_reason == "Frustrated")` sessions in
Build A against
`r sum(sessions$build == "B" & sessions$quit_reason == "Frustrated")` in Build B.
`x = NULL` removes an axis label that would only repeat the category names.

```{r}
#| label: demo-boxplot
#| fig-alt: "Four boxplots of score by archetype. Duelist, Engineer and Scout have similar compact boxes; Sentinel's box is much taller, reaching down to about 1,000, with its lower whisker near 500."
ggplot(sessions, aes(x = archetype, y = score)) +
  geom_boxplot(fill = "grey85") +
  labs(title = "Score by archetype", x = "Archetype", y = "Score")
```

`geom_boxplot()` is `boxplot(score ~ archetype)` with the formula replaced by the
mapping: group on x, numeric on y. The Sentinel's box is the tall one, with an IQR
of
`r round(IQR(sessions$score[sessions$archetype == "Sentinel"]))` points against at
most
`r round(max(tapply(sessions$score, sessions$archetype, IQR)[c("Duelist", "Engineer", "Scout")]))`
for the other three archetypes.

```{r}
#| label: demo-point
#| fig-alt: "Scatterplot of score against awareness rating with points coloured by cohort. Score rises with awareness; Veterans cluster towards the upper right and Novices towards the lower left, with a few low-scoring points along the bottom."
ggplot(sessions, aes(x = awareness, y = score, colour = cohort)) +
  geom_point() +
  labs(title = "Awareness rating against score",
       x = "Awareness rating", y = "Score", colour = "Cohort")
```

`geom_point()` is `plot(x, y)`. Mapping `colour = cohort` adds a third variable
without a second chart. The cloud rises from left to right: players rated higher
for awareness tend to score more. How strong that relationship is, and whether
awareness is the cause, is the subject of
[Relationships and Correlation](10-relationships-and-correlation.qmd), which starts
from this scatterplot.

```{r}
#| label: demo-facet
#| fig-alt: "Two faceted histograms of minutes played, Build A above and Build B below. Build A has a cluster of short sessions around 10 to 14 minutes and another near 30; most of Build B's sessions are between 25 and 32 minutes."
#| fig-height: 5
ggplot(sessions, aes(x = session_minutes)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Session length, by build",
       x = "Minutes played", y = "Sessions")
```

`facet_wrap(~ build)` is the line base R has no short equivalent for. Build A has a
cluster of sessions that end between minutes 11 and 13:
`r sum(sessions$build == "A" & sessions$session_minutes > 11 & sessions$session_minutes <= 13)`
of them, against
`r sum(sessions$build == "B" & sessions$session_minutes > 11 & sessions$session_minutes <= 13)`
in Build B. A single histogram of both builds would have blurred that cluster into
the overall spread.

```{r}
#| label: demo-base-pair
#| fig-alt: "Base R boxplots of score by archetype: the same four boxes as the ggplot2 version, with Sentinel's box much taller than the others."
boxplot(score ~ archetype, data = sessions)
```

For comparison, the base R version of the boxplot above. One line, no styling, and
entirely adequate for checking a hunch. Keep it for that.

---

## A Figure for Your Report

The Stage 3A target is **one figure from your team's own playtest data that you
would put in the final report**. Before it goes in, check it against this list:

1. **One question.** You can write the question the figure answers in one
   sentence, and the figure makes that comparison the easiest thing to see.
2. **The right geometry** for the variable types: histogram, bar, boxplot or
   point.
3. **Groups split for the comparison you mean**: dodged or faceted, never stacked
   when the groups are what matters.
4. **Colour only where it carries information.** Set it, outside `aes()`, if it is
   only styling.
5. **Labels for a reader.** A title that states the subject, axis labels in words
   with units, and no raw column names.
6. **n is visible** somewhere: in the title, a subtitle or the caption.
7. **A caption that states the takeaway**, not the name of the chart.

In your Quarto report the caption goes in the chunk options, and the code can be
hidden from the reader:

````markdown
```{{r}}
#| label: fig-wrong-turns
#| fig-cap: "Players made fewer wrong turns in Build B (n = 60 sessions per build)."
#| echo: false
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```
````

**For your project:** pick the result your team is most likely to lead the report
with, and build its figure now. It is much easier to fix a figure in December than
the night before the deadline.

---

## Interpretation Guidance

**What the data shows.** In this playtest, Build B's median was `r median(wt_b)`
wrong turns per session against Build A's `r median(wt_a)`. Build B's sessions
were also more spread out, and its worst session took `r max(wt_b)` wrong turns.

**What the analysis suggests.** The signposting guided most players but not all of
them. The long right-hand tail of Build B's panel is a group worth identifying: a
cohort, an archetype or a route that the new lighting does not reach.

**What can legitimately be concluded.** The testers who played Build B took fewer
wrong turns than the testers who played Build A. A figure cannot tell you whether
another group of players would show the same gap, or whether a gap this size could
arise by chance. That is what the t-test in
[Comparing Groups](08-comparing-groups.qmd) answers, and it is why
the figure comes before the test rather than instead of it.

**Every figure answers one question.** Before you draw, write the question the
reader is meant to answer in one sentence. Then check the finished figure makes
that comparison the easiest thing on the page to see. If it does not, change the
split, not the colours.

**Label for the reader, not for R.** `wrong_turns` is a column name. "Wrong turns"
is an axis label. A title that states the subject ("Wrong turns per session, by
build") lets the figure be understood when it is lifted out of your report and
pasted somewhere else, which it will be.

---

## Common Mistakes

**Putting a fixed colour inside `aes()`.** `geom_point(aes(colour = "blue"))` does
not draw blue points. It maps a column containing the word "blue" to colour, so
the points come out in `ggplot2`'s first default colour, with a legend reading
"blue". Fixed colours go outside `aes()`.

**Stacking what should be compared.** A stacked histogram or bar chart shows
totals well and the groups badly, because only the bottom group starts at zero.
Dodge the bars or facet the panels.

**Mapping a number where a category was meant.** A column of 1s and 2s mapped to
`colour` draws a continuous gradient, not two colours. Check the type with
`glimpse()` first ([rule 3](ten-rules.qmd)).

**Freeing the facet scales without saying so.** `facet_wrap(~ build, scales =
"free")` gives each panel its own axis range. Two panels can then look alike when
their values are very different. Keep the axes shared unless you have a reason and
say so on the figure.

**Colour for its own sake.** Every mapped colour is a claim that the grouping
matters. Four colours for four archetypes on a chart about builds is noise the
reader has to decode and ignore.

**Default labels in a report.** `session_minutes` on an axis tells the reader the
figure was never meant for them. Two lines of `labs()` fix it.

---

## Practice

Each takes two or three minutes.

1. Did the signposting make Build B less enjoyable? Draw the distribution of
   `enjoyment`, a 1–7 rating, for each build. Which geometry fits a 1–7 rating,
   and why?
2. This code is meant to draw blue points. What does it draw instead, and how do
   you fix it?

    ```r
    ggplot(sessions, aes(x = deaths, y = score)) +
      geom_point(aes(colour = "blue"))
    ```

3. Draw the quit reasons by build twice: once with `fill = build` and
   `position = "dodge"`, once with `facet_wrap(~ build)`. Which version makes it
   easier to see how Build A's players left, and which makes it easier to compare
   the builds on one reason?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
**1.**

```{r}
#| label: practice-enjoyment
#| fig-alt: "Two faceted bar charts of enjoyment rating, Build A above and Build B below. Both peak at 4 and have very similar shapes across ratings 2 to 6."
ggplot(sessions, aes(x = enjoyment)) +
  geom_bar(fill = "grey60") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Enjoyment rating, by build", x = "Enjoyment (1-7)", y = "Sessions")
```

`geom_bar()`, because a 1–7 rating has only seven possible values: it is ordinal,
and a bar per value shows exactly how many players chose it. Both panels peak at
the same rating,
`r names(which.max(table(sessions$enjoyment[sessions$build == "A"])))` in Build A
and `r names(which.max(table(sessions$enjoyment[sessions$build == "B"])))` in
Build B. Mean enjoyment was
`r sprintf("%.1f", mean(sessions$enjoyment[sessions$build == "A"]))` in Build A
and `r sprintf("%.1f", mean(sessions$enjoyment[sessions$build == "B"]))` in Build
B: fewer wrong turns, and no visible cost to enjoyment.

**2.**

```{r}
#| label: practice-mapping-bug
#| fig-alt: "Scatterplot of score against deaths in which every point is pinkish red and a legend reads colour: blue, the result of putting a fixed colour inside aes()."
ggplot(sessions, aes(x = deaths, y = score)) +
  geom_point(aes(colour = "blue"))
```

The points come out salmon, `ggplot2`'s first default colour, with a legend
reading "blue", because `colour = "blue"` inside `aes()` is a mapping. Move it
outside:

```{r}
#| label: practice-mapping-fix
#| fig-alt: "The same scatterplot of score against deaths with every point blue and no legend. Scores tend to fall as deaths rise."
ggplot(sessions, aes(x = deaths, y = score)) +
  geom_point(colour = "blue")
```

**3.**

```{r}
#| label: practice-dodge-facet
#| fig-alt: "Two horizontal bar charts side by side, one per build, counting how sessions ended. Completed test is the longest bar in both; Build A's frustrated bar is about four times as long as Build B's."
ggplot(sessions, aes(y = quit_reason)) +
  geom_bar(fill = "grey60") +
  facet_wrap(~ build) +
  labs(x = "Sessions", y = NULL)
```

The faceted version shows each build's full profile of quit reasons, so it answers
"how did Build A's players leave?". The dodged version earlier on this page puts
the two Frustrated bars next to each other, so it answers "which build frustrated
more players?". Mapping `quit_reason` to `y` instead of `x` turns the bars
sideways so the long labels fit. Neither version is wrong; they answer different
questions.
:::

---

## Summary

- Every `ggplot2` figure is data, mapped to aesthetics with `aes()`, drawn as a
  geometry. Pieces are added with `+`.
- Mapping (inside `aes()`) makes a property depend on the data. Setting (outside)
  fixes it.
- The geometry follows from the variable: histogram, bar, boxplot or point.
- Split by group with `fill`, `colour` or facets, and pick the one that makes the
  intended comparison easiest to see. Avoid stacking groups you mean to compare.
- Base R is for finding the finding; `ggplot2` is for showing it. Label every
  figure for a reader who was not in the room.
- Drawn properly, the wrong-turn data shows Build B lower **and** wider: the
  signposting helped most players, not all of them.

---

## Links

- [Ten Rules](ten-rules.qmd): rule 3 (know what kind of number you are holding),
  rule 4 (look at the shape before you quote the average) and rule 7 (compare
  distributions, not averages)
- [Distributions and Shape](05-distributions-and-shape.qmd): the base R
  versions of these plots, and what to look for in each
- [ggplot2 Reference](../reference/reference-ggplot2.qmd): every geometry, option
  and polishing step
- [Lab 03: A Figure for Your Report](../labs/clinic-figures.qmd), the Stage 3A
  clinic
- [Stage 3 demo](../demos/stage-3.qmd), Demo 1
- Next: [Relationships and Correlation](10-relationships-and-correlation.qmd)

---

## Reflective Questions

- Draft 2 and Draft 3 of the worked example use exactly the same data. One leaves
  the reader unable to answer the question and one answers it at a glance. What
  does that say about who is responsible when a reader misreads a chart?
- A teammate colours every figure in the report by archetype "for consistency".
  When is consistent colour a help to the reader, and when does it get in the way?
- Build B's wrong turns are lower and more spread out. Which of those two findings
  would you put in the headline figure for the level designer, and which for the
  producer? Would it be the same figure?
- You have ten minutes before a meeting and one question to check. Do you reach for
  `hist()` or `ggplot()`, and what would change your answer?
