---
title: "Relationships, Correlation and Regression"
subtitle: "When two things move together, and what that does and does not mean"
body-classes: "wf-at-describe wf-at-confounds"
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")
library(ggplot2)

abilities <- c("accuracy", "mobility", "survivability", "objective_control",
               "support", "awareness", "shots_fired")
cors <- sort(cor(sessions[, abilities], sessions$score)[, 1], decreasing = TRUE)
stopifnot(names(cors)[1] == "awareness", abs(cors[["shots_fired"]]) < 0.1)

r_dd <- cor(sessions$deaths, sessions$difficulty)

r_ms    <- cor(sessions$mobility, sessions$survivability)
r_ms_ar <- sapply(split(sessions, sessions$archetype),
                  function(d) cor(d$mobility, d$survivability))
stopifnot(r_ms < -0.5, all(r_ms_ar > 0))

aw_cohort <- round(tapply(sessions$awareness, sessions$cohort, mean), 1)
sc_cohort <- round(tapply(sessions$score, sessions$cohort, median))
r_within_cohort <- sapply(split(sessions, sessions$cohort),
                          function(d) cor(d$awareness, d$score))
stopifnot(aw_cohort[["Veteran"]] > aw_cohort[["Novice"]], all(r_within_cohort > 0.3))

fit   <- lm(score ~ awareness, data = sessions)
b0    <- unname(coef(fit)[1]); b1 <- unname(coef(fit)[2])
b1_ci <- confint(fit)["awareness", ]
r2    <- summary(fit)$r.squared
res   <- resid(fit)
worst <- sessions$archetype[order(res)][1:7]
stopifnot(sum(worst == "Sentinel") >= 5,
          isTRUE(all.equal(r2, cor(sessions$awareness, sessions$score)^2)))

ct_cq <- prop.table(table(sessions$cohort, sessions$quit_reason), 1)
stopifnot(ct_cq["Regular", "Frustrated"] > ct_cq["Novice", "Frustrated"])
```

> Most playtest questions are about relationships: do players who die more find
> the game harder? Do more aware players score more? This page covers how to see a
> relationship (the scatterplot), how to measure it (correlation), how to describe
> it as a line (simple linear regression), and the trap underneath all three:
> **two things moving together does not mean one causes the other.** The playtest
> question running through it: *"Awareness is the strongest predictor of score.
> Should we buff awareness?"*

::: {.at-a-glance}
**Time:** 30 minutes in class, shared with [Visualising with ggplot2](09-visualising-with-ggplot2.qmd);
25 more for regression and the Going further sections.
**Before this:** [Visualising with ggplot2](09-visualising-with-ggplot2.qmd).
**Key terms:** scatterplot, direction, form, strength, correlation, causation,
confounding variable, Simpson's paradox, simple linear regression, slope,
intercept, residual, R², extrapolation, cross-tabulation.
**R introduced:** `cor()`, `lm()`, `coef()`, `confint()`, `resid()`,
`geom_point()`, `geom_smooth(method = "lm")`, `geom_bar(position = "fill")`.
:::

## Learning Outcomes

By the end of these notes you can:

- read a scatterplot for direction, form and strength;
- compute and interpret a correlation, and say why context decides whether it is
  strong;
- name what correlation cannot see, including curves, outliers and groups;
- give three explanations other than cause for two things moving together;
- identify a confounding variable, and explain Simpson's paradox with a
  *Tidebreak* example;
- fit a simple linear regression with `lm()`, and interpret its slope, intercept,
  R² and residuals in plain English;
- describe the relationship between two categorical variables with a
  cross-tabulation;
- critique a playtest whose conclusion does not follow from its design.

---

## Why This Matters for Playtesting

Comparing two builds was a controlled question: the team chose who played which.
Most of what a playtest records is not controlled. Awareness, deaths, experience,
difficulty ratings: they all arrive together, and they all move together, because
they all come from the same players.

**A relationship is a lead, not a finding.** "Players with high awareness scored
more" is true in the *Tidebreak* data, and it is tempting to read it as "awareness
causes score, so buff awareness". The data cannot say that. Better players may
simply be more aware *and* score more. The difference between those two readings
decides whether a balance patch helps or does nothing, and only careful reasoning,
not more computation, tells them apart ([rule 7](ten-rules.qmd)).

---

## Seeing a Relationship: the Scatterplot

A **scatterplot** draws one point per observation, with one variable on each axis.
Do players who die more rate the game as more difficult?

```{r}
#| label: fig-deaths-difficulty
#| fig-cap: "Deaths against the player's difficulty rating, one point per session. Points are jittered slightly because both variables are whole numbers."
#| fig-alt: "Scatterplot of difficulty rating from 2 to 7 against deaths from 0 to 7. The cloud of points rises from lower left to upper right, with a lot of scatter."
#| fig-height: 3.8
ggplot(sessions, aes(x = deaths, y = difficulty)) +
  geom_jitter(width = 0.15, height = 0.15, colour = pal$line, alpha = 0.6) +
  labs(x = "Deaths in the session", y = "Difficulty rating (1–7)")
```

Read a scatterplot for three things:

- **Direction.** Do the points rise from left to right (**positive**: more of one,
  more of the other) or fall (**negative**)? Here they rise.
- **Form.** Do they follow a straight line, a curve, or no pattern at all? Here,
  roughly a line.
- **Strength.** How tightly do they cluster around that form? Here, loosely:
  plenty of players died often and still rated the game easy.

**In a report you would write:** "Players who died more rated the game as harder,
though with wide variation between players."

---

## Measuring a Relationship: Correlation

The **correlation** coefficient, *r*, puts a number on direction and strength for a
**straight-line** relationship. It runs from −1 to +1:

```{r}
#| label: fig-r-grid
#| echo: false
#| fig-cap: "What correlations look like. Each panel is 150 simulated points with the correlation shown."
#| fig-alt: "Six scatterplots in a row, labelled r = −0.9, −0.5, −0.2, 0.2, 0.5 and 0.9. At −0.9 the points form a tight falling band; at ±0.2 they are an almost shapeless cloud; at 0.9 a tight rising band."
#| fig-height: 2.6
#| fig-width: 9
set.seed(10)
rs <- c(-0.9, -0.5, -0.2, 0.2, 0.5, 0.9)
grid <- do.call(rbind, lapply(rs, function(r) {
  x <- rnorm(150)
  data.frame(r = sprintf("r = %.1f", r), x = x,
             y = r * x + sqrt(1 - r^2) * rnorm(150))
}))
grid$r <- factor(grid$r, levels = sprintf("r = %.1f", rs))
ggplot(grid, aes(x, y)) +
  geom_point(size = 0.7, colour = pal$line, alpha = 0.6) +
  facet_wrap(~ r, nrow = 1) +
  labs(x = NULL, y = NULL, caption = "Illustration: simulated data") +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

- **+1**: a perfect rising line. **−1**: a perfect falling line.
- **0**: no straight-line relationship.
- The further from 0, the more tightly the points follow a line.

`cor()` computes it:

```{r}
#| label: cor-dd
cor(sessions$deaths, sessions$difficulty)
```

r = `r round(r_dd, 2)`: a clear positive relationship with a lot of scatter, which
is what @fig-deaths-difficulty showed.

**Which abilities go with score?** Correlating each against score at once:

```{r}
#| label: cor-abilities
abilities <- c("accuracy", "mobility", "survivability", "objective_control",
               "support", "awareness", "shots_fired")
round(sort(cor(sessions[, abilities], sessions$score)[, 1], decreasing = TRUE), 2)
```

Awareness is the strongest, at `r round(cors[["awareness"]], 2)`. Shots fired, the
statistic everyone assumes matters, is `r round(cors[["shots_fired"]], 2)`: no
relationship at all.

**Context decides what counts as strong.** There is no universal cut-off. Between
two measurements of the same thing, such as two raters scoring the same session,
0.6 would be worryingly low. Between a single ability rating and a match outcome
shaped by teammates, opponents and luck, `r round(cors[["awareness"]], 2)` is large.
Say what the number means for the question, not just whether it is "strong".

::: {.callout-note collapse="true" title="Going further: Spearman's correlation"}
The usual correlation (Pearson's) measures straight-line relationships and is
pulled around by extreme values. **Spearman's correlation** ranks each variable
first and correlates the ranks, so it measures whether one goes up when the other
goes up, in any form, and is robust to outliers. It suits ordinal data such as
ratings:

```{r}
#| label: spearman
cor(sessions$deaths, sessions$difficulty, method = "spearman")
```
:::

---

## What Correlation Cannot See

A single number cannot describe a picture. In 1973 the statistician Francis
Anscombe published four small datasets that make the point. They have almost
identical means, SDs and correlations, and they look nothing alike:

```{r}
#| label: fig-anscombe
#| fig-cap: "Anscombe's quartet: four datasets with the same correlation, about 0.82, and four different stories."
#| fig-alt: "Four scatterplots, each with a straight fitted line of the same slope. Set 1 is a loose linear cloud. Set 2 is a smooth curve. Set 3 is a perfect line with one outlier pulling the fit. Set 4 is a vertical stack of points with one far-right point creating the whole relationship."
#| fig-height: 4.2
quartet <- do.call(rbind, lapply(1:4, function(i) {
  data.frame(set = paste("Set", i),
             x = anscombe[[paste0("x", i)]], y = anscombe[[paste0("y", i)]])
}))
round(sapply(split(quartet, quartet$set), function(d) cor(d$x, d$y)), 2)

ggplot(quartet, aes(x, y)) +
  geom_point(colour = pal$line, size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = pal$highlight, linewidth = 0.8) +
  facet_wrap(~ set, nrow = 1) +
  labs(x = NULL, y = NULL)
```

`anscombe` ships with R. Only Set 1 is what r = 0.82 suggests. Set 2 is a curve,
which r describes badly. In Set 3, one outlier drags the line away from an
otherwise perfect fit. In Set 4, a single point creates the entire relationship.

**Always draw the scatterplot before quoting r** ([rule 4](ten-rules.qmd)).
Correlation sees only straight lines, is moved by single extreme points, and, as
the next two sections show, cannot see groups.

---

## Correlation Is Not Causation

When two variables move together, there are always several explanations, and
"one causes the other" is only one of them:

| Explanation | Awareness and score |
| :- | :- |
| **A causes B** | Being aware of the map makes players score more |
| **B causes A** | Players who are scoring well have more time to look around |
| **Something else causes both** | Experienced players are both more aware and better at everything |
| **Chance** | In a small sample, unrelated measures can line up by luck |

A playtest observes; it does not intervene. Only a test in which you *change* one
variable and watch the other, such as giving half the players a better minimap at
random, can separate the first explanation from the rest. Until then, write "is
associated with", never "causes" or "drives".

---

## Confounding and Simpson's Paradox

A **confounding variable** is a third variable that drives both of the variables
you are comparing, creating or hiding a relationship between them.

```{mermaid}
%%| label: fig-confounder
%%| fig-cap: "A confounder drives both variables. The relationship you see between them may come entirely from it."
flowchart TD
  E["Experience<br/>(cohort)"] --> A["Awareness"]
  E --> S["Score"]
  A -.->|"the relationship<br/>you observe"| S
  classDef c fill:#fbeaea,stroke:#b22222,color:#212529
  classDef v fill:#f8f9fa,stroke:#6c757d,color:#212529
  class E c
  class A,S v
```

**Experience is a candidate confounder for awareness and score.** Veterans average
`r aw_cohort[["Veteran"]]` awareness against `r aw_cohort[["Novice"]]` for Novices,
and a median score of `r sc_cohort[["Veteran"]]` against `r sc_cohort[["Novice"]]`.
Some of the awareness–score relationship is experience. Checking within each
cohort, the relationship survives:

```{r}
#| label: within-cohort
sapply(split(sessions, sessions$cohort),
       function(d) cor(d$awareness, d$score))
```

From `r round(min(r_within_cohort), 2)` to `r round(max(r_within_cohort), 2)`
within each cohort: experience is part of the story, not all of it. That is how you
check a suspected confounder: **split by it, and see whether the relationship
holds inside each group.**

### When the relationship reverses

Sometimes splitting by the confounder does not just weaken a relationship; it
reverses it. That is **Simpson's paradox**, and *Tidebreak* has a clear case.
Across all players, mobility and survivability have a correlation of
`r round(r_ms, 2)`: faster players are more fragile. Within each archetype:

```{r}
#| label: fig-simpson
#| fig-cap: "Mobility against survivability. Left: all players, a falling trend. Right: the same points coloured by archetype, each with its own rising trend."
#| fig-alt: "Two scatterplots of survivability against mobility. Left: one black falling trend line through all the points. Right: the points form four separate clusters by archetype, Scout at high mobility and low survivability, Sentinel the reverse, and within each cluster the fitted line rises."
#| fig-height: 3.8
both <- rbind(
  data.frame(sessions[, c("mobility", "survivability", "archetype")], panel = "All players"),
  data.frame(sessions[, c("mobility", "survivability", "archetype")], panel = "By archetype")
)
ggplot(both, aes(mobility, survivability)) +
  geom_point(aes(colour = ifelse(panel == "All players", "All", archetype)), alpha = 0.7) +
  geom_smooth(data = subset(both, panel == "All players"), method = "lm",
              se = FALSE, colour = "black") +
  geom_smooth(data = subset(both, panel == "By archetype"),
              aes(colour = archetype), method = "lm", se = FALSE) +
  scale_colour_manual(values = c(All = "grey60", Duelist = "#3a5f8a",
                                 Engineer = "#6c9a3c", Scout = "#b22222",
                                 Sentinel = "#8a5a9c"),
                      breaks = c("Duelist", "Engineer", "Scout", "Sentinel")) +
  facet_wrap(~ panel) +
  labs(x = "Mobility rating", y = "Survivability rating", colour = "Archetype")
```

```{r}
#| label: within-archetype
sapply(split(sessions, sessions$archetype),
       function(d) cor(d$mobility, d$survivability))
```

Within every archetype the correlation is **positive**, from
`r round(min(r_ms_ar), 2)` to `r round(max(r_ms_ar), 2)`. The overall falling trend
is the gap *between* archetypes: Scouts are built fast and fragile, Sentinels slow
and tough. Within any one archetype, better players are better at both.
**Archetype is the confounder**, and ignoring it reverses the conclusion. A patch
that "makes fast players more fragile" would punish exactly the players who are
playing their role well. This is Lab 03, Exercise 07.

---

## Simple Linear Regression

Correlation says how tightly two variables follow a line. **Simple linear
regression** finds that line, and so answers a more useful question: *how much does
one variable change, on average, for each unit of the other?*

```{r}
#| label: lm
fit <- lm(score ~ awareness, data = sessions)
coef(fit)
confint(fit)
```

`lm()` stands for *linear model*. `score ~ awareness` reads "score as explained by
awareness", the same formula style as `boxplot(y ~ group)`. The result describes a
straight line:

$$\text{predicted score} = `r round(b0)` + `r round(b1, 1)` \times \text{awareness}$$

**The slope** is the number that matters: each extra point of awareness goes with
`r round(b1, 1)` more points of score, on average (95% CI
`r round(b1_ci[1], 1)` to `r round(b1_ci[2], 1)`). It is in the units of the
question, so a designer can use it: the
`r round(aw_cohort[["Veteran"]] - aw_cohort[["Novice"]])`-point gap in average
awareness between Novices and Veterans goes with about
`r round((aw_cohort[["Veteran"]] - aw_cohort[["Novice"]]) * b1, -1)` points of
score.

**The intercept** is the predicted score at awareness zero:
`r round(b0)`. No player has an awareness anywhere near zero (the lowest is
`r min(sessions$awareness)`), so it means nothing on its own; it just anchors the
line.

```{r}
#| label: fig-regression
#| fig-cap: "Score against awareness, with the regression line. Each grey segment is a residual: the gap between a session's actual score and the line's prediction."
#| fig-alt: "Scatterplot of score against awareness with a rising red line. Short grey vertical segments join each point to the line. A handful of points sit far below the line at the bottom, with long segments, coloured differently from the rest."
#| fig-height: 4.2
sessions$predicted <- fitted(fit)
sessions$residual  <- resid(fit)
ggplot(sessions, aes(awareness, score)) +
  geom_segment(aes(xend = awareness, yend = predicted), colour = "grey75") +
  geom_point(aes(colour = residual < -600), size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = pal$highlight) +
  scale_colour_manual(values = c(`FALSE` = pal$line, `TRUE` = "#8a5a9c"),
                      labels = c("Other sessions", "Residual below −600"), name = NULL) +
  labs(x = "Awareness rating", y = "Score") +
  theme(legend.position = "bottom")
```

**A residual** is the gap between what actually happened and what the line
predicted: actual minus predicted. The line is chosen to make the residuals, taken
together, as small as possible (it minimises the sum of their squares, hence
"least squares"). `geom_smooth(method = "lm")` draws the same line.

**Residuals are where the findings hide.** The seven sessions furthest *below* the
line, the ones that scored far less than their awareness predicted, include
`r sum(worst == "Sentinel")` Sentinels. The regression has found the
Sentinel collapses from [Describing Variables](04-describing-variables.qmd#worked-example)
on its own: those games are not explained by awareness at all.

**R²**, the **coefficient of determination**, is the share of the variation in score
that the line accounts for: `r round(r2, 2)`, or `r pct_fmt(r2)`. For a simple
regression it is exactly the correlation squared
(`r round(cors[["awareness"]], 2)`² = `r round(r2, 2)`). Awareness accounts for about
a third of the differences in score; the other two thirds are everything else.

**Predicting, carefully.** The line predicts a score of about
`r round(predict(fit, data.frame(awareness = 80)), -1)` for awareness 80. That
prediction is an average for players like that, not a promise for any one player,
whose residual could easily be `r round(sd(res), -1)` points either way. And
**never predict outside the data**: the line knows nothing about awareness below
`r min(sessions$awareness)` or above `r max(sessions$awareness)`. Using it there is
**extrapolation**, and the straight line may not continue.

**A regression is still a correlation.** A steep, significant slope says awareness
and score move together, by this much. It does not say that raising a player's
awareness would raise their score. Every warning in the previous two sections
applies unchanged.

::: {.callout-note collapse="true" title="Going further: reading the full summary"}
`summary(fit)` prints everything at once:

```{r}
#| label: lm-summary
summary(fit)
```

The `awareness` row of the **Coefficients** table gives the slope, its standard
error and a p-value testing whether the true slope could be zero. **Multiple
R-squared** is R². Regression with several predictors at once (multiple
regression) can adjust for a confounder such as cohort, but it is beyond this
module, and adjusting for the confounders you measured does not protect you from
the ones you did not.
:::

---

## Two Categorical Variables

When both variables are categories, the relationship is a **cross-tabulation**,
read through **row proportions** so that groups of different sizes compare fairly.
Do more experienced players quit in frustration less?

```{r}
#| label: crosstab
round(prop.table(table(sessions$cohort, sessions$quit_reason), 1), 2)
```

```{r}
#| label: fig-crosstab
#| fig-cap: "How sessions ended, by cohort. Each bar is one cohort, scaled to 100%."
#| fig-alt: "Three stacked bars, one per cohort, each split into completed test, frustrated, lost interest and technical fault. Regulars have the largest frustrated segment; Veterans the largest lost-interest segment."
#| fig-height: 3.5
ggplot(sessions, aes(x = cohort, fill = quit_reason)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = function(x) paste0(x * 100, "%")) +
  scale_fill_manual(values = c("grey80", "#b22222", "#9bb0c9", "grey40")) +
  labs(x = "Cohort", y = "Share of sessions", fill = "How the session ended")
```

`position = "fill"` scales every bar to 100%, the picture of row proportions.

Not in a straight line. **Regulars** quit in frustration most often,
`r pct_fmt(ct_cq["Regular", "Frustrated"])` of their sessions, against
`r pct_fmt(ct_cq["Novice", "Frustrated"])` for Novices and
`r pct_fmt(ct_cq["Veteran", "Frustrated"])` for Veterans. Veterans more often simply
lost interest. A relationship between categories does not have to be "more of A,
more of B", which is one reason correlation is the wrong tool here.

---

## Critiquing a Playtest

Every one of these ideas comes together when a conclusion is drawn from data that
cannot support it. Two scenarios to take apart.

**Scenario 1.** *A team adds an optional minimap. Players who turned it on scored
22% higher than players who left it off. The team makes the minimap mandatory.*

::: {.callout-note collapse="true" title="What is wrong"}
Players chose whether to use the minimap, so the two groups differ in more than the
minimap. Experienced, engaged players are more likely to explore the options menu
*and* more likely to score well: a confounder. The 22% may be entirely skill. To
know what the minimap does, assign it at random, or have every player try both,
counterbalanced ([Planning a Playtest](01-planning-a-playtest.qmd#random-assignment)).
:::

**Scenario 2.** *The new build was tested on Friday evening with fans recruited
from the studio Discord; the old build on Monday morning with first-year students.
Enjoyment was 1.5 points higher for the new build, p < 0.001. The team reports that
the new build is more enjoyable.*

::: {.callout-note collapse="true" title="What is wrong"}
Build is confounded with everything else that differed: the players (committed fans
against strangers), the time of day, and probably the room and who ran the session.
Any of them could produce the gap. The p-value says the gap is unlikely to be
chance; it says nothing about *which* of the differences caused it. The fix is to
recruit one group and assign builds at random within it, at the same times, in the
same room.
:::

---

## Worked Example

**Should we buff awareness?** The balance designer has seen that awareness is the
strongest predictor of score and proposes raising every archetype's base awareness.

**Step 1: look.** The scatterplot (@fig-regression) shows a clear rising
relationship, with a cluster of sessions far below the line.

**Step 2: measure.** r = `r round(cors[["awareness"]], 2)`, the strongest of the
ability ratings. The slope is `r round(b1, 1)` points of score per awareness point
(95% CI `r round(b1_ci[1], 1)` to `r round(b1_ci[2], 1)`), and awareness accounts
for `r pct_fmt(r2)` of the variation in score.

**Step 3: ask what else could explain it.** Experience: Veterans have more of both.
Within each cohort the relationship weakens but survives
(`r round(min(r_within_cohort), 2)` to `r round(max(r_within_cohort), 2)`), so
experience is part of the story. But a player's awareness *rating* may measure
skill as much as cause it: good players notice more because they are good.

**Step 4: look at the residuals.** The biggest misses are Sentinel collapses, which
awareness does not explain at all. A buff to awareness does nothing for them.

**Step 5: conclude.** The data shows an association, by a useful amount. It does not
show that raising awareness would raise score. The defensible recommendation is a
test, not a patch: give half the players a feature that raises awareness, such as a
clearer minimap or audio cues, at random, and measure score in both groups.

---

## Interpretation Guidance

**What the data shows.** Score rises with awareness: about `r round(b1, 1)` points
per awareness point, r = `r round(cors[["awareness"]], 2)`, across
`r nrow(sessions)` sessions.

**What the analysis suggests.** Awareness, or whatever awareness ratings reflect,
matters to how a match goes. Experience explains part of the relationship but not
all of it.

**What can legitimately be concluded.** An association, not a cause. A design
change aimed at awareness is worth testing; it is not yet supported as a fix.

**Use the right verbs.** "Is associated with", "goes with", "predicts" for
relationships in observed data. "Causes", "drives", "improves" only after an
experiment that changed one variable on purpose.

---

## Common Mistakes

**Quoting r without the scatterplot.** Anscombe's four datasets share a correlation
and nothing else ([rule 4](ten-rules.qmd)).

**Reading correlation as cause.** Two things moving together has at least four
explanations. Name the alternatives before choosing one.

**Ignoring groups.** A relationship across groups can vanish or reverse within
them. Split by the obvious confounders, such as archetype, cohort or build.

**Calling r "strong" or "weak" without context.** Say what it means for the
question.

**Interpreting the intercept when zero is impossible.** No player has zero
awareness. The intercept anchors the line; it is not a finding.

**Extrapolating.** The line describes the range of the data, not beyond it.

**Treating a prediction as a promise.** A regression line predicts an average;
individual players scatter around it by the residual SD.

**Using correlation on categories.** Two categorical variables need a
cross-tabulation, not `cor()`.

---

## Practice

Each takes two or three minutes.

**1. Predict.** Before computing it: is the correlation between `kills` and
`score` closer to 0, 0.5 or 0.9? Then check.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-kills
cor(sessions$kills, sessions$score)
```

Close to 0, `r round(cor(sessions$kills, sessions$score), 2)`, which surprises
most people. In *Tidebreak*, score tracks staying alive (time alive correlates at
`r round(cor(sessions$time_alive_s, sessions$score), 2)` with score) far more than
it tracks kills. A number you were sure about is exactly the one to check.
:::

**2. Compute.** Fit a regression of `score` on `deaths`. Interpret the slope in one
sentence a designer could use, and compare its R² with awareness's.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-deaths
fit_d <- lm(score ~ deaths, data = sessions)
coef(fit_d)
summary(fit_d)$r.squared
```

Each extra death goes with about `r abs(round(coef(lm(score ~ deaths, data = sessions))[2]))`
fewer points of score, on average. R² is
`r round(summary(lm(score ~ deaths, data = sessions))$r.squared, 2)`, a little below
awareness's `r round(r2, 2)`. The same caution applies: dying more and scoring less
may both come from being outplayed.
:::

**3. Critique.** "Wrong turns correlate at
`r round(cor(sessions$wrong_turns, sessions$enjoyment), 2)` with enjoyment, so
getting lost does not affect how much players enjoy the game." What is wrong with
that conclusion?

::: {.callout-note collapse="true" title="Check your answer"}
A correlation near zero means no *straight-line* relationship across all sessions.
It does not rule out a curve, a relationship within one build or cohort that is
cancelled out by another, or an effect on a different outcome (players who got lost
may have quit, and quitters' enjoyment could be missing or lower). Draw the
scatterplot, split by build, and look at quit reason too before concluding
anything.
:::

**4. Decide.** A designer reads the regression and says: "If we raise a Novice's
awareness by 20 points, they will score about `r round(20 * b1, -1)` more." Give two
reasons to doubt it.

::: {.callout-note collapse="true" title="Check your answer"}
First, the slope is an association across players, not the effect of changing one
player: awareness ratings may reflect skill rather than cause it. Second, it is an
average: individual players scatter around the line by about
`r round(sd(res), -1)` points. A third reason, if the Novice starts low: their raised
awareness may still be in a range where the data is thin.
:::

---

## Summary

- A scatterplot shows a relationship's direction, form and strength. Draw it first.
- Correlation, from −1 to +1, measures straight-line relationships. It misses
  curves, is moved by outliers, and cannot see groups. Context decides "strong".
- Correlation is not causation: the reverse direction, a third variable or chance
  can produce the same pattern.
- A confounder drives both variables. Check by splitting into groups; Simpson's
  paradox is when the relationship reverses inside them.
- Simple linear regression, `lm(y ~ x)`, fits a line. The slope is the change in y
  per unit of x; R² is the share of variation explained; residuals are where the
  unexplained cases, often the findings, sit.
- Do not interpret an impossible intercept or extrapolate beyond the data.
- Two categorical variables: cross-tabulation with row proportions.

| Key term | Meaning |
| :- | :- |
| **Scatterplot** | One point per observation, for two numeric variables |
| **Correlation (r)** | −1 to +1: direction and strength of a straight-line relationship |
| **Causation** | One variable changing another; needs an experiment to show |
| **Confounding variable** | A third variable that drives both variables you are comparing |
| **Simpson's paradox** | A relationship that reverses when the data is split into groups |
| **Simple linear regression** | Fitting the straight line that best predicts y from x |
| **Slope** | The change in y, on average, for each unit increase in x |
| **Intercept** | The line's predicted y when x is zero |
| **Residual** | Actual minus predicted |
| **R²** | The share of the variation in y the line accounts for |
| **Extrapolation** | Predicting outside the range of the data |
| **Cross-tabulation** | Counts of every combination of two categorical variables |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 4 (look at the shape before you quote the
  average) and rule 7 (compare distributions, not averages)
- [Statistics Reference: relationships](../reference/reference-statistics.qmd#relationships)
- [ggplot2 Reference](../reference/reference-ggplot2.qmd)
- [Lab 03: Complete Analysis](../labs/complete-analysis.qmd), Part A
- [Week 3 demo](../demos/week-3.qmd)
- Next: [Interpreting and Reporting Evidence](11-reporting-evidence.qmd)

---

## Reflective Questions

- Name a relationship in your own game's telemetry that you suspect is confounded.
  What is the confounder, and how would you check?
- The awareness regression explains about a third of the variation in score. Is
  that a lot or a little for a game outcome? What would you expect the other two
  thirds to be?
- Simpson's paradox reversed the mobility–survivability relationship. What other
  grouping in a game could hide or reverse a relationship in the same way?
- Your producer wants a single sentence for the balance meeting about awareness.
  Write it, using only what the data can support.
