---
title: "Relationships and Correlation"
subtitle: "When two things move together, and what that does and does not mean"
body-classes: "wf-at-compare wf-at-interpret"
aliases:
  - 10-relationships-and-regression.html
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")
library(dplyr)
library(ggplot2)

abilities <- c("accuracy", "mobility", "survivability", "objective_control",
               "support", "awareness", "shots_fired")
cors <- sort(cor(sessions[, abilities], sessions$score)[, 1], decreasing = TRUE)
stopifnot(names(cors)[1] == "awareness", abs(cors[["shots_fired"]]) < 0.1)

r_dd  <- cor(sessions$deaths, sessions$difficulty)
rs_dd <- cor(sessions$deaths, sessions$difficulty, method = "spearman")

metrics <- c("deaths", "wrong_turns", "difficulty", "enjoyment", "session_minutes")
cm <- cor(sessions[, metrics], method = "spearman")
stopifnot(cm["deaths", "difficulty"] > 0.5,
          cm["difficulty", "enjoyment"] < -0.25,
          all(abs(cm["wrong_turns", setdiff(metrics, "wrong_turns")]) < 0.15))

joined <- merge(sessions, survey, by = "session_id")
r_wt_nav <- cor(joined$wrong_turns, joined$level_navigable, method = "spearman")
nav_by_build <- tapply(joined$level_navigable, joined$build, mean)
wt_by_build  <- tapply(joined$wrong_turns, joined$build, mean)
stopifnot(abs(r_wt_nav) < 0.15, abs(diff(nav_by_build)) < 0.2,
          wt_by_build[["A"]] - wt_by_build[["B"]] > 2)

r_ms    <- cor(sessions$mobility, sessions$survivability)
r_ms_ar <- sapply(split(sessions, sessions$archetype),
                  function(d) cor(d$mobility, d$survivability))
stopifnot(r_ms < -0.5, all(r_ms_ar > 0))

aw_cohort <- round(tapply(sessions$awareness, sessions$cohort, mean), 1)
sc_cohort <- round(tapply(sessions$score, sessions$cohort, median))
r_within_cohort <- sapply(split(sessions, sessions$cohort),
                          function(d) cor(d$awareness, d$score))
stopifnot(aw_cohort[["Veteran"]] > aw_cohort[["Novice"]], all(r_within_cohort > 0.3))

ct_cq <- prop.table(table(sessions$cohort, sessions$quit_reason), 1)
stopifnot(ct_cq["Regular", "Frustrated"] > ct_cq["Novice", "Frustrated"])
```

> Most playtest questions are about relationships: do players who die more find
> the game harder? Do players who get lost rate the level as confusing? This page
> covers how to see a relationship (the scatterplot), how to measure it
> (correlation), how to scan several at once (a correlation matrix), and the trap
> underneath all of them: **two things moving together does not mean one causes
> the other.** The playtest question running through it: *"Awareness is the
> strongest predictor of score. Should we buff awareness?"*

::: {.at-a-glance}
**Stage:** 3B · Core. Taught in a clinic, January or February.
**Time:** 30 minutes, then the clinic lab on your own data.
**Before this:** [Visualising with ggplot2](09-visualising-with-ggplot2.qmd).
**Key terms:** scatterplot, direction, form, strength, correlation, Pearson,
Spearman, correlation matrix, causation, confounding variable.
**R you will write:** `cor()`, `cor(..., method = "spearman")`, `select()` with
`cor()` for a matrix, `geom_point()`.
:::

## Learning Outcomes

By the end of these notes you can:

- read a scatterplot for direction, form and strength;
- compute and interpret a Pearson correlation, and say why context decides whether
  it is strong;
- say when to use Spearman's correlation instead, and compute it;
- build and read a correlation matrix of 3–5 playtest measures;
- name what correlation cannot see, and give three explanations other than cause
  for two things moving together;
- check a suspected confounder by looking at the relationship within groups;
- relate telemetry to self-report, and say what it means when they disagree.

---

## Why This Matters for Playtesting

Comparing two builds was a controlled question: the team chose who played which.
Most of what a playtest records is not controlled. Awareness, deaths, experience,
difficulty ratings: they all arrive together, and they all move together, because
they all come from the same players.

**A relationship is a lead, not a finding.** "Players with high awareness scored
more" is true in the *Tidebreak* data, and it is tempting to read it as "awareness
causes score, so buff awareness". The data cannot say that. Better players may
simply be more aware *and* score more. The difference between those two readings
decides whether a balance patch helps or does nothing, and only careful reasoning,
not more computation, tells them apart ([rule 7](ten-rules.qmd)).

**For your project:** by now your team has its own playtest data. This page is
written so that every step works on it. The clinic lab ends with a correlation
matrix of your own measures.

---

## Seeing a Relationship: the Scatterplot

A **scatterplot** draws one point per observation, with one variable on each axis.
Do players who die more rate the game as more difficult?

```{r}
#| label: fig-deaths-difficulty
#| fig-cap: "Players who died more rated the game as harder, with a lot of variation. Points are jittered slightly because both variables are whole numbers."
#| fig-alt: "Scatterplot of difficulty rating from 2 to 7 against deaths from 0 to 7. The cloud of points rises from lower left to upper right, with a lot of scatter."
#| fig-height: 3.8
ggplot(sessions, aes(x = deaths, y = difficulty)) +
  geom_jitter(width = 0.15, height = 0.15, colour = pal$line, alpha = 0.6) +
  labs(x = "Deaths in the session", y = "Difficulty rating (1–7)")
```

`geom_jitter()` is `geom_point()` with a little random nudge, so that sessions with
the same values do not sit on top of each other. Read a scatterplot for three
things:

- **Direction.** Do the points rise from left to right (**positive**: more of one,
  more of the other) or fall (**negative**)? Here they rise.
- **Form.** Do they follow a straight line, a curve, or no pattern at all? Here,
  roughly a line.
- **Strength.** How tightly do they cluster around that form? Here, loosely:
  plenty of players died often and still rated the game easy.

**In a report you would write:** "Players who died more rated the game as harder,
though with wide variation between players."

---

## Measuring a Relationship: Correlation

The **correlation** coefficient, *r*, puts a number on direction and strength for a
**straight-line** relationship. It runs from −1 to +1:

```{r}
#| label: fig-r-grid
#| echo: false
#| fig-cap: "What correlations look like. At ±0.2 there is barely a pattern; at ±0.9 the points form a tight band."
#| fig-alt: "Six scatterplots in a row, labelled r = −0.9, −0.5, −0.2, 0.2, 0.5 and 0.9. At −0.9 the points form a tight falling band; at ±0.2 they are an almost shapeless cloud; at 0.9 a tight rising band."
#| fig-height: 2.6
#| fig-width: 9
set.seed(10)
rs <- c(-0.9, -0.5, -0.2, 0.2, 0.5, 0.9)
grid <- do.call(rbind, lapply(rs, function(r) {
  x <- rnorm(150)
  data.frame(r = sprintf("r = %.1f", r), x = x,
             y = r * x + sqrt(1 - r^2) * rnorm(150))
}))
grid$r <- factor(grid$r, levels = sprintf("r = %.1f", rs))
ggplot(grid, aes(x, y)) +
  geom_point(size = 0.7, colour = pal$line, alpha = 0.6) +
  facet_wrap(~ r, nrow = 1) +
  labs(x = NULL, y = NULL, caption = "Illustration: simulated data") +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

- **+1**: a perfect rising line. **−1**: a perfect falling line.
- **0**: no straight-line relationship.
- The further from 0, the more tightly the points follow a line.

`cor()` computes it:

```{r}
#| label: cor-dd
cor(sessions$deaths, sessions$difficulty)
```

r = `r round(r_dd, 2)`: a clear positive relationship with a lot of scatter, which
is what @fig-deaths-difficulty showed.

**Context decides what counts as strong.** There is no universal cut-off. Between
two measurements of the same thing, such as two raters scoring the same session,
0.6 would be worryingly low. Between a single rating and a match outcome shaped by
teammates, opponents and luck, 0.5 is large. Say what the number means for the
question, not just whether it is "strong".

### Pearson or Spearman?

The correlation above is **Pearson's**, the default. It measures how well the
points follow a straight line, and a few extreme values can pull it around.

**Spearman's correlation** ranks each variable first, then correlates the ranks.
It asks a simpler question: *when one goes up, does the other tend to go up?* It
does not need a straight line, and an outlier is just the highest rank, so it
cannot drag the result.

```{r}
#| label: spearman
cor(sessions$deaths, sessions$difficulty, method = "spearman")
```

`r round(rs_dd, 2)`, almost the same as Pearson's here, because the relationship is
roughly straight with no wild values.

**For now,** use this rule:

| Use | When |
| :- | :- |
| **Pearson** | Both variables are measurements or counts, the scatterplot looks roughly straight, and nothing sits far from the rest |
| **Spearman** | Either variable is a rating (ordinal), the relationship curves but keeps going one way, or a few extreme values sit apart |

Playtest data is full of ratings, so Spearman is often the safer choice. If the two
disagree a lot, draw the scatterplot: something is bending or pulling the line.

---

## Many Relationships at Once: the Correlation Matrix

A playtest records many measures, and the first useful question is often *which of
them go together?* A **correlation matrix** answers it in one table: the
correlation of every measure with every other.

Pick **3 to 5 measures that matter to your question**, not every column you have.
For *Tidebreak*'s navigation and difficulty questions:

```{r}
#| label: cor-matrix
sessions |>
  select(deaths, wrong_turns, difficulty, enjoyment, session_minutes) |>
  cor(method = "spearman") |>
  round(2)
```

`select()` keeps the five columns, `cor()` of a whole table correlates every pair,
and `round(2)` makes it readable. Spearman, because `difficulty` and `enjoyment`
are 1–7 ratings.

**How to read it:**

- **The diagonal is all 1s.** Every measure correlates perfectly with itself.
  Ignore it.
- **It is symmetric.** The value in row `deaths`, column `difficulty` is the same
  as row `difficulty`, column `deaths`. Read one triangle.
- **Read one row at a time** and look for the values furthest from 0.

**What this one says:**

- `deaths` and `difficulty` go together most strongly, `r round(cm["deaths", "difficulty"], 2)`:
  players who die more rate the game harder. Telemetry and self-report agree.
- `difficulty` and `enjoyment` move in opposite directions,
  `r round(cm["difficulty", "enjoyment"], 2)`: players who found it harder
  enjoyed it less.
- `wrong_turns` is close to 0 with everything, between
  `r round(min(cm["wrong_turns", setdiff(metrics, "wrong_turns")]), 2)` and
  `r round(max(cm["wrong_turns", setdiff(metrics, "wrong_turns")]), 2)`. Getting
  lost does not go with dying, with rating the game hard, or with enjoying it less.

That last line is the interesting one, and the next section explains why.

**Draw before you trust the matrix.** Each number stands in for a scatterplot you
have not drawn. Base R's `pairs()` draws them all at once:

```{r}
#| label: fig-pairs
#| fig-cap: "Every pair of the five measures as a scatterplot. Deaths against difficulty is the clearest pattern; wrong turns show none against anything."
#| fig-alt: "A five by five grid of small scatterplots for deaths, wrong turns, difficulty, enjoyment and session minutes. Most panels are shapeless clouds of whole-number points; deaths against difficulty shows a rising band."
#| fig-height: 6
pairs(sessions[, c("deaths", "wrong_turns", "difficulty", "enjoyment",
                   "session_minutes")],
      pch = 19, cex = 0.4, col = adjustcolor(pal$line, alpha.f = 0.4))
```

**You do not need to memorise this:** `pairs()` is a quick check, not a figure for
the report.

**For your project:** choose the 3–5 measures your design question depends on,
build their matrix, and read it row by row. The strongest relationship is a lead
to look at, and the one you expected but did not find is often the finding.

---

## Telemetry, Observation and Self-Report Together

The three sources of evidence from
[Planning a Playtest](01-planning-a-playtest.qmd#choose-your-measures) answer
different questions, and relating them is where correlation earns its keep.

**When they agree,** the finding is stronger. Deaths (telemetry) and the difficulty
rating (self-report) go together: what happened and what players felt point the
same way.

**When they disagree,** there is something to find. Build B cut wrong turns from
`r round(wt_by_build[["A"]], 1)` to `r round(wt_by_build[["B"]], 1)` per session on
average. So do players who got lost more rate the level as harder to navigate? The
survey item "the level was easy to navigate" (`level_navigable`) is in a separate
file, so join it on first:

```{r}
#| label: wt-nav
joined <- sessions |>
  left_join(survey, by = "session_id")

cor(joined$wrong_turns, joined$level_navigable, method = "spearman")

joined |>
  group_by(build) |>
  summarise(mean_wrong_turns = mean(wrong_turns),
            mean_navigable   = mean(level_navigable))
```

`left_join()` adds the survey columns to each session, matched on `session_id`.

The correlation is `r round(r_wt_nav, 2)`: no relationship. And the builds rated
navigation almost identically, `r round(nav_by_build[["A"]], 2)` against
`r round(nav_by_build[["B"]], 2)`, even though Build A's players got lost far more.
**Players did not notice that they were lost.** The telemetry saw a navigation
problem the self-report did not.

That is a finding, and it changes what you do with each source. A navigation
survey on its own would have said the level was fine. Wrong turns on their own say
what happened but not whether players minded. **Observation** is what would
connect them: watching where players hesitate or backtrack, and whether they seem
to know it.

**In a report you would write:** "Build A players made more wrong turns, but rated
the level as no harder to navigate (Spearman r = `r round(r_wt_nav, 2)` between
wrong turns and the navigation rating). Players appear not to notice when they are
lost, so self-reported navigation is not a reliable check on this level."

---

## What Correlation Cannot See

A single number cannot describe a picture. In 1973 the statistician Francis
Anscombe published four small datasets that make the point. They have almost
identical means, SDs and correlations, and they look nothing alike:

```{r}
#| label: fig-anscombe
#| code-fold: true
#| code-summary: "Show the code"
#| fig-cap: "Anscombe's quartet: four datasets with the same correlation, about 0.82, and four different stories."
#| fig-alt: "Four scatterplots, each with a straight fitted line of the same slope. Set 1 is a loose linear cloud. Set 2 is a smooth curve. Set 3 is a perfect line with one outlier pulling the fit. Set 4 is a vertical stack of points with one far-right point creating the whole relationship."
#| fig-height: 4.2
quartet <- do.call(rbind, lapply(1:4, function(i) {
  data.frame(set = paste("Set", i),
             x = anscombe[[paste0("x", i)]], y = anscombe[[paste0("y", i)]])
}))

ggplot(quartet, aes(x, y)) +
  geom_point(colour = pal$line, size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = pal$highlight, linewidth = 0.8) +
  facet_wrap(~ set, nrow = 1) +
  labs(x = NULL, y = NULL)
```

Only Set 1 is what r = 0.82 suggests. Set 2 is a curve, which r describes badly.
In Set 3, one outlier drags the line away from an otherwise perfect fit. In Set 4,
a single point creates the entire relationship.

**Always draw the scatterplot before quoting r** ([rule 4](ten-rules.qmd)).
Correlation sees only straight lines, is moved by single extreme points, and, as
the next two sections show, cannot see groups.

---

## Correlation Is Not Causation

When two variables move together, there are always several explanations, and
"one causes the other" is only one of them:

| Explanation | Awareness and score |
| :- | :- |
| **A causes B** | Being aware of the map makes players score more |
| **B causes A** | Players who are scoring well have more time to look around |
| **Something else causes both** | Experienced players are both more aware and better at everything |
| **Chance** | In a small sample, unrelated measures can line up by luck |

A playtest observes; it does not intervene. Only a test in which you *change* one
variable and watch the other, such as giving half the players a better minimap at
random, can separate the first explanation from the rest. Until then, write "is
associated with", never "causes" or "drives".

---

## Confounding: Look Within Groups

A **confounding variable** is a third variable that drives both of the variables
you are comparing, creating or hiding a relationship between them.

```{mermaid}
%%| label: fig-confounder
%%| fig-cap: "A confounder drives both variables. The relationship you see between them may come partly or entirely from it."
flowchart TD
  E["Experience<br/>(cohort)"] --> A["Awareness"]
  E --> S["Score"]
  A -.->|"what you observe"| S
  class E accent
```

**Which abilities go with score?**

```{r}
#| label: cor-abilities
sessions |>
  select(accuracy, mobility, survivability, objective_control, support,
         awareness, shots_fired) |>
  cor(sessions$score) |>
  round(2)
```

Awareness is the strongest, at `r round(cors[["awareness"]], 2)`. Shots fired, the
statistic everyone assumes matters, is `r round(cors[["shots_fired"]], 2)`: no
relationship at all.

**Experience is a candidate confounder for awareness and score.** Veterans average
`r aw_cohort[["Veteran"]]` awareness against `r aw_cohort[["Novice"]]` for Novices,
and a median score of `r sc_cohort[["Veteran"]]` against `r sc_cohort[["Novice"]]`.
Some of the awareness–score relationship is experience. Check within each cohort:

```{r}
#| label: within-cohort
sessions |>
  group_by(cohort) |>
  summarise(n = n(), r = cor(awareness, score))
```

From `r round(min(r_within_cohort), 2)` to `r round(max(r_within_cohort), 2)`
within each cohort: experience is part of the story, not all of it. That is how you
check a suspected confounder: **split by it, and see whether the relationship
holds inside each group.** It is `group_by()` and `summarise()` again, with `cor()`
as the summary.

### When the relationship reverses

Sometimes splitting by the confounder does not just weaken a relationship; it
reverses it. That is **Simpson's paradox**, and *Tidebreak* has a clear case. You do
not need to be able to produce this figure; it is here because it is the most
memorable demonstration of confounding there is.

Across all players, mobility and survivability have a correlation of
`r round(r_ms, 2)`: faster players are more fragile. Within each archetype:

```{r}
#| label: fig-simpson
#| code-fold: true
#| code-summary: "Show the code"
#| fig-cap: "Across all players (left) faster means more fragile. Within every archetype (right) the trend rises: the falling line is the gap between archetypes."
#| fig-alt: "Two scatterplots of survivability against mobility. Left: one black falling trend line through all the points. Right: the points form four separate clusters by archetype, Scout at high mobility and low survivability, Sentinel the reverse, and within each cluster the fitted line rises."
#| fig-height: 3.8
both <- rbind(
  data.frame(sessions[, c("mobility", "survivability", "archetype")], panel = "All players"),
  data.frame(sessions[, c("mobility", "survivability", "archetype")], panel = "By archetype")
)
ggplot(both, aes(mobility, survivability)) +
  geom_point(aes(colour = ifelse(panel == "All players", "All", archetype)), alpha = 0.7) +
  geom_smooth(data = subset(both, panel == "All players"), method = "lm",
              se = FALSE, colour = "black") +
  geom_smooth(data = subset(both, panel == "By archetype"),
              aes(colour = archetype), method = "lm", se = FALSE) +
  scale_colour_manual(values = c(All = "grey60", Duelist = "#3a5f8a",
                                 Engineer = "#6c9a3c", Scout = "#b22222",
                                 Sentinel = "#8a5a9c"),
                      breaks = c("Duelist", "Engineer", "Scout", "Sentinel")) +
  facet_wrap(~ panel) +
  labs(x = "Mobility rating", y = "Survivability rating", colour = "Archetype")
```

```{r}
#| label: within-archetype
sessions |>
  group_by(archetype) |>
  summarise(r = cor(mobility, survivability))
```

Within every archetype the correlation is **positive**, from
`r round(min(r_ms_ar), 2)` to `r round(max(r_ms_ar), 2)`. The overall falling trend
is the gap *between* archetypes: Scouts are built fast and fragile, Sentinels slow
and tough. Within any one archetype, better players are better at both.
**Archetype is the confounder**, and ignoring it reverses the conclusion. A patch
that "makes fast players more fragile" would punish exactly the players who are
playing their role well.

**For your project:** before you report any relationship, split it by the one or
two groupings most likely to matter in your game: build, level, character or
experience.

---

## Two Categorical Variables {.when-needed}

When both variables are categories, correlation is the wrong tool. Count every
combination, then turn each group's counts into shares, so that groups of
different sizes compare fairly. Do more experienced players quit in frustration
less?

```{r}
#| label: crosstab
sessions |>
  count(cohort, quit_reason) |>
  group_by(cohort) |>
  mutate(share = round(n / sum(n), 2))
```

```{r}
#| label: fig-crosstab
#| fig-cap: "Regulars quit in frustration most often; Veterans more often simply lost interest. Each bar is one cohort, scaled to 100%."
#| fig-alt: "Three stacked bars, one per cohort, each split into completed test, frustrated, lost interest and technical fault. Regulars have the largest frustrated segment; Veterans the largest lost-interest segment."
#| fig-height: 3.5
ggplot(sessions, aes(x = cohort, fill = quit_reason)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = function(x) paste0(x * 100, "%")) +
  scale_fill_manual(values = c("grey80", "#b22222", "#9bb0c9", "grey40")) +
  labs(x = "Cohort", y = "Share of sessions", fill = "How the session ended")
```

`position = "fill"` scales every bar to 100%. Not in a straight line: **Regulars**
quit in frustration most often, `r pct_fmt(ct_cq["Regular", "Frustrated"])` of
their sessions, against `r pct_fmt(ct_cq["Novice", "Frustrated"])` for Novices and
`r pct_fmt(ct_cq["Veteran", "Frustrated"])` for Veterans.

---

## Critiquing a Playtest

Every one of these ideas comes together when a conclusion is drawn from data that
cannot support it. Two scenarios to take apart.

1. *A team adds an optional minimap. Players who turned it on scored 22% higher
   than players who left it off. The team makes the minimap mandatory.*
2. *The new build was tested on Friday evening with fans recruited from the studio
   Discord; the old build on Monday morning with first-year students. Enjoyment was
   1.5 points higher for the new build, p < 0.001. The team reports that the new
   build is more enjoyable.*

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
1. Players chose whether to use the minimap, so the two groups differ in more than
   the minimap. Experienced, engaged players are more likely to explore the options
   menu *and* more likely to score well: a confounder. The 22% may be entirely
   skill. To know what the minimap does, assign it at random, or have every player
   try both, counterbalanced
   ([Planning a Playtest](01-planning-a-playtest.qmd#random-assignment)).
2. Build is confounded with everything else that differed: the players (committed
   fans against strangers), the time of day, and probably the room and who ran the
   session. Any of them could produce the gap. The p-value says the gap is unlikely
   to be chance; it says nothing about *which* of the differences caused it. The
   fix is to recruit one group and assign builds at random within it, at the same
   times, in the same room.
:::

---

## Worked Example

**Should we buff awareness?** The balance designer has seen that awareness is the
strongest predictor of score and proposes raising every archetype's base awareness.

**Step 1: look.** A scatterplot of score against awareness shows a clear rising
relationship, with a cluster of sessions far below the rest: the Sentinel
collapses from [Describing Variables](04-describing-variables.qmd#worked-example).

**Step 2: measure.** r = `r round(cors[["awareness"]], 2)`, the strongest of the
ability ratings. Shots fired, the obvious candidate, is
`r round(cors[["shots_fired"]], 2)`.

**Step 3: ask what else could explain it.** Experience: Veterans have more of both.
Within each cohort the relationship weakens but survives
(`r round(min(r_within_cohort), 2)` to `r round(max(r_within_cohort), 2)`), so
experience is part of the story. And a player's awareness *rating* may measure
skill as much as cause it: good players notice more because they are good.

**Step 4: conclude.** The data shows an association, and a fairly strong one. It
does not show that raising awareness would raise score.

**What should the team do?** Run a test, not a patch: give half the players a
feature that raises awareness, such as a clearer minimap or audio cues, at random,
and compare score between the groups. That turns the relationship into a Stage 2
comparison, which can show cause.

---

## Interpretation Guidance

**What the data shows.** Score rises with awareness, r =
`r round(cors[["awareness"]], 2)`, across `r nrow(sessions)` sessions, and the
relationship survives within each cohort.

**What the analysis suggests.** Awareness, or whatever awareness ratings reflect,
matters to how a match goes. Experience explains part of the relationship but not
all of it.

**What can legitimately be concluded.** An association, not a cause. A design
change aimed at awareness is worth testing; it is not yet supported as a fix.

**Use the right verbs.** "Is associated with", "goes with", "predicts" for
relationships in observed data. "Causes", "drives", "improves" only after an
experiment that changed one variable on purpose.

**Going further:** fitting the line itself, with slope, R² and residuals, and
reading what the residuals tell you, is on
[Regression](../going-further/regression.qmd).

---

## Common Mistakes

**Quoting r without the scatterplot.** Anscombe's four datasets share a correlation
and nothing else ([rule 4](ten-rules.qmd)).

**Reading correlation as cause.** Two things moving together has at least four
explanations. Name the alternatives before choosing one.

**Ignoring groups.** A relationship across groups can vanish or reverse within
them. Split by the obvious confounders, such as archetype, cohort or build.

**Calling r "strong" or "weak" without context.** Say what it means for the
question.

**Putting every column in the matrix.** A 20 × 20 matrix has 190 correlations and
some will look strong by luck. Choose the 3–5 that answer your question.

**Using Pearson on ratings without checking.** For 1–7 ratings, use Spearman, or at
least check that the two agree.

**Using correlation on categories.** Two categorical variables need counts and
shares, not `cor()`.

---

## Practice

Each takes two or three minutes.

1. **Predict.** Before computing it: is the correlation between `kills` and
   `score` closer to 0, 0.5 or 0.9? Then check.
2. **Compute.** Build a Spearman correlation matrix of `kills`, `deaths`,
   `time_alive_s` and `score`. Which measure goes with score most strongly?
3. **Critique.** "Wrong turns correlate at
   `r round(cor(sessions$wrong_turns, sessions$enjoyment), 2)` with enjoyment, so
   getting lost does not affect how much players enjoy the game." What is wrong
   with that conclusion?
4. **Decide.** The correlation between deaths and difficulty is about the same in
   both builds. Check it with `group_by()`. Why is that worth knowing before you
   use deaths as a stand-in for "how hard players found it"?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
**1.**

```{r}
#| label: practice-kills
cor(sessions$kills, sessions$score)
```

Close to 0, `r round(cor(sessions$kills, sessions$score), 2)`, which surprises
most people. A number you were sure about is exactly the one to check.

**2.**

```{r}
#| label: practice-matrix
sessions |>
  select(kills, deaths, time_alive_s, score) |>
  cor(method = "spearman") |>
  round(2)
```

```{r}
#| label: practice-matrix-check
#| include: false
pm <- cor(sessions[, c("kills", "deaths", "time_alive_s", "score")], method = "spearman")
stopifnot(abs(pm["deaths", "score"]) > abs(pm["time_alive_s", "score"]),
          abs(pm["time_alive_s", "score"]) > abs(pm["kills", "score"]))
```

Read the `score` row, and look at the size, not the sign. Deaths go with score
most strongly, `r round(pm["deaths", "score"], 2)`: negative, so fewer deaths, more
score. Time alive is next, `r round(pm["time_alive_s", "score"], 2)`. Kills barely
register, `r round(pm["kills", "score"], 2)`. In *Tidebreak*, staying alive is what
scores, not killing.

**3.** A correlation near zero means no *straight-line* relationship across all
sessions. It does not rule out a curve, a relationship within one build or cohort
that is cancelled out by another, or an effect on a different outcome: players who
got lost may have quit. Draw the scatterplot, split by build, and look at quit
reason too before concluding anything.

**4.**

```{r}
#| label: practice-by-build
sessions |>
  group_by(build) |>
  summarise(r = cor(deaths, difficulty, method = "spearman"))
```

The relationship holds in each build, so it is not produced by mixing two
different builds together. That makes deaths a more trustworthy stand-in for felt
difficulty in either build, although it is still only a stand-in: plenty of
players died often and rated the game easy.
:::

---

## Summary

- A scatterplot shows a relationship's direction, form and strength. Draw it first.
- Correlation, from −1 to +1, measures how two things move together. Pearson for
  straight-line relationships between measurements; Spearman for ratings, curves
  that keep going one way, or outliers.
- A correlation matrix of 3–5 chosen measures shows which go together. Read one row
  at a time, and draw the pairs before trusting it.
- Telemetry and self-report can disagree: *Tidebreak*'s players got lost without
  rating the level as harder to navigate. The disagreement is a finding.
- Correlation is not causation: the reverse direction, a third variable or chance
  can produce the same pattern.
- A confounder drives both variables. Check by splitting into groups; Simpson's
  paradox is when the relationship reverses inside them.

| Key term | Meaning |
| :- | :- |
| **Scatterplot** | One point per observation, for two numeric variables |
| **Correlation (r)** | −1 to +1: how strongly two variables move together |
| **Pearson** | The default correlation: how well the points follow a straight line |
| **Spearman** | Correlation of ranks: does one go up when the other does? |
| **Correlation matrix** | The correlation of every chosen measure with every other |
| **Causation** | One variable changing another; needs an experiment to show |
| **Confounding variable** | A third variable that drives both variables you are comparing |
| **Simpson's paradox** | A relationship that reverses when the data is split into groups |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 4 (look at the shape before you quote the
  average) and rule 7 (compare distributions, not averages)
- [Regression](../going-further/regression.qmd): the line, its slope, R² and
  residuals
- [Statistics Reference: relationships](../reference/reference-statistics.qmd#relationships)
- [ggplot2 Reference](../reference/reference-ggplot2.qmd)
- [Lab 04: Relationships in Your Data](../labs/clinic-relationships.qmd), the
  Stage 3B clinic
- [Stage 3 demo](../demos/stage-3.qmd), Demo 2
- Next, in Stage 4: [Interpreting and Reporting Evidence](11-reporting-evidence.qmd)

---

## Reflective Questions

- Name a relationship in your own game's telemetry that you suspect is confounded.
  What is the confounder, and how would you check?
- *Tidebreak*'s players got lost without noticing. Which of your own game's
  self-report questions might have the same problem, and what telemetry would you
  check it against?
- Simpson's paradox reversed the mobility–survivability relationship. What other
  grouping in a game could hide or reverse a relationship in the same way?
- Your producer wants a single sentence for the balance meeting about awareness.
  Write it, using only what the data can support.
