---
title: "Interpreting and Reporting Evidence"
subtitle: "From a set of results to a finding someone can act on"
body-classes: "wf-at-interpret wf-at-report"
aliases:
  - reporting-evidence.html
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")
library(dplyr)
library(tidyr)

long   <- pivot_longer(survey, -session_id, names_to = "item", values_to = "rating")
joined <- left_join(long, select(sessions, session_id, build, quit_reason),
                    by = "session_id")

items <- long |>
  group_by(item) |>
  summarise(mean = mean(rating), agree = mean(rating >= 4),
            disagree = mean(rating <= 2)) |>
  arrange(mean)

by_build <- joined |>
  group_by(item, build) |>
  summarise(mean = mean(rating), .groups = "drop") |>
  pivot_wider(names_from = build, values_from = mean)

obj    <- joined[joined$item == "objective_clear", ]
obj_a  <- mean(obj$rating[obj$build == "A"])
obj_b  <- mean(obj$rating[obj$build == "B"])
obj_fr <- mean(obj$rating[obj$quit_reason == "Frustrated"])
obj_nf <- mean(obj$rating[obj$quit_reason != "Frustrated"])
item_row <- function(i) items[items$item == i, ]

themed <- sessions |>
  mutate(theme = case_when(
    grepl("lost|circles|meant to go|objective", comment) ~ "Lost or unclear",
    grepl("route|exit|signposting", comment)             ~ "Found the way",
    grepl("Died|hard", comment)                          ~ "Difficulty",
    TRUE                                                 ~ "General"
  ))
lost_a  <- sum(themed$theme == "Lost or unclear" & themed$build == "A")
lost_b  <- sum(themed$theme == "Lost or unclear" & themed$build == "B")
found_a <- sum(themed$theme == "Found the way" & themed$build == "A")
found_b <- sum(themed$theme == "Found the way" & themed$build == "B")
helped_a <- sum(sessions$comment == "The signposting helped." & sessions$build == "A")

wt <- t.test(wrong_turns ~ build, data = sessions)

stopifnot(items$item[1] == "objective_clear",
          items$item[nrow(items)] == "visual_style",
          obj_b > obj_a, obj_fr < obj_nf,
          lost_b > lost_a, found_b < found_a, helped_a > 0,
          all(themed$enjoyment[themed$theme == "Found the way"] >= 5),
          all(themed$enjoyment[themed$theme == "Lost or unclear"] < 5),
          wt$estimate[[1]] > wt$estimate[[2]],
          by_build$item[which.max(abs(by_build$B - by_build$A))] == "objective_clear")

fr_check <- table(themed$theme[themed$quit_reason == "Frustrated"])
stopifnot(names(which.max(fr_check)) == "Difficulty")

# For the annotated report and the severity table
tt_enj  <- t.test(enjoyment ~ build, data = sessions)
frus_n  <- tapply(sessions$quit_reason == "Frustrated", sessions$build, sum)
pt_frus <- prop.test(frus_n, table(sessions$build))
b_ms    <- frames$ms[frames$build == "B"]
slow_b  <- frames$build == "B" & frames$ms > 1000 / 30
n_players <- length(unique(sessions$player_id))
stopifnot(max(abs(tt_enj$conf.int)) < 0.5, pt_frus$p.value < 0.01,
          all(frames$zone[slow_b] == "Corridor"))
```

> A playtest ends in a decision, and the report is the only part of your work the
> decision-maker will read. This page is about turning results into that report:
> the nine-part structure, reshaping a survey so every item can be reported at
> once, using players' comments honestly, writing each finding as **What → Pattern
> → Meaning → Action**, and ranking findings so the most important one leads. The
> playtest question running through it: *"Build B was meant to be better. What do
> we tell the team?"*

::: {.at-a-glance}
**Time:** 30 minutes in class; the R demonstrations support Lab 03 Part B.
**Before this:** [Relationships, Correlation and Regression](10-relationships-and-regression.qmd).
**Key terms:** nine-part report, wide, long, `pivot_longer()`, theme, What →
Pattern → Meaning → Action, severity, limitation.
**R introduced:** `pivot_longer()`, `pivot_wider()`, `left_join()` (from Week 2),
`grepl()`, `case_when()`, `knitr::inline_expr()`.
:::

## Learning Outcomes

By the end of these notes you can:

- lay out a playtest report in nine parts, and say what question each part answers;
- explain why survey exports arrive wide, and what makes a table tidy;
- reshape a whole survey with `pivot_longer()` and summarise or plot every item in
  one step;
- categorise free-text comments into a few themes, and say what they can and cannot
  support;
- write a finding as What → Pattern → Meaning → Action, and diagnose a finding that
  skips a step;
- rank findings by severity, frequency and cost, so the report leads with the one
  that matters most;
- explain why a Quarto report cannot quote a stale number.

---

## Why This Matters for Playtesting

A playtest ends in a decision: keep the change, cut it, or test again. The person
making that decision will not rerun your analysis. They will read the report, and
the report is the only part of your work that reaches them.

**A report is an argument, not a results dump.** Every table and figure you include
should move the reader towards the conclusion or warn them off it. A reader faced
with forty numbers will pick the one they already believed.

**The reader must be able to check you.** A report that says what was measured, how
the data was cleaned, what was tested and what was not, lets a sceptical producer
find the weak point themselves. That is what makes the conclusion trustworthy, and
it is why the limitations section is not optional.

---

## The Nine-Part Report

Every playtest report in this module uses the same nine parts, in this order. The
order is the analytical workflow: question, then data, then evidence, then what it
means.

| Part | Answers | In the *Tidebreak* report |
| :- | :- | :- |
| 1. Research question | What did we want to know? | Did Build B's guidance improve navigation without reducing enjoyment? |
| 2. Experimental design | Who played what, and how was the data cleaned? | `r nrow(sessions)` sessions, each player on one build; the cleaning log |
| 3. Descriptive evidence | What are the numbers? | Wrong turns per session: mean `r round(wt$estimate[[1]], 1)` in A, `r round(wt$estimate[[2]], 1)` in B |
| 4. Visual evidence | What does it look like? | The faceted histogram of wrong turns |
| 5. Statistical evidence | Could it be chance, and how big could it be? | The t-test and its confidence interval |
| 6. Practical significance | Is it big enough to matter? | Is the reduction in wrong turns worth the build's cost? |
| 7. Qualitative evidence | What did players say? | The comment themes |
| 8. Limitations | What could make this wrong? | Sample, design, confounds, what was not measured |
| 9. Design conclusion | What do we do next? | A decision and the next test |

**Parts 3 to 7 are evidence; parts 8 and 9 are judgement.** Keep them separate. A
reader should be able to disagree with your conclusion while agreeing with every
number above it.

**Part 8 is where confounds go.** Anything from
[Relationships, Correlation and Regression](10-relationships-and-regression.qmd#confounding-and-simpsons-paradox)
that you could not rule out belongs here, stated plainly. A limitation you name costs you less than one the reader finds.

---

## Wide, Long and Tidy

A survey tool exports one row per respondent and one column per question, because
that is how a form is filled in. That shape is **wide**. The *Tidebreak* survey has
`r nrow(survey)` rows and `r ncol(survey)` columns: a session ID and eight
agreement items, each rated 1 to 5.

**Tidy data**, defined in [Data, Variables and Types](03-data-and-types.qmd#tidy-data),
has one row per observation and one column per variable. In the survey, an
observation is one player's answer to one item. So the tidy shape has three
columns, `session_id`, `item` and `rating`, and one row per answer. That shape is
**long**: `r nrow(survey)` sessions × 8 items = `r nrow(long)` rows.

**Why long matters.** In the wide table, the item is a column *name*, and you cannot
group by, filter on or plot a column name. In the long table, the item is a column
*value*, so `group_by(item)` summarises every item at once, `filter(item == ...)`
picks one, and `ggplot2` can put every item on one axis. Eight separate calculations
become one.

**`pivot_longer()`** is the reshape. It takes the columns you name, stacks them, and
writes their old names into one new column and their values into another. It is the
last stage of the R cycle, **Tidy**, and it is usually needed exactly when you want
to report a whole questionnaire.

---

## Qualitative Evidence

The `comment` column holds one sentence per session. **Categorising** comments means
reading them, deciding on a handful of **themes**, and assigning each comment to
one. It is a coding decision, like a cleaning decision, and it goes in the report.

**Comments explain; they do not measure.** A comment tells you what a player
noticed and how they felt. It does not tell you how often something happened, and
it can contradict the telemetry from the same session. When it does, the telemetry
says what happened and the comment says how it felt. Both are true, and a report
that reports only one of them has thrown evidence away.

**Quote a comment to show a pattern, never to be the pattern.** One vivid quote
beside a chart of frustration quits makes the chart human. One vivid quote instead
of the chart is an anecdote.

---

## What → Pattern → Meaning → Action

Every finding in the report is written in four moves:

| Move | Says | Test |
| :- | :- | :- |
| **What** | What was measured, on whom, how many | Could a reader repeat the measurement? |
| **Pattern** | What the data shows, with its size | Is there a number, and is it the right one? |
| **Meaning** | What that suggests about the game, and how sure you are | Does it follow from the pattern, and no further? |
| **Action** | The one change or test you recommend | Could someone start work on it tomorrow? |

```{mermaid}
%%| label: fig-wpma
%%| fig-cap: "What → Pattern → Meaning → Action. Facts on the left, judgement on the right; the action is usually the next test, which starts the cycle again."
flowchart LR
  W["What<br/>measured, on whom,<br/>how many"] --> P["Pattern<br/>what the data shows,<br/>with its size"]
  P --> M["Meaning<br/>what it suggests,<br/>and how sure"]
  M --> A["Action<br/>one change<br/>or test"]
  A -.->|"the next playtest"| W
  classDef fact fill:#f8f9fa,stroke:#6c757d,color:#212529
  classDef judge fill:#fbeaea,stroke:#b22222,color:#212529
  class W,P fact
  class M,A judge
```

**Meaning is where findings go wrong.** What and Pattern are facts. Action is a
recommendation. Meaning is the step where you connect them, and it is the step most
often skipped, overstated or quietly replaced by what you hoped to find. A finding
that does not name an action is not finished ([rule 10](ten-rules.qmd)); a finding
whose action does not follow from its meaning is worse.

---

## Ranking Findings

A playtest rarely produces one finding. *Tidebreak* produced at least five, and a
report that lists them in the order they were found buries the one that matters.
**Rank them, and lead with the top one.**

Rate each finding on three things:

- **Severity**: how badly it hurts the player experience when it happens.
- **Frequency**: how many players it affects.
- **Confidence**: how strong the evidence is.

A common severity scale, adapted from usability testing, runs from 4 down to 0:

| Severity | Meaning |
| :-: | :- |
| 4 Critical | Stops play or makes players quit |
| 3 Major | Hurts the experience for many players, but they carry on |
| 2 Minor | A noticeable irritation |
| 1 Cosmetic | Fix if there is time |
| 0 None | Not a problem |

Applied to *Tidebreak*:

| Finding | Severity | Frequency | Evidence | Priority |
| :- | :-: | :- | :- | :- |
| Build A frustration quits | 4 | `r frus_n[["A"]]` of `r sum(sessions$build == "A")` Build A sessions, against `r frus_n[["B"]]` in Build B | Strong: `r p_fmt(pt_frus$p.value)`, and Build B already fixes it | Lead the report |
| Build B Corridor hitch | 3 | `r sum(slow_b)` frames over the 30 FPS budget, every one in the Corridor | Strong: a fixed interval, reproducible | Engineering ticket now |
| Objective unclear | 3 | `r round(100 * item_row("objective_clear")$disagree)`% of responses disagreed; better in Build B | Strong for the build difference; suggestive for the link to quits | Design change and retest |
| Build B rated harder | 1 | `r round(abs(diff(t.test(difficulty ~ build, data = sessions)$estimate)), 2)` points on a 1–7 scale | Weak: deaths did not change | Monitor in the next test |
| Visual style | 0 | `r round(100 * item_row("visual_style")$agree)`% agreed | Strong, but nothing to fix | Mention, do not lead with it |

**Severity is a judgement; write down why.** Two analysts may rate the Corridor hitch
differently. What matters is that the reasons are visible, so the team can argue
about the rating rather than about which finding the analyst happened to like.

---

## Worked Examples

**The data.** The same `r nrow(survey)` surveys, reshaped long and joined to the
session outcomes, feed both examples below. Every number in both is correct.

### A good finding

> **What.** `r nrow(survey)` players each rated eight statements about their session
> from 1 (strongly disagree) to 5 (strongly agree).
>
> **Pattern.** "I always knew what my objective was" was the lowest-rated of the
> eight, averaging `r round(item_row("objective_clear")$mean, 2)`, with
> `r round(100 * item_row("objective_clear")$disagree)`% disagreeing. It split
> sharply by build: `r round(obj_a, 2)` in Build A against `r round(obj_b, 2)` in
> Build B. Players who quit in frustration rated it `r round(obj_fr, 2)`, against
> `r round(obj_nf, 2)` for everyone else.
>
> **Meaning.** Build B's guidance made the objective clearer. The playtest was
> designed to compare the two builds, so this is the comparison it can best support.
> The link to frustration quits is an association: players who did not know what to
> do were the ones who left, which is consistent with unclear objectives causing
> the quits but does not prove it.
>
> **Action.** Keep Build B's guidance. For the next build, add an explicit on-screen
> objective in the first minute, and retest with objective clarity and the
> frustration-quit rate as the two target measures, with the success threshold
> agreed before the test.

**Why it works.** The What says who and how many. The Pattern gives the size and
the comparison, not only the ranking. The Meaning separates the claim the design
supports (the build changed clarity) from the one it only suggests (clarity drives
quits). The Action is specific, testable, and follows from the Meaning.

### A bad finding

> Players loved *Tidebreak*: visual style scored
> `r round(item_row("visual_style")$mean, 2)` out of 5, with
> `r round(100 * item_row("visual_style")$agree)`% agreeing. Build B scored higher
> than Build A on `r sum(by_build$B > by_build$A)` of the 8 survey items. Overall the
> playtest was a success, and the team should continue as planned.

**Why it fails.** Every number is true, and the finding is still useless.

- **What is missing.** Who answered, how many, and on what scale.
- **The Pattern is chosen, not found.** It leads with the best-rated item, which
  differs between the builds by
  `r sprintf("%.2f", abs(by_build$B[by_build$item == "visual_style"] - by_build$A[by_build$item == "visual_style"]))`
  of a point. "Higher on `r sum(by_build$B > by_build$A)` of 8" counts that as a
  win, the same as the objective item's gap of
  `r round(obj_b - obj_a, 2)`.
- **The Meaning is a verdict.** "A success" measured against what? Nothing in the
  question was about visual style.
- **The Action names nothing.** "Continue as planned" is not a decision anyone can
  start work on ([rule 10](ten-rules.qmd)).
- **It misses the finding.** The worst-rated item, the one tied to why players
  quit, is not mentioned. The bad version is not a weaker account of the good one.
  It is an account of a different, more comfortable playtest.

---

## An Annotated Report

What a finished report looks like, in miniature. Each part is one or two sentences
here; in your report each is a section. The notes in grey say why each part is
written the way it is.

::: {.callout-note appearance="simple" icon="false"}
**1. Research question.** Did Build B's environmental guidance improve navigation
without reducing enjoyment?

[*Stated as a question with a yes or no answer, using the measures that will
answer it.*]{style="color: #6c757d"}

**2. Experimental design.** `r nrow(sessions)` sessions from `r n_players` players,
`r unique(as.vector(table(sessions$player_id)))` sessions each; each player was
assigned to one build. The raw export was cleaned as recorded in the cleaning log
(Appendix A).

[*Gives both counts, and points to the cleaning log, so the reader can reproduce
n.*]{style="color: #6c757d"}

**3. Descriptive evidence.** Build A players averaged
`r round(wt$estimate[[1]], 1)` wrong turns per session; Build B players
`r round(wt$estimate[[2]], 1)`.

[*The numbers the test will compare, before the test.*]{style="color: #6c757d"}

**4. Visual evidence.** Figure 1 shows wrong turns by build: the middle halves of
the two distributions do not overlap.

[*Every figure is referred to in the text and earns its place.*]{style="color: #6c757d"}

**5. Statistical evidence.** Build B players took `r round(diff(rev(wt$estimate)), 1)`
fewer wrong turns per session (95% CI `r ci_fmt(wt$conf.int)`,
`r p_fmt(wt$p.value)`). Enjoyment did not differ: the 95% interval for the
difference, `r ci_fmt(tt_enj$conf.int, 2)` points on a 1–7 scale, rules out any
difference larger than half a point.

[*Size first, then interval, then p-value. The "without reducing enjoyment" half of
the question is answered with an interval, not with "not significant".*]{style="color: #6c757d"}

**6. Practical significance.** A reduction of `r round(diff(rev(wt$estimate)), 1)`
wrong turns per session is `r pct_fmt(diff(rev(wt$estimate)) / wt$estimate[[1]])`
of Build A's average: large enough for players to feel.

[*Translates the statistic into the player's experience.*]{style="color: #6c757d"}

**7. Qualitative evidence.** Comments did not track the builds: more Build B players
than Build A players said they got lost. Comments tracked enjoyment instead
(Table 3).

[*Reports the evidence that disagrees, and says what it is evidence of.*]{style="color: #6c757d"}

**8. Limitations.** Each player contributed four sessions, so sessions are not
independent; the result holds when each player counts once. Players were not
asked about prior experience with the genre. The separate paired study did not
record play order, so it cannot separate guidance from practice.

[*Names the weaknesses before the reader finds them, and says which ones were
checked.*]{style="color: #6c757d"}

**9. Design conclusion.** Keep Build B's guidance. Next: add an on-screen objective
in the first minute and retest objective clarity and the frustration-quit rate, with
the success threshold agreed before the test.

[*One decision and one test, following from parts 5 to 8.*]{style="color: #6c757d"}
:::

---

## R Demonstrations

```{r}
#| label: demo-load
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
survey   <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)
dim(survey)
```

### Wide to long

```{r}
#| label: demo-pivot
long <- survey |>
  pivot_longer(-session_id, names_to = "item", values_to = "rating")

dim(long)
head(long, 3)
```

`-session_id` means "every column except `session_id`". Those eight columns are
stacked: their names go into `item` and their values into `rating`. The session ID
is repeated on each of its eight rows, so nothing is lost.

### Every item at once

```{r}
#| label: demo-items
long |>
  group_by(item) |>
  summarise(mean     = mean(rating),
            agree    = mean(rating >= 4),
            disagree = mean(rating <= 2)) |>
  arrange(mean)
```

One `group_by()` summarises all eight items. `mean(rating >= 4)` is the proportion
who agreed: TRUE counts as 1. For a rating scale, report the proportion agreeing
and disagreeing alongside the mean; the mean of an ordinal scale is a convenience,
not a measurement ([rule 3](ten-rules.qmd)).

```{r}
#| label: demo-prop
round(prop.table(table(long$item, long$rating), margin = 1), 2)
```

`table()` counts each rating for each item; `prop.table(..., margin = 1)` turns each
row into proportions that sum to 1. This is the full distribution of every item in
one table.

### Every item in one figure

```{r}
#| label: demo-plot
#| fig-height: 4
long |>
  ggplot(aes(y = reorder(item, rating), fill = factor(rating))) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "RdBu") +
  labs(title = "Post-session survey, all items",
       x = "Proportion of players", y = NULL, fill = "Rating")
```

Only possible because the data is long: `item` is a column, so it can go on an
axis. `reorder(item, rating)` sorts the items by their mean rating, and
`position = "fill"` stretches every bar to the same length so the items compare as
proportions. The diverging palette puts disagreement in red and agreement in blue.

### Joining the survey to what happened

```{r}
#| label: demo-join
long |>
  left_join(select(sessions, session_id, build), by = "session_id") |>
  filter(item == "objective_clear") |>
  group_by(build) |>
  summarise(n = n(), mean = mean(rating), disagree = mean(rating <= 2))
```

`left_join()` from Week 2 adds each session's build to all eight of its rows. The
same pipeline with `quit_reason` in place of `build` gives the frustration-quit
comparison in the good finding above.

### Categorising comments

```{r}
#| label: demo-themes
themed <- sessions |>
  mutate(theme = case_when(
    grepl("lost|circles|meant to go|objective", comment) ~ "Lost or unclear",
    grepl("route|exit|signposting", comment)             ~ "Found the way",
    grepl("Died|hard", comment)                          ~ "Difficulty",
    TRUE                                                 ~ "General"
  ))

table(themed$theme, themed$build)
```

`grepl()` returns TRUE when a comment contains any of the words, and `case_when()`
assigns the first theme that matches. Read every comment before writing the
patterns; the themes are your decision, and a reader should see them.

The table contradicts the telemetry. Build B players took fewer wrong turns, yet
`r lost_b` of them left a "lost or unclear" comment against `r lost_a` in Build A,
and `r helped_a` Build A players wrote that "the signposting helped" about the
build without the new signage.

```{r}
#| label: demo-themes-enjoy
table(themed$theme, enjoyed = themed$enjoyment >= 5)
```

Split by enjoyment instead of build, the themes separate cleanly: every "found the
way" comment came from a player who rated enjoyment 5 or above, and every "lost or
unclear" comment from one who rated it lower. These comments record how the
session *felt*, not where it went wrong. That is a finding in its own right, and it
is why comments explain the numbers rather than replace them.

---

## Quarto: The Report Is the Analysis

A playtest report is a `.qmd` file: prose, code and output in one document, rendered
to HTML with `quarto render report.qmd`. The skeleton is the nine parts as
headings:

````markdown
---
title: "Tidebreak Build B: Playtest Report"
---

## 1. Research question

## 2. Experimental design

```{{r}}
#| echo: false
sessions <- readr::read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
```

Each of the `r knitr::inline_expr("nrow(sessions)")` sessions ...

## 3. Descriptive evidence
````

**Inline code keeps the prose honest.** `` `r knitr::inline_expr("nrow(sessions)")` `` in
the text is replaced by the number when the report renders. When next week's data
arrives, you render again and every figure, table and sentence updates together.
A number typed by hand into the prose is the one that will be wrong. Every page of
this site is built this way.

**`echo: false`** hides a chunk's code but keeps its output: the reader sees the
table or figure without the R that made it. Use it for the report; leave the code
visible in your own working file.

---

## Interpretation Guidance

**What the data shows.** The objective item is the worst-rated of eight and differs
most between the builds. The comment themes do not follow the build at all; they
follow enjoyment.

**What the analysis suggests.** Build B's guidance improved how clear the objective
felt. The comments are a record of mood, so they are weak evidence about navigation
in either direction.

**What can legitimately be concluded.** Build B made the objective clearer to the
players who tested it. It did not make players *say* they got lost less, and a
report that quoted only the comments would have reached the opposite conclusion to
the telemetry.

**Order the evidence by strength.** In part 9, lead with the conclusion the strongest
evidence supports, and name any weaker evidence that disagrees with it. A report
that hides a contradiction invites the reader to find it.

---

## Common Mistakes

**Reporting results in the order you computed them.** The reader needs the order of
the argument. Use the nine parts.

**Leading with the best-rated item.** It is usually the art, it rarely differs
between builds, and it is almost never what the playtest was run to find.

**Averaging a rating scale and stopping.** Give the proportion agreeing and
disagreeing. A mean of 3 can be everyone neutral or half the players furious.

**Counting comments as if they were measurements.** A count of complaints is a
theme, not a rate. Which comment a player writes depends on how they felt, not only
on what happened to them.

**An Action that is a sentiment.** "Improve onboarding" is a wish. "Add an on-screen
objective in the first minute and retest clarity" is an action.

**Copying numbers into the prose by hand.** They go stale the next time the data
changes, and nobody notices until a meeting.

---

## Practice

Each takes two or three minutes. Start from `long`, `sessions` and `themed` in the
demonstrations.

**1.** Which survey item has the largest share of players answering 1 or 2 in
Build A?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-disagree-a
disagree_a <- long |>
  left_join(select(sessions, session_id, build), by = "session_id") |>
  filter(build == "A") |>
  group_by(item) |>
  summarise(disagree = mean(rating <= 2)) |>
  arrange(desc(disagree))

head(disagree_a, 3)
```

The objective item: `r round(100 * disagree_a$disagree[1])`% of Build A players
answered 1 or 2, against `r round(100 * disagree_a$disagree[2])`% for the next
item. Filtering after the join keeps the whole
survey available for the next question.
:::

**2.** Without running it, how many rows would `pivot_longer()` give if the survey
had ten items instead of eight?

::: {.callout-note collapse="true" title="Check your answer"}

`r nrow(survey)` × 10 = `r nrow(survey) * 10`. One row per session per item. If a
reshape gives a row count you did not predict, check which columns you told it to
stack.
:::

**3.** Among the frustration quits only, which comment theme is most common?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-themes-frustrated
fr_themes <- table(themed$theme[themed$quit_reason == "Frustrated"])
fr_themes
```

Difficulty, with `r fr_themes[["Difficulty"]]`, ahead of lost or unclear with
`r fr_themes[["Lost or unclear"]]`. The survey says frustration quitters did not
know their objective; their comments mostly blame difficulty. Players name the
symptom they felt, not the cause. Read the counts alongside what they are counts
of: a handful of one-sentence comments, each chosen by how the session felt. They can illustrate why frustration
quits happened. They cannot tell you how often each cause occurs.
:::

**4. Rank.** A colleague's draft report leads with "93% of players liked the visual
style". Using the severity table above, rewrite the report's first sentence.

::: {.callout-note collapse="true" title="Check your answer"}
Lead with the highest-priority finding: "One in three Build A sessions
(`r frus_n[["A"]]` of `r sum(sessions$build == "A")`) ended in a frustration quit,
against `r frus_n[["B"]]` in Build B; Build B's guidance should be kept." Visual
style can appear later, as a strength that needs no action.
:::

---

## Summary

- A report has nine parts, in the order of the argument: question, design,
  evidence from descriptive to qualitative, limitations, conclusion.
- Surveys arrive wide. `pivot_longer()` makes them long, so every item can be
  grouped, filtered, joined and plotted at once. That is the Tidy stage of the R
  cycle.
- Comments explain and illustrate. They can contradict the telemetry, and when they
  do, report both.
- Write every finding as What → Pattern → Meaning → Action. Meaning is where
  findings go wrong; Action is where they become useful.
- Rank findings by severity, frequency and strength of evidence, and lead with the
  top one.
- A report built in Quarto computes its own numbers, so it cannot quote a stale
  one.

| Key term | Meaning |
| :- | :- |
| **Nine-part report** | Question, design, descriptive, visual, statistical, practical and qualitative evidence, limitations, conclusion |
| **Wide** | One row per respondent, one column per question |
| **Long** | One row per respondent per question |
| **Theme** | A category assigned to free-text comments |
| **What → Pattern → Meaning → Action** | The four moves of a finding |
| **Severity** | How badly a problem hurts the player experience, 0 to 4 |
| **Limitation** | Something that could make the conclusion wrong, stated by the author |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 2 (a number without a question attached is
  decoration), rule 3 (know what kind of number you are holding) and rule 10 (a
  finding that does not name an action is not finished)
- [Cleaning and Transforming](06-cleaning-and-transforming.qmd): the cleaning log that
  goes in part 2
- [Comparing Groups](08-comparing-groups.qmd): statistical against
  practical significance, parts 5 and 6
- [Statistics Reference: which tool for which question](../reference/reference-statistics.qmd#which-tool-for-which-question)
- [R and Quarto Reference: Quarto](../reference/r-quarto-reference.qmd#quarto)
  and [reshaping and joining](../reference/r-quarto-reference.qmd#reshaping-and-joining)
- [Lab 03: complete playtest analysis](../labs/complete-analysis.qmd#part-b-reshape-and-report),
  Part B
- [Week 3 demo](../demos/week-3.qmd)
- Back to the overview: [Start Here](00-start-here.qmd)

---

## Reflective Questions

- The bad finding uses only true numbers. What, if anything, should stop a true
  number going into a report?
- The comments say Build B players got lost more; the telemetry says they took fewer
  wrong turns. Write the one sentence you would put in part 7, and say why you
  chose that wording.
- A producer asks you to cut the limitations section because it "undermines the
  recommendation". What do you say?
- Your Action recommends a retest. What would you have to decide before that retest
  starts, so that its result can change your mind ([rule 1](ten-rules.qmd))?

---
