---
title: "Interpreting and Reporting Evidence"
subtitle: "From a set of results to a finding someone can act on"
body-classes: "wf-at-interpret wf-at-act"
aliases:
  - reporting-evidence.html
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")
library(dplyr)
library(tidyr)

long   <- pivot_longer(survey, -session_id, names_to = "item", values_to = "rating")
joined <- left_join(long, select(sessions, session_id, build, quit_reason),
                    by = "session_id")

items <- long |>
  group_by(item) |>
  summarise(mean = mean(rating), agree = mean(rating >= 4),
            disagree = mean(rating <= 2)) |>
  arrange(mean)

by_build <- joined |>
  group_by(item, build) |>
  summarise(mean = mean(rating), .groups = "drop") |>
  pivot_wider(names_from = build, values_from = mean)

obj    <- joined[joined$item == "objective_clear", ]
obj_a  <- mean(obj$rating[obj$build == "A"])
obj_b  <- mean(obj$rating[obj$build == "B"])
obj_fr <- mean(obj$rating[obj$quit_reason == "Frustrated"])
obj_nf <- mean(obj$rating[obj$quit_reason != "Frustrated"])
item_row <- function(i) items[items$item == i, ]

wt <- t.test(wrong_turns ~ build, data = sessions)

stopifnot(items$item[1] == "objective_clear",
          items$item[nrow(items)] == "visual_style",
          obj_b > obj_a, obj_fr < obj_nf,
          wt$estimate[[1]] > wt$estimate[[2]],
          by_build$item[which.max(abs(by_build$B - by_build$A))] == "objective_clear")

# For the annotated report and the ranking table
tt_enj  <- t.test(enjoyment ~ build, data = sessions)
frus_n  <- tapply(sessions$quit_reason == "Frustrated", sessions$build, sum)
pt_frus <- prop.test(frus_n, table(sessions$build))
slow_b  <- frames$build == "B" & frames$ms > 1000 / 30
n_players <- length(unique(sessions$player_id))
per_player <- unique(as.vector(table(sessions$player_id)))
stopifnot(max(abs(tt_enj$conf.int)) < 0.5, pt_frus$p.value < 0.01,
          all(frames$zone[slow_b] == "Corridor"))
```

> A playtest ends in a decision, and the report is the only part of your work the
> decision-maker will read. This page is about turning results into that report:
> what goes under each heading of the assessment report, writing each finding as
> **What → Pattern → Meaning → Action**, ranking findings so the most important
> one leads, and building the report in Quarto so its numbers cannot go stale. The
> playtest question running through it: *"Build B was meant to be better. What do
> we tell the team?"*

::: {.at-a-glance}
**Stage:** 4 · Core. Taught in the March workshop.
**Time:** 30 minutes, then the workshop lab on your own data.
**Before this:** everything from Stages 1–3, and
[Combining Evidence](12-combining-evidence.qmd) for the qualitative part.
**Key terms:** assessment report headings, Nine-Part Evidence Workflow, What →
Pattern → Meaning → Action, severity, limitation, inline R.
**R you will write:** inline `` `r ` `` code, chunk options `echo` and `fig-cap`;
`pivot_longer()` if your survey is wide.
:::

## Learning Outcomes

By the end of these notes you can:

- say what goes under each heading of the assessment report, and map each part of
  your analysis onto one;
- write a finding as What → Pattern → Meaning → Action, and diagnose a finding that
  skips a step;
- rank findings by severity, frequency, evidence and actionability, so the report
  leads with the one that matters most;
- build the report in Quarto with every number computed, so it cannot quote a
  stale number;
- reshape a wide survey with `pivot_longer()` when you need to report every item.

---

## Why This Matters for Playtesting

A playtest ends in a decision: keep the change, cut it, or test again. The person
making that decision will not rerun your analysis. They will read the report, and
the report is the only part of your work that reaches them.

**A report is an argument, not a results dump.** Every table and figure you include
should move the reader towards the conclusion or warn them off it. A reader faced
with forty numbers will pick the one they already believed.

**The reader must be able to check you.** A report that says what was measured, how
the data was cleaned, what was tested and what was not, lets a sceptical producer
find the weak point themselves. That is what makes the conclusion trustworthy, and
it is why limitations are not optional.

---

## The Report Headings

Your final assessment report has these headings, in this order. Every report
skeleton on this site uses them.

| Heading | What goes in it | In the *Tidebreak* report |
| :- | :- | :- |
| **1. Executive Summary** | The question, the one or two findings that matter, and what the team should do. Written last, read first | Build B's guidance cut wrong turns without costing enjoyment; keep it, and fix the Corridor hitch |
| **2. Introduction** | The game, the design problem and the research question | Players were getting lost and quitting; did Build B's guidance help? |
| **3. Methodology** | Who played what, what was measured, how the data was cleaned | `r nrow(sessions)` sessions from `r n_players` players, each on one build; telemetry, survey, comments; the cleaning log |
| **4. Results** | The evidence: descriptive, visual, statistical and qualitative, without interpretation | Wrong turns by build, with its figure, interval and p-value; the survey items; the comment codes |
| **5. Discussion** | What the results mean for the player: practical significance, where the evidence agrees and disagrees, limitations | Is the reduction big enough to matter? Why do comments disagree with telemetry? What could make this wrong? |
| **6. Conclusion** | The decisions and the next test, following from the Discussion | Keep the guidance; add an on-screen objective and retest |
| **7. References** | Anything you cite: questionnaires, methods, sources | The questionnaire your survey items came from |
| **8. Appendices** | The detail a checker needs but a reader does not: the cleaning log, the codebook, extra tables | Cleaning log; comment codebook; every survey item |

**Results reports; Discussion judges.** Keep them apart. A reader should be able to
disagree with your Discussion while agreeing with every number in your Results.

**Write the Executive Summary last.** It is the only section some readers will
read, so it must stand on its own: question, finding, action, in a few sentences.

### The Nine-Part Evidence Workflow

Earlier versions of these notes organised a report as nine parts. They are still a
useful **checklist** for whether your analysis is complete, but they are not the
report structure. Each part belongs under one of the assessment headings:

| Evidence workflow | Report section |
| :- | :- |
| 1 Research question | Introduction |
| 2 Experimental design | Methodology |
| 3 Descriptive evidence | Results |
| 4 Visual evidence | Results |
| 5 Statistical evidence | Results |
| 6 Practical significance | Discussion |
| 7 Qualitative evidence | Results / Discussion |
| 8 Limitations | Discussion |
| 9 Design conclusion | Conclusion |

**Use it as a check**, not as headings. Before you submit, tick off all nine: if
one is missing, a section of the report is thin.

**Qualitative evidence sits in two places.** The codes and their counts are
Results. What they mean, and how they agree or disagree with the telemetry, is
Discussion. [Combining Evidence](12-combining-evidence.qmd) shows how.

---

## What → Pattern → Meaning → Action

Every finding in the report is written in four moves:

| Move | Says | Test |
| :- | :- | :- |
| **What** | What was measured, on whom, how many | Could a reader repeat the measurement? |
| **Pattern** | What the data shows, with its size | Is there a number, and is it the right one? |
| **Meaning** | What that suggests about the game, and how sure you are | Does it follow from the pattern, and no further? |
| **Action** | The one change or test you recommend | Could someone start work on it tomorrow? |

```{mermaid}
%%| label: fig-wpma
%%| fig-cap: "What and Pattern are facts; Meaning and Action are judgement. The action is usually the next test, which starts the cycle again."
flowchart LR
  W["What"] --> P["Pattern"]
  P --> M["Meaning"]
  M --> A(["Action"])
  A -.->|"next playtest"| W
  class M accent
  class A action
```

**Meaning is where findings go wrong.** What and Pattern are facts. Action is a
recommendation. Meaning is the step where you connect them, and it is the step most
often skipped, overstated or quietly replaced by what you hoped to find. A finding
that does not name an action is not finished ([rule 10](ten-rules.qmd)); a finding
whose action does not follow from its meaning is worse.

**How it maps onto the report.** What goes in Methodology, Pattern in Results,
Meaning in Discussion, Action in Conclusion. For a comparison between builds, the
Pattern is the six questions from [Comparing Groups](08-comparing-groups.qmd):
what happened, how large, how uncertain, could it be chance.

---

## Ranking Findings

A playtest rarely produces one finding. *Tidebreak* produced at least five, and a
report that lists them in the order they were found buries the one that matters.
**Rank them, and lead with the top one.**

Rate each finding on four things:

- **Severity**: how badly it hurts the player experience when it happens.
- **Frequency**: how many players it affects.
- **Evidence**: how strong and how consistent the evidence is.
- **Actionability**: whether the team can do something about it, and what.

A common severity scale, adapted from usability testing, runs from 4 down to 0:

| Severity | Meaning |
| :-: | :- |
| 4 Critical | Stops play or makes players quit |
| 3 Major | Hurts the experience for many players, but they carry on |
| 2 Minor | A noticeable irritation |
| 1 Cosmetic | Fix if there is time |
| 0 None | Not a problem |

Applied to *Tidebreak*:

| Finding | Severity | Frequency | Evidence | Action |
| :- | :-: | :- | :- | :- |
| Build A frustration quits | 4 | `r frus_n[["A"]]` of `r sum(sessions$build == "A")` Build A sessions, against `r frus_n[["B"]]` in Build B | Strong: `r p_fmt(pt_frus$p.value)`, and Build B already fixes it | Keep Build B's guidance. Lead the report |
| Build B Corridor hitch | 3 | `r sum(slow_b)` frames over the 30 FPS budget, every one in the Corridor | Strong: a fixed interval, reproducible | Engineering ticket now |
| Objective unclear | 3 | `r round(100 * item_row("objective_clear")$disagree)`% of responses disagreed; better in Build B | Strong for the build difference; suggestive for the link to quits | Design change and retest |
| Build B rated harder | 1 | `r round(abs(diff(t.test(difficulty ~ build, data = sessions)$estimate)), 2)` points on a 1–7 scale | Weak: deaths did not change | Monitor in the next test |
| Visual style | 0 | `r round(100 * item_row("visual_style")$agree)`% agreed | Strong, but nothing to fix | Mention; do not lead with it |

**Severity is a judgement; write down why.** Two analysts may rate the Corridor
hitch differently. What matters is that the reasons are visible, so the team can
argue about the rating rather than about which finding the analyst happened to
like.

**For your project:** list every finding your team has, rate each, and put the
table in an appendix. The top one or two go in the Executive Summary.

---

## Worked Examples

**The data.** The `r nrow(survey)` post-session surveys, joined to the session
outcomes, feed both examples below. Every number in both is correct.

### A good finding

> **What.** `r nrow(survey)` players each rated eight statements about their session
> from 1 (strongly disagree) to 5 (strongly agree).
>
> **Pattern.** "I always knew what my objective was" was the lowest-rated of the
> eight, averaging `r round(item_row("objective_clear")$mean, 2)`, with
> `r round(100 * item_row("objective_clear")$disagree)`% disagreeing. It split
> sharply by build: `r round(obj_a, 2)` in Build A against `r round(obj_b, 2)` in
> Build B. Players who quit in frustration rated it `r round(obj_fr, 2)`, against
> `r round(obj_nf, 2)` for everyone else.
>
> **Meaning.** Build B's guidance made the objective clearer. The playtest was
> designed to compare the two builds, so this is the comparison it can best support.
> The link to frustration quits is an association: players who did not know what to
> do were the ones who left, which is consistent with unclear objectives causing
> the quits but does not prove it.
>
> **Action.** Keep Build B's guidance. For the next build, add an explicit on-screen
> objective in the first minute, and retest with objective clarity and the
> frustration-quit rate as the two target measures, with the success threshold
> agreed before the test.

**Why it works.** The What says who and how many. The Pattern gives the size and
the comparison, not only the ranking. The Meaning separates the claim the design
supports (the build changed clarity) from the one it only suggests (clarity drives
quits). The Action is specific, testable, and follows from the Meaning.

### A bad finding

> Players loved *Tidebreak*: visual style scored
> `r round(item_row("visual_style")$mean, 2)` out of 5, with
> `r round(100 * item_row("visual_style")$agree)`% agreeing. Build B scored higher
> than Build A on `r sum(by_build$B > by_build$A)` of the 8 survey items. Overall the
> playtest was a success, and the team should continue as planned.

**Why it fails.** Every number is true, and the finding is still useless.

- **What is missing.** Who answered, how many, and on what scale.
- **The Pattern is chosen, not found.** It leads with the best-rated item, which
  differs between the builds by
  `r sprintf("%.2f", abs(by_build$B[by_build$item == "visual_style"] - by_build$A[by_build$item == "visual_style"]))`
  of a point. "Higher on `r sum(by_build$B > by_build$A)` of 8" counts that as a
  win, the same as the objective item's gap of `r round(obj_b - obj_a, 2)`.
- **The Meaning is a verdict.** "A success" measured against what? Nothing in the
  question was about visual style.
- **The Action names nothing.** "Continue as planned" is not a decision anyone can
  start work on ([rule 10](ten-rules.qmd)).
- **It misses the finding.** The worst-rated item, the one tied to why players
  quit, is not mentioned. The bad version is not a weaker account of the good one.
  It is an account of a different, more comfortable playtest.

---

## An Annotated Report

What a finished report looks like, in miniature, under the assessment headings.
Each section is a sentence or two here; in your report each is a section. The notes
in grey say why each part is written the way it is.

::: {.callout-note appearance="simple" icon="false"}
**1. Executive Summary.** Build B's environmental guidance cut wrong turns by about
`r round(diff(rev(wt$estimate)), 1)` per session without reducing enjoyment, and
frustration quits fell from `r frus_n[["A"]]` to `r frus_n[["B"]]`. Keep the
guidance; fix a frame-rate hitch Build B introduced in the Corridor; add an
on-screen objective and retest.

[*Question, findings and actions in three sentences. Written last.*]{style="color: #6c757d"}

**2. Introduction.** Build A players were getting lost and quitting early. Build B
added lighting and signage. Did it improve navigation without reducing enjoyment?

[*The design problem, then the question with a yes or no answer, using the
measures that will answer it.*]{style="color: #6c757d"}

**3. Methodology.** `r nrow(sessions)` sessions from `r n_players` players,
`r per_player` sessions each; each player was assigned to one build. Measures:
wrong turns and quit reason (telemetry), an eight-item survey (self-report), and a
one-sentence comment. The raw export was cleaned as recorded in Appendix A.

[*Gives both counts, names the source of each measure, and points to the cleaning
log so the reader can reproduce n.*]{style="color: #6c757d"}

**4. Results.** Build B players took `r round(diff(rev(wt$estimate)), 1)` fewer
wrong turns per session (Figure 1; 95% CI `r ci_fmt(wt$conf.int)`,
`r p_fmt(wt$p.value)`). Enjoyment did not differ (95% CI for the difference
`r ci_fmt(tt_enj$conf.int, 2)` points on a 1–7 scale). Comments coded "lost or
unclear" were more common in Build B (Table 3).

[*Size first, then interval, then p-value. Every figure is referred to in the text.
No interpretation yet.*]{style="color: #6c757d"}

**5. Discussion.** A reduction of `r round(diff(rev(wt$estimate)), 1)` wrong turns
is `r pct_fmt(diff(rev(wt$estimate)) / wt$estimate[[1]])` of Build A's average:
large enough for players to feel. The comments disagree with the telemetry because
they track how the session felt, not where the player went. Limitations: each
player contributed `r per_player` sessions (the result holds when each counts
once); genre experience was not recorded.

[*Practical significance, the evidence that disagrees and why, and the weaknesses,
before the reader finds them.*]{style="color: #6c757d"}

**6. Conclusion.** Keep Build B's guidance. Fix the Corridor hitch before release.
Add an on-screen objective in the first minute and retest objective clarity and
the frustration-quit rate, with the success threshold agreed before the test.

[*Decisions and the next test, each following from the Discussion.*]{style="color: #6c757d"}

**7. References** and **8. Appendices.** The questionnaire source; Appendix A, the
cleaning log; Appendix B, the comment codebook.
:::

---

## Building the Report in Quarto

Your report is a `.qmd` file: prose, code and output in one document, rendered to
HTML with `quarto render report.qmd`. The skeleton is the assessment headings:

````markdown
---
title: "Tidebreak Build B: Playtest Report"
---

```{{r}}
#| echo: false
library(readr)
library(dplyr)
library(ggplot2)
sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
```

## Executive Summary

## Introduction

## Methodology

We analysed `r knitr::inline_expr("nrow(sessions)")` sessions from
`r knitr::inline_expr("length(unique(sessions$player_id))")` players.

## Results

## Discussion

## Conclusion

## References

## Appendices
````

**Inline code keeps the prose honest.** `` `r knitr::inline_expr("nrow(sessions)")` `` in
the text is replaced by the number when the report renders. When next week's data
arrives, you render again and every figure, table and sentence updates together.
A number typed by hand into the prose is the one that will be wrong. Every page of
this site is built this way.

**`echo: false`** hides a chunk's code but keeps its output: the reader sees the
table or figure without the R that made it. Use it for the report; leave the code
visible in your own working file.

**`fig-cap`** gives a figure a numbered caption. State the takeaway in it, as in
[A Figure for Your Report](09-visualising-with-ggplot2.qmd#a-figure-for-your-report).

### A Quarto document in one page

A Quarto document is a text file with three kinds of content:

````markdown
---
title: "Build B playtest report"
---

## Session length

Build B sessions were longer. The median was `r knitr::inline_expr("median(b$session_minutes)")` minutes.

```{{r}}
#| label: session-hist
hist(b$session_minutes)
```
````

**The header** between the `---` lines sets the title and options. **Prose** is
written in Markdown: `##` for a heading, `**bold**`, `*italic*`, `-` for a list.
**Code chunks** start with ```` ```{r} ```` and end with ```` ``` ````; lines
starting `#|` are chunk options, such as a label or whether to show the code.

**Rendering** runs every chunk from the top, in order, in a fresh session, and
writes the web page: the Render button in either editor, or `quarto render` in a
terminal. Because it starts fresh, a document that renders proves the analysis
works from the data alone, not from something you happened to type in the console
earlier.

**For your project:** start `report.qmd` now, with the eight headings and a setup
chunk that loads your data, and render it while it is empty. Every finding you add
from now on goes straight into it.

---

## Reporting a Whole Survey {.when-needed}

A survey tool exports one row per respondent and one column per question, because
that is how a form is filled in. That shape is **wide**. The *Tidebreak* survey has
`r nrow(survey)` rows and `r ncol(survey)` columns: a session ID and eight
agreement items, each rated 1 to 5.

To report every item at once you need it **long**: one row per answer, with
columns `session_id`, `item` and `rating`
([Tidy Data](03-data-and-types.qmd#tidy-data)). `pivot_longer()`, from `tidyr`,
does the reshape:

```{r}
#| label: demo-pivot
library(tidyr)

long <- survey |>
  pivot_longer(-session_id, names_to = "item", values_to = "rating")

dim(long)
head(long, 3)
```

`-session_id` means "every column except `session_id`". Those eight columns are
stacked: their names go into `item` and their values into `rating`.
`r nrow(survey)` sessions × 8 items = `r nrow(long)` rows.

Now `item` is a column, so one `group_by()` summarises every item:

```{r}
#| label: demo-items
long |>
  group_by(item) |>
  summarise(mean     = mean(rating),
            agree    = mean(rating >= 4),
            disagree = mean(rating <= 2)) |>
  arrange(mean)
```

For a rating scale, report the proportion agreeing and disagreeing alongside the
mean; the mean of an ordinal scale is a convenience, not a measurement
([rule 3](ten-rules.qmd)). And `item` can go on an axis:

```{r}
#| label: demo-plot
#| fig-cap: "Every survey item at once. Objective clarity has the most disagreement; visual style is almost all agreement."
#| fig-alt: "Horizontal stacked bars for the eight survey items, each split by rating from 1 in red to 5 in dark blue and ordered by mean rating. Visual style is almost entirely 4s and 5s; objective clear has the most red."
#| fig-height: 4
library(ggplot2)
long |>
  ggplot(aes(y = reorder(item, rating), fill = factor(rating))) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "RdBu") +
  labs(title = "Post-session survey, all items",
       x = "Proportion of players", y = NULL, fill = "Rating")
```

`reorder(item, rating)` sorts the items by their mean rating, and
`position = "fill"` stretches every bar to the same length.

**Going further:** `pivot_wider()`, the other kinds of join, and how to check a
join did what you meant are on
[Reshaping and Joining](../going-further/reshaping-and-joining.qmd).

---

## Interpretation Guidance

**What the data shows.** The objective item is the worst-rated of eight and differs
most between the builds. Wrong turns fell in Build B; enjoyment did not change.

**What the analysis suggests.** Build B's guidance improved navigation and how
clear the objective felt, at no cost to enjoyment.

**What can legitimately be concluded.** Build B made navigation better for the
players who tested it. Whether the clearer objective is what reduced frustration
quits is a hypothesis for the next test, not a finding.

**Order the evidence by strength.** In the Conclusion, lead with what the strongest
evidence supports, and say in the Discussion where weaker evidence disagrees. A
report that hides a contradiction invites the reader to find it.

---

## Common Mistakes

**Reporting results in the order you computed them.** The reader needs the order of
the argument. Use the headings.

**Interpreting in the Results.** "Wrong turns fell, showing the guidance works"
mixes a result with a judgement. Results says the first half; Discussion says the
second.

**Using the nine-part workflow as headings.** It is a checklist. The report has the
assessment headings.

**Leading with the best-rated item.** It is usually the art, it rarely differs
between builds, and it is almost never what the playtest was run to find.

**An Action that is a sentiment.** "Improve onboarding" is a wish. "Add an on-screen
objective in the first minute and retest clarity" is an action.

**A generic limitation.** "The sample was small" tells the reader nothing. "Each
player contributed four sessions, so sessions are not independent; the result holds
when each player counts once" tells them what you checked.

**Copying numbers into the prose by hand.** They go stale the next time the data
changes, and nobody notices until a meeting.

---

## Practice

Each takes two or three minutes.

1. **Place it.** Which report section does each of these belong in? (a) "Each
   player contributed four sessions." (b) "Players took 3.4 fewer wrong turns in
   Build B." (c) "A reduction of this size is large enough for players to notice."
   (d) "Keep the guidance and retest with an on-screen objective."
2. **Diagnose.** "Build B is better. Players liked it more and got lost less. We
   should ship it." Which of What, Pattern, Meaning and Action is missing or weak?
3. **Rank.** A colleague's draft report leads with "93% of players liked the visual
   style". Using the ranking table above, rewrite the report's first sentence.
4. **Compute.** Which survey item has the largest share of players answering 1 or 2
   in Build A?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
**1.** (a) Methodology, and again in Discussion if it is a limitation. (b) Results.
(c) Discussion: it is a judgement about practical significance. (d) Conclusion.

**2.** What is missing entirely: no measure, no players, no n. The Pattern has no
sizes and no interval. The Meaning ("better") is a verdict. The Action ("ship it")
does not say what was shown to work. Rewrite it with the wrong-turn difference and
its interval as the Pattern.

**3.** Lead with the highest-priority finding: "One in three Build A sessions
(`r frus_n[["A"]]` of `r sum(sessions$build == "A")`) ended in a frustration quit,
against `r frus_n[["B"]]` in Build B; Build B's guidance should be kept." Visual
style can appear later, as a strength that needs no action.

**4.**

```{r}
#| label: practice-disagree-a
disagree_a <- long |>
  left_join(select(sessions, session_id, build), by = "session_id") |>
  filter(build == "A") |>
  group_by(item) |>
  summarise(disagree = mean(rating <= 2)) |>
  arrange(desc(disagree))

head(disagree_a, 3)
```

The objective item: `r round(100 * disagree_a$disagree[1])`% of Build A players
answered 1 or 2, against `r round(100 * disagree_a$disagree[2])`% for the next
item.
:::

---

## Summary

- The report has eight headings: Executive Summary, Introduction, Methodology,
  Results, Discussion, Conclusion, References, Appendices. Results reports;
  Discussion judges.
- The Nine-Part Evidence Workflow is a checklist that maps onto those headings,
  not a replacement for them.
- Write every finding as What → Pattern → Meaning → Action. Meaning is where
  findings go wrong; Action is where they become useful.
- Rank findings by severity, frequency, evidence and actionability, and lead with
  the top one.
- A report built in Quarto computes its own numbers, so it cannot quote a stale
  one.
- When you need every survey item, `pivot_longer()` makes the survey long.

| Key term | Meaning |
| :- | :- |
| **Executive Summary** | The question, the findings that matter and the actions, written last and read first |
| **Nine-Part Evidence Workflow** | A checklist of the nine kinds of evidence a complete analysis covers |
| **What → Pattern → Meaning → Action** | The four moves of a finding |
| **Severity** | How badly a problem hurts the player experience, 0 to 4 |
| **Limitation** | Something that could make the conclusion wrong, stated by the author |
| **Wide, long** | One row per respondent, or one row per respondent per question |

---

## Links

- [Ten Rules](ten-rules.qmd): rule 2 (a number without a question attached is
  decoration) and rule 10 (a finding that does not name an action is not
  finished)
- [Combining Evidence](12-combining-evidence.qmd): telemetry, observation and
  self-report together, coding comments, limitations
- [Cleaning and Transforming](06-cleaning-and-transforming.qmd): the cleaning log
  for Methodology and the Appendices
- [Comparing Groups](08-comparing-groups.qmd): the six questions behind a Results
  paragraph
- [R and Quarto Reference: Quarto](../reference/r-quarto-reference.qmd#quarto)
- [Lab 05: From Evidence to Report](../labs/complete-analysis.qmd)
- [Stage 4 demo](../demos/stage-4.qmd)

---

## Reflective Questions

- The bad finding uses only true numbers. What, if anything, should stop a true
  number going into a report?
- A producer asks you to cut the limitations because they "undermine the
  recommendation". What do you say?
- Write your own project's Executive Summary now, in three sentences, before you
  have finished the analysis. What does that tell you about which analysis is
  still missing?
- Your Conclusion recommends a retest. What would you have to decide before that
  retest starts, so that its result can change your mind ([rule 1](ten-rules.qmd))?
