---
title: "Combining Evidence"
subtitle: "Telemetry, observation and self-report, coded, compared and reported honestly"
body-classes: "wf-at-interpret wf-at-act"
---

{{< include _workflow.qmd >}}

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")
library(dplyr)

coded <- sessions |>
  mutate(code = case_when(
    grepl("lost|circles|meant to go|objective", comment) ~ "Lost or unclear",
    grepl("route|exit|signposting", comment)             ~ "Found the way",
    grepl("Died|hard", comment)                          ~ "Difficulty",
    TRUE                                                 ~ "General"
  ))
lost_a   <- sum(coded$code == "Lost or unclear" & coded$build == "A")
lost_b   <- sum(coded$code == "Lost or unclear" & coded$build == "B")
found_a  <- sum(coded$code == "Found the way" & coded$build == "A")
found_b  <- sum(coded$code == "Found the way" & coded$build == "B")
helped_a <- sum(sessions$comment == "The signposting helped." & sessions$build == "A")
n_comments <- length(unique(sessions$comment))

fr_codes <- table(coded$code[coded$quit_reason == "Frustrated"])

joined   <- merge(sessions, survey, by = "session_id")
wt_build <- tapply(sessions$wrong_turns, sessions$build, mean)
nav_build <- tapply(joined$level_navigable, joined$build, mean)
obj_build <- tapply(joined$objective_clear, joined$build, mean)
r_wt_nav <- cor(joined$wrong_turns, joined$level_navigable, method = "spearman")

stopifnot(lost_b > lost_a, found_b < found_a, helped_a > 0,
          all(coded$enjoyment[coded$code == "Found the way"] >= 5),
          all(coded$enjoyment[coded$code == "Lost or unclear"] < 5),
          names(which.max(fr_codes)) == "Difficulty",
          wt_build[["A"]] - wt_build[["B"]] > 2,
          abs(diff(nav_build)) < 0.2, obj_build[["B"]] > obj_build[["A"]],
          abs(r_wt_nav) < 0.15)
```

> A single measure can mislead you. This page is about putting the three sources
> of playtest evidence side by side: what the game recorded, what you saw, and
> what the players said. It covers turning free-text comments into codes you can
> count, checking a finding against more than one source, and writing honest
> limitations. The playtest question running through it: *"The telemetry says
> Build B fixed navigation. Do the players agree?"*

::: {.at-a-glance}
**Stage:** 4 · Core. Read before the March workshop.
**Time:** 25 minutes, then the workshop lab on your own data.
**Before this:** [Planning a Playtest](01-planning-a-playtest.qmd#choose-your-measures)
for the three sources, and
[Relationships and Correlation](10-relationships-and-correlation.qmd) for the
navigation-rating finding.
**Key terms:** telemetry, observation, self-report, triangulation, code, codebook,
limitation.
**R you will write:** `case_when()` with `grepl()` inside `mutate()`, then
`count()`.
:::

## Learning Outcomes

By the end of these notes you can:

- say what each of telemetry, observation and self-report can and cannot tell you;
- turn free-text comments into a small set of codes with a written codebook, and
  count them;
- use quotes as evidence without letting one quote become the finding;
- triangulate a finding across sources, and say what it means when they disagree;
- write limitations that are specific to your playtest.

---

## Why This Matters for Playtesting

By the end of a project your team has numbers from the game, notes from watching
people play, survey ratings and a pile of comments. It is tempting to report the
source that tells the best story. The reader cannot see what you left out.

**The strongest findings use more than one source.** "Players got lost less" from
telemetry alone is a count. Add an observer who saw players stop backtracking at
the junction, and players who wrote that the route was obvious, and it is a
finding nobody will argue with.

**The most interesting findings are where the sources disagree.** *Tidebreak* has
one. It is the worked example at the end of this page.

---

## Three Sources, Three Questions

| Source | Answers | Strength | Weakness |
| :- | :- | :- | :- |
| **Telemetry** | What happened in the game? | Complete, precise, cheap at scale | Says nothing about why, or how it felt |
| **Observation** | How did the player's behaviour unfold? | Sees hesitation, confusion and workarounds that leave no trace in a log | Slow; one observer sees one player; observers see what they expect |
| **Self-report** | What does the player say they experienced? | The only source for feelings, intentions and opinions | Players misremember, are polite, and do not notice everything |

**Choose measures from at least two sources for every question that matters.**
That decision is made when you plan the playtest, not when you write the report
([Planning a Playtest](01-planning-a-playtest.qmd)).

### Recording observation

Observation is only evidence if it is recorded the same way for every player. Write
a short **observation sheet** before the playtest, with the behaviours you expect
to matter, and tick or time-stamp them as they happen:

| Time | Location | Behaviour | Note |
| :- | :- | :- | :- |
| 02:10 | First junction | Backtracked | Went left, returned |
| 02:45 | First junction | Hesitated > 5 s | Looked at the ceiling lights |
| 06:30 | Corridor | Asked a question | "Where am I meant to go?" |

Fix the behaviour list in advance: *backtracked*, *hesitated more than 5 seconds*,
*asked for help*, *ignored a prompt*. Then observations can be counted per player,
exactly like telemetry, and joined to it on the player ID.

**For your project:** if your team watched playtests but did not write anything
down, say so in the limitations. Next time, take an observation sheet with three to
five behaviours on it.

---

## Coding Comments

Free text is the richest self-report you have, and the hardest to use. **Coding**
turns it into something you can count: read the comments, decide on a small set of
**codes**, and give each comment one.

### Step 1: read everything first

Before writing a single code, read every comment. *Tidebreak* has
`r nrow(sessions)` comments, but only `r n_comments` different sentences:

```{r}
#| label: read-comments
sessions |> count(comment, sort = TRUE)
```

Real comments will be messier, longer and more varied. Read them anyway.

### Step 2: write a codebook

A **codebook** lists each code, what counts as that code, and an example. Keep it
to four to six codes; more than that and nobody can apply them consistently.

| Code | Use when the comment... | Example |
| :- | :- | :- |
| **Lost or unclear** | says the player did not know where to go or what to do | "Ran in circles for a while." |
| **Found the way** | says navigation or guidance worked | "Lighting made the exit easy to spot." |
| **Difficulty** | is about dying, challenge or fairness | "Too hard for a first attempt." |
| **General** | is anything else | "Enjoyed it, would play again." |

**The codebook goes in an appendix.** A reader needs to see your codes to judge
your counts, exactly as they need your cleaning log to judge your *n*.

### Step 3: apply the codes in R

`grepl()` returns TRUE when a comment contains any of the words you give it, and
`case_when()` gives each comment the first code that matches:

```{r}
#| label: code-comments
coded <- sessions |>
  mutate(code = case_when(
    grepl("lost|circles|meant to go|objective", comment) ~ "Lost or unclear",
    grepl("route|exit|signposting", comment)             ~ "Found the way",
    grepl("Died|hard", comment)                          ~ "Difficulty",
    TRUE                                                 ~ "General"
  ))

coded |> count(code)
```

`|` inside the quotes means "or". `TRUE ~ "General"` catches everything left over.

### Step 4: check by hand

Keyword matching is fast and blunt. Read a sample of comments next to their codes,
and fix the codebook or the keywords until they agree with what you would have
decided:

```{r}
#| label: check-codes
coded |> distinct(comment, code)
```

With real comments, check at least twenty. **If two people on the team code the
same twenty comments separately and disagree often, the codebook is not clear
enough yet.**

### Step 5: count by group, then read what you counted

```{r}
#| label: codes-by-build
coded |> count(code, build)
```

Codes are counts of *what players chose to write*, not rates of what happened.
Report them as "`r lost_b` of `r sum(sessions$build == "B")` Build B comments
mentioned being lost or unclear", not as
"`r round(100 * lost_b / sum(sessions$build == "B"))`% of Build B players got lost".

### Quotes as evidence

**Quote a comment to show a pattern, never to be the pattern.** One vivid quote
beside a chart of frustration quits makes the chart human. One vivid quote instead
of the chart is an anecdote. Pick quotes that are typical of their code, say how
many comments the code covers, and never quote only the comments that agree with
you.

---

## Triangulation

**Triangulation** means checking a finding against more than one source. Three
outcomes are possible, and each tells you something different:

| The sources... | What it means | What to write |
| :- | :- | :- |
| **Agree** | The finding is strong | Lead with it, and cite every source |
| **Disagree** | One source is measuring something different from what you thought | Report both, and explain which to trust for which question |
| **Only one source has anything** | The finding rests on one kind of evidence | Report it with that caveat, and plan the missing source for the next test |

```{mermaid}
%%| label: fig-triangulation
%%| fig-cap: "Agreement makes a finding strong. Disagreement is not a failure: it tells you what each source is really measuring."
flowchart TD
  F["One finding"] --> T["Telemetry"]
  F --> O["Observation"]
  F --> S["Self-report"]
  T --> Q{"Do they agree?"}
  O --> Q
  S --> Q
  Q -->|Yes| A(["Strong finding"])
  Q -->|No| D(["Explain the gap"])
  class D accent
```

---

## Worked Example

**The telemetry says Build B fixed navigation. Do the players agree?** Put every
source on the navigation question side by side.

**Telemetry.** Build A players averaged `r round(wt_build[["A"]], 1)` wrong turns
per session; Build B players `r round(wt_build[["B"]], 1)`. Large, and very
unlikely to be chance ([Comparing Groups](08-comparing-groups.qmd)).

**Self-report, ratings.** "The level was easy to navigate" averaged
`r round(nav_build[["A"]], 2)` in Build A and `r round(nav_build[["B"]], 2)` in
Build B: no change. Across sessions it does not correlate with wrong turns at all
(Spearman r = `r round(r_wt_nav, 2)`). "I always knew what my objective was" did
improve, from `r round(obj_build[["A"]], 2)` to `r round(obj_build[["B"]], 2)`.

**Self-report, comments.** `r lost_b` Build B comments were coded *lost or
unclear*, against `r lost_a` in Build A: the opposite of the telemetry. And
`r helped_a` Build A players wrote that "the signposting helped", about the build
without the new signage. Split by enjoyment instead of build, the codes separate
cleanly:

```{r}
#| label: codes-by-enjoyment
coded |> count(code, enjoyed = enjoyment >= 5)
```

Every *found the way* comment came from a player who rated enjoyment 5 or above,
and every *lost or unclear* comment from one who rated it lower. The comments
record how the session *felt*, not where the player went.

**Observation.** *Tidebreak* did not record any. That is the gap this example
exposes.

**Putting it together.**

- What happened: Build B cut wrong turns. Only telemetry measured it directly, and
  it is the source to trust for *what happened*.
- What players noticed: not much. The navigation rating did not move, and players
  appear not to know when they are lost.
- What the comments measure: mood. Useful for knowing how the session felt; weak
  evidence about navigation in either direction.
- What is missing: someone watching the junctions, to see whether players hesitate
  and backtrack, and whether they seem aware of it.

**What should the team do?** Keep Build B's guidance on the strength of the
telemetry. Stop relying on the navigation rating to check navigation on this
level. In the next playtest, add an observation sheet for the first two junctions,
so the telemetry and the observation can be compared directly.

**In the Discussion you would write:** "The reduction in wrong turns was not
reflected in players' navigation ratings or comments. The comments tracked
enjoyment rather than build, which suggests they describe how the session felt
rather than where the player went. We treat the telemetry as the evidence for
navigation, and recommend structured observation at the junctions in the next
playtest to check whether players notice when they are lost."

---

## Writing Limitations

Every playtest has weaknesses. Naming them is not an admission of failure: it is
what lets the reader trust everything else. **A limitation you name costs you less
than one the reader finds.**

A good limitation is **specific** to your playtest, says **what it might change**,
and says **whether you checked it**.

| Kind | Generic (weak) | Specific (useful) |
| :- | :- | :- |
| Sample | "The sample was small." | "`r length(unique(sessions$player_id))` players; how they were recruited was not recorded, so we cannot say how closely they match the game's audience." |
| Design | "There may be confounds." | "Each player played one build, so differences between the two groups of players could explain part of the gap. The cohort mix in each build is in Appendix C." |
| Independence | "Data may not be independent." | "Each player contributed `r unique(as.vector(table(sessions$player_id)))` sessions. The result holds when each player counts once." |
| Measurement | "Surveys are subjective." | "The navigation rating did not track wrong turns, so it is not a reliable check on navigation for this level." |
| Coding | "Comments were interpreted." | "Comments were coded by keyword into four codes (Appendix B); a second coder was not used." |
| Not measured | "More research is needed." | "No observation was recorded, so we cannot say whether players noticed when they were lost." |

**Limitations go in the Discussion**, close to the claims they qualify. The
Conclusion should already take them into account.

**For your project:** write your limitations as you go, not on the last night. Each
time you make a decision you are not sure about (a row dropped, a code merged, a
measure you could not collect), add a line.

---

## Common Mistakes

**Reporting only the source that tells the best story.** The reader cannot see what
you left out. Report every source you collected on the question.

**Treating a count of comments as a rate.** A comment is what a player chose to
write. "Ten players mentioned the lighting" is not "ten players noticed the
lighting".

**Coding without a codebook.** If the reader cannot see what each code means, they
cannot judge the counts.

**Too many codes.** Fifteen codes on sixty comments gives counts of two and three.
Four to six is plenty.

**One quote as the finding.** A quote illustrates a pattern; it cannot be one.

**Explaining away a disagreement.** When telemetry and self-report disagree, the
disagreement is a finding. Report it and say what each source is measuring.

**Generic limitations.** "The sample was small" tells the reader nothing they can
use.

---

## Practice

Each takes two or three minutes.

1. **Classify.** Which source is each of these? (a) A death count. (b) "Player
   paused for 20 seconds at the bridge." (c) A 1–7 enjoyment rating. (d) The time
   at which a player first opened the map. (e) "I didn't know I could sprint."
2. **Compute.** Among frustration quits only, which comment code is most common?
3. **Critique.** "12 of 60 players complained about the camera, so the camera is a
   problem for 20% of players." What is wrong?
4. **Rewrite.** Make this limitation specific: "There may be other factors that
   affected the results."

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
**1.** (a) Telemetry. (b) Observation. (c) Self-report. (d) Telemetry. (e)
Self-report.

**2.**

```{r}
#| label: practice-frustrated
coded |>
  filter(quit_reason == "Frustrated") |>
  count(code, sort = TRUE)
```

Difficulty, with `r fr_codes[["Difficulty"]]`, ahead of lost or unclear with
`r fr_codes[["Lost or unclear"]]`. Yet the survey says frustration quitters did
not know their objective. Players name the symptom they felt, not the cause, which
is why comments illustrate a finding rather than measure it.

**3.** A comment is what a player chose to write, not a measurement. Players who
did not mention the camera may still have had trouble with it, and players who
mentioned it may have been mildly irritated. Report "12 of 60 comments mentioned
the camera", and check it against telemetry or observation, such as how often the
camera was reset.

**4.** Name the factor, what it might change, and whether you checked it. For
example: "Experienced players may navigate better whichever build they play. We
checked the wrong-turns result within each cohort." Any specific factor, checked or
not, is better than "other factors".
:::

---

## Summary

- Telemetry says what happened, observation how it unfolded, self-report what
  players experienced. Choose measures from at least two for every question that
  matters.
- Observation is evidence only when it is recorded the same way for every player:
  use an observation sheet with a fixed list of behaviours.
- Code comments with a short, written codebook; apply it in R with `case_when()`
  and `grepl()`; check a sample by hand; report counts as counts of comments.
- Quote to illustrate a pattern, never to be the pattern.
- Triangulate. Agreement makes a finding strong; disagreement tells you what each
  source is really measuring.
- Write limitations that are specific, say what they might change, and say whether
  you checked.

| Key term | Meaning |
| :- | :- |
| **Triangulation** | Checking a finding against more than one source of evidence |
| **Observation sheet** | A fixed list of behaviours recorded the same way for every player |
| **Code** | A category given to a piece of free text |
| **Codebook** | The list of codes, what each means, and an example of each |
| **Limitation** | A specific weakness that could change the conclusion, stated by the author |

---

## Links

- [Planning a Playtest](01-planning-a-playtest.qmd#choose-your-measures): the three
  sources, chosen before the playtest
- [Relationships and Correlation](10-relationships-and-correlation.qmd#telemetry-observation-and-self-report-together):
  the navigation rating that did not track wrong turns
- [Interpreting and Reporting Evidence](11-reporting-evidence.qmd): where all of
  this goes in the report
- [Lab 05: From Evidence to Report](../labs/complete-analysis.qmd)
- Back to the overview: [Start Here](00-start-here.qmd)

---

## Reflective Questions

- Which of your team's measures would you trust most for *what happened*, and which
  for *how it felt*? Are they the same?
- Your comments say players loved a feature that the telemetry says almost nobody
  used. Write the sentence you would put in the Discussion.
- What three behaviours would you put on an observation sheet for your next
  playtest, and why those?
- Which of your own limitations would most change your Conclusion if it turned out
  to matter?
