---
title: "Cleaning and Transforming"
subtitle: "Every cleaning choice is part of the analysis"
---

> Week 2, first cycle, for everyone, before Part A of Lab 02. It covers getting a
> raw export into a shape that can answer a question: finding what is missing,
> doubled, mislabelled or impossible, and deciding what to do about each. Then it
> introduces the six `dplyr` verbs and the pipe, and shows what they are doing by
> setting them next to the base R you already know. The worked example is the one
> to remember: eight blank cells that were hiding the finding.

```{r}
#| label: setup
#| include: false
library(readr)
library(dplyr)

raw      <- read_csv("data/tidebreak_sessions_raw.csv",
                     trim_ws = FALSE, show_col_types = FALSE)
original <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)

deduped <- raw[!duplicated(raw), ]
fixed   <- tolower(trimws(deduped$cohort))
mislabelled <- sum(!deduped$cohort %in% c("Novice", "Regular", "Veteran"))
blank   <- deduped[is.na(deduped$session_minutes), ]
kept    <- deduped[complete.cases(deduped), ]

frus <- function(d, b) sum(d$build == b & d$quit_reason == "Frustrated")
med_a <- function(d) median(d$session_minutes[d$build == "A"], na.rm = TRUE)

stopifnot(sum(duplicated(raw)) == sum(duplicated(raw$session_id)),
          nrow(deduped) == nrow(original),
          all(blank$build == "A"),
          all(blank$quit_reason == "Frustrated"),
          length(unique(fixed)) == 3,
          med_a(deduped) > med_a(original),
          sum(deduped$session_minutes <= 0, na.rm = TRUE) == 0,
          all(deduped$enjoyment %in% 1:7),
          all(head(deduped$archetype[order(deduped$score)], 5) == "Sentinel"))
```

---

## Learning Outcomes

By the end of these notes you can:

- say what an `NA` means in a playtest export, and check whether the pattern of
  missing values is itself a finding;
- find exact duplicates and duplicates on the key, and remove them;
- tell an impossible value from an unusual one, and treat each correctly;
- repair inconsistent category labels and convert text to numbers without losing
  data silently;
- keep a cleaning log that records every row dropped and what it did to *n*;
- use `select()`, `filter()`, `mutate()`, `arrange()`, `group_by()` and
  `summarise()`, composed with the pipe;
- explain what `group_by()` + `summarise()` does by writing the same answer with
  `tapply()`.

---

## Why This Matters for Playtesting

No playtest export arrives ready. Telemetry drops events when a build crashes.
Spreadsheets get pasted into twice. Someone types "veteran" where the form said
"Veteran". Every one of these has to be dealt with before any summary means
anything, and every way of dealing with it is a choice.

**Cleaning choices change conclusions.** Dropping the rows with a blank cell is one
line of R. If the blanks are not scattered at random, that line removes a
particular kind of player, and the summary you compute afterwards describes the
players who were left. The worked example below does exactly this to the strongest
finding in the *Tidebreak* data.

**That is why cleaning goes in the report.** A reader who cannot see what you
removed cannot judge what you found. Cleaning is analysis
([rule 6](ten-rules.qmd)): record every choice where the reader can see it, not
hidden in a script.

---

## What `NA` Actually Means

`NA` is R's marker for a **missing value**: a cell with nothing in it. It does not
mean zero, and it does not mean the event did not happen. It means the export does
not know. A blank `deaths` cell is not a player who never died.

**`NA` is contagious.** Any calculation that touches an `NA` returns `NA`, because
R will not guess what the missing value was. `mean(deaths)` on a column with one
blank returns `NA`. Adding `na.rm = TRUE` tells R to leave the blanks out, which is
itself a decision: it assumes the missing players look like the ones you can see.

Three questions, in this order, before you do anything to a missing value:

1. **How much is missing?** A handful of blanks in 120 rows is different from half
   a column.
2. **Where is it missing?** Split the count by build, cohort and outcome. If the
   blanks cluster, something produced them.
3. **Is the pattern the finding?** A build whose quit path never writes the session
   length has a telemetry bug, and the players on that path are the ones the
   playtest was run to find ([rule 9](ten-rules.qmd)).

---

## Duplicates

A **duplicate** is a row that appears more than once. It inflates *n*, and it
counts one player's experience twice.

An **exact duplicate** repeats every column. It is almost always an export or
copy-paste fault, and removing it is safe. A **duplicate on the key** repeats the
identifier, here `session_id`, but may differ elsewhere. That is two records
claiming to be the same session, and you have to work out which one is right before
you remove either.

**Check both.** If the two counts match, every key duplicate is an exact copy. If
they do not, look at the rows that differ before deciding.

---

## Impossible Versus Unusual

An **impossible value** cannot happen in the game: a negative session length, a
rating of 9 on a 1–7 scale, more objectives captured than the map has. It is an
error. Fix it if you can trace the cause, otherwise set it to `NA`, and log it
either way.

An **unusual value** is extreme but possible: a player who never died, a Sentinel
who scored almost nothing. It is data. Removing it because it spoils a chart is
deleting the players you most need to hear from.

**The test is the game's rules, not the chart.** A boxplot flags points by distance
from the rest. Only you know whether the game allows that value. Write the rule
down, such as `session_minutes > 0` or `enjoyment` in 1–7, then check the data
against it.

---

## Labels and Types

**Inconsistent labels.** R compares text exactly. `"Veteran"`, `"veteran"` and
`"Veteran "` with a trailing space are three different cohorts to `table()`, and a
grouped summary will report three groups where the design had one. `trimws()`
strips leading and trailing spaces and `tolower()` makes the case consistent. Use
them together, and always check the result with `table()`. `read_csv()` trims
spaces on import by default, which hides them in R but not in the file anyone else
opens.

**Numbers stored as text.** When one cell in a numeric column contains `"n/a"` or a
comma decimal, `read_csv()` reads the whole column as text. `as.numeric()` converts
it back, and turns every value it cannot read into `NA` with only a warning. Count
the `NA`s before and after converting: any increase is data you just lost.

**Factors.** A **factor** is a categorical variable that knows its possible values
and their order. `factor(cohort, levels = c("novice", "regular", "veteran"))`
fixes the order to the order of experience, so tables and plots read left to right
from least to most experienced, not alphabetically. It also makes a misspelled
label visible: anything not in `levels` becomes `NA` rather than a fourth group
([rule 3](ten-rules.qmd)).

---

## Keeping a Cleaning Log

A **cleaning log** is a list of every change made to the data, in order, with the
number of rows before and after. It is short, and it goes in the report.

| Step | Rows after | Why |
| :- | -: | :- |
| Raw export | `r nrow(raw)` | As delivered |
| Remove exact duplicates | `r nrow(deduped)` | `r sum(duplicated(raw))` sessions exported twice |
| Standardise `cohort` labels | `r nrow(deduped)` | `r mislabelled` values re-spaced or re-cased; no rows removed |
| Keep blank `session_minutes` | `r nrow(deduped)` | All `r nrow(blank)` are Build A frustration quits; see the worked example |

**A step that removes no rows still goes in the log.** Relabelling changes which
group a player is counted in, and the reader needs to know that happened.

---

## The Six Verbs and the Pipe

The `dplyr` package, part of the tidyverse, reduces most data manipulation to six
**verbs**. Each takes a data frame and returns a data frame.

| Verb | What it does | Base R you already know |
| :- | :- | :- |
| `select()` | Keep some columns | `sessions[, c("build", "score")]` |
| `filter()` | Keep some rows | `sessions[sessions$build == "A", ]` |
| `mutate()` | Add or change a column | `sessions$kd <- sessions$kills / sessions$deaths` |
| `arrange()` | Sort the rows | `sessions[order(sessions$score), ]` |
| `group_by()` | Mark the groups | The second argument of `tapply()` |
| `summarise()` | One row of results per group | The function inside `tapply()` |

The **pipe**, `|>`, passes the result on its left in as the first argument of the
function on its right. `sessions |> filter(build == "A")` means
`filter(sessions, build == "A")`. Its value is in chaining: a sequence of verbs
reads top to bottom, in the order the steps happen, instead of inside out.

**You will also see `%>%`.** It is the older tidyverse pipe and does the same job
in everything you will write here. [Distributions and Variation](distributions-and-variation.qmd)
used it; these notes use `|>`, which is built into R and needs no package.

---

## Worked Examples

**Which build keeps players longer?** The raw export has `r nrow(raw)` rows. After
removing the `r sum(duplicated(raw))` duplicates, `r nrow(blank)` sessions have a
blank `session_minutes`. Every one of them is Build A. Every one of them ended in a
frustration quit.

```{r}
#| label: worked-where-missing
table(build   = deduped$build,
      missing = is.na(deduped$session_minutes))
```

That pattern is not random, so both obvious ways of handling it are wrong.

**Choice 1: drop incomplete rows.** `complete.cases()` keeps only rows with no
blank anywhere. It keeps `r nrow(kept)` of `r nrow(deduped)` sessions. Before,
Build A had `r frus(deduped, "A")` frustration quits against Build B's
`r frus(deduped, "B")`. After, it has `r frus(kept, "A")` against
`r frus(kept, "B")`. One line of cleaning removed
`r round(100 * (frus(deduped, "A") - frus(kept, "A")) / frus(deduped, "A"))`% of
the evidence for Build A's biggest problem, and nothing warned you.

**Choice 2: leave them out of the median.** With `na.rm = TRUE`, Build A's median
session is `r round(med_a(deduped), 1)` minutes. But the Build A frustration quits
that *were* recorded have a median of
`r round(median(deduped$session_minutes[deduped$build == "A" & deduped$quit_reason == "Frustrated"], na.rm = TRUE), 1)`
minutes. The missing sessions are almost certainly short ones, so leaving them out
makes Build A look better than it is.

**What the complete data says.** For this exercise the unbroken export exists, so
the damage can be measured. Build A's true median is
`r round(med_a(original), 1)` minutes, not `r round(med_a(deduped), 1)`. The blank
cells were never noise. They were the shortest, angriest sessions in the playtest,
lost by a telemetry bug on exactly the path the playtest was meant to examine.

**The defensible handling** keeps the rows, counts them from `quit_reason`, which
survived, reports the median with a note that `r nrow(blank)` short sessions are
missing from it, and files a bug against Build A's session logging.

---

## R Demonstrations

Base R first for each cleaning step, then the six verbs, then `tapply()` and
`group_by()` answering the same question.

```{r}
#| label: demo-load
library(readr)
library(dplyr)

raw <- read_csv("data/tidebreak_sessions_raw.csv",
                trim_ws = FALSE, show_col_types = FALSE)
nrow(raw)
```

`read_csv()` normally strips spaces from the edges of text values without telling
you. `trim_ws = FALSE` switches that off, so you see the file the way Excel or a
database import would.

### Missing values

```{r}
#| label: demo-na
na_counts <- colSums(is.na(raw))
na_counts[na_counts > 0]

mean(raw$deaths)
mean(raw$deaths, na.rm = TRUE)
```

`is.na()` returns `TRUE` for each blank cell, and `colSums()` counts them per
column. Printing only the columns with blanks keeps the output readable. The first
`mean()` shows the contagion; the second is a decision.

```{r}
#| label: demo-complete
sum(complete.cases(raw))
```

`complete.cases()` returns `TRUE` for each row with no blanks anywhere. Count it
before you ever filter by it.

### Duplicates

```{r}
#| label: demo-duplicated
sum(duplicated(raw))
sum(duplicated(raw$session_id))
raw$session_id[duplicated(raw)]

sessions <- raw[!duplicated(raw), ]
nrow(sessions)
```

`duplicated()` marks the second and later copies of a row, or of a value when you
give it one column. Keeping `!duplicated()` keeps the first copy of each.

### Labels, types and impossible values

```{r}
#| label: demo-labels
table(sessions$cohort)

sessions$cohort <- tolower(trimws(sessions$cohort))
sessions$cohort <- factor(sessions$cohort,
                          levels = c("novice", "regular", "veteran"))
table(sessions$cohort)
```

`r length(unique(deduped$cohort))` spellings become three cohorts, in order of experience.

```{r}
#| label: demo-as-numeric
#| warning: true
minutes_text <- c("12.9", "24.6", "n/a", "8,6")
as.numeric(minutes_text)
```

A constructed example, since this export's numbers all arrived as numbers. Two of
four values became `NA`, and the only sign was a warning.

```{r}
#| label: demo-impossible
sum(sessions$session_minutes <= 0, na.rm = TRUE)
sum(!sessions$enjoyment %in% 1:7)
head(sessions[order(sessions$score), c("session_id", "archetype", "score")], 5)
```

No impossible lengths or ratings. The five lowest scores are all Sentinels. They
are unusual, not impossible, and they are the bimodal Sentinel finding from
[Describing What You Have](describing-what-you-have.qmd). They stay.

### The six verbs

```{r}
#| label: demo-verbs
sessions |>
  filter(build == "A", quit_reason == "Frustrated") |>
  mutate(logged = !is.na(session_minutes)) |>
  select(session_id, cohort, session_minutes, logged) |>
  arrange(session_minutes) |>
  head(4)
```

Read it as a sentence: take the sessions, keep Build A's frustration quits, add a
column saying whether the length was logged, keep four columns, sort by length,
show the first four. Each line is one step, and removing a line removes one step.
`arrange()` puts `NA` last, so the shortest *recorded* quits come first.

### The same question, twice

*What is the average number of wrong turns in each build?*

::: {layout-ncol=2}

```{r}
#| label: demo-tapply-one
tapply(sessions$wrong_turns,
       sessions$build,
       mean)
```

```{r}
#| label: demo-group-one
sessions |>
  group_by(build) |>
  summarise(mean_wrong = mean(wrong_turns))
```

:::

Same numbers: the tibble prints three significant figures, but stores the same
values. Both do the same three things, which together are called
**split–apply–combine**: split `wrong_turns` into one piece per build, apply
`mean()` to each piece, combine the results. `tapply()` takes all three as
arguments in one call and returns a named vector. `group_by()` does the split,
`summarise()` does the apply, and the combine is a tibble with one row per group.

The difference shows when the question grows. *How many sessions does each build
have, how many are missing a length, and what is the median length of the rest?*

::: {layout-ncol=2}

```{r}
#| label: demo-tapply-many
cbind(
  n       = tapply(sessions$session_minutes,
                   sessions$build, length),
  missing = tapply(is.na(sessions$session_minutes),
                   sessions$build, sum),
  median  = tapply(sessions$session_minutes,
                   sessions$build, median,
                   na.rm = TRUE)
)
```

```{r}
#| label: demo-group-many
sessions |>
  group_by(build) |>
  summarise(
    n       = n(),
    missing = sum(is.na(session_minutes)),
    median  = median(session_minutes,
                     na.rm = TRUE)
  )
```

:::

`tapply()` needs one call per statistic, each repeating the column and the
grouping, and `cbind()` to stick them together. `summarise()` takes as many
statistics as you like against one `group_by()`. The answer is the worked example
in one table: all `r nrow(blank)` missing lengths are Build A's, and the median
beside them is computed without them.

---

## Interpretation Guidance

**What the data shows.** The raw export has `r sum(duplicated(raw))` duplicated
sessions, `r mislabelled` mislabelled cohorts and
`r nrow(blank)` blank session lengths, all in Build A frustration quits.

**What the analysis suggests.** Build A's quit path does not write the session
length. The blanks are a telemetry fault concentrated in the sessions that matter
most.

**What can legitimately be concluded.** Build A's frustration-quit count is
reliable, because it comes from `quit_reason`, which is complete. Its median
session length is an overestimate by an amount you cannot know from this export.
Say both.

**Report what you did, not only what you found.** The cleaning log goes in the
report's experimental design section. A reader should be able to reproduce your
*n* from it.

---

## Common Mistakes

**Dropping incomplete rows by reflex.** `na.omit()` and `complete.cases()` are one
line each and can remove the finding. Check where the blanks are first.

**Treating `NA` as zero.** A blank deaths cell is unknown, not a player who never
died. Replacing it with 0 invents a very good player.

**Using `na.rm = TRUE` without saying so.** It is the right tool, and it changes
what the number describes. Report how many values it left out.

**Fixing labels without checking.** Run `table()` before and after. A label you did
not know about survives `tolower()`.

**Deleting unusual values as if they were impossible.** The game's rules decide
what is impossible. The chart does not.

**Cleaning in a spreadsheet.** Edits made by hand cannot be seen, repeated or
reviewed. Every change belongs in code, where it runs again on the next export.

---

## Short Practice Activities

Each takes two or three minutes. Start from `sessions`, the de-duplicated data with
fixed labels from the demonstrations.

**1.** Which rows would `complete.cases()` remove, split by build and quit reason?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-incomplete
dropped <- sessions[!complete.cases(sessions), ]
table(dropped$build, dropped$quit_reason)
```

`r nrow(dropped)` rows. The `r nrow(blank)` Build A frustration quits are the
blank lengths; the other `r nrow(dropped) - nrow(blank)` are blank `deaths` cells
scattered across other outcomes. Those two groups need different handling, and
`complete.cases()` treats them the same.
:::

**2.** Rewrite `tapply(sessions$score, sessions$archetype, median)` with
`group_by()` and `summarise()`, and add the number of sessions per archetype.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-archetype
sessions |>
  group_by(archetype) |>
  summarise(n = n(), median_score = median(score))
```

Adding `n = n()` costs one argument in `summarise()` and a second `tapply()` call
in base R. Always report *n* next to a group summary
([rule 8](ten-rules.qmd)).
:::

**3.** Using the pipe, list the sessions with a blank `deaths` cell, showing only
`session_id`, `build` and `quit_reason`.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-deaths
sessions |>
  filter(is.na(deaths)) |>
  select(session_id, build, quit_reason)
```

No pattern across builds or outcomes: these blanks look scattered, so leaving them
out of a deaths summary with `na.rm = TRUE` is defensible. That is a conclusion
about these three rows, not a rule for every blank.
:::

---

## Summary

- `NA` means unknown. Count it, find where it is, and ask whether the pattern is
  the finding before doing anything to it.
- Check duplicates both exactly and on the key.
- Impossible values are errors; unusual values are data. The game's rules decide
  which is which.
- `trimws()` and `tolower()` fix labels; `factor()` fixes their order; `table()`
  checks the result.
- Keep a cleaning log with *n* at every step. It goes in the report.
- The six verbs, chained with `|>`, are split–apply–combine made readable.
  `tapply()` does the same thing in one call.
- In *Tidebreak*, the `r nrow(blank)` blank cells were the frustration quits.
  Dropping them would have hidden Build A's worst problem.

---

## Links to Relevant Reference Material

- [Ten Rules](ten-rules.qmd): rule 3 (know what kind of number you are holding),
  rule 6 (cleaning is analysis), rule 8 (small samples lie confidently) and rule 9
  (sometimes the rare value is the whole story)
- [Distributions and Variation](distributions-and-variation.qmd): the three kinds of
  outlier
- [R and Quarto Reference: getting at parts of a data frame](../reference/r-quarto-reference.qmd#getting-at-parts-of-a-data-frame)
  and [describing by group](../reference/r-quarto-reference.qmd#describing-by-group)
- [R and Quarto Reference: cleaning](../reference/r-quarto-reference.qmd#cleaning)
  and [transforming with dplyr](../reference/r-quarto-reference.qmd#transforming-with-dplyr)
- [Lab 02: cleaning, transforming and comparing](../labs/comparing-builds.qmd#part-a-clean-and-transform), Part A

---

## Reflective Questions

- A colleague removes every row with a blank cell "to be safe". What would you need
  to know about those rows before you agreed with them?
- Build A's median session length is wrong in a direction you know but by an amount
  you do not. Should the report quote it at all? What would you write instead?
- Two rows share a `session_id` but have different scores. What evidence would you
  use to decide which one to keep, and what do you do if there is none?
- `tapply()` and `group_by()` give the same answer. When would you choose the base R
  version in your own analysis?

---
