---
title: "Comparing Playtest Conditions"
subtitle: "Is the difference real, and does it matter?"
---

> Week 2, second cycle, for everyone, before Part B of Lab 02. It takes the
> question every A/B playtest ends with, "is Build B better?", and splits it into
> the two questions it actually contains: would the difference hold up with other
> players, and is it big enough to act on? The worked example is the one to
> remember: two results that are both statistically significant, one of which you
> should act on and one of which you should not.

```{r}
#| label: setup
#| include: false
library(readr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
paired   <- read_csv("data/tidebreak_paired.csv", show_col_types = FALSE)
frames   <- read_csv("data/tidebreak_frame_times.csv", show_col_types = FALSE)

p_fmt  <- function(p) if (p < 0.001) "p < 0.001" else paste("p =", round(p, 3))
ci_fmt <- function(ci, d = 1) paste(round(ci[1], d), "to", round(ci[2], d))

tt_wt   <- t.test(wrong_turns ~ build, data = sessions)
gap_wt  <- unname(tt_wt$estimate[1] - tt_wt$estimate[2])

tt_diff <- t.test(difficulty ~ build, data = sessions)
gap_dif <- unname(tt_diff$estimate[2] - tt_diff$estimate[1])
ci_dif  <- -rev(tt_diff$conf.int)
tt_dth  <- t.test(deaths ~ build, data = sessions)

tt_enj  <- t.test(enjoyment ~ build, data = sessions)

tt_pair <- t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)
tt_unp  <- t.test(paired$wrong_turns_a, paired$wrong_turns_b)
d_pair  <- paired$wrong_turns_a - paired$wrong_turns_b

atrium  <- frames[frames$zone == "Atrium", ]
tt_atr  <- t.test(ms ~ build, data = atrium)
gap_atr <- unname(tt_atr$estimate[1] - tt_atr$estimate[2])

per_player <- aggregate(wrong_turns ~ player_id + build, data = sessions, FUN = mean)
tt_player  <- t.test(wrong_turns ~ build, data = per_player)

n_players <- length(unique(sessions$player_id))
n_per     <- unique(as.vector(table(sessions$player_id)))

wt_out <- tapply(sessions$wrong_turns, sessions$build,
                 function(x) length(boxplot.stats(x)$out))

stopifnot(tt_wt$p.value < 0.001, tt_wt$conf.int[1] > 0,
          tt_diff$p.value < 0.05, gap_dif > 0, gap_dif < 0.5, ci_dif[1] < 0.1,
          tt_dth$p.value > 0.5,
          tt_enj$p.value > 0.5, max(abs(tt_enj$conf.int)) < 0.5,
          tt_pair$p.value < 0.001,
          diff(tt_pair$conf.int) < diff(tt_unp$conf.int),
          tt_atr$p.value < 0.001, abs(gap_atr) < 0.5,
          tt_player$p.value < 0.001,
          isTRUE(all.equal(unname(diff(rev(tt_player$estimate))), gap_wt)),
          length(n_per) == 1,
          all(boxplot.stats(sessions$wrong_turns[sessions$build == "A"])$out <
                median(sessions$wrong_turns[sessions$build == "A"])))
```

---

## Learning Outcomes

By the end of these notes you can:

- explain why a second group of players would give a different result, and why
  that matters before you act on any difference;
- report an observed difference with its confidence interval, and read the
  interval as a range of plausible sizes;
- state what a p-value is, and what it is not;
- tell statistical significance apart from practical significance, with a
  *Tidebreak* example of each;
- choose between an independent-samples and a paired t-test from how the playtest
  was designed;
- check the assumptions of a t-test at a practical level, including with a Q-Q
  plot;
- work through **visualise → describe → estimate difference → confidence interval
  → statistical test → interpret** in that order.

---

## Why This Matters for Playtesting

Every build comparison ends with a difference. Build B's players took fewer wrong
turns, or scored more, or quit less. The difference is always there, because two
groups of people never produce identical numbers. The question is never "is there
a difference?" but **"is this difference telling me something about the build, or
only about the players who happened to turn up?"**

**A playtest is a sample.** The `r n_players` people who played *Tidebreak* stand
in for everyone who will play it after launch. Run the test again next week with
different players and every number would move. Before you rebuild a level on the
strength of a difference, you need to know whether that difference would survive
a second playtest.

**And a real difference can still be worthless.** With enough data, a change too
small for any player to notice becomes statistically detectable. A result can be
certain and irrelevant at the same time. Deciding whether it matters is a design
judgement, and no test makes it for you.

---

## The Workflow

Comparing two conditions always follows the same six steps, in this order:

> **visualise → describe → estimate difference → confidence interval → statistical
> test → interpret**

| Step | Question | Tool |
| :- | :- | :- |
| Visualise | What do the two distributions look like? | Side-by-side boxplots or faceted histograms |
| Describe | Where is each centre, how wide is each spread? | Medians, means, IQRs, standard deviations |
| Estimate difference | How far apart are the groups in this sample? | Difference in means |
| Confidence interval | How far apart could they plausibly be, for players in general? | `t.test()` |
| Statistical test | Could a gap this size be chance alone? | `t.test()` |
| Interpret | Is it real, is it big enough, and what do we do? | Judgement |

**The test comes fifth, not first.** A p-value with no picture and no size
attached cannot be interpreted. If the boxplots show a gap nobody would care
about, no p-value changes that ([rule 7](ten-rules.qmd)).

---

## Sampling Variability

**Sampling variability** is the fact that a different sample gives different
numbers, even when nothing about the game has changed. Some of the difference
between two groups is the build. Some of it is which players ended up in which
group: a few more veterans in one, a bad night's sleep in the other.

```{r}
#| label: sampling-variability
#| include: false
set.seed(2026)
resampled <- replicate(6, {
  a <- sample(sessions$wrong_turns[sessions$build == "A"], 15)
  b <- sample(sessions$wrong_turns[sessions$build == "B"], 15)
  mean(a) - mean(b)
})
```

To see it, pretend the *Tidebreak* team could only afford 15 sessions per build,
and draw six such playtests at random from the sessions they actually have. The
difference in mean wrong turns comes out as
`r paste(sprintf("%.1f", resampled), collapse = ", ")`. Same builds, same kind of
player, six different answers.

**Smaller samples swing further.** The fewer players you test, the more one
unusual player moves the result ([rule 8](ten-rules.qmd)). The statistics in the
rest of these notes exist to put a size on that swing.

---

## Estimates and Confidence Intervals

The **observed difference** is the gap between the groups in your sample: Build
A's mean minus Build B's. It is your best single guess at the true difference, and
it is almost certainly not exactly right.

A **confidence interval** is a range of plausible values for the true difference,
given the sample. A 95% confidence interval is built by a method that captures the
true value in 95% of the playtests you could have run.

**Read the interval as a range of sizes.** "Build B's players took
`r ci_fmt(tt_wt$conf.int)` fewer wrong turns" tells a designer both that the effect
is real and how big it might be. An interval that runs from almost nothing to something large says the
direction is clear and the size is not.

**Width is information.** A narrow interval means the sample pins the difference
down. A wide one means more players are needed before the size can be trusted.
More players, or less variation between them, narrows it.

**If the interval includes zero, "no difference" is still plausible.** That is the
link between confidence intervals and the test that follows.

---

## Hypothesis Testing

A hypothesis test asks one narrow question: if the builds really made no
difference, how surprising would this data be?

The **null hypothesis** is the claim of no difference: Build A and Build B produce
the same mean number of wrong turns. The **alternative hypothesis** is that they
do not. The test looks for evidence against the null.

The **p-value** is the probability of seeing a difference at least as large as the
one observed, *if the null hypothesis were true*. A small p-value means the data
would be surprising in a world where the builds did not differ.

::: {.callout-important}
**A p-value is not the probability that the null hypothesis is true.**

It is calculated by *assuming* the null is true, so it cannot also measure how
likely the null is. p = 0.03 does not mean "a 3% chance the builds are the same".
It means "if the builds were the same, a gap this large would turn up about 3% of
the time".
:::

**Statistical significance** means the p-value fell below a threshold chosen
before looking, conventionally 0.05. It means "unlikely to be chance alone" and
nothing more. Choose the threshold in advance for the same reason you decide what
would change your mind before looking at the data ([rule 1](ten-rules.qmd)).

**Not significant does not mean no difference.** A large p-value means this sample
could not rule out chance. That can be because there is no difference, or because
there were too few players to see one. The confidence interval tells you which.

---

## Statistical and Practical Significance

**Practical significance** is whether a difference is large enough to change a
design decision. Statistical significance cannot tell you this, because it depends
on how much data you have as well as on how big the difference is.

The *Tidebreak* frame-time capture has `r nrow(atrium)` frames from the Atrium
zone. In the Atrium, Build B's frames are `r sprintf("%.2f", abs(gap_atr))` ms
faster than Build A's, `r p_fmt(tt_atr$p.value)`. That is about as statistically
significant as a result gets, and a player would need to perceive a difference of
less than half a millisecond per frame to notice it. Nobody should schedule work
around it.

| | Practically small | Practically large |
| :- | :- | :- |
| **Statistically significant** | Real but not worth acting on | Act on it |
| **Not significant** | Nothing to see, if the interval is narrow | Test more players: the sample was too small to tell |

**Always report the size of the difference.** "Significant" alone hides which row
of that table you are in.

---

## Two Designs: Independent and Paired

How the playtest was run decides which test you use.

**Independent samples.** Different players test each build. That is how the
*Tidebreak* sessions file was collected: `r n_players` players, each of whom played
only one build. The comparison is between two separate groups of people, so
differences between the people are mixed in with differences between the builds.
Use the **independent-samples t-test**:

```r
t.test(wrong_turns ~ build, data = sessions)
```

**Paired.** The same players test both builds. `tidebreak_paired.csv` records a
second study in which `r nrow(paired)` players each played both builds. Each player
is compared with themselves, so their skill cancels out. Use the **paired
t-test**:

```r
t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)
```

**Player ability is the biggest source of variation in most playtests.** A
confident player takes few wrong turns in any build; a lost one takes many in any
build. In the paired data, a player's wrong turns in Build A and in Build B have a
correlation of `r round(cor(paired$wrong_turns_a, paired$wrong_turns_b), 2)`: the
same people do well or badly in both. The paired design removes that variation,
which is why it can detect a difference with far fewer players.

**Paired designs bring their own problems.** A player who plays both builds plays
the second one with more experience (**learning effects**), in a fixed position
(**order effects**) and more tired (**fatigue effects**). If everyone plays Build A
first, some of Build B's improvement is just practice. The fix is to give half the
players Build A first and half Build B first. `tidebreak_paired.csv` does not record
which build each player played first, so this study cannot rule it out.

---

## Assumptions, Practically

A t-test is reliable when these hold closely enough. None needs a formal test; each
needs a look.

| Assumption | What it means | How to check | *Tidebreak* |
| :- | :- | :- | :- |
| **Independence** | One observation does not influence another | Think about how the data was collected | Each player contributed `r n_per` sessions, so the `r nrow(sessions)` sessions are really `r n_players` players |
| **No strong outliers** | No few values dominate the means | Boxplots | `r sum(wt_out)` flagged wrong-turn counts across `r nrow(sessions)` sessions |
| **Approximate normality** | Each group roughly bell-shaped | A Q-Q plot | Close enough; counts are whole numbers, which shows as steps |
| **Appropriate design** | Independent test for separate groups, paired test for the same people | How the playtest was run | Sessions: independent. Paired file: paired |

**With a few dozen players per group, the normality assumption matters least.** The
t-test tolerates moderate departures. Strong outliers and a mismatched design
matter far more, and no amount of data fixes the wrong design.

**The independence problem has a simple check.** Average each player's four
sessions into one value, then compare players rather than sessions. If the
conclusion survives, the repeated sessions were not driving it. The R
Demonstrations below do exactly that.

---

## Worked Examples

**Did the signposting reduce wrong turns?** The six steps, in order.

*Visualise.*

```{r}
#| label: worked-visualise
#| fig-height: 3
ggplot(sessions, aes(x = build, y = wrong_turns)) +
  geom_boxplot(fill = "grey85") +
  labs(title = "Wrong turns per session, by build", x = "Build", y = "Wrong turns")
```

Build B's box sits lower and is taller. The boxplot flags `r wt_out[["A"]]`
outlying sessions in Build A and `r wt_out[["B"]]` in Build B, too few to dominate
either mean.

*Describe.* Build A's median is
`r median(sessions$wrong_turns[sessions$build == "A"])` wrong turns and Build B's
is `r median(sessions$wrong_turns[sessions$build == "B"])`. Build B's spread is
wider: an IQR of `r IQR(sessions$wrong_turns[sessions$build == "B"])` against
`r IQR(sessions$wrong_turns[sessions$build == "A"])`.

*Estimate the difference.* Build A's players averaged
`r round(tt_wt$estimate[[1]], 1)` wrong turns and Build B's
`r round(tt_wt$estimate[[2]], 1)`: a difference of `r round(gap_wt, 1)`.

*Confidence interval.* The 95% interval for the difference runs from
`r ci_fmt(tt_wt$conf.int)` wrong turns. It is nowhere near zero.

*Test.* `r p_fmt(tt_wt$p.value)`. If the builds made no difference, a gap this
size would almost never occur.

*Interpret.* The signposting reduced wrong turns, and the smallest plausible
reduction, `r round(tt_wt$conf.int[1], 1)` per session, is still large enough for a
player to feel. This is a result to act on: keep the signposting.

**Was Build B harder?** The same six steps on the 1–7 difficulty rating give a
second significant result. Build B was rated `r round(gap_dif, 2)` points harder,
`r p_fmt(tt_diff$p.value)`, with a confidence interval from
`r ci_fmt(ci_dif, 2)` points.

Three things say not to act on it. The size is small: under half a point on a
seven-point scale. The interval's lower end, `r round(ci_dif[1], 2)`, is a
difference no player would report. And the play data does not back it up: deaths
did not differ between the builds (`r p_fmt(tt_dth$p.value)`). Statistically
significant, practically marginal, and unsupported by what players actually did.
Note it, and check it in the next playtest.

---

## R Demonstrations

```{r}
#| label: demo-load
library(readr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
paired   <- read_csv("data/tidebreak_paired.csv", show_col_types = FALSE)
```

```{r}
#| label: demo-describe
tapply(sessions$wrong_turns, sessions$build, mean)
tapply(sessions$wrong_turns, sessions$build, sd)
```

Describe before testing. These are the numbers the test will compare.

```{r}
#| label: demo-ttest
result <- t.test(wrong_turns ~ build, data = sessions)
result
```

`t.test(y ~ group)` is the independent-samples t-test. R runs the Welch version by
default, which does not assume the two builds have the same spread, and that is the
right default. Read three lines of the output: the **sample estimates** (the two
means), the **95 percent confidence interval** (for Build A minus Build B) and the
**p-value**.

```{r}
#| label: demo-ttest-parts
result$estimate
result$conf.int
result$p.value
```

`$` pulls single parts out of the result, which is how you get numbers into a
sentence with inline R instead of copying them by hand.

```{r}
#| label: demo-paired
t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)
```

`paired = TRUE` compares each player's two values. The estimate is the **mean
difference**: on average, a player took `r round(tt_pair$estimate, 1)` fewer wrong
turns in Build B than in Build A.

```{r}
#| label: demo-paired-wrong
t.test(paired$wrong_turns_a, paired$wrong_turns_b)
```

The same data tested as if the players were different people. The estimate is the
same, but the interval is `r round(diff(tt_unp$conf.int), 1)` wrong turns wide
against `r round(diff(tt_pair$conf.int), 1)` for the paired test. Ignoring the
pairing throws away what the design bought you.

```{r}
#| label: demo-paired-plot
#| fig-height: 3
paired$improvement <- paired$wrong_turns_a - paired$wrong_turns_b

ggplot(paired, aes(x = improvement)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "Wrong turns saved per player, Build A minus Build B",
       x = "Improvement (wrong turns)", y = "Players")
```

For paired data, draw the differences. `r sum(d_pair > 0)` of `r length(d_pair)`
players took fewer wrong turns in Build B, `r sum(d_pair == 0)` took the same
number and `r sum(d_pair < 0)` took more. `geom_vline()` marks zero, the line
between better and worse.

```{r}
#| label: demo-qq
#| fig-height: 3
ggplot(sessions, aes(sample = wrong_turns)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ build)
```

A **Q-Q plot** places each value against where it would sit if the data were
perfectly normal. Points close to the line mean close to normal. Here both builds
follow it through the middle, in horizontal steps because wrong turns are whole
numbers. Build A's lowest few points drop away below the line: those are the
`r wt_out[["A"]]` sessions its boxplot flagged, players who barely got lost at all.
A handful of points bending off at one end is close enough for a t-test with
`r min(table(sessions$build))` sessions per build. A whole curve bending away from
the line would not be.

```{r}
#| label: demo-per-player
per_player <- aggregate(wrong_turns ~ player_id + build,
                        data = sessions, FUN = mean)
nrow(per_player)
t.test(wrong_turns ~ build, data = per_player)
```

The independence check. `aggregate()` averages each player's sessions into one
row, leaving `r nrow(per_player)` players instead of `r nrow(sessions)` sessions.
The difference is the same and the interval is a little wider, from
`r ci_fmt(tt_player$conf.int)`, because there are fewer independent observations.
The conclusion survives.

---

## Interpretation Guidance

**What the data shows.** In this playtest, Build B's players averaged
`r round(gap_wt, 1)` fewer wrong turns per session than Build A's (95% CI
`r ci_fmt(tt_wt$conf.int)`, `r p_fmt(tt_wt$p.value)`). In a separate study where
`r nrow(paired)` players played both builds, they took
`r round(tt_pair$estimate, 1)` fewer in Build B (95% CI
`r ci_fmt(tt_pair$conf.int)`).

**What the analysis suggests.** The signposting helps players find their way, and
by enough to matter. Two different designs, with different players, point the same
way.

**What can legitimately be concluded.** The reduction is very unlikely to be
chance, and even its smallest plausible size is a noticeable improvement. The
paired study cannot separate the signposting from practice, because play order was
not recorded; the independent study does not have that problem, which is why the
two together are stronger than either alone.

**Write the result as a sentence a designer can use.** Lead with the size, give
the interval, then the p-value:

> Build B's players took `r round(gap_wt, 1)` fewer wrong turns per session on
> average (95% CI `r ci_fmt(tt_wt$conf.int)`, `r p_fmt(tt_wt$p.value)`).

"The result was significant" is not a finding. It names no size and no action
([rule 10](ten-rules.qmd)).

---

## Common Mistakes

**Reading the p-value as the chance the null is true.** It is the chance of data
this extreme *if* the null is true. The two are different, and only the first one
is what R reports.

**Reading "not significant" as "no difference".** It means this sample could not
tell. Check the confidence interval: narrow and around zero is evidence of no
meaningful difference; wide is a sample too small to say.

**Reading "significant" as "important".** The Atrium frame times are among the
most significant results in the whole playtest and among the least important.

**Testing before looking.** A t-test on data you have not drawn can hide outliers,
skew or two groups mixed together. Visualise first, every time.

**Using the independent test on paired data.** It ignores what the pairing bought
you and gives an interval far wider than the data supports. Match the test to the
design.

**Testing every column and reporting what came up.** At a 0.05 threshold, roughly
one test in twenty on builds that do not differ will come up significant anyway.
Decide which comparisons answer your question before you run any of them.

**Counting sessions as players.** Four sessions from one player are not four
independent players. Check that the result survives when each player counts once.

---

## Short Practice Activities

Each takes two or three minutes.

**1.** Did the signposting cost enjoyment? Run the t-test on `enjoyment` by build
and use the confidence interval, not only the p-value, to answer.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-enjoyment
t.test(enjoyment ~ build, data = sessions)
```

The p-value is `r round(tt_enj$p.value, 2)`, so there is no evidence of a
difference. The interval says more: it runs from `r ci_fmt(tt_enj$conf.int, 2)`
points on a seven-point scale, so any true difference is smaller than half a
point either way. That is evidence the signposting did not cost enjoyment, which
is a stronger statement than "not significant".
:::

**2.** Run the paired data through both the paired and the unpaired t-test. Why is
the paired interval narrower?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-paired
t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)$conf.int
t.test(paired$wrong_turns_a, paired$wrong_turns_b)$conf.int
cor(paired$wrong_turns_a, paired$wrong_turns_b)
```

Players who take many wrong turns in one build take many in the other: the
correlation is `r round(cor(paired$wrong_turns_a, paired$wrong_turns_b), 2)`. The
paired test compares each player with themselves, so that player-to-player
variation drops out. The unpaired test leaves it in and has to allow for it.
:::

**3.** Build B's sessions were longer on average. Run the test on
`session_minutes`. Is the difference statistically significant? Is it practically
significant?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-minutes
t.test(session_minutes ~ build, data = sessions)
```

```{r}
#| label: practice-minutes-numbers
#| include: false
tt_min <- t.test(session_minutes ~ build, data = sessions)
ci_min <- -rev(tt_min$conf.int)
```

Statistically, yes: `r p_fmt(tt_min$p.value)`. Practically, it is hard to say. The
interval says Build B's sessions were `r ci_fmt(ci_min)` minutes longer:
anywhere from `r round(ci_min[1] * 60)` seconds to `r round(ci_min[2], 1)`
minutes. The direction is clear and the size
is not, which is a case for a larger playtest rather than a decision.
:::

---

## Summary

- A playtest is a sample. Another group of players would give different numbers,
  so every difference needs a measure of how far it could swing.
- Work in order: visualise, describe, estimate the difference, compute the
  confidence interval, test, interpret.
- A confidence interval gives a range of plausible sizes. Report it with every
  difference.
- A p-value is how surprising the data would be if there were no difference. It is
  not the probability that there is no difference.
- Statistically significant means unlikely to be chance. Practically significant
  means big enough to act on. You need both.
- Different players on each build: independent-samples t-test. The same players on
  both: paired t-test, with the play order varied.
- The signposting reduced wrong turns by `r round(gap_wt, 1)` per session, and the
  smallest plausible reduction is still worth having.

---

## Links to Relevant Reference Material

- [Ten Rules](ten-rules.qmd): rule 1 (decide what would change your mind before you
  look), rule 7 (compare distributions, not averages), rule 8 (small samples lie
  confidently) and rule 10 (a finding that does not name an action is not
  finished)
- [Statistics Reference: from sample to population](../reference/reference-statistics.qmd#from-sample-to-population)
  and [testing a difference](../reference/reference-statistics.qmd#testing-a-difference)
- [Visualising with ggplot2](visualising-with-ggplot2.qmd): the boxplots and
  faceted histograms that come before every test
- [ggplot2 Reference](../reference/reference-ggplot2.qmd)

---

## Reflective Questions

- Build B's difficulty rating was significantly higher, but deaths did not change.
  What would you need to see in the next playtest to take the difficulty result
  seriously?
- A colleague says the paired study proves the signposting works because its
  p-value is smaller. What is wrong with that argument, and what does the paired
  study add that the sessions data does not?
- You can afford 30 testers for the next build comparison. Would you run them as
  two independent groups of 15 or as one paired group of 30? What would you have to
  control in the paired design?
- A producer asks, "is it significant?". What do you tell them instead, and why is
  that a more useful answer?

---
