---
title: "Describing What You Have"
subtitle: "Variable types, tidy data, centre, spread and shape"
---

> Week 1, for everyone, before Part B of Lab 01. This is the vocabulary for
> describing a single variable: what kind of number it is, where its centre sits,
> how spread out it is and what shape it has. Every later page assumes it.

---

## Learning Outcomes

By the end of these notes you can:

- classify a variable as nominal, ordinal, discrete or continuous, and say which
  operations that allows;
- recognise a tidy table and name the most common way data fails to be one;
- say what a sample is, and why every number you compute describes the sample
  rather than your audience;
- choose between mean and median, and read the gap between them as evidence of
  skew;
- describe spread with the standard deviation, quartiles and IQR, and say why the
  range is rarely enough;
- draw a histogram and a boxplot in base R and say what each shows that a summary
  hides.

---

## Why This Matters for Playtesting

Before comparing anything to anything, you have to be able to say what a single
variable looks like. This section is the vocabulary for that, and it is the part
students most often skip and most often get caught by.

---

## What Kind of Number Is This?

Four types, and the type decides which operations are legal.

**Nominal** — labels with no order. Character class, platform, quit reason. You
can count them. You cannot average them. The mean of `Scout` and `Sentinel` does
not exist, and if your code produces one, something has gone wrong upstream.

**Ordinal** — ordered labels with unknown gaps. Difficulty setting, player
experience level, and every agreement scale ever printed. Novice, Regular and
Veteran are in an order, but Regular is not "halfway" between the other two in any
meaningful sense.

**Discrete** — counts. Deaths, kills, items collected. Whole numbers, and there is
no such thing as 2.4 deaths in a single session — though the *average* number of
deaths across many sessions is perfectly meaningful at 2.4.

**Continuous** — measurements. Completion time, frame time, distance travelled.
Any value in a range, limited only by the precision of your instrument.

**The awkward case is the rating scale**, and it is awkward every time. A 1–5
agreement scale is strictly **ordinal**: you know Agree is above Neutral, but you
do not know that the step from Disagree to Neutral is the same size as the step
from Agree to Strongly Agree. In practice, when a scale has five or more points
and the responses are reasonably spread, people treat it as numeric and average
it. That convention is defensible and universal, and you should still know you are
making an assumption. Where it matters — where the whole finding rests on a mean
of 3.2 against 3.4 — report the distribution instead.

---

## Tidy Data: One Row Per Observation

Data arrives in whatever shape the tool that produced it happened to use. Almost
all analysis assumes one particular shape, and it is worth naming.

A table is **tidy** when each row is one observation, each column is one variable,
and each cell is one value. For a playtest: one row per player per session, one
column per thing you recorded.

The most common violation is a survey export, where the eight questions become
eight columns. That is convenient to read and inconvenient to analyse, because
"question" is really a variable and its eight values have been spread across the
column headings. Reshaping it into one row per player *per question* is what makes
the whole survey analysable in one operation, and you will do exactly that in
Lab 03.

---

## Sample and Population

```{r}
#| label: data
#| include: false
library(readr)
sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
survey   <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)
```

The `r nrow(sessions)` *Tidebreak* sessions come from
`r length(unique(sessions$player_id))` players who happened to attend this
playtest. They are a **sample**. The **population** is everybody you are making
the game for, and you never get to measure it.

**Every number describes the sample.** The question you care about is always about
the population: not "did these players struggle?" but "will players struggle?".
Bring in a different group next week and every figure on this page moves. How far
it can move is the subject of Week 2; until then, report how many people a number
came from every time you quote it ([rule 8](ten-rules.qmd)).

**Who is in the sample matters as much as how many.** A playtest recruited from the
studio's own Discord samples the game's most committed fans, not its audience. No
amount of analysis afterwards corrects for who walked in the door.

---

## Centre

Two summaries of "the typical value", and choosing between them is a real
decision, not a formality.

The **mean** adds everything and divides by the count. It uses every value, which
is its strength, and it is dragged by extremes, which is its weakness.

The **median** is the middle value when sorted. Half above, half below. It ignores
how extreme the extremes are, which is its strength and its weakness.

**When they disagree, the data is skewed, and the gap tells you which way.** If
the mean sits well above the median, a few large values are pulling it up: a long
right tail. If it sits below, a group of small values is pulling it down.

For game data this is not an edge case, it is the norm. Completion times, session
lengths and frame times are all skewed almost by construction — there is a floor
on how fast anyone can finish and no ceiling on how slow. **Default to the median
for anything time-like** and reach for the mean only when you have looked at the
shape and satisfied yourself it is roughly symmetric.

When you compare the two, compare them relative to the size of the variable. A gap
of 40 points on a score in the thousands is nothing; a gap of 3 minutes on a median
of 26 is a real skew.

---

## Spread

The centre alone is nearly useless. Two levels with identical average completion
times, one where everybody takes four minutes and one where half take one minute
and half take seven, are not the same level.

**Range** is minimum to maximum. Easy to quote and almost always misleading, since
a single unusual value defines it entirely.

**Standard deviation** is the typical distance from the mean. It is the right
measure for roughly symmetric data and it inherits the mean's sensitivity to
extremes.

**Quartiles** split the sorted data into four. The first quartile has a quarter
below it, the third has three quarters. The **interquartile range** — the distance
between them — is the width of the middle half of your players, and it is the most
robust description of spread you have. For skewed game data, quote the median and
the IQR together.

**Percentiles** generalise this: the 95th percentile has 95% of values below it.
[Distributions and Variation](distributions-and-variation.qmd) will show you why
the 99th percentile of frame time is the single most
important number in a performance capture.

---

## Look at the Shape First

Every summary statistic throws information away. That is what a summary is for.
The risk is that it throws away the finding.

A **histogram** shows you the whole distribution: where the bulk sits, whether it
is symmetric, whether there is a second hump, whether there are outliers. Draw one
before quoting any average. It costs one line.

The number of bins changes what you see, sometimes dramatically. Too few and a
second hump disappears; too many and noise looks like structure. Try several. That
a chart changes with its settings is not a flaw to hide — it is the first evidence
that a chart is an argument rather than a fact, and you are responsible for the
argument you make.

A **boxplot** compresses the same information into median, quartiles and outliers.
It is less informative than a histogram for one variable and much better for
comparing several groups side by side, which is what Section 4 is about.

**Two humps are always worth chasing.** A distribution with two peaks usually means
two populations have been mixed: players who found the key and players who did
not, sessions that completed and sessions that crashed. The average of a bimodal
distribution describes nobody at all — it names a value that sits in the valley
between the two groups, where the fewest players actually are.

---

## Worked Examples

```{r}
#| label: worked-sentinel
#| include: false
arch_mean   <- tapply(sessions$score, sessions$archetype, mean)
arch_median <- tapply(sessions$score, sessions$archetype, median)
arch_sd     <- tapply(sessions$score, sessions$archetype, sd)
stopifnot(names(which.min(arch_mean)) == "Sentinel",
          names(which.max(arch_sd))   == "Sentinel")

rank_word  <- c("highest", "second-highest", "third-highest", "lowest")
sen_median <- rank_word[rank(-arch_median)[["Sentinel"]]]

sentinel <- sessions$score[sessions$archetype == "Sentinel"]
others   <- arch_sd[names(arch_sd) != "Sentinel"]

obj_low <- sum(survey$objective_clear <= 2)
obj_top  <- sum(survey$objective_clear == 5)
```

**Is the Sentinel underpowered?** Sort the four archetypes by mean score and the
Sentinel comes last, at `r round(arch_mean[["Sentinel"]])` points. A balance pass
built on that number buffs the Sentinel.

Sort by median instead and the Sentinel is the `r sen_median`, at
`r round(arch_median[["Sentinel"]])`. The mean and the median disagree by
`r round(arch_median[["Sentinel"]] - arch_mean[["Sentinel"]])` points, and the
standard deviation is `r round(arch_sd[["Sentinel"]])` against
`r round(min(others))` to `r round(max(others))` for the other three. Something is
dragging the mean down without moving the middle.

```{r}
#| label: worked-sentinel-hist
hist(sentinel,
     breaks = 12,
     main   = "Sentinel score, one bar per band",
     xlab   = "Score",
     col    = "grey85",
     border = "white")
```

The histogram answers it. `r sum(sentinel < 700)` of the `r length(sentinel)`
Sentinel sessions scored under 700 and sit in a clump of their own at the far left,
`r min(sentinel[sentinel >= 700]) - max(sentinel[sentinel < 700])` points below the
next-lowest Sentinel game. The Sentinel either anchors the round or collapses early.
It is not underpowered; it is inconsistent. A flat buff raises the scores of the
players already doing well, and the design question becomes what the collapsing
games have in common ([rule 4](ten-rules.qmd)).

**How clear was the objective?** The survey item "I always knew what my objective
was" averages `r round(mean(survey$objective_clear), 2)` on a 1 to 5 scale. That is
a mean of an ordinal rating, and the distribution says more:

```{r}
#| label: worked-objective
table(survey$objective_clear)
```

`r obj_low` of `r nrow(survey)` players chose 1 or 2, and `r obj_top` chose 5.
"`r round(100 * obj_low / nrow(survey))`% of players disagreed that they always
knew their objective" is a sentence a designer acts on. "Mean
`r round(mean(survey$objective_clear), 2)`" is not ([rule 3](ten-rules.qmd)).

---

## R Demonstrations

All base R. The variable here is `time_alive_s`, seconds survived in the match.

```{r}
#| label: demo-centre-spread
summary(sessions$time_alive_s)
sd(sessions$time_alive_s)
IQR(sessions$time_alive_s)
quantile(sessions$time_alive_s, c(0.1, 0.9))
```

`summary()` gives centre and quartiles in one call; `sd()` and `IQR()` give the two
measures of spread; `quantile()` with a vector of proportions gives any percentiles
you ask for, here the 10th and 90th.

```{r}
#| label: demo-hist
hist(sessions$time_alive_s,
     breaks = 15,
     main   = "Time alive per session",
     xlab   = "Seconds",
     col    = "grey85",
     border = "white")
```

`breaks` sets roughly how many bars. Change it and look again before believing any
shape.

```{r}
#| label: demo-table
table(sessions$cohort)
```

`table()` counts a categorical variable. It is the only summary a nominal variable
has.

```{r}
#| label: demo-by-group
tapply(sessions$time_alive_s, sessions$cohort, median)
boxplot(time_alive_s ~ cohort,
        data = sessions,
        main = "Time alive by cohort",
        ylab = "Seconds",
        col  = "grey85")
```

`tapply()` applies one function to one variable, split by a group. The boxplot
shows the same split as a picture: the thick line is the median, the box is the
middle half, and the whiskers and points show the rest.

---

## Interpretation Guidance

**A description has four parts: centre, spread, shape and size.** One sentence can
hold all four:

> Median time alive was `r median(sessions$time_alive_s)` seconds (IQR
> `r IQR(sessions$time_alive_s)`, n = `r nrow(sessions)`). The mean sits
> `r round(abs(mean(sessions$time_alive_s) - median(sessions$time_alive_s)), 1)`
> seconds from the median, so the centre is not being pulled by a tail.
> `r sum(sessions$time_alive_s == max(sessions$time_alive_s))` sessions sit at
> exactly `r max(sessions$time_alive_s)` seconds, the full match: those players
> never died, and the tallest bar on the right is a ceiling, not a peak.

**Pick the summary to fit the shape, and say why.** Median and IQR for anything
skewed or with outliers; mean and standard deviation only when the histogram is
roughly symmetric.

**Keep what the data shows apart from what it suggests.** "The Sentinel's scores
fall into two groups" is what the data shows. "Some Sentinel players do not know
how to play the role" is a suggestion, and it needs a different kind of evidence
before it becomes a conclusion.

---

## Common Mistakes

**Averaging a label.** `mean(sessions$cohort)` returns `NA` with a warning, which
is R protecting you. Code the cohorts as 1, 2 and 3 and R will average them
happily, and the answer will mean nothing.

**Trusting R's idea of the type.** R stores `cohort` as text and every survey item
as a number. Neither is wrong, but neither tells R that cohorts come in an order or
that a rating is ordinal. The type R prints is where your checking starts
([rule 3](ten-rules.qmd)).

**Quoting a mean without looking.** The Sentinel's mean is a correct calculation
that describes no Sentinel game. Draw the histogram first
([rule 4](ten-rules.qmd)).

**Quoting the range as the spread.** One extreme player sets the range on their
own. Quote the IQR.

**Trusting one bin setting.** A histogram drawn once is one argument among several
the data could make. Redraw it with fewer and more bars.

**Leaving out n.** "Players rated it 2.6" means nothing without "out of how many"
([rule 8](ten-rules.qmd)).

---

## Short Practice Activities

Each takes two or three minutes.

**1.** Classify each of these as nominal, ordinal, discrete or continuous: `kills`,
`quit_reason`, `cohort`, `distance_m`, `visual_style`.

::: {.callout-note collapse="true" title="Check your answer"}
`kills` discrete; `quit_reason` nominal; `cohort` ordinal; `distance_m`
continuous; `visual_style` ordinal (a 1 to 5 rating), whatever type R gives it.
:::

**2.** Describe `distance_m` in one sentence with centre, spread and size. Does the
mean sit above or below the median?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-distance
median(sessions$distance_m)
IQR(sessions$distance_m)
mean(sessions$distance_m)
```

Median distance was `r median(sessions$distance_m)` m (IQR
`r IQR(sessions$distance_m)`, n = `r nrow(sessions)`). The mean sits
`r if (mean(sessions$distance_m) > median(sessions$distance_m)) "above" else "below"`
the median by
`r round(abs(mean(sessions$distance_m) - median(sessions$distance_m)), 1)` m, which
is small against a spread of that size.
:::

**3.** Use `table()` to find the most common number of deaths in a session. Why is
`table()` sensible for `deaths` but not for `distance_m`?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-deaths
table(sessions$deaths)
```

The most common count is `r names(which.max(table(sessions$deaths)))` deaths. Deaths
is discrete with only a handful of possible values, so counting each one is
informative. `r length(unique(sessions$distance_m))` of the `r nrow(sessions)`
`distance_m` values are different, so its table is mostly a list of ones; a
continuous variable needs a histogram instead.
:::

---

## Summary

- The type of a variable decides which summaries are legal. A 1 to 5 rating is
  ordinal, and averaging it is an assumption you should know you are making.
- Tidy data has one row per observation and one column per variable.
- Every number describes the sample. Report n every time.
- When mean and median disagree, the data is skewed. Default to the median for
  anything time-like.
- Describe spread with the IQR for skewed data and the standard deviation for
  symmetric data. The range is rarely enough.
- Look at the histogram before quoting any average. Two humps mean two groups.

---

## Links to Relevant Reference Material

- [Ten Rules](ten-rules.qmd): rule 3 (know what kind of number you are holding),
  rule 4 (look at the shape before you quote the average), rule 5 (the mean is the
  wrong summary more often than you think) and rule 8 (small samples lie
  confidently)
- [R and Quarto Reference: describing one variable](../reference/r-quarto-reference.qmd#describing-one-variable),
  [describing by group](../reference/r-quarto-reference.qmd#describing-by-group)
  and [plots](../reference/r-quarto-reference.qmd#plots)
- [Lab 01: exploring playtest data](../labs/reading-data.qmd), Part B

---

## Reflective Questions

- Take a variable from your own project's telemetry. Which of the four types is
  it, and what would break if you treated it as the type next door?
- Why is the median usually the right summary for completion time, but not for
  something like accuracy percentage?
- You draw a histogram of level completion times and see two clear humps. List
  three things that could have caused that, and say how you would tell them apart.
- A teammate reports "average session length is 22 minutes". What two further
  numbers would you ask for before letting that into a report?

---
