---
title: "Distributions and Variation"
subtitle: "Why the average is never the whole story"
---

## Learning Outcomes

By the end of these notes you can:

- explain why an average on its own cannot describe a playtest;
- read a standard deviation as the typical distance of a player from the mean;
- describe a distribution by its centre, spread, skew and outliers;
- say whether an outlier is an error, an extreme player or the finding itself;
- read the mean, median, standard deviation and high percentiles together and say
  what each adds;
- draw `hist()`, `boxplot()`, `barplot()` and `plot()` in base R and say what each
  one is for.

---

## Why This Matters for Playtesting

Every design question is really a question about players, plural, and players
differ. A level that takes eight minutes on average can be a level everyone
finishes in eight minutes, or a level half the testers finish in four and the other
half abandon. Those are different games with different fixes, and the average
cannot tell them apart.

**The players you design for are rarely the average player.** The ones who quit,
the ones who get stuck, the ones whose machine stutters: they live in the edges of
the distribution, and the edges are exactly what a mean averages away.

**Variation is information, not noise.** When two builds differ in how spread out
their players are, that difference is often the finding. A build that tightens the
spread has made the experience more consistent, even if the average has not moved.

---

## Variation Between Players

The **variation** in a variable is how much its values differ from one another.
Some of it is the game: a harder route produces more deaths. Some of it is the
player: skill, experience and patience vary before anyone touches your build. You
cannot remove the second kind, so you have to measure it.

The **standard deviation** is the typical distance of a value from the mean. A
small standard deviation means players cluster tightly around the average; a large
one means the average is a poor guide to any individual. It is in the same units as
the data, so a standard deviation of 3 ms on a frame time is directly comparable to
the 16.7 ms a 60 FPS frame is allowed.

**Standard deviation belongs to the mean.** It measures distance from the mean, so
it inherits the mean's weakness: a few extreme values inflate it. For skewed data,
the interquartile range from the previous notes is the better description of
spread.

---

## Distributions and Skew

A **distribution** is the full pattern of values: which ones occur, and how often.
A histogram draws it. Every summary statistic is an attempt to describe a
distribution in one number, and every one of them fails for some distribution.

**Skew** is a distribution leaning to one side. A **right skew** has a long tail of
large values: most frames are quick, a few are very slow. A **left skew** has a
long tail of small values: most players stay for the full session, a few leave
early. The mean is pulled towards the tail; the median is not. That is why the gap
between them tells you which way the data leans.

**Game data is rarely symmetric.** Times have a floor and no ceiling. Match lengths
have a ceiling set by the timer. Ratings bunch against the ends of the scale. Expect
skew and check for it, rather than assuming the bell shape from school statistics.

---

## Outliers

An **outlier** is a value far from the rest. The boxplot's convention is any point
more than one and a half box-widths beyond the edge of the box, and R draws those
points individually. That rule is a prompt to look, not a verdict.

An outlier is one of three things, and the response to each is different:

| The outlier is | Example | What you do |
| :- | :- | :- |
| An error | A session length of −3 minutes; a rating of 7 on a 1–5 scale | Fix it or remove it, and record that you did |
| An extreme but ordinary player | One tester who never died | Keep it; report the median rather than the mean |
| The finding | A cluster of games that collapse early | Keep it and investigate it; it is why you ran the test |

**Never delete an outlier because it is inconvenient.** Every row you drop is a
decision with a consequence ([rule 6](ten-rules.qmd)), and the value most tempting
to drop is often the one the playtest was run to find ([rule 9](ten-rules.qmd)).

---

## Reading Several Summaries Together

No single statistic describes a distribution, but a small set of them, read
together, usually does.

| Summary | What it tells you | What it misses |
| :- | :- | :- |
| Mean | The total spread evenly across players | Skew, clusters, tails |
| Median | The middle player | How extreme the extremes are |
| Standard deviation | Typical distance from the mean | Which direction the spread goes |
| IQR | Width of the middle half | Everything outside the middle half |
| 95th or 99th percentile | How bad the worst few per cent get | The typical case |

**Where the summaries disagree, look at the picture.** Mean and median apart means
skew. A large standard deviation with a small IQR means a few extreme values. A
normal median with a bad 99th percentile means a rare problem that the average will
never show. Each disagreement is a pointer to a specific shape, and the histogram
confirms which one ([rule 4](ten-rules.qmd)).

---

## Worked Examples

```{r}
#| label: worked-example-data
#| include: false
library(readr)
frames   <- read_csv("data/tidebreak_frame_times.csv", show_col_types = FALSE)
sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)

a <- frames$ms[frames$build == "A"]
b <- frames$ms[frames$build == "B"]

fps <- function(x) round(1000 / mean(x), 1)
pct <- function(x, q) round(unname(quantile(x, q)), 1)

slow   <- 1000 / 30
slow_b <- frames[frames$build == "B" & frames$ms > slow, ]
gap    <- unique(diff(sort(slow_b$frame)))

stopifnot(mean(b) < mean(a), median(b) < median(a),
          quantile(b, 0.95) < quantile(a, 0.95),
          sum(a > slow) == 0,
          all(slow_b$zone == "Corridor"),
          length(gap) == 1)

corridor_b <- frames$ms[frames$build == "B" & frames$zone == "Corridor"]
```

**Which build runs better?** The *Tidebreak* team captured `r length(a)` frames
from each build, split across `r length(unique(frames$zone))` level zones. The frame-rate counter reports
Build A at `r fps(a)` FPS and Build B at `r fps(b)` FPS. By the number everyone
quotes, Build B is marginally the faster build.

The median agrees: `r round(median(a), 2)` ms per frame for Build A,
`r round(median(b), 2)` ms for Build B. So does the 95th percentile:
`r pct(a, 0.95)` ms against `r pct(b, 0.95)` ms. Three summaries, one verdict.

The 99th percentile disagrees. Build A's is `r pct(a, 0.99)` ms; Build B's is
`r pct(b, 0.99)` ms, `r sprintf("%.1f", quantile(b, 0.99) / quantile(a, 0.99))` times
as long. The standard deviation says the same thing more quietly:
`r round(sd(a), 2)` ms for Build A, `r round(sd(b), 2)` ms for Build B.

Count the frames slower than 30 FPS (`r round(slow, 1)` ms). Build A has
`r sum(a > slow)`. Build B has `r nrow(slow_b)`, every one of them in the Corridor
zone, and exactly `r gap` frames apart: at 60 FPS, a hitch
`r round(60 / gap, 1)` times a second for as long as the player is in the corridor.

```{r}
#| label: worked-frame-trace
corridor_a <- frames$ms[frames$build == "A" & frames$zone == "Corridor"]

plot(corridor_b[1:200], type = "l", col = "firebrick",
     ylim = c(10, 65),
     main = "Corridor frame times, first 200 frames",
     xlab = "Frame", ylab = "Frame time (ms)")
lines(corridor_a[1:200], col = "grey40")
abline(h = 1000 / 60, lty = 2)
legend("top", legend = c("Build A", "Build B"), horiz = TRUE,
       col = c("grey40", "firebrick"), lty = 1, bty = "n")
```

**Why the mean cannot see it.** The slow frames are
`r round(100 * nrow(slow_b) / length(b), 1)`% of Build B's total. Spread across
every frame of the capture, they barely move the mean, and Build B's faster frames
elsewhere more than make up for them. The 95th percentile sits below the problem
because the problem is rarer than one frame in twenty. Only a summary aimed at the
worst one per cent can see it, which is why the 99th percentile is the standard
figure in performance capture. The mean is not approximately wrong here. It is
structurally incapable of showing the thing players experience.

---

## R Demonstrations

Base R first, then the tidyverse equivalent for the grouped summary.

```{r}
#| label: demo-load
library(readr)
library(dplyr)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
frames   <- read_csv("data/tidebreak_frame_times.csv", show_col_types = FALSE)
```

```{r}
#| label: demo-hist
hist(sessions$session_minutes, breaks = 20,
     main = "Session length", xlab = "Minutes",
     col = "grey85", border = "white")
```

`hist()` draws the distribution of one numeric variable. Look for the bulk, the
tails, and any second hump.

```{r}
#| label: demo-box-build
boxplot(session_minutes ~ build, data = sessions,
        main = "Session length by build", ylab = "Minutes", col = "grey85")
tapply(sessions$session_minutes, sessions$build, sd)
tapply(sessions$session_minutes, sessions$build, IQR)
```

`boxplot()` with `y ~ group` puts one box per group on the same axis, which is the
quickest way to compare distributions. Build A's box is wider: its IQR is
`r round(IQR(sessions$session_minutes[sessions$build == "A"]), 1)` minutes against
`r round(IQR(sessions$session_minutes[sessions$build == "B"]), 1)` for Build B.
`r round(100 * mean(sessions$session_minutes[sessions$build == "A"] >= 11))`% of
Build A sessions were still going at minute 11, and
`r round(100 * mean(sessions$session_minutes[sessions$build == "A"] >= 13))`% at
minute 13. The spread is where Build A loses its players.

```{r}
#| label: demo-barplot
barplot(table(sessions$quit_reason),
        main = "How sessions ended", ylab = "Sessions", col = "grey85")
```

`barplot()` draws counts of a categorical variable. Give it the output of
`table()`, not the raw column.

```{r}
#| label: demo-scatter
plot(sessions$deaths, sessions$score,
     main = "Deaths against score", xlab = "Deaths", ylab = "Score",
     pch = 19, col = "grey40")
```

`plot()` with two numeric variables draws a scatterplot: one point per session.
Read it for direction and for scatter. How strongly two variables are related is
Week 3's subject.

```{r}
#| label: demo-quantile
quantile(frames$ms[frames$build == "B"], c(0.5, 0.95, 0.99))
```

`quantile()` with a vector of proportions returns any percentiles you ask for. For
performance data, always ask for the 99th.

```{r}
#| label: demo-summarise
frames %>%
  group_by(build) %>%
  summarise(mean_ms   = mean(ms),
            median_ms = median(ms),
            sd_ms     = sd(ms),
            p99_ms    = quantile(ms, 0.99),
            fps       = 1000 / mean(ms))
```

`group_by()` splits the data and `summarise()` computes one row of summaries per
group. This is the several-summaries-together table from the worked example in six
lines, and it is the base R `tapply()` repeated five times.

---

## Interpretation Guidance

**What the data shows.** Build A and Build B have near-identical mean frame times.
Build B has `r nrow(slow_b)` frames slower than 30 FPS, all in the Corridor, and
Build A has `r sum(a > slow)`.

**What the analysis suggests.** Something in Build B's Corridor costs a frame at a
fixed interval. The regular spacing points at a periodic job, such as streaming,
garbage collection or a timed effect, rather than at scene complexity.

**What can legitimately be concluded.** Build B introduced a hitch confined to the
Corridor, and a report that quotes average FPS alone would have shipped it. The
cause is not established by frame times; that needs a profiler.

**Quote the tail with the centre.** For anything where the worst case is what
players remember, such as frame time, load time or matchmaking wait, report the
median and a high percentile together. "Median `r round(median(b), 1)` ms, 99th
percentile `r round(quantile(b, 0.99))` ms" is a performance report.
"`r round(1000 / mean(b))` FPS" is a slogan.

---

## Common Mistakes

**Reporting average FPS as the performance result.** It is the number the engine
shows and the number least able to show a hitch.

**Comparing groups by their means alone.** Two builds with the same mean can be
different games ([rule 7](ten-rules.qmd)). Draw the boxplots side by side.

**Deleting outliers to clean the chart.** Find out which of the three kinds each one
is first. The Sentinel's low scores in
[Describing What You Have](describing-what-you-have.qmd) are the finding, not the
noise.

**Quoting a standard deviation for skewed data.** It measures distance from the
mean, and on skewed data the mean is in the wrong place. Use the IQR.

**Stopping at the 95th percentile.** A problem affecting one frame in a hundred is
invisible at the 95th. Match the percentile to how rare the problem could be.

**Reading a scatterplot as proof.** A downward cloud of points shows that two things
move together in this sample. It does not say which one causes the other.

---

## Short Practice Activities

Each takes two or three minutes.

**1.** Compute the 99th percentile of frame time for each zone in Build B. Which
zone is responsible?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-zone
frames_b <- frames[frames$build == "B", ]
tapply(frames_b$ms, frames_b$zone, quantile, 0.99)
```

The Corridor, at `r round(quantile(corridor_b, 0.99), 1)` ms. The other two
zones are at most
`r round(max(tapply(frames_b$ms, frames_b$zone, quantile, 0.99)[c("Atrium", "Gantry")]), 1)` ms. Splitting a summary by where it happened turns "Build B stutters"
into "Build B stutters in one place", which is a ticket someone can fix.
:::

**2.** Redraw the quit-reason bar chart with one bar per build side by side. What
does it add to the single bar chart?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-quit-build
barplot(table(sessions$build, sessions$quit_reason),
        beside = TRUE, legend.text = TRUE,
        main = "How sessions ended, by build", ylab = "Sessions")
```

`r sum(sessions$build == "A" & sessions$quit_reason == "Frustrated")` Build A
sessions ended in frustration against
`r sum(sessions$build == "B" & sessions$quit_reason == "Frustrated")` in Build B.
The single chart showed how often players left frustrated; the split shows which
build they left.
:::

**3.** Using `tapply()`, find the median score for each number of deaths. Does the
pattern in the scatterplot hold up?

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-deaths-score
tapply(sessions$score, sessions$deaths, median)
```

Broadly, yes: sessions with no deaths have a median score of
`r median(sessions$score[sessions$deaths == 0])`, and sessions with
`r max(sessions$deaths)` deaths have
`r median(sessions$score[sessions$deaths == max(sessions$deaths)])`. Check how
many sessions sit behind each median with `table(sessions$deaths)` before trusting
the ends of that list ([rule 8](ten-rules.qmd)).
:::

---

## Summary

- An average describes a group; it does not describe any player in it.
- The standard deviation is the typical distance from the mean. Use the IQR when
  the data is skewed.
- Skew pulls the mean towards the tail. Game data is rarely symmetric.
- An outlier is an error, an extreme player or the finding. Work out which before
  doing anything to it.
- Read several summaries together, and where they disagree, draw the picture.
- For performance data, quote a high percentile. Both builds run at about
  `r round(1000 / mean(frames$ms))` FPS, and only one of them stutters.

---

## Links to Relevant Reference Material

- [Ten Rules](ten-rules.qmd): rule 4 (look at the shape before you quote the
  average), rule 7 (compare distributions, not averages) and rule 9 (sometimes the
  rare value is the whole story)
- [Describing What You Have](describing-what-you-have.qmd): centre, spread and the
  mean against the median
- [R and Quarto Reference: describing one variable](../reference/r-quarto-reference.qmd#describing-one-variable),
  [describing by group](../reference/r-quarto-reference.qmd#describing-by-group)
  and [plots](../reference/r-quarto-reference.qmd#plots)
- [Lab 01: exploring playtest data](../labs/reading-data.qmd), Part B

---

## Reflective Questions

- A producer asks for "one number" to show the build's performance has improved.
  Which number do you give, and what do you say about the numbers you left out?
- Build A's session lengths are more spread out than Build B's. Is a narrower spread
  always better for a game? Give a case where it would not be.
- You find three sessions with a score of zero. What would you check before
  deciding they are errors, and what would change your mind?
- The Corridor hitch affects
  `r round(100 * nrow(slow_b) / length(b), 1)`% of Build B's frames. At what point does a rare
  problem stop being worth fixing, and who should decide?

---
