---
title: "Why Developers Measure Anything"
subtitle: "And what is actually worth measuring"
---

## Learning Outcomes

By the end of these notes you can:

- name the three cases where a team's judgement of its own game reliably fails;
- distinguish telemetry, performance capture and asking people, and say what each
  can and cannot tell you;
- rewrite a vague design question as a measure you could record for a single
  player;
- write down, before collecting anything, the result that would change your mind;
- load a playtest CSV into R and take a first look at it with `read_csv()`,
  `glimpse()` and `summary()`.

---

## Why This Matters for Playtesting

Every team makes decisions about their game constantly. Most of those decisions
are made from memory, taste and the loudest voice in the room, and most of the
time that is fine — taste is why anyone plays your game rather than a spreadsheet.

Measurement earns its place in the cases where taste reliably fails. There are
three of them, and they recur in every project.

**You cannot see your own game any more.** You have played your level five hundred
times. You know where the key is. You cannot un-know it, and no amount of care
will let you experience your own tutorial as a stranger. This is the single most
common cause of a game that the team finds obvious and the player finds
impenetrable.

**You cannot feel a rare event.** A stutter that happens once every two seconds
occupies less than one per cent of frames. Your average frame rate will not show
it. Your players will talk about nothing else.

**You cannot remember twenty playtests.** After a session you remember the two
people who struggled most and the last person you watched. You do not remember the
distribution, and you will confidently reconstruct one that never existed.

---

## The Three Sources

Almost everything you will analyse in a game project comes from one of three
places, and they have very different characters.

**Telemetry** is what the game records about itself while it runs: positions,
events, deaths, times, item pickups. It is objective, it arrives in enormous
quantities, and it tells you *what* happened and never *why*. Telemetry is cheap
to collect and easy to collect badly — a log with no session identifier or no
timestamp is close to worthless.

**Performance capture** is a special case of telemetry with its own conventions
and its own failure modes. Frame times, memory, draw calls. It is the one area
where the standard summary — average frame rate — is actively misleading, which is
why Section 5 spends its time there.

**Asking people** is surveys, interviews and think-aloud sessions. It is the only
source that tells you *why*, and the least reliable. People misremember, they are
polite, they rationalise, and they answer the question they think you meant. A
survey tells you what a player believes about their experience, which is a real
and useful thing, but it is not the same as what happened.

**The three disagree, and that is the point.** The most valuable finding in any
playtest is usually the place where telemetry and opinion contradict each other:
players say the level is fine and the heat map shows nobody finds the side route;
players say the game runs smoothly and the frame log shows a hitch every two
seconds. A team with only one source never finds these.

---

## Turning a Vague Question into a Measurable One

The question a designer asks is almost never answerable as asked.

| What you are asked | What you can actually measure |
| :- | :- |
| "Is the level too hard?" | Deaths per attempt; time to first completion; proportion who never finish; where attempts end |
| "Does the game feel good?" | Time to first successful interaction; retry rate after failure; rated responsiveness |
| "Is the tutorial working?" | Proportion who use the taught mechanic later, unprompted |
| "Do people like the art?" | Rated appeal; which screenshots they choose; where they stop to look |

Notice that each right-hand cell names a **specific number you could write down
after watching one person play**. That is the test. If you cannot describe how you
would record it for a single player, you have not finished turning the question
into a measurement.

Notice also that the right-hand column is often not one measure but several, and
that they can disagree. This is not a failure of the method. A tutorial that
everybody completes and nobody learns from is a real and common outcome, and only
two measures together can find it.

---

## Deciding in Advance

Before you collect anything, write down what result would change your mind. "If
more than a third of testers never find the second key, I will add a light cue."

This costs nothing and protects you from the most comfortable mistake in the
discipline: running the test, getting an ambiguous result, and finding in it
whatever you already believed. If the threshold is written down beforehand, the
data gets to answer. If it is not, you will answer and attribute it to the data.

---

## Worked Examples

```{r}
#| label: worked-example-data
#| include: false
library(readr)
sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
survey   <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)

n_a    <- sum(sessions$build == "A")
n_b    <- sum(sessions$build == "B")
frus_a <- sum(sessions$build == "A" & sessions$quit_reason == "Frustrated")
frus_b <- sum(sessions$build == "B" & sessions$quit_reason == "Frustrated")
pct    <- function(k, n) paste0(round(100 * k / n), "%")

item_means <- sort(colMeans(survey[, -1]))
stopifnot(names(item_means)[1] == "objective_clear")
```

**The question as asked.** The *Tidebreak* team ran `r nrow(sessions)` playtest
sessions, `r n_a` on Build A and `r n_b` on Build B. The designer's question is
"is Build A worse?", which cannot be answered as asked: worse at what, for whom,
by how much?

**The question as measured.** Two measures bear on it, one from each end of the
three sources:

| Source | Measure | Recorded per player as |
| :- | :- | :- |
| Telemetry | Did the session end in a frustration quit? | `quit_reason` is `Frustrated` |
| Asking people | "I always knew what my objective was" | A rating from 1 to 5 |

**The threshold, written first.** Before anyone opens the file, the team writes:
*"If more than a quarter of the sessions on either build end in a frustration
quit, that build does not ship."*

**The result.** `r frus_a` of `r n_a` Build A sessions ended in a frustration quit
(`r pct(frus_a, n_a)`), against `r frus_b` of `r n_b` on Build B
(`r pct(frus_b, n_b)`). Build A crosses the line; Build B does not. Nobody has to
argue about it, because the argument happened before the data arrived.

**The second source.** Of the `r ncol(survey) - 1` survey items, the objective item
is the lowest rated, with a mean of `r round(item_means[["objective_clear"]], 2)`
on the 1 to 5 scale. Telemetry says *what* happened: players on Build A left in
frustration. The survey offers a candidate for *why*: players did not know what
they were meant to be doing. Neither source alone would have pointed a designer at
the objective marker.

---

## R Demonstrations

Everything below runs as written, provided the data files sit in the `data/`
folder of your project as set up in [Lab 00](../labs/setting-up.qmd).

```{r}
#| label: demo-load
library(readr)
library(dplyr)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
survey   <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)
```

`read_csv()` reads a file into a **tibble**, R's table of rows and columns. The
path is relative to the project folder, which is why it starts `data/`.

```{r}
#| label: demo-glimpse
glimpse(sessions)
```

`glimpse()` prints one line per column: its name, its type and its first few
values. `<chr>` is text and `<dbl>` is a number. This is the first check on any
new file: does every column have the type you expected? Here `cohort` is text,
which is correct, but R has no idea that Novice, Regular and Veteran come in an
order. That knowledge is yours to supply ([rule 3](ten-rules.qmd)).

```{r}
#| label: demo-summary
summary(survey)
```

`summary()` gives the minimum, quartiles, median, mean and maximum of every
numeric column. Read down the `Mean` rows and the weakest survey item stands out
without any further code.

```{r}
#| label: demo-count
sum(sessions$build == "A" & sessions$quit_reason == "Frustrated")
sum(sessions$build == "B" & sessions$quit_reason == "Frustrated")
```

A condition such as `sessions$build == "A"` produces one `TRUE` or `FALSE` per row,
and `sum()` counts the `TRUE`s. Combining two conditions with `&` counts the rows
where both hold. This is how every number in the worked example was produced.

---

## Interpretation Guidance

Every finding is read at three levels, and the discipline is keeping them apart.

**What the data shows.** `r frus_a` of `r n_a` Build A sessions ended in a
frustration quit, against `r frus_b` of `r n_b` on Build B. This is a count. It is
not in dispute.

**What the analysis suggests.** Something in Build A frustrates players that Build
B avoided, and the survey's weakest item points at objective clarity as a
candidate.

**What can legitimately be concluded.** Build A fails the threshold the team set
in advance. That is all. The survey figure is averaged across both builds, so on
its own it cannot say that the unclear objective belongs to Build A. Linking the
two is a hypothesis for the next test, not a finding.

Telemetry reports behaviour and surveys report belief. When they agree, you have a
lead. When they disagree, you have the most valuable finding in the playtest, and
the disagreement itself is what goes in the report.

---

## Common Mistakes

**Collecting first, asking later.** Logging everything and hoping a question
emerges produces large files and no decisions. Write the question down first
([rule 2](ten-rules.qmd)).

**Trusting one source.** A survey on its own tells you what players believe; a log
on its own tells you what they did. Treat either alone as half the evidence.

**Setting the threshold after seeing the result.** A threshold chosen after the
data has been seen is always met by exactly the right margin. If it was not written
down beforehand, it is not a threshold ([rule 1](ten-rules.qmd)).

**Logging without identifiers.** A death event with no session or player ID cannot
be joined to anything, and a survey response with no session ID cannot be joined
to the telemetry from the same match.

**Reading `glimpse()` and moving on.** The types it prints are R's guess. A 1 to 5
rating arrives as `<dbl>`, and R will average it without complaint. Whether it
should be averaged is a decision R cannot make for you.

---

## Short Practice Activities

Each takes two or three minutes.

**1.** Run `glimpse(survey)`. How many survey items are there, and what type has R
given them?

::: {.callout-note collapse="true" title="Check your answer"}
There are `r ncol(survey) - 1` items after `session_id`, and every one is `<dbl>`.
R treats them as ordinary numbers; they are ordinal ratings, and the difference
matters as soon as you average them.
:::

**2.** Count the sessions that ended with `quit_reason` equal to
`"Technical fault"`, then write the threshold that would make you delay a release
over it.

::: {.callout-note collapse="true" title="Check your answer"}

```{r}
#| label: practice-fault
sum(sessions$quit_reason == "Technical fault")
```

Any threshold is defensible if it was written first and names a number: "if more
than one session in twenty ends in a technical fault, we do not release". "If it
seems like a lot" is not a threshold.
:::

**3.** Rewrite "is the corridor too dangerous?" as one telemetry measure and one
survey measure, each recordable for a single player.

::: {.callout-note collapse="true" title="Check your answer"}
Telemetry: deaths in the corridor per session. Asking people: agreement with "I
understood why I died". Either alone is weaker than both together: players who die
there often *and* do not understand why describe a different problem from players
who die there often and accept it.
:::

---

## Summary

- Measurement earns its place where taste fails: you cannot see your own game, feel
  a rare event, or remember twenty playtests.
- Telemetry says what happened, performance capture says how the machine coped,
  and asking people says what players believe. The disagreements are the findings.
- A question is measurable when you can say how you would record it for one
  player.
- Write the threshold that would change your mind before you look.
- `read_csv()`, `glimpse()` and `summary()` are the first three things you run on
  any new file.

---

## Links to Relevant Reference Material

- [Ten Rules](ten-rules.qmd): rule 1 (decide what would change your mind) and
  rule 2 (a number without a question is decoration)
- [R and Quarto Reference: getting data in](../reference/r-quarto-reference.qmd#getting-data-in)
- [Lab 00: setting up](../labs/setting-up.qmd) and
  [Lab 01: exploring playtest data](../labs/reading-data.qmd), Part B

---

## Reflective Questions

- Pick a mechanic in your own project. Write the designer's question, then three
  different measures that would bear on it, then say which one you would trust
  most and why.
- Name something about your own game that you can no longer judge, because you
  have played it too often.
- Your team disagrees about whether a puzzle is too obscure. What number, recorded
  from a single playtester, would settle it?

---
