---
title: "Stage 4 Demo"
subtitle: "Coding comments, triangulating, and a report that recomputes, live"
---

::: {.callout-note appearance="simple" icon="false" title="Get the files"}
- **Demo code:** [stage-4-demo.qmd](../downloads/stage-4-demo.qmd){download="stage-4-demo.qmd"}, every chunk on this page in one file you can run.
- **Data:** [tidebreak_sessions.csv](../data/tidebreak_sessions.csv){download="tidebreak_sessions.csv"}, [tidebreak_survey.csv](../data/tidebreak_survey.csv){download="tidebreak_survey.csv"}, [tidebreak_sessions_raw.csv](../data/tidebreak_sessions_raw.csv){download="tidebreak_sessions_raw.csv"}. Save them in a `data/` folder next to the demo file.
:::

## Demo 1 — From evidence to a report that recomputes

**Step 1. The survey arrives wide**

```{r}
#| label: d2-wide
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
survey <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)
head(survey, 2)   # one column per question: the item is a column NAME
```

**Question.** How would you get the mean of all eight items, by build,
from this shape?

**Step 2. Pivot it long**

```{r}
#| label: d2-long
# Stack every column except the key: names into item, values into rating.
long <- survey |>
  pivot_longer(-session_id, names_to = "item", values_to = "rating")

dim(survey)
dim(long)   # 120 sessions x 8 items = 960 rows
```

**Step 3. Every item at once, now that item is a column**

```{r}
#| label: d2-items
long |>
  group_by(item) |>
  summarise(agree = mean(rating >= 4)) |>   # TRUE counts as 1
  arrange(agree)
```

**Step 4. Code the comments, then ask whether the sources agree**

```{r}
#| label: d2-code
coded <- sessions |>
  mutate(code = case_when(
    grepl("lost|circles|meant to go|objective", comment) ~ "Lost or unclear",
    grepl("route|exit|signposting", comment)             ~ "Found the way",
    grepl("Died|hard", comment)                          ~ "Difficulty",
    TRUE                                                 ~ "General"
  ))

coded |> count(code, build)                      # against the telemetry...
coded |> count(code, enjoyed = enjoyment >= 5)   # ...and against enjoyment
```

**Question.** Build B took far fewer wrong turns, yet more of its players
wrote that they were lost. Which evidence do we trust for what? Where does this go
in the report: Results, Discussion, or both?

**Step 5. The report skeleton**

The report is a `.qmd` file: the eight assessment headings, a hidden setup chunk,
and numbers written as inline R.

````markdown
---
title: "Tidebreak Build B: Playtest Report"
---

```{{r}}
#| echo: false
sessions <- readr::read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
```

## Executive Summary
## Introduction
## Methodology

We analysed `r knitr::inline_expr("nrow(sessions)")` sessions.

## Results
## Discussion
## Conclusion
## References
## Appendices
````

**Step 6. Watch the inline numbers recompute**

```{r}
#| label: d2-render-helper
#| include: false
# Replaces each inline `r ...` in some text with its value, as quarto render does.
# knit_child() moves the working directory, so put it back afterwards.
render_text <- function(text) {
  env <- parent.frame()
  wd  <- getwd()
  on.exit(setwd(wd))
  cat(knitr::knit_child(text = text, quiet = TRUE, envir = env))
}
```

```{r}
#| label: d2-recompute
# One sentence of the Methodology, with its numbers as inline R.
part2 <- paste(
  "We analysed `r nrow(sessions)` sessions. Build A's median session lasted",
  "`r median(sessions$session_minutes[sessions$build == 'A'], na.rm = TRUE)` minutes."
)

# render_text() does to this sentence what quarto render does to the report.
render_text(part2)   # against the full data...

# ...then against Stage 2's raw export with incomplete rows dropped.
raw <- read_csv("data/tidebreak_sessions_raw.csv", show_col_types = FALSE)
sessions <- raw[!duplicated(raw), ]
sessions <- sessions[complete.cases(sessions), ]

render_text(part2)
```

**Question.** Which of those two sentences would have gone into the report
if the numbers were typed by hand? Then: open your own `report.qmd`.
