---
title: "Stage 3 Demo"
subtitle: "Figures and relationships, live"
aliases:
  - week-3.html
---

::: {.callout-note appearance="simple" icon="false" title="Get the files"}
- **Demo code:** [stage-3-demo.qmd](../downloads/stage-3-demo.qmd){download="stage-3-demo.qmd"}, every chunk on this page in one file you can run.
- **Data:** [tidebreak_sessions.csv](../data/tidebreak_sessions.csv){download="tidebreak_sessions.csv"}, [tidebreak_survey.csv](../data/tidebreak_survey.csv){download="tidebreak_survey.csv"}. Save them in a `data/` folder next to the demo file.
:::

Two short demos, one per clinic. Demo 1 opens the 3A figures clinic; Demo 2 opens
the 3B relationships clinic.

## Demo 1 — A figure, one layer at a time (3A)

**Step 1. Data and mapping, no geometry**

```{r}
#| label: d1-empty
#| fig-alt: "An empty grey ggplot2 panel with an x axis labelled wrong_turns and no data drawn."
library(readr)
library(dplyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)

# Data plus mapping gives axes and nothing else: no geometry, no marks.
ggplot(sessions, aes(x = wrong_turns))
```

**Step 2. Add a geometry**

```{r}
#| label: d1-geom
#| fig-alt: "Default dark grey ggplot2 histogram of wrong turns for all sessions."
# + adds a layer. binwidth = 1 because wrong turns are whole numbers.
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1)
```

**Question.** This is correct. Why does it not answer "did Build B help"?

**Step 3. Map a group, and watch it mislead**

```{r}
#| label: d1-fill
#| fig-alt: "Histogram of wrong turns with bars filled by build, Build A in red stacked on Build B in teal."
# fill = build inside aes() is a MAPPING: colour follows the data, legend appears.
# But histograms stack by default, so Build A floats on top of Build B.
ggplot(sessions, aes(x = wrong_turns, fill = build)) +
  geom_histogram(binwidth = 1)
```

**Step 4. Facet instead, and label it for a reader**

```{r}
#| label: d1-facet
#| fig-alt: "Faceted histograms of wrong turns per session, Build A above and Build B below. Build B's sessions sit further to the left."
#| fig-height: 5
# One panel per build, stacked so the x axes line up.
# fill = "grey60" OUTSIDE aes() is SETTING: fixed colour, no legend needed.
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```

**Question.** Build B is lower. What else is different about its shape?
What caption would you write under this in the report?

**Step 5. The mapping bug, on purpose**

```{r}
#| label: d1-bug
#| fig-alt: "Scatterplot of score against awareness in which every point is pinkish red and a legend reads colour: blue, the mapping bug."
# "blue" inside aes() is treated as data: salmon points, and a legend saying "blue".
ggplot(sessions, aes(x = awareness, y = score)) +
  geom_point(aes(colour = "blue"))
```

Move `colour = "blue"` outside `aes()` live, then open your own data. That is the
rest of the clinic.

---

## Demo 2 — Relationships, and when to distrust them (3B)

**Step 6. The scatterplot, then the number**

```{r}
#| label: d2-scatter
#| fig-alt: "Scatterplot of score against awareness rating with a rising black straight-line fit."
# colour SET outside aes(); geom_smooth(method = "lm") adds a straight-line trend.
ggplot(sessions, aes(x = awareness, y = score)) +
  geom_point(colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  labs(title = "Awareness against score", x = "Awareness rating", y = "Score")

cor(sessions$awareness, sessions$score)    # the picture, as one number
cor(sessions$shots_fired, sessions$score)  # the stat everyone assumes matters
```

**Question.** So should we buff awareness? Hold the answer until Step 9.

**Step 7. Many at once: the correlation matrix**

```{r}
#| label: d2-matrix
# Spearman, because difficulty and enjoyment are 1-7 ratings.
sessions |>
  select(deaths, wrong_turns, difficulty, enjoyment, session_minutes) |>
  cor(method = "spearman") |>
  round(2)
```

**Question.** Read the `deaths` row. Which pair goes together most? Which
measure goes with nothing at all?

**Step 8. Telemetry against self-report**

```{r}
#| label: d2-nav
survey <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)

joined <- sessions |> left_join(survey, by = "session_id")

# Did players who got lost rate the level as harder to navigate?
cor(joined$wrong_turns, joined$level_navigable, method = "spearman")

joined |>
  group_by(build) |>
  summarise(wrong_turns = mean(wrong_turns), navigable = mean(level_navigable))
```

**Question.** Build A's players got lost far more, and rated navigation the
same. What does that say about asking players whether they got lost?

**Step 9. A strong correlation that means the opposite**

```{r}
#| label: d2-reversal
#| fig-alt: "Survivability against mobility coloured by archetype. Each archetype has a rising trend line, while the clusters together fall from the top left to the bottom right."
cor(sessions$mobility, sessions$survivability)   # faster players are more fragile?

# Map colour to archetype: one cloud and one trend line per group.
ggplot(sessions, aes(x = mobility, y = survivability, colour = archetype)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Mobility rating", y = "Survivability rating", colour = "Archetype")

sessions |>
  group_by(archetype) |>
  summarise(r = cor(mobility, survivability))     # the same, within each group
```

**Question.** The overall correlation is negative; every archetype's line
goes up. Which one answers "should a Scout lose health as they get faster"?
Archetype is the confounder. Back to Step 6: what would you need to do before
buffing awareness? (A test, not a patch.) This is Lab 04, Exercise 06; then open
your own data.
