---
title: "Week 3 Demo"
subtitle: "Building figures and reporting, live"
---

> The lecturer's script for the two 15-minute demo slots in Week 3. Every chunk
> runs and shows its output. Step through it top to bottom; the bold headings are
> the places to pause. `ggplot2` from here on.

---

## Demo 1 — A figure, one layer at a time

**Step 1. Data and mapping, no geometry**

```{r}
#| label: d1-empty
library(readr)
library(dplyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)

# Data plus mapping gives axes and nothing else: no geometry, no marks.
ggplot(sessions, aes(x = wrong_turns))
```

**Step 2. Add a geometry**

```{r}
#| label: d1-geom
# + adds a layer. binwidth = 1 because wrong turns are whole numbers.
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1)
```

**Pause here.** Ask: this is correct. Why does it not answer "did Build B help"?

**Step 3. Map a group, and watch it mislead**

```{r}
#| label: d1-fill
# fill = build inside aes() is a MAPPING: colour follows the data, legend appears.
# But histograms stack by default, so Build A floats on top of Build B.
ggplot(sessions, aes(x = wrong_turns, fill = build)) +
  geom_histogram(binwidth = 1)
```

**Step 4. Facet instead, and label it for a reader**

```{r}
#| label: d1-facet
#| fig-height: 5
# One panel per build, stacked so the x axes line up.
# fill = "grey60" OUTSIDE aes() is SETTING: fixed colour, no legend needed.
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```

**Pause here.** Ask: Build B is lower. What else is different about its shape?

**Step 5. The mapping bug, on purpose**

```{r}
#| label: d1-bug
# "blue" inside aes() is treated as data: salmon points, and a legend saying "blue".
ggplot(sessions, aes(x = awareness, y = score)) +
  geom_point(aes(colour = "blue"))
```
