---
title: "Week 2 Demo"
subtitle: "Building figures and comparing builds, live"
---

> The lecturer's script for the two 15-minute demo slots in Week 2. Every chunk
> runs and shows its output. Step through it top to bottom; the bold headings are
> the places to pause. `ggplot2` from here on.

---

## Demo 1 — A figure, one layer at a time

**Step 1. Data and mapping, no geometry**

```{r}
#| label: d1-empty
library(readr)
library(dplyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)

# Data plus mapping gives axes and nothing else: no geometry, no marks.
ggplot(sessions, aes(x = wrong_turns))
```

**Step 2. Add a geometry**

```{r}
#| label: d1-geom
# + adds a layer. binwidth = 1 because wrong turns are whole numbers.
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1)
```

**Pause here.** Ask: this is correct. Why does it not answer "did Build B help"?

**Step 3. Map a group, and watch it mislead**

```{r}
#| label: d1-fill
# fill = build inside aes() is a MAPPING: colour follows the data, legend appears.
# But histograms stack by default, so Build A floats on top of Build B.
ggplot(sessions, aes(x = wrong_turns, fill = build)) +
  geom_histogram(binwidth = 1)
```

**Step 4. Facet instead, and label it for a reader**

```{r}
#| label: d1-facet
#| fig-height: 5
# One panel per build, stacked so the x axes line up.
# fill = "grey60" OUTSIDE aes() is SETTING: fixed colour, no legend needed.
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```

**Pause here.** Ask: Build B is lower. What else is different about its shape?

**Step 5. The mapping bug, on purpose**

```{r}
#| label: d1-bug
# "blue" inside aes() is treated as data: salmon points, and a legend saying "blue".
ggplot(sessions, aes(x = awareness, y = score)) +
  geom_point(aes(colour = "blue"))
```

---

## Demo 2 — Is the difference real, and does it matter?

**Step 6. Another playtest, another answer**

```{r}
#| label: d2-resample
# Pretend we could only afford 15 sessions per build. Draw five such playtests.
set.seed(8015)
replicate(5, {
  a <- sample(sessions$wrong_turns[sessions$build == "A"], 15)
  b <- sample(sessions$wrong_turns[sessions$build == "B"], 15)
  mean(a) - mean(b)  # the difference this playtest would have reported
})
```

**Pause here.** Ask: same builds, five answers. Which one is "the" difference?

**Step 7. The t-test: size, interval, p-value**

```{r}
#| label: d2-ttest
# Different players on each build, so the independent-samples test.
result <- t.test(wrong_turns ~ build, data = sessions)
result

result$conf.int  # read this first: how big could the difference plausibly be?
result$p.value   # then this: could a gap this big be chance alone?
```

**Step 8. Same players, both builds**

```{r}
#| label: d2-paired
paired <- read_csv("data/tidebreak_paired.csv", show_col_types = FALSE)

# Paired: each player compared with themselves, so player skill cancels out.
t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)$conf.int

# Same data, wrongly treated as two separate groups: a much wider interval.
t.test(paired$wrong_turns_a, paired$wrong_turns_b)$conf.int
```

**Step 9. Significant is not the same as important**

```{r}
#| label: d2-difficulty
# Build B was rated harder, and the result is significant...
harder <- t.test(difficulty ~ build, data = sessions)
harder

# ...but the players did not actually die more.
t.test(deaths ~ build, data = sessions)$p.value
```

**Finish here.** Wrong turns: a big, certain difference worth acting on.
Difficulty: a significant difference of
`r round(unname(diff(harder$estimate)), 2)` points on a seven-point scale, with
nothing in the play data behind it. Same test, opposite decisions. The number that decides is the
size, not the p-value.
