---
title: "Week 3 Demo"
subtitle: "Figures, relationships, reshaping and a report that recomputes, live"
---

> The lecturer's script for the two 15-minute demo slots in Week 3. Every chunk
> runs and shows its output. Step through it top to bottom; the bold headings are
> the places to pause. `ggplot2` from here on.

---

## Demo 1 — A figure, one layer at a time

**Step 1. Data and mapping, no geometry**

```{r}
#| label: d1-empty
library(readr)
library(dplyr)
library(ggplot2)

sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)

# Data plus mapping gives axes and nothing else: no geometry, no marks.
ggplot(sessions, aes(x = wrong_turns))
```

**Step 2. Add a geometry**

```{r}
#| label: d1-geom
# + adds a layer. binwidth = 1 because wrong turns are whole numbers.
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1)
```

**Pause here.** Ask: this is correct. Why does it not answer "did Build B help"?

**Step 3. Map a group, and watch it mislead**

```{r}
#| label: d1-fill
# fill = build inside aes() is a MAPPING: colour follows the data, legend appears.
# But histograms stack by default, so Build A floats on top of Build B.
ggplot(sessions, aes(x = wrong_turns, fill = build)) +
  geom_histogram(binwidth = 1)
```

**Step 4. Facet instead, and label it for a reader**

```{r}
#| label: d1-facet
#| fig-height: 5
# One panel per build, stacked so the x axes line up.
# fill = "grey60" OUTSIDE aes() is SETTING: fixed colour, no legend needed.
ggplot(sessions, aes(x = wrong_turns)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~ build, ncol = 1) +
  labs(title = "Wrong turns per session, by build",
       x = "Wrong turns", y = "Sessions")
```

**Pause here.** Ask: Build B is lower. What else is different about its shape?

**Step 5. The mapping bug, on purpose**

```{r}
#| label: d1-bug
# "blue" inside aes() is treated as data: salmon points, and a legend saying "blue".
ggplot(sessions, aes(x = awareness, y = score)) +
  geom_point(aes(colour = "blue"))
```

**Step 6. The scatterplot, done properly**

```{r}
#| label: d1-scatter
# colour SET outside aes(); geom_smooth(method = "lm") adds the straight-line fit.
ggplot(sessions, aes(x = awareness, y = score)) +
  geom_point(colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  labs(title = "Awareness against score", x = "Awareness rating", y = "Score")

cor(sessions$awareness, sessions$score)    # the line, as one number
cor(sessions$shots_fired, sessions$score)  # the stat everyone assumes matters
```

**Step 7. A strong correlation that means the opposite**

```{r}
#| label: d1-reversal
cor(sessions$mobility, sessions$survivability)   # faster players are more fragile?

# Map colour to archetype: one cloud and one trend line per group.
ggplot(sessions, aes(x = mobility, y = survivability, colour = archetype)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Mobility rating", y = "Survivability rating", colour = "Archetype")
```

**Finish here.** Ask: the overall correlation is negative; every archetype's line
goes up.
Which one answers "should a Scout lose health as they get faster"? Archetype is the
confound. This is Lab 03, Exercise 06.

---

## Demo 2 — From a survey to a report that recomputes

**Step 8. The survey arrives wide**

```{r}
#| label: d2-wide
library(tidyr)

survey <- read_csv("data/tidebreak_survey.csv", show_col_types = FALSE)
head(survey, 2)   # one column per question: the item is a column NAME
```

**Pause here.** Ask: how would you get the mean of all eight items, by build,
from this shape?

**Step 9. Pivot it long**

```{r}
#| label: d2-long
# Stack every column except the key: names into item, values into rating.
long <- survey |>
  pivot_longer(-session_id, names_to = "item", values_to = "rating")

dim(survey)
dim(long)   # 120 sessions x 8 items = 960 rows
```

**Step 10. Every item at once, now that item is a column**

```{r}
#| label: d2-items
long |>
  group_by(item) |>
  summarise(agree = mean(rating >= 4)) |>   # TRUE counts as 1
  arrange(agree)
```

**Step 11. The report skeleton**

The report is a `.qmd` file: the nine part headings, a hidden setup chunk, and
numbers written as inline R.

````markdown
---
title: "Tidebreak Build B: Playtest Report"
---

```{{r}}
#| echo: false
sessions <- readr::read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
```

## 1. Research question
## 2. Experimental design

We analysed `r knitr::inline_expr("nrow(sessions)")` sessions.

## 3. Descriptive evidence
...
## 9. Design conclusion
````

**Step 12. Watch the inline numbers recompute**

```{r}
#| label: d2-render-helper
#| include: false
# Replaces each inline `r ...` in some text with its value, as quarto render does.
# knit_child() moves the working directory, so put it back afterwards.
render_text <- function(text) {
  env <- parent.frame()
  wd  <- getwd()
  on.exit(setwd(wd))
  cat(knitr::knit_child(text = text, quiet = TRUE, envir = env))
}
```

```{r}
#| label: d2-recompute
# One sentence of the report, with its numbers as inline R.
part2 <- paste(
  "We analysed `r nrow(sessions)` sessions. Build A's median session lasted",
  "`r median(sessions$session_minutes[sessions$build == 'A'], na.rm = TRUE)` minutes."
)

# render_text() does to this sentence what quarto render does to the report.
render_text(part2)   # against the full data...

# ...then against Week 2's raw export with incomplete rows dropped.
raw <- read_csv("data/tidebreak_sessions_raw.csv", show_col_types = FALSE)
sessions <- raw[!duplicated(raw), ]
sessions <- sessions[complete.cases(sessions), ]

render_text(part2)
```

**Finish here.** Same sentence, nobody retyped a number, and both figures changed.
The second one also shows why the cleaning log goes in part 2: the report is only
as honest as the data it was rendered against.
