---
title: "Week 2 Demo"
subtitle: "Cleaning, the six verbs and comparing builds, live"
---

> The lecturer's script for the two 15-minute demo slots in Week 2. Every chunk
> runs and shows its output. Step through it top to bottom; the bold headings are
> the places to pause. Base R plotting only this week.

---

## Demo 1 — Cleaning is analysis

**Step 1. Read the export as delivered**

```{r}
#| label: d1-read
library(readr)
library(dplyr)

# trim_ws = FALSE: see the stray spaces Excel would see. read_csv hides them.
raw <- read_csv("data/tidebreak_sessions_raw.csv",
                trim_ws = FALSE, show_col_types = FALSE)
nrow(raw)   # 120 sessions were played
```

**Step 2. Rows exported twice**

```{r}
#| label: d1-dupes
sum(duplicated(raw))             # rows repeating every column
sum(duplicated(raw$session_id))  # rows repeating the key: same count, so all exact

clean <- raw[!duplicated(raw), ] # keep the first copy of each
nrow(clean)
```

**Step 3. Labels that should match**

```{r}
#| label: d1-labels
table(clean$cohort)   # spaces and capitals make extra "cohorts"

clean$cohort <- tolower(trimws(clean$cohort))
table(clean$cohort)   # always check the fix with table()
```

**Step 4. Where the gaps are**

```{r}
#| label: d1-gaps
na_counts <- colSums(is.na(clean))
na_counts[na_counts > 0]

# Split the gaps by something that might explain them.
table(build = clean$build, quit = clean$quit_reason,
      missing = is.na(clean$session_minutes))["A", , ]
```

**Pause here.** Ask: every blank session length is a Build A frustration quit.
Noise, or a finding?

**Step 5. The one-line fix that removes the finding**

```{r}
#| label: d1-complete
kept <- clean[complete.cases(clean), ]   # "just drop the NAs"
nrow(kept)

table(clean$build[clean$quit_reason == "Frustrated"])  # before
table(kept$build[kept$quit_reason == "Frustrated"])    # after
```

**Pause here.** Ask: nothing warned us. What goes in the cleaning log?

**Step 6. The six verbs, read as a sentence**

```{r}
#| label: d1-verbs
# The pipe passes the left-hand result in as the first argument on the right.
clean |>
  filter(build == "A", quit_reason == "Frustrated") |>
  mutate(logged = !is.na(session_minutes)) |>
  select(session_id, cohort, session_minutes, logged) |>
  arrange(session_minutes) |>
  head(4)
```

**Step 7. Same question, base R then tidyverse**

```{r}
#| label: d1-same
# Split wrong_turns by build, apply mean, combine: split-apply-combine.
tapply(clean$wrong_turns, clean$build, mean)

clean |>
  group_by(build) |>
  summarise(n = n(), mean_wrong = mean(wrong_turns))
```

**Finish here.** `tapply()` does it in one call. `group_by()` + `summarise()` does
the same thing in two named steps, and takes a second statistic for one more
argument.

---

## Demo 2 — Is the difference real, and does it matter?

**Step 8. Look before you test**

```{r}
#| label: d2-look
# The complete export: same sessions, logging fault fixed.
sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)

boxplot(wrong_turns ~ build, data = sessions,
        main = "Wrong turns by build", ylab = "Wrong turns", col = "grey85")
```

**Step 9. Another playtest, another answer**

```{r}
#| label: d2-resample
# Pretend we could only afford 15 sessions per build. Draw five such playtests.
set.seed(8015)
replicate(5, {
  a <- sample(sessions$wrong_turns[sessions$build == "A"], 15)
  b <- sample(sessions$wrong_turns[sessions$build == "B"], 15)
  mean(a) - mean(b)  # the difference this playtest would have reported
})
```

**Pause here.** Ask: same builds, five answers. Which one is "the" difference?

**Step 10. The t-test: size, interval, p-value**

```{r}
#| label: d2-ttest
# Different players on each build, so the independent-samples test.
result <- t.test(wrong_turns ~ build, data = sessions)
result

result$conf.int  # read this first: how big could the difference plausibly be?
result$p.value   # then this: could a gap this big be chance alone?
```

**Step 11. Same players, both builds**

```{r}
#| label: d2-paired
paired <- read_csv("data/tidebreak_paired.csv", show_col_types = FALSE)

# Paired: each player compared with themselves, so player skill cancels out.
t.test(paired$wrong_turns_a, paired$wrong_turns_b, paired = TRUE)$conf.int

# Same data, wrongly treated as two separate groups: a much wider interval.
t.test(paired$wrong_turns_a, paired$wrong_turns_b)$conf.int
```

**Step 12. Significant is not the same as important**

```{r}
#| label: d2-difficulty
# Build B was rated harder, and the result is significant...
harder <- t.test(difficulty ~ build, data = sessions)
harder

# ...but the players did not actually die more.
t.test(deaths ~ build, data = sessions)$p.value
```

**Finish here.** Wrong turns: a big, certain difference worth acting on.
Difficulty: a significant difference of
`r round(unname(diff(harder$estimate)), 2)` points on a seven-point scale, with
nothing in the play data behind it. Same test, opposite decisions. The number that decides is the
size, not the p-value.
