---
title: "Week 1 Demo"
subtitle: "Describing playtest data, live"
---

> The lecturer's script for the two 15-minute demo slots in Week 1. Every chunk
> runs and shows its output. Step through it top to bottom; the bold headings are
> the places to pause. Base R plotting only this week.

---

## Demo 1 — From values to a data frame

**Step 1. A vector is a column**

```{r}
#| label: d1-vector
# Five players' deaths, typed in by hand. c() combines values into a vector.
deaths <- c(2, 0, 5, 3, 1)

deaths          # print it
length(deaths)  # how many players
deaths[3]       # the third player: square brackets pick by position
deaths > 2      # a condition gives one TRUE/FALSE per player
sum(deaths > 2) # TRUE counts as 1, so this is a count, not a total
```

**Step 2. A missing value poisons the arithmetic**

```{r}
#| label: d1-na
# The fourth player's log was lost. NA means "not recorded", not zero.
deaths <- c(2, 0, 5, NA, 1)

mean(deaths)               # NA: R will not guess
mean(deaths, na.rm = TRUE) # say out loud what you are leaving out
```

**Step 3. Read the real file**

```{r}
#| label: d1-read
library(readr)
library(dplyr)

# The path is relative to the project folder.
sessions <- read_csv("data/tidebreak_sessions.csv", show_col_types = FALSE)
```

**Step 4. Look before you touch**

```{r}
#| label: d1-look
glimpse(sessions)          # every column: name, type, first values
summary(sessions$deaths)   # $ pulls out one column as a vector
```

**Pause here.** Ask the room: which columns are text, and is `cohort` just text?

---

## Demo 2 — Describing, then looking

**Step 5. Centre and spread**

```{r}
#| label: d2-centre
kills <- sessions$kills

mean(kills)
median(kills)
sd(kills)                     # typical distance from the mean
IQR(kills)                    # width of the middle half
quantile(kills, c(0.1, 0.9))  # any percentiles you ask for
```

**Step 6. The shape, the groups, the counts**

```{r}
#| label: d2-plots
hist(sessions$time_alive_s, breaks = 15,
     main = "Seconds alive", xlab = "Seconds", col = "grey85")

boxplot(session_minutes ~ build, data = sessions,
        main = "Session length by build", ylab = "Minutes", col = "grey85")

barplot(table(sessions$quit_reason),
        main = "How sessions ended", col = "grey85")

plot(sessions$deaths, sessions$score, pch = 19,
     main = "Deaths against score", xlab = "Deaths", ylab = "Score")
```

**Pause here.** Ask: which build loses players, and which plot told you?

**Step 7. The frame-rate trap**

```{r}
#| label: d2-frames
frames <- read_csv("data/tidebreak_frame_times.csv", show_col_types = FALSE)

# Base R: one summary per build, one call per summary.
tapply(frames$ms, frames$build, mean)
tapply(frames$ms, frames$build, quantile, 0.99)

# Tidyverse: every summary per build in one table.
frames %>%
  group_by(build) %>%
  summarise(fps    = 1000 / mean(ms),
            median = median(ms),
            p99    = quantile(ms, 0.99))
```

**Finish here.** Same frame rate, one build stutters. The mean could never have
shown it.
