---
title: "Is It Normal Enough?"
subtitle: "z-scores, checking the bell shape, and Q-Q plots"
---

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")

within_sd <- function(x, k) mean(abs(x - mean(x)) < k * sd(x))
aw <- sessions$awareness
sm <- sessions$session_minutes
b  <- frames$ms[frames$build == "B"]
```

> **Going further.** This page extends
> [Distributions and Shape](../notes/05-distributions-and-shape.qmd#the-normal-distribution).
> Nothing here is needed for the labs or the assessment. Read it if you want to
> know how to check a shape properly, or when a later page or a paper you are
> reading assumes data is "approximately normal".

---

## z-scores

A **z-score** says how many standard deviations a value sits from the mean:

$$z = \frac{\text{value} - \text{mean}}{\text{SD}}$$

A player with an awareness rating of 90 has
$z = (90 - `r round(mean(aw), 1)`) / `r round(sd(aw), 1)` = `r round((90 - mean(aw)) / sd(aw), 1)`$:
nearly 2 SDs above average. If awareness is normal-shaped, the 68–95–99.7 rule
puts that in roughly the top few per cent.

z-scores let you compare values measured on different scales: "two SDs above
average" means the same thing for a score, a time or a rating.

**The catch** is the "if". For a skewed variable the SD is inflated by the tail, and
the rule no longer tells you how unusual a value is. Score is a good example:

```{r}
#| label: z-score
(2000 - mean(sessions$score)) / sd(sessions$score)
mean(sessions$score < 2000)
```

A score of 2,000 is about `r round((2000 - mean(sessions$score)) / sd(sessions$score), 1)`
SDs above the mean. But the Sentinel collapses give score a left tail, which
inflates the SD and drags the mean down. The percentile is more honest:
`r pct_fmt(mean(sessions$score < 2000))` of sessions scored below 2,000.

---

## Checking the Bell Shape

Real data is never exactly normal. The useful question is whether it is close
enough for the 68–95–99.7 rule to work. Draw the histogram on a density scale and
lay the normal curve with the same mean and SD over it:

```{r}
#| label: fig-normal-check
#| fig-cap: "Awareness follows the normal curve closely; session length, piled up against its 32-minute cap, does not."
#| fig-alt: "Two histograms. Left: awareness, with a bell curve that follows the bars closely. Right: session length, where the bell curve overestimates the middle and misses the tall bars piled up near 30 minutes."
#| fig-height: 3.5
op <- par(mfrow = c(1, 2), mar = c(4, 3, 2, 1))
hist(aw, breaks = 12, freq = FALSE, col = pal$fill, border = "white",
     main = "Awareness", xlab = "Rating", ylab = "")
curve(dnorm(x, mean(aw), sd(aw)), add = TRUE, lwd = 2, col = pal$highlight)
hist(sm, breaks = 12, freq = FALSE, col = pal$fill, border = "white",
     main = "Session length", xlab = "Minutes", ylab = "")
curve(dnorm(x, mean(sm), sd(sm)), add = TRUE, lwd = 2, col = pal$highlight)
par(op)
```

`freq = FALSE` draws the histogram as a density, so the bars and the curve share a
scale. `curve()` draws a function of `x`; `dnorm()` is the normal curve.

The rule makes a checkable prediction, so check it. `abs(x - mean(x)) < k * sd(x)`
is `TRUE` for every value within `k` SDs of the mean, and its `mean()` is the
proportion:

```{r}
#| label: within-sd
within_sd <- function(x, k) mean(abs(x - mean(x)) < k * sd(x))
within_sd(sessions$awareness, 1)
within_sd(sessions$session_minutes, 1)
```

| Variable | Within 1 SD | Within 2 SDs | Verdict |
| :- | :-: | :-: | :- |
| Normal distribution | 68% | 95% | |
| `awareness` | `r pct_fmt(within_sd(aw, 1))` | `r pct_fmt(within_sd(aw, 2))` | close: the rule works |
| `distance_m` | `r pct_fmt(within_sd(sessions$distance_m, 1))` | `r pct_fmt(within_sd(sessions$distance_m, 2))` | symmetric but flatter than a bell |
| `session_minutes` | `r pct_fmt(within_sd(sm, 1))` | `r pct_fmt(within_sd(sm, 2))` | not normal: skewed and capped |
| Build B frame ms | `r pct_fmt(within_sd(b, 1))` | `r pct_fmt(within_sd(b, 2))` | not normal: a spike and a long tail |

For session length, "mean ± 1 SD" covers only `r pct_fmt(within_sd(sm, 1))` of
sessions, so quoting the mean and SD as if they described a bell would mislead.
For frame times, the SD is inflated by a handful of very slow frames, and the
mean ± 2 SDs covers almost every frame while saying nothing about the ones that
matter.

---

## Q-Q Plots

A **Q-Q plot** is a sharper check than a histogram. It plots each value against
where it would fall if the data were perfectly normal. Normal data lies along a
straight line; skew bends the ends away from it.

```{r}
#| label: fig-qq
#| fig-cap: "Awareness hugs the line; session length flattens out at the top, where values pile up against the 32-minute cap."
#| fig-alt: "Two Q-Q plots. Left, awareness: the points lie close to the straight reference line. Right, session length: the points follow the line in the middle but flatten out along the top."
#| fig-height: 3.2
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
qqnorm(aw, main = "Awareness", pch = 19, cex = 0.6); qqline(aw, col = pal$highlight)
qqnorm(sm, main = "Session length", pch = 19, cex = 0.6); qqline(sm, col = pal$highlight)
par(op)
```

You will see Q-Q plots in discussions of the assumptions behind a t-test.

---

## Why It Matters Later

Even when individual values are far from normal, the *averages* of samples tend to
be close to normal. That fact, the **central limit theorem**, is what makes
confidence intervals and t-tests work on data like session length. It is why the
Stage 2 notes can use a t-test on skewed playtest data, as long as the sample is
not tiny and there is no extreme tail. For the Stage 2 version, see
[From Sample to Population](../notes/07-sample-to-population.qmd).

**Formal normality tests** (Shapiro–Wilk and others) exist, and you will see them in
papers. They are not used in this course. With small samples they rarely detect
anything; with large ones they flag differences too small to matter. The
histogram and the Q-Q plot tell you more.

---

## Try It

Check the 68–95 rule against `kills` and `distance_m`, then draw a Q-Q plot of each.
Which one would you describe with a mean and SD, and which with a median and IQR?

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answer"}

```{r}
#| label: try-it
#| fig-alt: "Two Q-Q plots, for kills and distance travelled."
#| fig-height: 3.2
within_sd(sessions$kills, 1)
within_sd(sessions$distance_m, 1)
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
qqnorm(sessions$kills, main = "Kills", pch = 19, cex = 0.6)
qqline(sessions$kills, col = pal$highlight)
qqnorm(sessions$distance_m, main = "Distance", pch = 19, cex = 0.6)
qqline(sessions$distance_m, col = pal$highlight)
par(op)
```

Neither matches the rule exactly. `kills` puts
`r pct_fmt(within_sd(sessions$kills, 1))` within 1 SD: it is a small count, so
its values bunch on a few whole numbers and its Q-Q plot is a staircase. `distance_m`
puts `r pct_fmt(within_sd(sessions$distance_m, 1))` there: symmetric, but flatter
than a bell, so the points follow the line in the middle and bend at the ends.

Distance is the one to describe with a mean and SD: it is symmetric, its mean and
median almost coincide, and the SD is a fair guide to spread. For kills, report the
median and IQR, or simply the count at each value: with so few possible values, the
counts are the clearest description.
:::
