---
title: "Regression"
subtitle: "The line through a scatterplot: slope, R² and residuals"
---

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")
library(ggplot2)

aw_cohort <- round(tapply(sessions$awareness, sessions$cohort, mean), 1)
r_aw  <- cor(sessions$awareness, sessions$score)
fit   <- lm(score ~ awareness, data = sessions)
b0    <- unname(coef(fit)[1]); b1 <- unname(coef(fit)[2])
b1_ci <- confint(fit)["awareness", ]
r2    <- summary(fit)$r.squared
res   <- resid(fit)
worst <- sessions$archetype[order(res)][1:7]
stopifnot(sum(worst == "Sentinel") >= 5,
          isTRUE(all.equal(r2, r_aw^2)))
```

> **Going further.** This page extends
> [Relationships and Correlation](../notes/10-relationships-and-correlation.qmd).
> Nothing here is needed for the labs or the assessment. Read it if your team wants
> to say *how much* one measure changes with another, or if a residual might be
> hiding your finding.

---

## Simple Linear Regression

Correlation says how tightly two variables follow a line. **Simple linear
regression** finds that line, and so answers a more useful question: *how much does
one variable change, on average, for each unit of the other?*

```{r}
#| label: lm
fit <- lm(score ~ awareness, data = sessions)
coef(fit)
confint(fit)
```

`lm()` stands for *linear model*. `score ~ awareness` reads "score as explained by
awareness", the same formula style as `boxplot(y ~ group)`. The result describes a
straight line:

$$\text{predicted score} = `r round(b0)` + `r round(b1, 1)` \times \text{awareness}$$

**The slope** is the number that matters: each extra point of awareness goes with
`r round(b1, 1)` more points of score, on average (95% CI
`r round(b1_ci[1], 1)` to `r round(b1_ci[2], 1)`). It is in the units of the
question, so a designer can use it: the
`r round(aw_cohort[["Veteran"]] - aw_cohort[["Novice"]])`-point gap in average
awareness between Novices and Veterans goes with about
`r round((aw_cohort[["Veteran"]] - aw_cohort[["Novice"]]) * b1, -1)` points of
score.

**The intercept** is the predicted score at awareness zero: `r round(b0)`. No player
has an awareness anywhere near zero (the lowest is `r min(sessions$awareness)`), so
it means nothing on its own; it just anchors the line.

---

## Residuals

```{r}
#| label: fig-regression
#| fig-cap: "Score against awareness, with the regression line. Each grey segment is a residual; the sessions furthest below the line are mostly Sentinel collapses."
#| fig-alt: "Scatterplot of score against awareness with a rising red line. Short grey vertical segments join each point to the line. A handful of points sit far below the line at the bottom, with long segments, coloured differently from the rest."
#| fig-height: 4.2
sessions$predicted <- fitted(fit)
sessions$residual  <- resid(fit)
ggplot(sessions, aes(awareness, score)) +
  geom_segment(aes(xend = awareness, yend = predicted), colour = "grey75") +
  geom_point(aes(colour = residual < -600), size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = pal$highlight) +
  scale_colour_manual(values = c(`FALSE` = pal$line, `TRUE` = "#8a5a9c"),
                      labels = c("Other sessions", "Residual below −600"), name = NULL) +
  labs(x = "Awareness rating", y = "Score") +
  theme(legend.position = "bottom")
```

**A residual** is the gap between what actually happened and what the line
predicted: actual minus predicted. The line is chosen to make the residuals, taken
together, as small as possible (it minimises the sum of their squares, hence
"least squares"). `geom_smooth(method = "lm")` draws the same line.

**Residuals are where the findings hide.** The seven sessions furthest *below* the
line, the ones that scored far less than their awareness predicted, include
`r sum(worst == "Sentinel")` Sentinels. The regression has found the Sentinel
collapses from [Describing Variables](../notes/04-describing-variables.qmd#worked-example)
on its own: those games are not explained by awareness at all.

```{r}
#| label: worst-residuals
sessions[order(resid(fit))[1:7], c("session_id", "archetype", "awareness", "score")]
```

---

## R²

**R²**, the **coefficient of determination**, is the share of the variation in
score that the line accounts for: `r round(r2, 2)`, or `r pct_fmt(r2)`. For a simple
regression it is exactly the correlation squared
(`r round(r_aw, 2)`² = `r round(r2, 2)`). Awareness accounts for about a third of the
differences in score; the other two thirds are everything else.

---

## Predicting, Carefully

The line predicts a score of about
`r round(predict(fit, data.frame(awareness = 80)), -1)` for awareness 80. That
prediction is an average for players like that, not a promise for any one player,
whose residual could easily be `r round(sd(res), -1)` points either way. And
**never predict outside the data**: the line knows nothing about awareness below
`r min(sessions$awareness)` or above `r max(sessions$awareness)`. Using it there is
**extrapolation**, and the straight line may not continue.

**A regression is still a correlation.** A steep, significant slope says awareness
and score move together, by this much. It does not say that raising a player's
awareness would raise their score. Every warning about confounding applies
unchanged.

---

## Reading the Full Summary

`summary(fit)` prints everything at once:

```{r}
#| label: lm-summary
summary(fit)
```

The `awareness` row of the **Coefficients** table gives the slope, its standard
error and a p-value testing whether the true slope could be zero. **Multiple
R-squared** is R². Regression with several predictors at once (multiple
regression) can adjust for a confounder such as cohort, but it is beyond this
course, and adjusting for the confounders you measured does not protect you from
the ones you did not.

---

## Try It

1. Fit a regression of `score` on `deaths`. Interpret the slope in one sentence a
   designer could use, and compare its R² with awareness's.
2. A designer reads the awareness regression and says: "If we raise a Novice's
   awareness by 20 points, they will score about `r round(20 * b1, -1)` more." Give
   two reasons to doubt it.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="Show answers"}
**1.**

```{r}
#| label: try-deaths
fit_d <- lm(score ~ deaths, data = sessions)
coef(fit_d)
summary(fit_d)$r.squared
```

Each extra death goes with about
`r abs(round(coef(lm(score ~ deaths, data = sessions))[2]))` fewer points of score,
on average. R² is `r round(summary(lm(score ~ deaths, data = sessions))$r.squared, 2)`,
a little below awareness's `r round(r2, 2)`. The same caution applies: dying more
and scoring less may both come from being outplayed.

**2.** First, the slope is an association across players, not the effect of
changing one player: awareness ratings may reflect skill rather than cause it.
Second, it is an average: individual players scatter around the line by about
`r round(sd(res), -1)` points. A third reason, if the Novice starts low: their
raised awareness may still be in a range where the data is thin.
:::
