---
title: "Reshaping and Joining"
subtitle: "pivot_longer, pivot_wider, and joins that do what you meant"
---

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")
library(dplyr)
library(tidyr)
```

> **Going further.** This page extends
> [Interpreting and Reporting Evidence](../notes/11-reporting-evidence.qmd#reporting-a-whole-survey).
> You need it when your data arrives in more than one file, or in a shape that
> does not suit the question. Nothing here is required for the labs.

---

## Long and Back Again

`pivot_longer()` stacks columns into rows; `pivot_wider()` spreads rows back into
columns. Long suits `group_by()` and `ggplot2`; wide suits a table a reader
compares across.

```{r}
#| label: pivot
long <- survey |>
  pivot_longer(-session_id, names_to = "item", values_to = "rating")

# Long: one row per item per build. Wide: builds side by side for reading.
long |>
  left_join(select(sessions, session_id, build), by = "session_id") |>
  group_by(item, build) |>
  summarise(agree = round(mean(rating >= 4), 2), .groups = "drop") |>
  pivot_wider(names_from = build, values_from = agree) |>
  mutate(change = B - A) |>
  arrange(desc(abs(change)))
```

`names_from` says which column's values become the new column names;
`values_from` says what fills them. With the builds side by side, the change is one
subtraction. `objective_clear` moved most; `level_navigable`, the item you might
expect Build B's signage to move, barely changed.

---

## Joins

A **join** adds columns from one table to another by matching on a **key**, a
column both tables share, such as `session_id`.

| Join | Keeps | Use when |
| :- | :- | :- |
| `left_join(x, y)` | Every row of `x`, with `y`'s columns where there is a match, `NA` where not | Adding information to your main table. The usual choice |
| `inner_join(x, y)` | Only rows with a match in both | You only want sessions that have both a log and a survey |
| `anti_join(x, y)` | Rows of `x` with **no** match in `y` | Finding what is missing: sessions with no survey |

```{r}
#| label: joins
nrow(sessions)
nrow(survey)

sessions |> left_join(survey, by = "session_id") |> nrow()
sessions |> anti_join(survey, by = "session_id") |> nrow()
```

Every session has a survey here, so the join keeps all `r nrow(sessions)` rows and
the anti-join finds none. Real exports are rarely that tidy.

### Checking a join

Run these three checks every time you join:

1. **Did the row count change?** A `left_join()` should keep the rows of `x`. If it
   gains rows, the key is repeated in `y`.
2. **Are there new `NA`s?** Rows of `x` with no match get `NA` in every new column.
   Count them, and find them with `anti_join()`.
3. **Is the key really unique?** `count(key) |> filter(n > 1)` on each table.

```{r}
#| label: join-checks
survey |> count(session_id) |> filter(n > 1)   # no repeated keys: good

joined <- sessions |> left_join(survey, by = "session_id")
sum(is.na(joined$objective_clear))             # no unmatched sessions: good
```

**A join that silently duplicates rows** is one of the commonest ways for a
playtest analysis to go wrong without an error. Two survey rows for one session
double that session in every summary. The checks take a minute.
