---
title: "Sampling Theory"
subtitle: "The standard error formula, the central limit theorem, and the bootstrap"
---

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")

pop      <- rep(sessions$session_minutes, times = 1000)
pop_mean <- mean(pop)
pop_sd   <- sd(pop)

set.seed(7)
draw_means <- function(n, reps = 1000) replicate(reps, mean(sample(pop, n)))
means_10  <- draw_means(10)
means_30  <- draw_means(30)
means_100 <- draw_means(100)
se_30 <- pop_sd / sqrt(30)

mins_b <- sessions$session_minutes[sessions$build == "B"]
n_b    <- length(mins_b)
ci_b   <- t.test(mins_b)$conf.int
```

> **Going further.** This page extends
> [From Sample to Population](../notes/07-sample-to-population.qmd). Nothing here is
> needed for the labs or the assessment. Read it if you want to know *why* the
> standard error and the confidence interval work, or if you need an interval for
> something other than a mean.

---

## The Standard Error Formula

The standard error can be estimated from a single sample:

$$\text{SE} = \frac{\text{SD}}{\sqrt{n}}$$

Check it against the simulation on the notes page. The population SD is
`r round(pop_sd, 2)`, so for n = 30 the formula gives `r round(pop_sd, 2)` / √30 =
`r round(se_30, 2)`. The SD of a thousand simulated playtest means is
`r round(sd(means_30), 2)`. The formula works.

**Bigger samples give smaller standard errors, but slowly.** Because n is under a
square root, you need **four times** the players to **halve** the standard error:

```{r}
#| label: fig-sample-size-se
#| fig-cap: "Means of 1,000 simulated playtests at three sizes. Each quadrupling of n roughly halves the spread."
#| fig-alt: "Three horizontal strips of points. With n = 10 the playtest means spread widely around the dashed true mean; with n = 30 they are tighter; with n = 100 they cluster closely around it."
#| fig-height: 3.4
par(mar = c(4, 6, 1, 1))
stripchart(list("n = 10" = means_10, "n = 30" = means_30, "n = 100" = means_100),
           method = "jitter", jitter = 0.25, pch = 16, cex = 0.4, las = 1,
           col = adjustcolor(pal$line, alpha.f = 0.3),
           xlab = "Playtest mean session length (minutes)")
abline(v = pop_mean, lty = 2, lwd = 2, col = pal$highlight)
```

| Players per playtest | SE from the formula | SD of 1,000 simulated means |
| :-: | :-: | :-: |
| 10 | `r round(pop_sd / sqrt(10), 2)` | `r round(sd(means_10), 2)` |
| 30 | `r round(pop_sd / sqrt(30), 2)` | `r round(sd(means_30), 2)` |
| 100 | `r round(pop_sd / sqrt(100), 2)` | `r round(sd(means_100), 2)` |

---

## The Central Limit Theorem

Session length is lopsided and two-humped, nothing like a bell. Yet the means of
playtests drawn from it form a bell almost perfectly:

```{r}
#| label: fig-clt
#| fig-cap: "The means of 1,000 playtests of 30, with a normal curve laid over them. The individual sessions are far from normal; their means are not."
#| fig-alt: "Histogram of 1,000 sample means of session length, a narrow symmetric bell centred near 23 minutes, with a red normal curve that follows it closely."
#| fig-height: 3.4
par(mar = c(4, 3, 1, 1))
hist(means_30, breaks = 30, freq = FALSE, col = pal$fill, border = "white",
     main = "", xlab = "Mean session length (minutes)", ylab = "", yaxt = "n")
curve(dnorm(x, mean(means_30), sd(means_30)), add = TRUE, lwd = 2,
      col = pal$highlight)
```

That is the **central limit theorem**: **whatever shape the individual values
have, the means of reasonably large samples are approximately normal.**

It is why the normal distribution matters even for data that is not normal. The
68–95–99.7 rule does not describe the session lengths, but it does describe the
sampling distribution of their mean. So about 95% of playtest means fall within 2
standard errors of the true mean. That single fact is what confidence intervals and
t-tests are built on.

**How large is "reasonably large"?** For mildly skewed data, 30 is usually plenty.
For heavily skewed data, such as frame times with rare hitches, it takes more. And
it applies to *means*: it does not rescue a 99th percentile.

---

## The Interval by Hand

The "about 2" in "mean ± about 2 SE" is 1.96 for large samples. For smaller ones R
uses a slightly larger number from the **t distribution**, which allows for the SD
itself being estimated from the sample. The part after the ± is the **margin of
error**.

```{r}
#| label: ci-by-hand
m  <- mean(mins_b)
se <- sd(mins_b) / sqrt(length(mins_b))
multiplier <- qt(0.975, df = length(mins_b) - 1)

c(mean = m, se = se, multiplier = multiplier)
m + c(-1, 1) * multiplier * se
```

The same as `t.test(mins_b)$conf.int`: `r ci_fmt(ci_b)`.

$$\text{95\% CI} = \bar{x} \pm t_{0.975,\,n-1} \times \frac{s}{\sqrt{n}}$$

A 99% interval uses a larger multiplier, so it is wider: more confidence costs
precision.

---

## The Bootstrap

The formula relies on the central limit theorem. The **bootstrap** gets an interval
without it, by treating your sample as a stand-in for the population and resampling
it. Draw `r n_b` sessions *with replacement* from Build B's `r n_b`, take the mean,
and repeat a thousand times; the middle 95% of those means is the interval.

```{r}
#| label: bootstrap
set.seed(3)
boot_means <- replicate(1000, mean(sample(mins_b, replace = TRUE)))
quantile(boot_means, c(0.025, 0.975))
```

It agrees closely with `t.test()`'s `r ci_fmt(ci_b)`. The bootstrap earns its keep
for summaries with no simple formula, such as a median or a 99th percentile:
replace `mean` with `median` in the line above and it still works.

```{r}
#| label: bootstrap-median
set.seed(3)
boot_medians <- replicate(1000, median(sample(mins_b, replace = TRUE)))
quantile(boot_medians, c(0.025, 0.975))
```
