---
title: "Testing in Depth"
subtitle: "Errors and power, many tests at once, Q-Q plots, and a shuffle for any summary"
---

```{r}
#| label: setup
#| include: false
source("notes/_setup.R")

wt_a <- sessions$wrong_turns[sessions$build == "A"]
wt_b <- sessions$wrong_turns[sessions$build == "B"]

# Errors and power, by simulation
pop_wt <- rep(sessions$wrong_turns, times = 1000)
set.seed(5)
type1 <- mean(replicate(2000, t.test(sample(pop_wt, 30), sample(pop_wt, 30))$p.value < 0.05))
power_at <- function(n) {
  set.seed(n)
  mean(replicate(2000, t.test(sample(wt_a, n, TRUE), sample(wt_b, n, TRUE))$p.value < 0.05))
}
power_tab <- sapply(c(5, 10, 15, 30), power_at)
pow_fmt   <- function(p) if (p >= 0.995) ">99%" else pct_fmt(p)

# Multiple comparisons
items  <- names(survey)[-1]
sv     <- merge(survey, sessions[, c("session_id", "build")])
item_p <- sapply(items, function(i) t.test(sv[[i]] ~ sv$build)$p.value)
fwer8  <- 1 - 0.95^length(items)

# A shuffle for the 99th percentile
p99_gap <- function(ms, g) quantile(ms[g == "B"], 0.99) - quantile(ms[g == "A"], 0.99)
corr    <- frames[frames$zone == "Corridor", ]
obs_p99 <- p99_gap(corr$ms, corr$build)
set.seed(9)
shuf_p99 <- replicate(2000, p99_gap(corr$ms, sample(corr$build)))

stopifnot(type1 > 0.03, type1 < 0.07,
          all(diff(power_tab) > 0), power_tab[1] < 0.5,
          sum(item_p < 0.05) == 1, names(which.min(item_p)) == "objective_clear",
          item_p[["objective_clear"]] < 0.05 / length(items),
          max(abs(shuf_p99)) < abs(obs_p99))
```

> **Going further.** This page extends
> [Comparing Groups](../notes/08-comparing-groups.qmd). Nothing here is needed for
> the labs or the assessment. Read it when you are planning how many players a
> playtest needs, when you are tempted to test every survey item, or when the mean
> is the wrong thing to compare.

---

## Two Ways to Be Wrong

A test makes a decision from incomplete information, so it can be wrong in two
directions.

| | The build really makes no difference | The build really makes a difference |
| :- | :- | :- |
| **Test says "significant"** | **Type I error**: a false alarm | Correct |
| **Test says "not significant"** | Correct | **Type II error**: a missed effect |

**A Type I error** is acting on a difference that is not there: rebuilding a level
because of a fluke. Its rate is α. To check, draw two groups of 30 from the *same*
population of wrong turns, so there is no real difference, and test them two
thousand times: `r pct_fmt(type1, 1)` come out significant, close to the 5% that
α = 0.05 promises.

**A Type II error** is missing a difference that is there: shipping a build with a
problem because the playtest was too small to see it. The chance of *avoiding* it
is the test's **power**. Power grows with sample size. Drawing playtests of
different sizes from *Tidebreak*'s real Build A and Build B sessions, and counting
how often the t-test detects the signposting effect:

| Sessions per build | 5 | 10 | 15 | 30 |
| :- | :-: | :-: | :-: | :-: |
| Chance of detecting the effect | `r pow_fmt(power_tab[1])` | `r pow_fmt(power_tab[2])` | `r pow_fmt(power_tab[3])` | `r pow_fmt(power_tab[4])` |

Even this large effect is missed more often than not with five sessions per build.
A smaller effect needs far more players. **Plan the playtest size around the
smallest difference that would change your decision**, which is the "how many
players" question from [Planning a Playtest](../notes/01-planning-a-playtest.qmd#how-many-players),
now with numbers.

**The two errors trade off.** A stricter α, such as 0.01, means fewer false alarms
and more missed effects. Which mistake is worse depends on the decision: a false
alarm that costs a week of rework, or a missed problem that ships to every player.

::: {.callout-note collapse="true" appearance="simple" icon="false" title="The simulation code"}

```{r}
#| label: power-code
#| eval: false
# Type I: two samples from the same population, tested 2,000 times
pop_wt <- rep(sessions$wrong_turns, times = 1000)
mean(replicate(2000, t.test(sample(pop_wt, 30), sample(pop_wt, 30))$p.value < 0.05))

# Power: samples of n from each real build, tested 2,000 times
power_at <- function(n) {
  mean(replicate(2000, t.test(sample(wt_a, n, TRUE), sample(wt_b, n, TRUE))$p.value < 0.05))
}
sapply(c(5, 10, 15, 30), power_at)
```

:::

---

## Testing Many Things at Once

At α = 0.05, each test on a difference that is not real has a 5% chance of a false
alarm. Run several and the chances add up. With eight independent tests, the chance
that **at least one** comes up significant by luck alone is
1 − 0.95^`r length(items)` = `r pct_fmt(fwer8)`.

The *Tidebreak* survey has `r length(items)` items. Testing each one by build:

```{r}
#| label: many-tests
sv <- merge(survey, sessions[, c("session_id", "build")])
round(sort(sapply(names(survey)[-1],
                  function(i) t.test(sv[[i]] ~ sv$build)$p.value)), 3)
```

One item, `objective_clear`, is significant. Is it real, or the one-in-three false
alarm? Three things say real. It survives the simplest correction, **Bonferroni**:
divide α by the number of tests, here 0.05 / `r length(items)` =
`r round(0.05 / length(items), 4)`, and its p-value is still below that. It was the
item the team expected to move, written down before the test. And it agrees with
the telemetry: Build A had the frustration quits.

**The defence is to decide in advance.** Name the comparisons that answer your
question before you run any of them. A result you went looking for after scanning
twenty tests needs a second playtest before anyone acts on it
([rule 1](../notes/ten-rules.qmd)).

---

## Q-Q Plots and the t-Test

A t-test assumes the means are roughly normal. With a few dozen players per group
that is nearly always true, thanks to the central limit theorem
([Sampling Theory](sampling-theory.qmd#the-central-limit-theorem)). A **Q-Q plot**
is the standard visual check on the data itself. Close to the line means close to
normal:

```{r}
#| label: fig-qq
#| fig-cap: "Wrong turns (left) follow the line in steps, because they are whole numbers: fine for a t-test. Build B's frame times (right) bend sharply away at the top: a long tail the mean cannot describe."
#| fig-alt: "Two Q-Q plots. Left: wrong turns, with points in horizontal steps lying along the diagonal reference line. Right: Build B frame times, with points along the line in the middle but curving steeply upward at the right end."
#| fig-height: 3.4
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
qqnorm(sessions$wrong_turns, main = "Wrong turns", pch = 19, cex = 0.6,
       col = pal$line)
qqline(sessions$wrong_turns, col = pal$highlight, lwd = 2)
b_ms <- frames$ms[frames$build == "B"]
qqnorm(b_ms, main = "Build B frame times", pch = 19, cex = 0.4, col = pal$line)
qqline(b_ms, col = pal$highlight, lwd = 2)
par(op)
```

**Strong outliers and a mismatched design matter far more than normality**, and
no amount of data fixes the wrong design.

---

## A Shuffle for Any Summary

The permutation test on the notes page shuffled the build labels and recomputed a
difference in **means**. Nothing about the shuffle needs a mean. Replace it with
any summary that answers your question, and the logic is unchanged. That makes it
the fallback when the mean is the wrong thing to compare.

Frame times are the case. The question is not "is Build B's average frame slower?"
but "is Build B's worst one per cent slower?". In the Corridor:

```{r}
#| label: shuffle-p99
corr <- frames[frames$zone == "Corridor", ]

p99_gap <- function(ms, g) quantile(ms[g == "B"], 0.99) - quantile(ms[g == "A"], 0.99)

observed <- p99_gap(corr$ms, corr$build)
set.seed(9)
shuffled <- replicate(2000, p99_gap(corr$ms, sample(corr$build)))

observed
mean(abs(shuffled) >= abs(observed))
```

Build B's 99th percentile in the Corridor is `r round(obs_p99, 1)` ms slower than
Build A's, and none of 2,000 shuffles produced a gap that large: the hitch is not
chance. A t-test on the same frames would answer a different question, about the
mean, which is the one that hides the hitch.
